### 2020-11-28 ext/ext.Rnw
setting up files

\documentclass[11pt]{article}

<<debug,echo=F,results=F>>=
#DEBUG <- TRUE
DEBUG <- FALSE
if(DEBUG){
  ### set run levels low
  gammaSlice_run_level <-3
  r0slice_run_level <- 1
} else {
  ### for longer runs
  gammaSlice_run_level <- 3
  r0slice_run_level <- 2
}
@

<<packages,include=F,cache=F>>=
#library("coda")
#library("foreach") #included in doParallel or doMPI
#library("doMC")
library("ggplot2")
#library("grid")
#library("magrittr")
#library("plyr")
#library("reshape2")
#library("xtable")
library("spatPomp")
#ibrary("dplyr")
#stopifnot(packageVersion("pomp")>="2.0")
library(doParallel)
library(doRNG)

cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()  
registerDoParallel(cores)

gl_cores <- 36
doob_cores <- 40
ito_cores <- 8
registerDoParallel(cores)

@

<<set-opts,include=F,cache=F>>=

options(
        scipen=2,
        help_type="html",
        stringsAsFactors=FALSE,
        prompt="R> ",
        continue="+  ",
        width=70,
        useFancyQuotes=FALSE,
        reindent.spaces=2,
        xtable.comment=FALSE
        )
@

<<knitr-opts,include=F,cache=F,purl=F>>=
library("knitr")
opts_knit$set(concordance=TRUE)
opts_chunk$set(
    progress=TRUE,prompt=TRUE,highlight=FALSE,
    tidy=TRUE,
    tidy.opts=list(
        keep.blank.line=FALSE
    ),
    comment="",
    warning=FALSE,
    message=FALSE,
    error=TRUE,
    echo=TRUE,
    cache=FALSE,
    strip.white=TRUE,
    results="markup",
    background="#FFFFFF00",
    size="normalsize",
    fig.path="figure/",
    fig.lp="fig:",
    fig.align="left",
    fig.show="asis",
#    dpi=300,
    dev="pdf",
    dev.args=list(
        bg="transparent",
        pointsize=9
    )
)

myround<- function (x, digits = 1) {
  # taken from the broman package
  if (digits < 1) 
    stop("This is intended for the case digits >= 1.")
  if (length(digits) > 1) {
    digits <- digits[1]
    warning("Using only digits[1]")
  }
  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
  tmp[tmp == paste0("-", zero)] <- zero
  tmp
}

@

\usepackage{amsmath,amsfonts,amssymb,graphicx}

\input{../header-ms.tex}
\input{../theorems.tex}

\renewcommand\thefigure{E-\arabic{figure}}
\renewcommand\thetable{E-\arabic{table}}
\renewcommand\thepage{E-\arabic{page}}
\renewcommand\thesection{E\arabic{section}}
\renewcommand\theequation{E\arabic{equation}}
\renewcommand\theprop{E\arabic{prop}}
\renewcommand\thelemma{E\arabic{lemma}}
\renewcommand{\contentsname}{Contents}
\renewcommand{\refname}{References}

\newcommand\altx{x^\prime}
\newcommand\alty{y^\prime}

\bibliographystyle{apalike}
%% added or modified from pnas-new.cls
\usepackage{natbib} 
\usepackage{fullpage}
\renewcommand\myeqref[1]{(\ref{#1})}

\renewcommand\siOnly[1]{#1}
\newcommand\secTitleSpace{\hspace{3mm}}

\begin{document}

%\begin{center}
\title{Extensions to ``{\mytitle}''}
\date{\today}
\author{E. L. Ionides, K. Asfaw, J. Park and A. A. King}
\maketitle

%{\Large \bf Extensions to ``Adapted simulation island filters for partially observed spatiotemporal systems''}

%\medskip

%\authorList

%\end{center}

\medskip

This is a place to write additional material, not necessarily intended for publication, supplying comments or additional details or other related material.

\medskip

\tableofcontents

\newpage

\section{\secTitleSpace A bootstrap guide function within ASIF-IR}\label{sec:bootguide_abfir}

Brainstorming the naming of guide functions, it seems to me that the previous mean/variance based guide function could be called a {\it simulated moments} guide, and the {\it quantile guide function} is a variation on something that could be called a {\it bootstrap guide function} which arises most naturally when all quantiles are considered in the \citet{park20} quantile guide function.

In the context of ASIF-IR, here is some pseudocode.
There are two possibilities given for constructing the bootstrap guide residuals.
One is to calculate the residuals at time $t_{n}$ and scale them according to $\sqrt{\frac{t_{\time}-t_{\time,\ninter}}{t_{\time}-t_{\time,0}}}$, as done in the quantile guide function.
Another, given here as \eic{ALT}, is to save residuals at each intermediate time point and use these.
That could have advantages if the variability in the latent process changes considerably over one observation interval.
For measles growing after a trough, there might perhaps be substantially more cases, and correspondingly more variability due to log scale noise, in the second half of a 2-week observation interval.
This requires more memory, but not more calls to rprocess or flow, and if the mememory is available it might take comparable time. 
It may require more care to code efficiently, so perhaps \eic{ALT} can wait.
\begin{center}
\noindent\begin{tabular}{l}
\hline
{\bf {\AIRSIF}. Adapted bagged filter with intermediate resampling.} 
\firstLineSpace \\  
{\bf \hspace{13mm} With updated guide function}
\vspace{0.4mm} \\
\hline
\firstLineSpace
Initialize adapted simulation: $\myvec{X}^{\IF}_{0,\island} \sim f_{\myvec{X}_0}(\myvec{x}_0)$
\\
For $\time\ \mathrm{in}\ \seq{1}{\Time}$
\\
\asp Guide simulations:
    $\myvec{X}_{\time,\island,\npgir}^{G} \sim 
    f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}} 
    \big( \myvec{x}_{\time}\given \myvec{X}^{\IF}_{\time-1,\island} \big)$
\\
\asp Guide residuals: $\myvec{\epsilon}_{\time,\island,\npgir}=
 \myvec{X}_{\time,\island,\npgir}^{G} - \myvec{\mu}\big( \myvec{X}^{\IF}_{\time-1,\island},t_{\time-1},t_{\time} \big)$
\\
\asp \eic{ALT} Guide residuals: $\myvec{\epsilon}_{\time,\ninter,\island,\npgir}=
 \myvec{X}_{\time,\ninter,\island,\npgir}^{G} - \myvec{\mu}\big( \myvec{X}^{\IF}_{\time-1,\island},t_{\time-1},t_{\time,\ninter} \big)$ 
\\
\asp $\guideFunc^{\resample}_{\time,0,\island,\np}=1 \; \; $ and
$\; \myvec{X}_{\time,0,\island,\np}^{\GR}=\myvec{X}^{\IF}_{\time-1,\island}$
\\
\asp For $\ninter  \,\, \mathrm{in} \,\, \seq{1}{\Ninter}$
\\
\asp\asp Intermediate proposals:
        ${\myvec{X}}_{\time,\ninter,\island,\np}^{\GP}
          \sim {f}_{{\myvec{X}}_{\time,\ninter}|{\myvec{X}}_{\time,\ninter-1}}
          \big(\mydot|{\myvec{X}}_{\time,\ninter-1,\island,\np}^{\GR}\big)$ 
\\
\asp\asp 
        $\myvec{\mu}^{\GP}_{\time,\ninter,\island,\np} 
           = \myvec{\mu}\big( \myvec{X}^{\GP}_{\time,\ninter,\island,\np},t_{\time,\ninter},t_{\time} \big)$
\\
\asp\asp  % Guide function: 
        $
\guideFunc_{\time,\ninter,\island,\np}=
\frac{1}{\Npgir^{\Unit}}
          \prod_{\unit=1}^{\Unit}
	  \sum_{\npgir=1}^{\Npgir}
          f_{Y_{\unit,\time}|X_{\unit,\time}}
          \Big(
            \data{y}_{\unit,\time}\given \mu^{\GP}_{\unit,\time,\ninter,\island,\np} + \sqrt{\frac{t_{\time}-t_{\time,\ninter}}{t_{\time}-t_{\time,0}}} \myvec{\epsilon}_{\time,\island,\npgir}
	     \giventh \theta
          \Big)$
\\
\asp\asp  \eic{ALT} % Guide function: 
        $
\guideFunc_{\time,\ninter,\island,\np}=
\frac{1}{\Npgir^{\Unit}}
          \prod_{\unit=1}^{\Unit}
	  \sum_{\npgir=1}^{\Npgir}
          f_{Y_{\unit,\time}|X_{\unit,\time}}
          \Big(
            \data{y}_{\unit,\time}\given \mu^{\GP}_{\unit,\time,\ninter,\island,\np} + \myvec{\epsilon}_{\time,\Ninter,\island,\npgir} - \myvec{\epsilon}_{\time,\ninter,\island,\npgir}
	    \giventh \theta 
          \Big)$
\\
\asp\asp Guide weights:
$w^G_{\time,\ninter,\island,\np}= \guideFunc^{}_{\time,\ninter,\island,\np}
         \big/ \guideFunc^{\resample}_{\time,\ninter-1,\island,\np}$
\\
\asp\asp
      Resampling:
        $\prob\big[\resampleIndex({\island,\np})=a \big] = w^G_{\time,\ninter,\island,a}
\Big( \sum_{\altNp=1}^{\Np}w^G_{\time,\ninter,\island,\altNp}\Big)^{-1}$
\\
\asp\asp
        $\myvec{X}_{\time,\ninter,\island,\np}^{\GR}=\myvec{X}_{\time,\ninter,\island,\resampleIndex({\island,\np})}^{\GP}\; \; $ and
        $\; \guideFunc^{\resample}_{\time,\ninter,\island,\np}= \guideFunc^{}_{\time,\ninter,\island,\resampleIndex({\island,\np})}\,$
\\
\asp
End For
\\
\asp
  Set $\myvec{X}^{\IF}_{\time,\island}=\myvec{X}^{\GR}_{\time,\Ninter,\island,1}$ 
\\ 
\asp Measurement weights:
  $w^M_{\unit,\time,\island,\npgir} = 
    f_{Y_{\unit,\time}|X_{\unit,\time}} 
    \big (\data{y}_{\unit,\time}\given X^{G}_{\unit,\time,\island,\npgir} \big)$
\\
\asp % Prediction weights:
  $w^{\LCP}_{\unit,\time,\island,\npgir}= \displaystyle
  \prod_{\altTime=1}^{\time-1}
  \Big[
    \frac{1}{\Npgir}\sum_{a=1}^{\Npgir}
    \hspace{1mm}
       \prod_{(\altUnit,\altTime)\in B^{[\altTime]}_{\unit,\time}} 
    \hspace{-1mm}
        w^M_{\altUnit,\altTime,\island,a}
  \Big] \prod_{(\altUnit,\time)\in B^{[\time]}_{\unit,\time}} 
    \hspace{-1mm}
        w^M_{\altUnit,\time,\island,\npgir}$
\\
End for
\\
$\displaystyle \MC{\loglik}_{\unit,\time}= 
\log\Bigg(
\frac{
\sum_{\island=1}^\Island \sum_{\npgir=1}^{\Npgir} w^M_{\unit,\time,\island,\npgir}w^P_{\unit,\time,\island,\npgir}
}{
\sum_{\island=1}^\Island \sum_{\npgir=1}^{\Npgir} w^P_{\unit,\time,\island,\npgir}
}
\Bigg)
$
\vspace{1mm}
\\
\hline
\end{tabular}
\end{center}

\section{\secTitleSpace A bootstrap guide function within GIRF: numerical results on the measles example}
<<bootgirf,echo=T,eval=F>>=
UU <- 40; nobs <- 30; Np <- 2000; lookahead <- 2; Nguide <- 10; tol <- 1e-300
set.seed(123321)
msls_girfd <- girfd_measles(U=UU, N = nobs, Np = Np, Nguide = Nguide, lookahead = lookahead)
system.time(
    bootgirf(msls_girfd, method='adams', Np=Np, Ninter = UU, lookahead = lookahead, Nguide = Nguide, tol = tol) -> bootgirf_out 
) -> bootgirf_time
system.time(
    bootgirf2(msls_girfd, method='adams', Np=Np, Ninter = UU, lookahead = lookahead, Nguide = Nguide, tol = tol) -> bootgirf2_out 
) -> bootgirf2_time
system.time(
    girf(msls_girfd, method='adams', Np=Np, Ninter = UU, lookahead = lookahead, Nguide = Nguide, tol = tol) -> girf_out 
) -> girf_time
@

The above code was run using the recently developed bootgirf() and bootgirf2() functions (today's date: sep 24 2020)
The bootgirf() function implemented the [ALT] approach described in \ref{sec:bootguide_abfir}.
The bootgirf2() function implemented the ``quantile-based'' approach described in \citet{park20}.
The bootgirf2() method is based on the assumption that the process noise per unit time is roughly constant over the guide simulation time interval.
However, the memory requirement of bootgirf2() can be substantially lower than that for bootgirf().
This is because bootgirf2() stores the residuals (i.e., guide simulations minus skeleton simulations) only at the lookahead observation time points, whereas bootgirf() stores the residuals at all intermediate time points in the guided simulation time interval.
The memory cost for bootgirf() can be reduced by storing the residual simulations only once in every few intermediate simulation time points.

In the above example, I ran for all forty UK cities, for 30 biweeks, with the number of particles = 2000, the number of guide simulations per particle = 10, the number of lookahead observations = 2.
All simulations were carried out on the greatlakes computing cluster at the University of Michigan.
The bootgirf2() run used 8gb of RAM, and the girf() run used 4gb of RAM.
The bootgirf() was run with 8gb of RAM, but was terminated due to insufficient memory.

\begin{verbatim}
Results for U=40
girf_time: 4875.384 (s),  logLik(girf) per unit per obs: -5.52
bootgirf2_time:  5134.528 (s),  logLik(girf_out) per unit per obs: -4.15
\end{verbatim}

The three methods were also run for the same measles model with $U=20$, and the number of particles = 1000.
All other settings were the same.
The results were as follows.

\begin{verbatim}
bootgirf_time:  744.541 (s) ,  logLik(bootgirf):  -4.36
bootgirf2_time:  539.719 (s) , logLik(bootgirf2): -4.41
girf_time:  523.253 (s) , logLik(girf): -5.37
\end{verbatim}

For both examples ($U=20$ and 40), the log likelihood estimates by bootgirf2() was comparable to that of the EnKf in Figure~3 of the main text.

%%%%% ggggggggggggggggggggggggggg

\section{\secTitleSpace A likelihood slice across recovery rate for the measles model}

<<gammaSlice_settings,echo=F,eval=T>>=

gammaSlice_files_dir <- paste0("gammaSlice_",gammaSlice_run_level,"/")
if(!dir.exists(gammaSlice_files_dir)) dir.create(gammaSlice_files_dir)

stew(file=paste0(gammaSlice_files_dir,"gammaSlice_settings.rda"),{

  # copy variables that should be included in the stew
  gammaSlice_run_level <- gammaSlice_run_level 
  gammaSlice_cores <- cores

  gamma_lo <- 40
  gamma_hi <- 64
  gamma_bpf_lo <- 30
  gamma_bpf_hi <- 75
#  gamma_bpf_lo <- gamma_lo
#  gamma_bpf_hi <- gamma_hi

  if(gammaSlice_run_level==1){
    gammaSlice_Ntheta <- 5
    gammaSlice_U <- 5
    gammaSlice_N <- 5
    gammaSlice_replicates <- 2
    gammaSlice_girf_Np <- 50
    gammaSlice_girf_lookahead <- 2
    gammaSlice_girf_nguide <- 10
    gammaSlice_asif_islands <- 3
    gammaSlice_asif_Np_per_island <- 10
    gammaSlice_asifir_islands <- 3
    gammaSlice_asifir_Np_per_island <- 10
    gammaSlice_bpf_units_per_block <- 1
    gammaSlice_bpf_Np <- 50
    gammaSlice_bif_islands <- 10
  }

  if(gammaSlice_run_level==2){
    gammaSlice_Ntheta <- 10
    gammaSlice_U <- 10
    gammaSlice_N <- 5*26
    gammaSlice_replicates <- 1 # number of Monte Carlo replicates
    gammaSlice_girf_Np <- 50
    gammaSlice_girf_lookahead <- 2
    gammaSlice_girf_nguide <- 20
    gammaSlice_asif_islands <- 20
    gammaSlice_asif_Np_per_island <- 50
    gammaSlice_asifir_islands <- 10
    gammaSlice_asifir_Np_per_island <- 100
    gammaSlice_bpf_units_per_block <- 1
    gammaSlice_bpf_Np <- 700
    gammaSlice_bif_islands <- 100
  }

  if(gammaSlice_run_level==3){
    gammaSlice_Ntheta <- 10
    gammaSlice_replicates <- 3 # number of Monte Carlo replicates
    gammaSlice_U <- 40
    gammaSlice_N <- 15*26
    gammaSlice_girf_lookahead <- 1
    gammaSlice_girf_Np <- 200*5
    gammaSlice_girf_nguide <- 50*2
    gammaSlice_asif_islands <- 1000
    gammaSlice_asif_Np_per_island <- 100
    gammaSlice_asifir_islands <- 50*5
    gammaSlice_asifir_Np_per_island <- 50*2
    gammaSlice_bpf_units_per_block <- 2
    gammaSlice_bpf_Np <- 40000
    gammaSlice_bif_islands <- 40000
  }

  if(gammaSlice_run_level==4){
    gammaSlice_Ntheta <- 9
    gammaSlice_replicates <- 3 # number of Monte Carlo replicates
    gammaSlice_U <- 40
    gammaSlice_N <- 15*26
    gammaSlice_girf_lookahead <- 1
    gammaSlice_girf_Np <- 500
    gammaSlice_girf_nguide <- 100
    gammaSlice_asif_islands <- 1000
    gammaSlice_asif_Np_per_island <- 200
    gammaSlice_asifir_islands <- 50*5
    gammaSlice_asifir_Np_per_island <- 50*2
    gammaSlice_bpf_units_per_block <- 1
    gammaSlice_bpf_Np <- 5000
    gammaSlice_bif_islands <- 50000    
  }

})
@

<<gammaSlice_spatPomp,eval=T,echo=F>>=

  gammaSlice_model_dir <- "../measlesModel/"
  # note: care is required if gammaSlice_model_dir is set to something other than the default measlesModel
  # however, it may be useful to do that while testing model variations.

  load(file=paste0(gammaSlice_model_dir,"measles_spatPomp.rda"))

gammaSlice_spatPomp <- stew(file=paste0(gammaSlice_files_dir,"gammaSlice_spatPomp.rda"),{

  gammaSlice_sim <- measles_subset(m_U=gammaSlice_U, m_N=gammaSlice_N)

  gammaSlice_design <- slice_design(
    center=coef(gammaSlice_sim),
    gamma=rep(
      seq(from=gamma_lo,to=gamma_hi,length=gammaSlice_Ntheta),
      each=gammaSlice_replicates)
  )

  gammaSlice_bpf_design <- slice_design(
    center=coef(gammaSlice_sim),
    gamma=rep(
      seq(from=gamma_bpf_lo,to=gamma_bpf_hi,length=gammaSlice_Ntheta),
      each=gammaSlice_replicates)
  )

})

@

As discussed in the main text, we expect parameters of within-city transmission that are shared between cities to be well identified.
Figure~\ref{fig:gamma_slice_asif_plot} shows a likelihood slice for the per-capita recovery rate, $\mu_{IR}$ (in units of $\mbox{ year}^{-1}$) on simulated data from the coupled model for $\Unit=40$ cities.
As expected, the likelihood has peak in the vicinity of the truth at $\mu_{IR}=\Sexpr{myround(measles_params["gamma"],1)}\mbox{ year}^{-1}$.
The rapid decrease in the log likelihood away from this maximum dwarfs the Monte Carlo error in computing this log likelihood.

<<gammaSlice_nbhd,echo=F>>=
gammaSlice_nbhd <- function(object, time, unit) {
  nbhd_list <- list()
  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
  return(nbhd_list)
}
@

<<gammaSlice_girf,eval=F,cache=F,echo=F>>=
theta_tmp <- NA
gammaSlice_girf_files <- paste0(gammaSlice_files_dir,"gammaSlice_girf",1:nrow(gammaSlice_design),".rda")
foreach (gammaSlice_row=1:nrow(gammaSlice_design),theta=iter(gammaSlice_design,"row"), .combine=rbind,
  .inorder=FALSE) %dopar% {
  theta_tmp <- unlist(theta[names(coef(gammaSlice_sim))])
  gammaSlice_girf <- stew(file=gammaSlice_girf_files[gammaSlice_row],seed=572817+gammaSlice_row,{
    system.time(
      girf(gammaSlice_sim,params=theta_tmp,  method='adams',
        Np=gammaSlice_girf_Np,
        Ninter = gammaSlice_U,
        lookahead = gammaSlice_girf_lookahead,
        Nguide = gammaSlice_girf_nguide, 
        tol = 1e-300
      ) -> girf1
    ) -> gammaSlice_girf_time
    girf_logLik <- logLik(girf1)
    rm(girf1)
  })
  c(logLik=girf_logLik,theta_tmp, Np=gammaSlice_girf_Np,
    Ninter = length(unit_names(gammaSlice_sim)), lookahead = gammaSlice_girf_lookahead,
    Nguide = gammaSlice_girf_nguide, U=length(unit_names(gammaSlice_sim)), N=length(time(gammaSlice_sim)),
    time=gammaSlice_girf_time["elapsed"]
  )
} -> gammaSlice_girf_results

@

<<gamma_DEBUGGING,eval=F,echo=F>>=
th <- unlist(gammaSlice_design[1,names(coef(gammaSlice_sim))])
      girf(gammaSlice_sim,params=th,
        Np=50,
        Ninter = 5, lookahead = 1,
        Nguide = 50,
        tol = 1e-300
      ) -> g1

x <- girfd_measles()

g2 <- girf(x)

@

<<gammaSlice_asif, eval=T, cache=F,echo=F>>=
gammaSlice_asif_files <- paste0(gammaSlice_files_dir,"gammaSlice_asif",1:nrow(gammaSlice_design),".rda")
# parallelization is carried out within asif so we use a serial foreach
foreach (gammaSlice_row=1:nrow(gammaSlice_design), .combine=rbind) %do% {
  gammaSlice_asif <- stew(file=gammaSlice_asif_files[gammaSlice_row],seed=4000817+gammaSlice_row,{
    theta <- unlist(gammaSlice_design[gammaSlice_row,names(coef(gammaSlice_sim))])
    system.time(
      asif(gammaSlice_sim,params=theta, 
        islands = gammaSlice_asif_islands,
        Np=gammaSlice_asif_Np_per_island,
        nbhd = gammaSlice_nbhd,tol=1e-300
      ) -> asif1
    ) -> gammaSlice_asif_time
    asif_logLik <- logLik(asif1)
    rm(asif1)
  })
  c(logLik=asif_logLik,theta, Np=gammaSlice_asif_Np_per_island,
    islands = gammaSlice_asif_islands, U=length(unit_names(gammaSlice_sim)), N=length(time(gammaSlice_sim)),
    time=gammaSlice_asif_time["elapsed"]
  )
} -> gammaSlice_asif_results    
@

<<gammaSlice_bif, eval=T, cache=F,echo=F>>=
gammaSlice_bif_files <- paste0(gammaSlice_files_dir,"gammaSlice_bif",1:nrow(gammaSlice_design),".rda")
# parallelization is carried out within asif so we use a serial foreach
foreach (gammaSlice_row=1:nrow(gammaSlice_design), .combine=rbind) %do% {
  gammaSlice_bif <- stew(file=gammaSlice_bif_files[gammaSlice_row],seed=4000817+gammaSlice_row,{
    theta <- unlist(gammaSlice_design[gammaSlice_row,names(coef(gammaSlice_sim))])
    system.time(
      asif(gammaSlice_sim,params=theta, 
        islands = gammaSlice_bif_islands,
        Np=1,
        nbhd = gammaSlice_nbhd,tol=1e-300
      ) -> bif1
    ) -> gammaSlice_bif_time
    bif_logLik <- logLik(bif1)
    rm(bif1)
  })
  c(logLik=bif_logLik,theta,
    islands = gammaSlice_bif_islands, U=length(unit_names(gammaSlice_sim)), N=length(time(gammaSlice_sim)),
    time=gammaSlice_bif_time["elapsed"]
  )
} -> gammaSlice_bif_results    
@

<<gammaSlice_bpf,cache=F,echo=F,eval=T>>=
gammaSlice_bpf_files <- paste0(gammaSlice_files_dir,"gammaSlice_bpf",1:nrow(gammaSlice_bpf_design),".rda")
foreach (gammaSlice_row=1:nrow(gammaSlice_bpf_design), .combine=rbind) %dopar% {
  theta_eval <<- unlist(gammaSlice_bpf_design[gammaSlice_row,names(coef(gammaSlice_sim))])  
  gammaSlice_bpf <- stew(file=gammaSlice_bpf_files[gammaSlice_row],seed=400818+gammaSlice_row,{
    theta <- theta_eval
    system.time(
      bpfilter(gammaSlice_sim,params=theta,
        Np=gammaSlice_bpf_Np,
	num_partitions=ceiling(gammaSlice_U/gammaSlice_bpf_units_per_block)
      ) -> gammaSlice_bpf_out
    ) -> gammaSlice_bpf_time
    bpf_logLik <- gammaSlice_bpf_out@loglik
    rm(gammaSlice_bpf_out)
  })
  c(logLik=bpf_logLik,theta, Np=gammaSlice_bpf_Np,
    units_per_block = gammaSlice_bpf_units_per_block,
    U=length(unit_names(gammaSlice_sim)), N=length(time(gammaSlice_sim)),
    time=gammaSlice_bpf_time["elapsed"]
  )
} -> gammaSlice_bpf_results    
@

<<gammaSlice_asif_plot, eval = T, echo=F,  fig.align='center', fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Likelihood slice of the recovery rate parameter, computed via {\\ASIF} on the measles model with $U=40$ cities. The true parameter value for the simulation is marked with a blue dashed line.')>>=

if(0){
gammaSlice_results <- as.data.frame(rbind(
  gammaSlice_asif_results[,c("logLik","gamma")],
  gammaSlice_asifir_results[,c("logLik","gamma")],
  gammaSlice_girf_results[,c("logLik","gamma")]
))
gammaSlice_results$Method <- rep( c("ABF", "ABF-IR", "GIRF"),each=nrow(gammaSlice_design))
} else {
  gammaSlice_results <-  as.data.frame(gammaSlice_asif_results[,c("logLik","gamma")])
  gammaSlice_results$Method <- rep("ABF",each=nrow(gammaSlice_design))
}

save(file=paste0(gammaSlice_files_dir,"gammaSlice_results.rda"),gammaSlice_asif_results,gammaSlice_results)

# ggplot(gammaSlice_results[gammaSlice_results$Method %in% c("ABF", "ABF-IR"),],mapping = aes(x = g, y = logLik)) +
# ggplot(gammaSlice_results,mapping = aes(x = gamma, y = logLik)) +
#   geom_point() +
#   labs(x = expression(gamma), y = "log likelihood") +
# #  scale_y_continuous(limits=c(-20,NA))
#  scale_y_continuous(limits=c(-66600,-66100)) +
#  scale_x_continuous(limits=c(25,40))

plot.profile2 <- function(mcap1,ylab,xline=2.5,yline=2.5,xlab="",quadratic=FALSE,...){
  if(missing(ylab)) ylab <- "profile log likelihood"
  # par(mai=c(0.7,0.7,0.3,0.3))
  ggplot() + geom_point(aes(x = mcap1$parameter, y = mcap1$lp)) + 
    geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$smoothed),
              color = "red") +
    {if(quadratic) geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$quadratic),
                             color = "blue",
                             lwd = 1.25)} +
    labs(x = xlab, y = ylab) +
    geom_vline(xintercept = mcap1$ci, color = "red") + 
    geom_hline(yintercept = max(mcap1$fit$smoothed, na.rm = T) - mcap1$delta, color = "red") +
    theme(panel.border = element_rect(colour = "black", fill=NA))
  }

gammaSlice_asif_mcap <- panelPomp::mcap(
  lp=gammaSlice_asif_results[,"logLik"],
  parameter=gammaSlice_asif_results[,"gamma"],
  lambda = 1
)

plot.profile2(gammaSlice_asif_mcap,xlab="recovery rate (/year)", quadratic = FALSE) +
  geom_vline(xintercept=measles_params["gamma"],color="blue",linetype="dashed")

@

<<gammaSlice_bpf_plot, eval = T, echo=F,  fig.align='center', fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Likelihood slice of the recovery rate parameter, computed via BPF on the measles model with $U=40$ cities. The true parameter value for the simulation is marked with a blue dashed line.')>>=

gammaSlice_bpf_mcap <- panelPomp::mcap(
  lp=gammaSlice_bpf_results[,"logLik"],
  parameter=gammaSlice_bpf_results[,"gamma"],
  lambda = 1
)

plot.profile2(gammaSlice_bpf_mcap,xlab="recovery rate (/year)", quadratic = FALSE) +
  geom_vline(xintercept=measles_params["gamma"],color="blue",linetype="dashed")

@

<<gammaSlice_bif_plot, eval = T, echo=F,  fig.align='center', fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Likelihood slice of the recovery rate parameter, computed via {\\TIF} on the measles model with $U=40$ cities. The true parameter value for the simulation is marked with a blue dashed line.')>>=

gammaSlice_bif_mcap <- panelPomp::mcap(
  lp=gammaSlice_bif_results[,"logLik"],
  parameter=gammaSlice_bif_results[,"gamma"],
  lambda = 1
)

plot.profile2(gammaSlice_bif_mcap,xlab="recovery rate (/year)", quadratic = FALSE) +
  geom_vline(xintercept=measles_params["gamma"],color="blue",linetype="dashed")

@

Fig~\ref{fig:gammaSlice_asif_plot} took $\Sexpr{myround(mean(gammaSlice_asif_results[,"time.elapsed"]/60),1)}$ minutes to evaluate each point using a parallelized implementation on a computing node with $\Sexpr{gl_cores}$ cores.
By contrast, Fig~\ref{fig:gammaSlice_bpf_plot} took $\Sexpr{myround(mean(gammaSlice_bpf_results[,"time.elapsed"]/60),1)}$ minutes, and Fig~\ref{fig:gammaSlice_bif_plot} took $\Sexpr{myround(mean(gammaSlice_bif_results[,"time.elapsed"]/60),1)}$ minutes.
No doubt, all the implementations could be sped up with addition attention to the code.

\clearpage

\section{\secTitleSpace
Slices over $R_0$}

<<slice_settings,echo=F,eval=T>>=

r0slice_files_dir <- paste0("r0slice_",r0slice_run_level,"/")
if(!dir.exists(r0slice_files_dir)) dir.create(r0slice_files_dir)

stew(file=paste0(r0slice_files_dir,"r0slice_settings.rda"),{

  # copy variables that should be included in the stew
  r0slice_run_level <- r0slice_run_level 
  r0slice_cores <- cores

  R0_lo <- 25
  R0_hi <- 40

#  R0_bpf_lo <- 50
#  R0_bpf_hi <- 750
  R0_bpf_lo <- R0_lo
  R0_bpf_hi <- R0_hi

  if(r0slice_run_level==1){
    r0slice_Ntheta <- 5
    r0slice_U <- 5
    r0slice_N <- 3
    r0slice_replicates <- 2
    r0slice_girf_Np <- 20
    r0slice_girf_lookahead <- 2
    r0slice_girf_nguide <- 10
    r0slice_asif_islands <- 3
    r0slice_asif_Np_per_island <- 10
    r0slice_asifir_islands <- 3
    r0slice_asifir_Np_per_island <- 10
    r0slice_bpf_units_per_block <- 1
    r0slice_bpf_Np <- 20
    r0slice_bif_islands <- 10
}

  if(r0slice_run_level==2){
    r0slice_Ntheta <- 10
    r0slice_U <- 10
    r0slice_N <- 5*26
    r0slice_replicates <- 2 # number of Monte Carlo replicates
    r0slice_girf_Np <- 100
    r0slice_girf_lookahead <- 2
    r0slice_girf_nguide <- 50
    r0slice_asif_islands <- 100
    r0slice_asif_Np_per_island <- 50
    r0slice_asifir_islands <- 50
    r0slice_asifir_Np_per_island <- 100
    r0slice_bpf_units_per_block <- 1
    r0slice_bpf_Np <- 5000
    r0slice_bif_islands <- 200
  }

  if(r0slice_run_level==3){
    r0slice_Ntheta <- 10
    r0slice_replicates <- 4 # number of Monte Carlo replicates
    r0slice_U <- 40
    r0slice_N <- 15*26
    r0slice_asif_islands <- 1000
    r0slice_asif_Np_per_island <- 100
    r0slice_asifir_islands <- 50*5
    r0slice_asifir_Np_per_island <- 50*2
    r0slice_bpf_units_per_block <- 2
    r0slice_bpf_Np <- 20000
    r0slice_bif_islands <- 40000
  }

  if(r0slice_run_level==4){
    r0slice_Ntheta <- 12
    r0slice_replicates <- 5 # number of Monte Carlo replicates
    r0slice_U <- 40
    r0slice_N <- 15*26
    r0slice_asif_islands <- 5000
    r0slice_asif_Np_per_island <- 100
    r0slice_asifir_islands <- 50*5
    r0slice_asifir_Np_per_island <- 50*2
    r0slice_bpf_units_per_block <- 2
    r0slice_bpf_Np <- 40000
    r0slice_bif_islands <- 40000
  }

})
@

<<r0slice_spatPomp,eval=T,echo=F>>=
r0slice_spatPomp <- stew(file=paste0(r0slice_files_dir,"r0slice_spatPomp.rda"),{

  r0slice_model_dir <-  "../measlesModel/"

  # note: care is required if r0slice_model_dir is set to something other than measles_model_dir
  # however, it may be useful to do that while testing model variations.

  load(file=paste0(r0slice_model_dir,"measles_spatPomp.rda"))
  r0slice_sim <- measles_subset(m_U=r0slice_U, m_N=r0slice_N)

  r0slice_sliceJobs <- slice_design(
    center=coef(r0slice_sim),
    R0=rep(seq(from=R0_lo,to=R0_hi,length=r0slice_Ntheta),each=r0slice_replicates)
  )

  r0slice_bpf_sliceJobs <- slice_design(
    center=coef(r0slice_sim),
    R0=rep(seq(from=R0_bpf_lo,to=R0_bpf_hi,length=r0slice_Ntheta),each=r0slice_replicates)
  )

})
@

<<r0slice_nbhd,echo=F>>=
r0slice_nbhd <- function(object, time, unit) {
  nbhd_list <- list()
  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
  return(nbhd_list)
}
@

<<r0slice_asif, eval=T, cache=F,echo=F>>=
r0slice_asif_files <- paste0(r0slice_files_dir,"r0slice_asif",1:nrow(r0slice_sliceJobs),".rda")
# parallelization is carried out within asif so we use a serial foreach
foreach (r0slice_row=1:nrow(r0slice_sliceJobs), .combine=rbind) %do% {
  r0slice_asif <- stew(file=r0slice_asif_files[r0slice_row],seed=4000817+r0slice_row,{
    theta <- unlist(r0slice_sliceJobs[r0slice_row,names(coef(r0slice_sim))])
    system.time(
      asif(r0slice_sim,params=theta, 
        islands = r0slice_asif_islands,
        Np=r0slice_asif_Np_per_island,
        nbhd = r0slice_nbhd,tol=1e-300
      ) -> asif1
    ) -> r0slice_asif_time
    asif_logLik <- logLik(asif1)
    rm(asif1)
  })
  c(logLik=asif_logLik,theta, Np=r0slice_asif_Np_per_island,
    islands = r0slice_asif_islands, U=length(unit_names(r0slice_sim)), N=length(time(r0slice_sim)),
    time=r0slice_asif_time["elapsed"]
  )
} -> r0slice_asif_results

@

<<r0slice_bif, eval=T, cache=F,echo=F>>=
r0slice_bif_files <- paste0(r0slice_files_dir,"r0slice_bif",1:nrow(r0slice_sliceJobs),".rda")
# parallelization is carried out within bif so we use a serial foreach
foreach (r0slice_row=1:nrow(r0slice_sliceJobs), .combine=rbind) %do% {
  r0slice_bif <- stew(file=r0slice_bif_files[r0slice_row],seed=4000817+r0slice_row,{
    theta <- unlist(r0slice_sliceJobs[r0slice_row,names(coef(r0slice_sim))])
    system.time(
      asif(r0slice_sim,params=theta, 
        islands = r0slice_bif_islands,
        Np=1,
        nbhd = r0slice_nbhd,tol=1e-300
      ) -> bif1
    ) -> r0slice_bif_time
    bif_logLik <- logLik(bif1)
    rm(bif1)
  })
  c(logLik=bif_logLik,theta,
    islands = r0slice_bif_islands, U=length(unit_names(r0slice_sim)), N=length(time(r0slice_sim)),
    time=r0slice_bif_time["elapsed"]
  )
} -> r0slice_bif_results

@

<<r0slice_asifir,eval=F,cache=F,echo=F>>=
r0slice_asifir_files <- paste0(r0slice_files_dir,"r0slice_asifir",1:nrow(r0slice_sliceJobs),".rda")
# parallelization is carried out within asifir so we use a serial foreach
foreach (r0slice_row=1:nrow(r0slice_sliceJobs), .combine=rbind) %do% {  
  r0slice_asifir <- stew(file=r0slice_asifir_files[r0slice_row],seed=4000817+r0slice_row,{
    theta <- unlist(r0slice_sliceJobs[r0slice_row,names(coef(r0slice_sim))])
    system.time(
      asifir(r0slice_sim,params=theta,  method='adams',
        islands = r0slice_asifir_islands,
        Np=r0slice_asifir_Np_per_island,
        nbhd = r0slice_nbhd,
        Ninter = as.integer(r0slice_U/2),
        tol = 1e-300
      ) -> asifir1
    ) -> r0slice_asifir_time
    asifir_logLik <- logLik(asifir1)
    rm(asifir1)
  })    
  c(logLik=asifir_logLik,theta, Np=r0slice_asifir_Np_per_island,
      islands = r0slice_asifir_islands, U=length(unit_names(r0slice_sim)), N=length(time(r0slice_sim)),
      time=r0slice_asifir_time["elapsed"]
  )
} -> r0slice_asifir_results 
@

<<r0slice_bpf,cache=F,echo=F,eval=T>>=

r0slice_bpf_files <- paste0(r0slice_files_dir,"r0slice_bpf",1:nrow(r0slice_bpf_sliceJobs),".rda")

r0slice_bpf_file2 <- paste0(r0slice_files_dir,"r0slice_bpf_new.rda")

stew(file=r0slice_bpf_file2,seed=4000817,{
 foreach (r0slice_row=1:nrow(r0slice_bpf_sliceJobs), .combine=rbind) %dopar% {
  theta <- unlist(r0slice_bpf_sliceJobs[r0slice_row,names(coef(r0slice_sim))])
    system.time(
      bpfilter(r0slice_sim,params=theta,
        Np=r0slice_bpf_Np,
	num_partitions=ceiling(r0slice_U/r0slice_bpf_units_per_block)
      ) -> r0slice_bpf_out
    ) -> r0slice_bpf_time
    bpf_logLik <- r0slice_bpf_out@loglik
    rm(r0slice_bpf_out)
  c(logLik=bpf_logLik,theta, Np=r0slice_bpf_Np,
    units_per_block = r0slice_bpf_units_per_block,
    U=length(unit_names(r0slice_sim)), N=length(time(r0slice_sim)),
    time=r0slice_bpf_time["elapsed"]
  )
 } -> r0slice_bpf_results
})
@

<<r0slice_asif_plot, eval = T, echo=F,  fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Likelihood slice varying the coupling parameter, computed via {\\ASIF} with $U=40$ cities. The solid lines construct a 95\\% Monte Carlo adjusted confidence interval \\citep{ionides17}. For this model realization, this interval includes the true parameter value identified by a blue dashed line.')>>=

#ASIF_ONLY <- FALSE
ASIF_ONLY <- TRUE
if(!ASIF_ONLY){
r0slice_results <- as.data.frame(rbind(
  r0slice_asif_results[,c("logLik","R0")],
  r0slice_bpf_results[,c("logLik","R0")]
))
r0slice_results$Method <- rep( c("ASIF", "BPF"),each=nrow(r0slice_sliceJobs))
} else {
  r0slice_results <-  as.data.frame(r0slice_asif_results[,c("logLik","R0")])
  r0slice_results$Method <- rep("ASIF",each=nrow(r0slice_sliceJobs))
}

# save(file=paste0(r0slice_files_dir,"r0slice_results.rda"),r0slice_asif_results,r0slice_results)

plot.profile <- function(mcap1,ylab,xline=2.5,yline=2.5,xlab="",quadratic=FALSE,...){
  #  if(missing(ylab)) ylab <- expression("\u2113"^"P")
  if(missing(ylab)) ylab <- "log likelihood"
  par(mai=c(0.7,0.7,0.3,0.3))
  plot(mcap1$lp ~ mcap1$parameter, xlab="",ylab="",...)
  mtext(side=2,text=ylab,line=yline)
  mtext(side=1,text=xlab,line=yline)
  lines(mcap1$fit$parameter,mcap1$fit$smoothed, col = "red", lwd = 1.5)
  abline(v=mcap1$ci,col="red")
  abline(h=max(mcap1$fit$smoothed,na.rm=T)-mcap1$delta,col="red")
  if(quadratic) lines(mcap1$fit$parameter,mcap1$fit$quadratic, col = "blue", lwd = 1.5,
                      lty="dotted")
  }

plot.profile2 <- function(mcap1,ylab,xline=2.5,yline=2.5,xlab="",quadratic=FALSE,...){
  if(missing(ylab)) ylab <- "log likelihood"
  # par(mai=c(0.7,0.7,0.3,0.3))
  ggplot() + geom_point(aes(x = mcap1$parameter, y = mcap1$lp)) + 
    geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$smoothed),
              color = "red") +
    {if(quadratic) geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$quadratic),
                             color = "blue",
                             lwd = 1.25)} +
    labs(x = xlab, y = ylab) +
    geom_vline(xintercept = mcap1$ci, color = "red") + 
    geom_hline(yintercept = max(mcap1$fit$smoothed, na.rm = T) - mcap1$delta, color = "red") +
    theme(panel.border = element_rect(colour = "black", fill=NA))
  }

r0slice_asif_mcap <- panelPomp::mcap(
  lp=r0slice_asif_results[,"logLik"],
  parameter=r0slice_asif_results[,"R0"],
  lambda = 1
)

plot.profile2(r0slice_asif_mcap,xlab="R0", quadratic = FALSE) +
  geom_vline(xintercept=measles_params["R0"],color="blue",linetype="dashed")

@

<<r0slice_bpf_plot, eval = T, echo=F,  fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Likelihood slice varying the coupling parameter, computed via {BPF} with $U=40$ cities. The solid lines construct a 95\\% Monte Carlo adjusted confidence interval \\citep{ionides17}. For this model realization, this interval includes the true parameter value identified by a blue dashed line.')>>=

r0slice_bpf_mcap <- panelPomp::mcap(
  lp=r0slice_bpf_results[,"logLik"],
  parameter=r0slice_bpf_results[,"R0"],
  lambda = 1
)

plot.profile2(r0slice_bpf_mcap,xlab="R0", quadratic = FALSE) +
  geom_vline(xintercept=measles_params["R0"],color="blue",linetype="dashed")

@

<<r0slice_bif_plot, eval = T, echo=F,  fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Likelihood slice varying the coupling parameter, computed via {\\TIF} with $U=40$ cities. The solid lines construct a 95\\% Monte Carlo adjusted confidence interval \\citep{ionides17}. For this model realization, this interval includes the true parameter value identified by a blue dashed line.')>>=

r0slice_bif_mcap <- panelPomp::mcap(
  lp=r0slice_bif_results[,"logLik"],
  parameter=r0slice_bif_results[,"R0"],
  lambda = 0.8
)

plot.profile2(r0slice_bif_mcap,xlab="R0", quadratic = FALSE) +
  geom_vline(xintercept=measles_params["R0"],color="blue",linetype="dashed")

@

\clearpage

\section{Comments on assumptions}

\begin{enumerate}
\item The same $\epsilon$ is used in various different bounds. In principle, these could all be different bounds. However, it is conceptually simpler to have just a single small quantity. Also, the proof shows that these errors all have additive consequences for the overall error, so they may as well be on comparable scales.

\item Recall the following assumptions:

\Ai

\setcounter{assumption}{3}

\Aiv

\Bi

\setcounter{assumptionB}{3}

\BivA

The specification of $B^{2c}_{\unit\comma\time}$ in Assumptions~\ref{A:unconditional:mix} and~\ref{BivA} does not have to be built using the same sets $B^{}_{\unit\comma\time}$ from Assumptions~\ref{A1} and~\ref{B1}. It may be simpler to imagine that one set of neighborhoods carries out both functions, and if different neighborhoods are specified then the union of these neighborhoods provides one set satisfying both conditions.

\end{enumerate}

\section{Deriving Assumption~\ref{B:girf} from GIRF results}

\setcounter{assumptionB}{4}

Recall the statement:

\Bv

\medskip

%Recall also the statement of Theorem~2 of \citet{park20}:
\noindent We check this is derived from the following result.

\medskip

\noindent{\bf Theorem~3} \citep{park20}.
{\it 
Suppose multinomial resampling is used in GIRF, and that regularity conditions specify $C_1>0$, $C_2>0$, $0<\rho<1$.
We don't need the exact conditions to check that the conclusion implies Assumption~B5.
If $h$ is a measurable function such that $\lVert h \rVert_\infty \leq 1$ and $a>1$ is an arbitrary constant, then we have
\begin{equation}
  \left\lvert \frac{1}{J} \sum_{j=1}^J h(X_{t_N}^{F,j}) - \mathbb{E} \big[ h(X_{t_N}) | Y_{1:N}=y_{1:N} \big] \right\rvert
  \leq \frac{ 4 a C_2(C_1+1)}{\rho\sqrt{J}} (NS+1)
  \label{extension:eqn:mainbound}
\end{equation}
with probability at least 
\begin{equation}
\label{extension:prob:bound}
1-\frac{(2NS+1)(NS+1)}{a^2},
\end{equation}
given that $\sqrt J \geq 8 \rho^{-2} a C_2 (C_1+1) NS$.
}

In our context, $N=1$ since we are carrying out guided sampling over one timestep. 
Now, set 
\begin{equation}
\frac{(2NS+1)(NS+1)}{a^2} < \epsilon/2,
\end{equation}
say, 
\begin{equation}
\label{eq:extension:1}
a = \sqrt{\frac{2(2S+1)(S+1)}{\epsilon}}.
\end{equation}
We want  $J$ large enough that
\begin{equation}
\label{eq:extension:2}
\frac{ 4 a C_2(C_1+1)}{\rho\sqrt{J}} (NS+1) < \epsilon/2,
\end{equation}
since then the bounds in \myeqref{extension:eqn:mainbound} and \myeqref{extension:prob:bound}, together with the bound on $h$, give the bound in expectation in Assumption~\ref{B:girf}.
Putting \myeqref{eq:extension:1} into \myeqref{eq:extension:2} gives
\begin{equation}
\label{eq:extension:3}
\frac{ 4 C_2(C_1+1)\sqrt{2(2S+1)(S+1)}}{\rho\sqrt{\epsilon J}} (S+1) < \epsilon/2.
\end{equation}
and so
\begin{equation}
\label{eq:extension:4}
J > \frac{128(2S+1)(S+1)^3C^2_2(C_1+1)^2}{\epsilon^3 \rho^2}
\end{equation}
which implies that we can pick
\begin{equation}
C_0 =  128(2S+1)(S+1)^3 C^2_2(C_1+1)^2 \rho^{-2}
\end{equation}
in Assumption~\ref{B:girf}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% iiiiiiiiiiiiiii

\section{Some adapted simulation identities}

The following proposition derives some results relating the adapted process to the filtering, prediction and likelihood evaluation tasks.
The first two results, \myeqref{eq:simpler:identity:1} and  \myeqref{eq:simpler:identity:2}, are simpler identities for temporal dependence ignoring sequential calculations over space.
They are provided primarily to give simpler analogs to the subsequent results.
\begin{prop}

Let $g(\vec{x})=f_{\vec{X}_0} \left[ \prod_{\time=1}^\Time f_{\vec{X}_{\time}|\vec{X}_{\time-1},\vec{Y}_{\time}}\right]$ be the joint density of $\vec{X}=\vec{X}_{0:\Time}$ for the adapted process, and let $\adapted_\unitTimeSubset(\vec{x})= \prod_{\altTime=1}^\Time f_{Y_{\unitTimeSubset^{[\altTime]}}|\vec{X}^{}_{\altTime-1}}$ be the proper adapted weight on $\unitTimeSubset$. Then,
\begin{eqnarray}
\label{eq:simpler:identity:1}
f_{\vec{X}_{\time},\vec{Y}_{1:\time}} &=& \int f_{\vec{X}_0} 
\left[
  \prod_{\altAltTime=1}^{\time}
  f_{\vec{X}_{\altAltTime}|\vec{X}_{\altAltTime-1},\vec{Y_{\altAltTime}}} \,
  f_{\vec{Y}_{\altAltTime}|\vec{X}_{\altAltTime-1}}
\right] d\vec{x}_{0:\time-1}
\\
\nonumber
f_{\vec{X}_{\time+1}|\vec{Y}_{1:\time}} &=& 
\frac{
\int f_{\vec{X}_0} 
\left[
  \prod_{\altAltTime=1}^{\time}
  f_{\vec{X}_{\altAltTime}|\vec{X}_{\altAltTime-1},\vec{Y_{\altAltTime}}} \,
  f_{\vec{Y}_{\altAltTime}|\vec{X}_{\altAltTime-1}}
\right] f_{\vec{X}_{\time+1}|\vec{X}_{\time}}\, d\vec{x}_{0:\time}
}{
\int f_{\vec{X}_0} 
\left[
  \prod_{\altAltTime=1}^{\time}
  f_{\vec{X}_{\altAltTime}|\vec{X}_{\altAltTime-1},\vec{Y_{\altAltTime}}} \,
  f_{\vec{Y}_{\altAltTime}|\vec{X}_{\altAltTime-1}}
\right] d\vec{x}_{0:\time}
}
\\
\label{eq:simpler:identity:2}
&=& \frac{
  \E_g\big[\adapted^{}_{\UnitSet\times\{1:\time\}}(\vec{X})\,f_{\vec{X}_{\time+1}|\vec{X}_{\time}}(\vec{x}_{\time+1}\given \vec{X}_{\time}) \big]
  }{
  \E_g\big[\adapted^{}_{\UnitSet\times\{1:\time\}}(\vec{X})\big]
  }
\\
\label{eq:adapted2}
f_{\vec{X}^{}_{\time},Y_{A_{\unit,\time}}} &=& \int f_{\vec{X}_0} \left[
  \prod_{\altAltTime=1}^{\time}
  f_{\vec{X}_{\altAltTime}|\vec{X}_{\altAltTime-1},\vec{Y_{\altAltTime}}} \,
  f_{Y_{A^{[\altAltTime]}_{\unit,\time}}|\vec{X}_{\altAltTime-1}}
\right] d\vec{x}_{0:\time-1}
\\
\nonumber
f_{Y_{A_{\unit,\time}}} &=& \int f_{\vec{X}_0} \left[
  \prod_{\altAltTime=1}^{\time}
  f_{\vec{X}_{\altAltTime}|\vec{X}_{\altAltTime-1},\vec{Y_{\altAltTime}}} \,
  f_{Y_{A^{[\altAltTime]}_{\unit,\time}}|\vec{X}_{\altAltTime-1}}
\right] d\vec{x}_{0:\time}
\\
\label{eq:adapted}
&=&
\E_g\big[ \adapted^{}_{A_{\unit,\time}}(\vec{X}) \big]
\\
\nonumber
f_{{X}^{}_{\unit:\Unit,\time}|Y^{}_{A_{\unit,\time}}} &=& \frac{
  \int f_{\vec{X}_0} \left[
    \prod_{\altAltTime=1}^{\time}
    f_{\vec{X}_{\altAltTime}|\vec{X}_{\altAltTime-1},\vec{Y_{\altAltTime}}} \,
    f_{Y_{A^{[\altAltTime]}_{\unit,\time}}|\vec{X}_{\altAltTime-1}}
  \right] dx^{}_{A_{\unit,\time}}
}{
  \int f_{\vec{X}_0} \left[
    \prod_{\altAltTime=1}^{\time}
    f_{\vec{X}_{\altAltTime}|\vec{X}_{\altAltTime-1},\vec{Y_{\altAltTime}}} \,
    f_{Y_{A^{[\altAltTime]}_{\unit,\time}}|\vec{X}_{\altAltTime-1}}
  \right] d\vec{x}_{0:\time}
}
\\
&=&
\label{eq:adapted3}
\frac{
  \E_g\big[\adapted^{}_{A_{\unit,\time}}(\vec{X}) \, f_{X^{}_{\unit:\Unit,\time}|X_{A_{\unit,\time}}}(x^{}_{\unit:\Unit,\time}|X_{A_{\unit,\time}}) \big]
  }{
  \E_g\big[\adapted^{}_{A_{\unit,\time}}(\vec{X})\big]
  }
\end{eqnarray}
\end{prop}
\begin{proof} Suppose inductively that \myeqref{eq:simpler:identity:1} holds for some $n$. 
Using the Markov property,
\begin{equation}
\label{eq:simpler:identity:1.1}
f_{\vec{Y}_{1:\time+1},\vec{X}_{\time+1}}=\int 
f_{\vec{Y}_{1:\time},\vec{X}_{\time}} \, f_{\vec{Y}_{\time+1},\vec{X}_{\time+1}|\vec{X}_{\time}} d\vec{x}_{\time}
\end{equation}
Substituting \myeqref{eq:simpler:identity:1} into \myeqref{eq:simpler:identity:1.1} and checking the case $n=1$ shows that \myeqref{eq:simpler:identity:1} holds also for $n+1$, and therefore \myeqref{eq:simpler:identity:1} holds for all $n\ge 1$ by induction.
The first equality in \myeqref{eq:simpler:identity:2} follows from \myeqref{eq:simpler:identity:1} by standard manipulation of condition densities.
The second equality in \myeqref{eq:simpler:identity:2} employs the definition 
$\adapted^{}_{\UnitSet\times\{1:\time\}}=\adapted^{}_{\UnitSet\times\{1:\time\}}(\vec{X}_{0:\Time})=
  \prod_{\altAltTime=1}^{\time} f_{\vec{Y}_{\altAltTime}|\vec{X}_{\altAltTime-1}}$.
\myeqref{eq:adapted2} is analogous to \myeqref{eq:simpler:identity:1}.
Integrating \myeqref{eq:adapted2} over $\vec{x}_{\time}$ gives the joint likelihood expression in \myeqref{eq:adapted}.
Integrating only over $x_{1:\unit-1,\time}$ and conditioning gives \myeqref{eq:adapted3}.
\end{proof}

\section{Variations on local weights}

These are some ideas that were written up before we settled on {\ABF}/{\ABFIR} as the main focus for this manuscript.

\subsection{Moving from weak weights to proper local weights}

Consider a bivariate situation where we draw $(X^{(1)}_j,Y^{(1)}_j)\sim f^{(1)}_{XY}(x,y)$ and resample with weights $g_X(x)g_Y(y)$ to obtain $(X^{(2)}_j,Y^{(2)}_j)$ targeting 
\begin{equation}
 f^{(2)}_{XY}(x,y) = \frac{ g_X(x)g_Y(y)f^{(1)}_{XY}(x,y)}{\int g_X(\altx)g_Y(\alty)f^{(1)}_{XY}(\altx,\alty)\, d\altx d\alty}.
\end{equation}
Now suppose we want to use  $\{ (X^{(2)}_j,Y^{(2)}_j), j\in 1\mycolon J\}$ to obtain a sample from the marginal $X$-distribution of
\begin{equation}
 f^{(3)}_{XY}(x,y) = \frac{ h_X(x)h_Y(y)f^{(1)}_{XY}(x,y)}{\int h_X(\altx)h_Y(\alty)f^{(1)}_{XY}(\altx,\alty)\, d\altx d\alty}.
\end{equation}
It appears we should resample $X^{(2)}_{1:J}$ with weights 
$$
W_j = \frac{h^{}_X(X^{(2)}_j)}{g^{}_X(X^{(2)}_j)}.
$$
We should do an analogous calculation in the IIF algorithm.
Note that the local reweighting (here, local to the $X$-component) is, in fact, a proper reweighting even though we ignore the $Y$-component. 
This follows, despite the dependence in $f_{XY}$ because of the product structure of the weights $g_X(x)g_Y(y)$ and $h_X(x)h_Y(y)$.

\subsection{Using the weakly weighted fixed lag smoothing distribution as a proposal}

For the time being, we suppose we are working with a single generic island $\island$, and we suppress the identity of this island from the notation. 
Write $B_{\unit,\time}^+ = B_{\unit,\time}\cup (\unit,\time)$ and $B_{\unit,\time}^- = B_{\unit,\time}\cap \{(\altUnit,\altTime): \altTime<\time\}$.
We need a proposal for $X_{B_{\unit,\time}^+}$ that can be locally reweighted at $(\unit,\time)$ to solve the proper local spatiotemporal filtering problem. 
Let a weakly smoothed particle approximation $X^{WS}_{[\unit,\time],\altUnit,\altTime,1:\Np}$ be defined as follows, for $(\altUnit,\altTime)\in {B_{\unit,\time}^+}$.
\begin{equation}
X^{WS}_{[\unit,\time],\altUnit,\altTime,\np}
=
\left\{
\begin{array}{ll}
X^{WF}_{\altUnit,\altTime,\langle\time,\altTime,\np\rangle}
& \mbox{for $\altTime < \time$} \\
X^{WP}_{\altUnit,\altTime,\np}
& \mbox{for $\altTime = \time$} 
\end{array}
\right.
\end{equation}
%Then, for $h: \Xspace^{B_{\unit,\time}^+}\to\R$, and a random variable $X$ taking values in $\Xspace^{B_{\unit,\time}^+}$, suppose
%\begin{equation}\label{eq:eh}
%\E[h(X)]\approx\frac{1}{\Np}\sum_{\np=1}^{\Np} 
%h(X^{WS}_{[\unit,\time],B_{\unit,\time}^+,\np}).
%\end{equation}
%This corresponds to $X$ having density 
This collection of particles, by construction, approximates a weakly weighted smoothed density,
\begin{equation}
f_{X_{B_{\unit,\time}^+}|\vec{Y}_{1:\time-1}}(x_{B_{\unit,\time}^+}\given\data{\vec{y}}_{1:\time-1}\param \sigma_{1:\Unit}\Omega_{1:\Unit}).
\end{equation}
We can appeal to the structure of the relaxed measurement process to give
\begin{eqnarray}
&&\hspace{-2cm}f_{X_{B_{\unit,\time}^-}|\vec{Y}_{B_{\unit,\time}^-}}(x_{B_{\unit,\time}^-}\given\data{y}_{B_{\unit,\time}^-}\param \sigma_{\altUnit})
\\
&\propto& f_{X_{B_{\unit,\time}^-}|Y_{1:\time-1}}(x_{B_{\unit,\time}^-}\given\data{y}_{1:\time-1}\param \sigma_{\altUnit}\Omega_{\altUnit})
\prod_{(\altUnit,\altTime)\in B_{\unit,\time}^-}
\frac{
f_{Y_{\altUnit,\altTime}|X_{\altUnit,\altTime}}(\data{y}_{\altUnit,\altTime}|x_{\altUnit,\altTime}\param \sigma_{1:\Unit})
}{
f_{Y_{\altUnit,\altTime}|X_{\altUnit,\altTime}}(\data{y}_{\altUnit,\altTime}|x_{\altUnit,\altTime}\param \sigma_{1:\Unit}\Omega_{1:\Unit})
}.
\end{eqnarray}
Now, $X^{WS}_{[\unit,\time],B^+_{\unit,\time},\np}$, considered as a particle taking values in $\Xspace^{B_{\unit,\time}^+}$ and intended to represent the target density $f_{X_{B^+_{\unit,\time}}|Y_{B_{\unit,\time}}}$, has proper weight given by
\begin{equation} \label{eq:extensionPredWeights}
w^P_{\unit,\time,\np}=
\prod_{(\altUnit,\altTime)\in B_{\unit,\time}^-}
\frac{
f_{Y_{\altUnit,\altTime}|X_{\altUnit,\altTime}}
\big(
\data{y}_{\altUnit,\altTime}|X^{WS}_{[\unit,\time],\altUnit,\altTime,\np}\param \sigma_{\unit}
\big)
}{
f_{Y_{\altUnit,\altTime}|X_{\altUnit,\altTime}}
\big(
\data{y}_{\altUnit,\altTime}|X^{WS}_{[\unit,\time],\altUnit,\altTime,\np}\param \sigma_{\unit}\Omega_{\unit}
\big)
}
\prod_{(\altUnit,\altTime)\in B_{\unit,\time}^\sharp}
f_{Y_{\altUnit,\altTime}|X_{\altUnit,\altTime}}
\big(
\data{y}_{\altUnit,\altTime}|X^{WS}_{[\unit,\time],\altUnit,\altTime,\np}\param \sigma_{\unit}
\big).
\end{equation}
The $(\unit,\time)$ marginal of these particles therefore represents
\begin{equation}
f_{X_{\unit,\time}|Y_{B_{\unit,\time}}}(x_{\unit,\time}\given \data{y}_{B_{\unit,\time}} \param \sigma_{1:\Unit})
\approx
f_{X_{\unit,\time}|Y_{A_{\unit,\time}}}(x_{\unit,\time}\given \data{y}_{A_{\unit,\time}} \param \sigma_{1:\Unit}).
\end{equation}
Therefore,
\begin{equation}\label{eq:condLikSum}
%\frac{1}{\Np} 
\sum_{\np=1}^{\Np}
\left\{
\frac{w^P_{\unit,\time,\np}}{\sum_{\altNp=1}^{\Np}w^P_{\unit,\time,\altNp}}
\right\}
f_{Y_{\unit,\time}|X_{\unit,\time}}(\data{y}_{\unit,\time}\given X^{WS}_{[\unit,\time],\unit,\time,\np} \param \sigma_\unit)
\approx
f_{Y_{\unit,\time}|Y_{A_{\unit,\time}}}(\data{y}_{\unit,\time}\given \data{y}_{A_{\unit,\time}} \param \sigma_{1:\Unit}).
\end{equation}
It might seem from \eqref{eq:condLikSum} that we need to build particles solving a different fixed lag smoothing problem at each spatiotemporal location $(\unit,\time)$. 
However, we see in the IIF algorithm that these particles are constructed as weighted collections of a single collection of weakly filtered particles and their ancestors defined over all space and time.
A simple computational approach is to carry out the entire weak filtering, saving all the particles and the identity of their parent, and then carry out the proper local weighting computations.
In principle, one can carry out the weak filtering and reweighting synchronously, enabling memory saving by discarding weakly filtered particles once they no longer show up in any current or future spatiotemporal neighborhood $B_{\unit,\time}$.
Doing this requires additional attention to communication in a multi-processor implementation of IIF.
So long as storage is not prohibitive (say, 1000 particles at 100 locations for 1000 time points, each with an 8-byte double precision real value, requiring 800Mb storage) it is tempting to treat the weak filtering and reweighting as separate tasks, solved sequentially.

The importance sampling in \eqref{eq:extensionPredWeights} occurs in dimension $\big| B_{\unit,\time}\big|$. 
If this dimension is large, the weights may be numerically stable. 
The curse of dimensionality is addressed since this dimension is assumed not to depend on $\Unit$ or $\Time$.

There is a tradeoff as the weakly filtered distribution approaches the target distribution, the weights in \eqref{eq:extensionPredWeights} become more balanced but the weak filtering itself becomes more problematic.

\subsection{Weak weights versus other weak filters}

The core idea of the island filter is that a collection of weaker filters, each of which is numerically feasible despite the dimensionality, can be combined in a neighborhood of each spatiotemporal location.

An alternative approach to weak filtering is to carry out local filtering, where each island focuses on making one draw from the conditional distribution of $X_{1:\Unit,\time}$ given $X_{1:\Unit,\time-1}$ and $Y_{1:\Unit,\time}$.
A particle filter using $f_{X_{1:\Unit,\time}|Y_{1:\Unit,\time},X_{1:\Unit,\time-1}}$ as its proposal is said to be {\it fully adapted}, with some intuition that this might be the ideal (usually inaccessible) proposal choice given information available at time $\time$, even though it is not necessarily better than a basic unadapted proposal \citep{johansen08}.
Our weak adapted filter (WAP) will circumvent some issues for the COD by using the collection of particles only to make a sample from $f_{X_{1:\Unit,\time}|Y_{1:\Unit,\time},X_{1:\Unit,\time-1}}$ rather than representing the whole distribution.
This is an easier problem since it requires only comparison of nearby particles, which will necessarily have similar weights (assuming the measurement density is uniformly continuous) and will therefore be less susceptible to the unbalanced weights causing the COD for importance sampling.
To  obtain the full filtering distribution by reweighting these draws, the proper weight is $f_{Y_{1:\Unit,\time}|X_{1:\Unit,\time-1}}$ which could itself be hard to compute for large $\Unit$. 
However, in the context of IIF, we need to compute these weights only locally.

In practice, the WAP can be combined with the weak weights.
An advantage of weak weights is that one can see how, at least in principle, allowing islands to ignore all but a fixed number of observations can lead to algorithms formally beating the COD.

It may be difficult to see how a one-good-particle GIRF can have local properties beating the COD. 
However, a filter such as GIRF with some good properties should be able to be used with weak weights to improve scaling.

%Let's write down the algorithm to make discussion more concrete.

A continuous time approximation may be helpful. 
In continuous time, supposing that massive amounts of information never show up instantaneously, the curse of too much information can be avoided. 
This enables importance sampling to give Monte Carlo simulation of $f_{\vec{X}(t+\delta)|\vec{Y}(t+\delta),\vec{X}(t)}$ and evaluation of $f_{\vec{Y}(t+\delta)|\vec{X}(t)}$ for sufficiently small $\delta$.
However, we may need small $\delta$, and hence a large number of importance sampling steps, as $\Unit$ increases. 
In general, to avoid COD, one expects $\delta\propto 1/\Unit$, as studied in the context of GIRF by \citet{park20}.
This produces a problem for SMC if we look for a  proper global reweighting of the islands.
One then expects the Monte Carlo variance of the log weights for each island to grow linearly with $1/\Unit$.
Put another way, the Monte Carlo variance in the importance sampling estimate of the likelihood should grow linearly with the information available.
We are saved by the local reweighting. 
Specifically, let $1\mycolon\Time$ be the discretization of the continuous system, so $t_\time=\time\delta$.
Then, we aim to sample from $f_{X_{\unit,\time}|Y_{B_{\unit,\time}}}$ (i.e., the localized version of the filtering problem) by taking the $(\unit,\time)$ component from an adapted filter proposal, combined with an unadapted prediction step (we can do this, since the prediction step is constructed by the filter particles before they are weighted and resampled).
%We compute the required weights conditional on $X_{B_{\unit,\time}^c}$ since, by assumption, these values have negligible impact on $X_{\unit,\time}$ condition on $Y_{B_{\unit,\time}}$.
We write $B_{\unit,\time}^{[m]}=\{(\altUnit,\altTime)\in B_{\unit,\time} : \altTime=m\}$. 
    Then, our proposal density for $X^+_{B_{\unit,\time}}$ is the marginal distribution on $\Xspace^+_{B_{\unit,\time}}$ of a joint density on $\Xspace^+_{A_{\unit,\time}}$ given by
\begin{equation} \label{eq:conditionalProposal}
%f_{X_{B^{[\time]}_{\time,\unit}}|\vec{Y}_{B^{[\time]}_{\time,\unit}},\vec{X}_{\time-1}}
%f_{X_{B^{[\time]}_{\time,\unit}}|\vec{X}_{\time-1}}(x_{B^{[\time]}_{\time,\unit}}\given \vec{x}_{\time-1})
f_{X_{\time,1:\unit}|\vec{X}_{\time-1}}(x_{\time,1:\unit}\given \vec{x}_{\time-1})
\prod_{\altTime=1}^{\time-1} f_{\vec{X}_{\altTime}|\vec{Y}_{\altTime},\vec{X}_{\altTime-1}}(\vec{x}_{\altTime}\given \data{\vec{y}}_{\altTime},\vec{x}_{\altTime-1}),
\end{equation}
Note that the proposal for $X_{\time,\unit}$ doesn't condition on the data $\data{y}_{\time,1:\unit-1}$ at time $\time$ but preceding in the spatial ording. 
This differs from a regular particle filter applied to the spatiotemporal ordering.
The density of a proposal for $X^+_{B_{\unit,\time}}$ from \eqref{eq:conditionalProposal} involves integrating out the proposal density for $X_{B_{\unit,\time}^c}$.
However, since we are interested in the distribution at $(\unit,\time)$, by the weak coupling assumption we incur only a small error by instead setting $x_{B_{\unit,\time}^c}$ to any arbitrary value.
It is natural to choose for this purpose the value of $X_{B_{\unit,\time}^c}$ for the filter particles, a choice which is convenient and approximates the omitted integral.
The resulting weight to target a draw from the filter density, $f_{X_{\unit,\time}|Y_{A_{\unit,\time}}}(x_{\unit,\time}\given \data{y}_{A_{\unit,\time}})$, via a draw $X^P_{A_{\unit,\time}}$ from \eqref{eq:conditionalProposal} is
\begin{eqnarray}
&& \hspace{-10mm}
%\prod_{\altUnit=1}^{\unit-1} f_{Y_{\altUnit,\time}|X_{\altUnit,\time}}
%\big(
%\data{y}_{\altUnit,\time}\given X^P_{\altUnit,\time}
%\big)
\frac{\displaystyle
f_{X_{\time,1:\unit},Y_{\time,1:\unit}|\vec{X}_{\time-1}}(x_{\time,1:\unit},\data{y}_{\time,1:\unit}\given \vec{x}_{\time-1})
\prod_{m=1}^{\time-1} 
f_{X_{B^{[m]}_{\unit,\time}},Y_{B^{[m]}_{\unit,\time}}|\vec{X}_{m-1}}
\Big(
X^P_{B^{[m]}_{\unit,\time}},\data{y}_{B^{[m]}_{\unit,\time}}\, \Big\vert \, \vec{X}^P_{m-1}
\Big)
}{\displaystyle
f_{X_{\time,1:\unit}|\vec{X}_{\time-1}}(x_{\time,1:\unit}\given \vec{x}_{\time-1})
\prod_{m=1}^{\time-1} 
f_{X_{B^{[m]}_{\unit,\time}}|Y_{B^{[m]}_{\unit,\time}},\vec{X}_{m-1}}
\Big(
X^P_{B^{[m]}_{\unit,\time}} \, \Big\vert\,  \data{y}_{B^{[m]}_{\unit,\time}},\vec{X}^P_{m-1}
\Big)
}
\\
&& \hspace{15mm} = f_{Y_{1:\unit-1,\time}|X_{1:\unit-1,\time}}(\data{y}_{1:\unit-1,\time}\given X^P_{1:\unit-1,\time})
\prod_{m=1}^{\time-1} f_{Y_{B^{[m]}_{\unit,\time}}|X_{B^{[m]}_{\unit,\time}}}
\Big(
\data{y}_{B^{[m]}_{\unit,\time}} \, \Big\vert \, X^P_{B^{[m]}_{\unit,\time}}
\Big)
\end{eqnarray}

%SMC is numerically stable only when the inevitable Monte Carlo errors have consequences that shrink (due to stability of the filtered system) faster than they build up. 
%If errors at $t$ have negligible consequences at $t+M$ then $M$ is called the mixing timescale.
%Here, however, we anticipate $M\delta^{-1}\propto M\Unit$ resampling events, and therefore growing amounts of Monte Carlo error within the mixing time (supposing that the mixing time is fixed as $\Unit$ grows).

\section{Comments on the guide function}

The plug-and-play guide function of \citet{park20} involves an additive moment-based approximation which is suitable if the process and measurement models are close to Gaussian. 
Here, we explore this assumption in the case of the measles model.

The guide function uses an approximation to $\var(Y_n|X_{n-1,s}=x_{n-1,s})$ where $X_{n-1,s}=X\big(t_{n-1,s})$ is the latent process at intermediate time $t_{n-1}=t_{n-1,0} <t_{n-1,s}<t_{n-1,S}=t_n$ for $1\le s<S$. 
Then, the predictive likelihood at $t_n$ given $X_{n-1,s}=x_{n-1,s}$ is approximated using the measurement model with center at an approximation $\hat{x_n}\approx \E\big[ Y_n\given X_{n-1,s}=x_{n-1,s} \big]$ and variance approximated by 
\begin{equation}
\var[ Y_n\given X_{n}=\hat{x_{n}} \big] + \var\big[ h(X_n)\given X_{n-1,s}=x_{n-1,s} \big]
\label{eq:guide:1}
\end{equation}
where $h(x)=\E[Y_n|X_n=x]$. We use \myeqref{eq:guide:1} in place of 
\begin{equation}
\var[ Y_n\given X_{n-1,s}=x_{n-1,s} \big] 
\label{eq:guide:2}
\end{equation}
since \myeqref{eq:guide:2} is hard to calculate.

For a toy analog to measles, consider the situation where $X=e^U$ and $Y=Xe^V = e^{U+V}$ where $U\sim\normal(\sigma^2)$ and $V\sim\normal(\tau^2)$. 
We have rescaled cases and reports for this analogy, but that is unimportant here. 
The approximation \myeqref{eq:guide:1} becomes
\begin{equation}
\var[ Y_n\given X_{n}=\hat{x_{n}} \big] + \var\big[ h(X_n)\given X_{n-1,s}=x_{n-1,s} \big] = \big(e^{2\sigma^2}-e^{\sigma^2}\big) + \big(e^{2\tau^2}-e^{\tau^2}\big),
\label{eq:guide:3}
\end{equation}
with the exact calculation being
\begin{equation}
\var[ Y_n\given X_{n-1,s}=x_{n-1,s} \big] = \big(e^{2(\sigma^2+\tau^2)}-e^{(\sigma^2+\tau^2)}\big).
\label{eq:guide:4}
\end{equation}
The ratio of \myeqref{eq:guide:3} to \myeqref{eq:guide:4} in the case $\sigma=\tau$ is
\begin{equation}
\frac{2}{e^{2\sigma^2}+e^{\sigma^2}}.
\end{equation}
This approaches 1 for $\sigma$ small, since in this case the measurement model is essentially additive. However, the ratio rapidly decays as $\sigma$ heads past 1.
\begin{verbatim}
> f <- function(sigma) 2/(exp(sigma^2)+exp(2*sigma^2))
> sigma <- (0:10)/10
> cbind(sigma,f=f(sigma))
      sigma   f
 [1,]   0.0 1.0000000
 [2,]   0.1 0.9850996
 [3,]   0.2 0.9415762
 [4,]   0.3 0.8728320
 [5,]   0.4 0.7841173
 [6,]   0.5 0.6819546
 [7,]   0.6 0.5734335
 [8,]   0.7 0.4654657
 [9,]   0.8 0.3640918
[10,]   0.9 0.2739351
[11,]   1.0 0.1978760
\end{verbatim}
In the measles context, rescaling to match this toy example, we consider 
$\tau^2=\psi^2=0.116$ and $\sigma^2=\sigma_{\mathrm{SE}}^2\times 14 = 0.0878^2 \times 14 = 0.108$. This gives a ratio of

% sigma <- sqrt(0.108) ; tau <- 0.116
% ( exp(2*sigma^2)+exp(2*tau^2)-exp(sigma^2)-exp(tau^2) ) / 
%  ( exp(2*(sigma^2+tau^2)) -  exp(sigma^2+tau^2) )

\begin{verbatim}
> sigma <- sqrt(0.108) ; tau <- 0.116
>  ( exp(2*sigma^2)+exp(2*tau^2)-exp(sigma^2)-exp(tau^2) ) / 
+   ( exp(2*(sigma^2+tau^2)) -  exp(sigma^2+tau^2) )
[1] 0.9654933
>
\end{verbatim}

%%%%%%% I vs J %%%%% asifn %%%%% ijijijijijijijij

\clearpage

\section{The islands vs particle per island trade-off for measles}

<<asifn_settings,cache=FALSE,echo=F>>=

#asifn_files_dir <- "bak/asifn-1dec19/"
asifn_files_dir <- "asifn/"

stew(file=paste0(asifn_files_dir,"asifn_settings.rda"),{

  asifn_tol <- 1e-300
  asifn_cores <- cores
  asifn_nbhd <- function(object, time, unit) {
      nbhd_list <- list()
      if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
      if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
      return(nbhd_list)
  }

  # run_level 1 for debugging; 2 for quick run; 3 for long run; 4 for production(?)
  asifn_run_level <- 2

  if(asifn_run_level==1){ #######
    asifn_U <- 2
    asifn_N <- 5
    asifn_replicates <- 2 
    asifn1_Np <- 20
    asifn1_I <- 5
    asifn2_Np <- 10
    asifn2_I <- 10
    asifn3_Np <- 5
    asifn3_I <- 20
    asifn4_Np <- 4
    asifn4_I <- 25
  } else if(asifn_run_level==2){ #######
    asifn_U <- 40
    asifn_N <- 104
    asifn_replicates <- 5
    asifn1_Np <- 2000
    asifn1_I <- 500
#    asifn1_name <- as.character(asifn1_Np)
    asifn1_name <- asifn1_Np
    asifn2_Np <- 1000
    asifn2_I <- 1000
#    asifn2_name <- as.character(asifn2_Np)
    asifn2_name <- asifn2_Np
    asifn3_Np <- 500
    asifn3_I <- 2000
#    asifn3_name <- as.character(asifn3_Np)
    asifn3_name <- asifn3_Np
    asifn4_Np <- 250
    asifn4_I <- 4000
#    asifn4_name <- as.character(asifn4_Np)
    asifn4_name <- asifn4_Np
    asifn5_Np <- 100
    asifn5_I <- 10000
#    asifn5_name <- as.character(asifn5_Np)
    asifn5_name <- asifn5_Np
    asifn6_Np <- 50
    asifn6_I <- 20000
#    asifn6_name <- as.character(asifn6_Np)
    asifn6_name <- asifn6_Np
    asifn7_Np <- 25
    asifn7_I <- 40000
#    asifn7_name <- as.character(asifn7_Np)
    asifn7_name <- asifn7_Np
    asifn8_Np <- 10
    asifn8_I <- 40000
#    asifn8_name <- as.character(asifn8_Np)
    asifn8_name <- asifn8_Np
    asifn9_Np <- 5
    asifn9_I <- 40000
#    asifn9_name <- as.character(asifn9_Np)
    asifn9_name <- asifn9_Np
    asifn10_Np <- 1
    asifn10_I <- 40000
#    asifn10_name <- as.character(asifn10_Np)
    asifn10_name <- asifn10_Np
    asifn11_Np <- 2
    asifn11_I <- 40000
#    asifn11_name <- as.character(asifn11_Np)
    asifn11_name <- asifn11_Np
  } else if(asifn_run_level==3){
  } else if(asifn_run_level==4){
  } else if(asifn_run_level==5){
  }

})

asifn_jobs <- data.frame(reps=1:asifn_replicates,U=asifn_U,N=asifn_N)

@

<<asifn_spatPomp,eval=T,echo=F>>=
asifn_spatPomp <- stew(file=paste0(asifn_files_dir,"asifn_spatPomp.rda"),{
  asifn_model_dir <- "../measlesModel/"
  # note: care is required if asifn_model_dir is set to something other than
  # the default measlesModel.
  # However, it may be useful to do that while testing model variations.  
  load(file=paste0(asifn_model_dir,"measles_spatPomp.rda"))
  asifn_measles <- measles_subset(m_U=asifn_U, m_N=asifn_N)
})

@

<<asifn_nbhd,echo=F>>=
asifn_nbhd <- function(object, time, unit) {
  nbhd_list <- list()
  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
  return(nbhd_list)
}
@

<<asifn1,cache=F,echo=F>>=
asifn1 <- stew(file=paste0(asifn_files_dir,"asifn1.rda"),seed=844424,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles, 
        islands = asifn1_I,
        Np=asifn1_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn1_out 
    ) -> asifn1_time
#    uncomment for debugging 
#    list(logLik=logLik(asifn1_out),time=asifn1_time["elapsed"],asifn1_out)
    list(logLik=logLik(asifn1_out),time=asifn1_time["elapsed"])
  } -> asifn1_list
})
asifn1_jobs <- asifn_jobs
asifn1_jobs$logLik <- vapply(asifn1_list,function(x)x$logLik,numeric(1))
asifn1_jobs$time <- vapply(asifn1_list,function(x) x$time,numeric(1))
asifn1_jobs$Method <- asifn1_name
@

<<asifn2,cache=F,echo=F>>=
asifn2 <- stew(file=paste0(asifn_files_dir,"asifn2.rda"),seed=389855,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles, 
        islands = asifn2_I,
        Np=asifn2_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn2_out 
    ) -> asifn2_time
#    uncomment for debugging 
#    list(logLik=logLik(asifn2_out),time=asifn2_time["elapsed"],asifn2_out)
    list(logLik=logLik(asifn2_out),time=asifn2_time["elapsed"])
  } -> asifn2_list
})
asifn2_jobs <- asifn_jobs
asifn2_jobs$logLik <- vapply(asifn2_list,function(x)x$logLik,numeric(1))
asifn2_jobs$time <- vapply(asifn2_list,function(x) x$time,numeric(1))
asifn2_jobs$Method <-  asifn2_name
@

<<asifn3,cache=F,echo=F>>=
asifn3 <- stew(file=paste0(asifn_files_dir,"asifn3.rda"),seed=186711,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles,
        islands = asifn3_I,
        Np=asifn3_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn3_out
    ) -> asifn3_time
#    uncomment for debugging
#    list(logLik=logLik(asifn3_out),time=asifn3_time["elapsed"],asifn3_out)
    list(logLik=logLik(asifn3_out),time=asifn3_time["elapsed"])
  } -> asifn3_list
})
asifn3_jobs <- asifn_jobs
asifn3_jobs$logLik <- vapply(asifn3_list,function(x)x$logLik,numeric(1))
asifn3_jobs$time <- vapply(asifn3_list,function(x) x$time,numeric(1))
asifn3_jobs$Method <-  asifn3_name
@

<<asifn4,cache=F,echo=F>>=
asifn4 <- stew(file=paste0(asifn_files_dir,"asifn4.rda"),seed=186711,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles,
        islands = asifn4_I,
        Np=asifn4_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn4_out
    ) -> asifn4_time
#    uncomment for debugging
#    list(logLik=logLik(asifn4_out),time=asifn4_time["elapsed"],asifn4_out)
    list(logLik=logLik(asifn4_out),time=asifn4_time["elapsed"])
  } -> asifn4_list
})
asifn4_jobs <- asifn_jobs
asifn4_jobs$logLik <- vapply(asifn4_list,function(x)x$logLik,numeric(1))
asifn4_jobs$time <- vapply(asifn4_list,function(x) x$time,numeric(1))
asifn4_jobs$Method <-  asifn4_name
@

<<asifn5,cache=F,echo=F>>=
asifn5 <- stew(file=paste0(asifn_files_dir,"asifn5.rda"),seed=186711,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles,
        islands = asifn5_I,
        Np=asifn5_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn5_out
    ) -> asifn5_time
#    uncomment for debugging
#    list(logLik=logLik(asifn5_out),time=asifn5_time["elapsed"],asifn5_out)
    list(logLik=logLik(asifn5_out),time=asifn5_time["elapsed"])
  } -> asifn5_list
})
asifn5_jobs <- asifn_jobs
asifn5_jobs$logLik <- vapply(asifn5_list,function(x)x$logLik,numeric(1))
asifn5_jobs$time <- vapply(asifn5_list,function(x) x$time,numeric(1))
asifn5_jobs$Method <- asifn5_name
@

<<asifn6,cache=F,echo=F>>=
asifn6 <- stew(file=paste0(asifn_files_dir,"asifn6.rda"),seed=186711,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles,
        islands = asifn6_I,
        Np=asifn6_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn6_out
    ) -> asifn6_time
#    uncomment for debugging
#    list(logLik=logLik(asifn6_out),time=asifn6_time["elapsed"],asifn6_out)
    list(logLik=logLik(asifn6_out),time=asifn6_time["elapsed"])
  } -> asifn6_list
})
asifn6_jobs <- asifn_jobs
asifn6_jobs$logLik <- vapply(asifn6_list,function(x)x$logLik,numeric(1))
asifn6_jobs$time <- vapply(asifn6_list,function(x) x$time,numeric(1))
asifn6_jobs$Method <- asifn6_name
@

<<asifn7,cache=F,echo=F>>=
asifn7 <- stew(file=paste0(asifn_files_dir,"asifn7.rda"),seed=186711,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles,
        islands = asifn7_I,
        Np=asifn7_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn7_out
    ) -> asifn7_time
#    uncomment for debugging
#    list(logLik=logLik(asifn7_out),time=asifn7_time["elapsed"],asifn7_out)
    list(logLik=logLik(asifn7_out),time=asifn7_time["elapsed"])
  } -> asifn7_list
})
asifn7_jobs <- asifn_jobs
asifn7_jobs$logLik <- vapply(asifn7_list,function(x)x$logLik,numeric(1))
asifn7_jobs$time <- vapply(asifn7_list,function(x) x$time,numeric(1))
asifn7_jobs$Method <- asifn7_name
@

<<asifn8,cache=F,echo=F>>=
asifn8 <- stew(file=paste0(asifn_files_dir,"asifn8.rda"),seed=186711,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles,
        islands = asifn8_I,
        Np=asifn8_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn8_out
    ) -> asifn8_time
#    uncomment for debugging
#    list(logLik=logLik(asifn8_out),time=asifn8_time["elapsed"],asifn8_out)
    list(logLik=logLik(asifn8_out),time=asifn8_time["elapsed"])
  } -> asifn8_list
})
asifn8_jobs <- asifn_jobs
asifn8_jobs$logLik <- vapply(asifn8_list,function(x)x$logLik,numeric(1))
asifn8_jobs$time <- vapply(asifn8_list,function(x) x$time,numeric(1))
asifn8_jobs$Method <- asifn8_name
@

<<asifn9,cache=F,echo=F>>=
asifn9 <- stew(file=paste0(asifn_files_dir,"asifn9.rda"),seed=186711,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles,
        islands = asifn9_I,
        Np=asifn9_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn9_out
    ) -> asifn9_time
#    uncomment for debugging
#    list(logLik=logLik(asifn9_out),time=asifn9_time["elapsed"],asifn9_out)
    list(logLik=logLik(asifn9_out),time=asifn9_time["elapsed"])
  } -> asifn9_list
})
asifn9_jobs <- asifn_jobs
asifn9_jobs$logLik <- vapply(asifn9_list,function(x)x$logLik,numeric(1))
asifn9_jobs$time <- vapply(asifn9_list,function(x) x$time,numeric(1))
asifn9_jobs$Method <- asifn9_name
@

<<asifn10,cache=F,echo=F>>=
asifn10 <- stew(file=paste0(asifn_files_dir,"asifn10.rda"),seed=186711,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles,
        islands = asifn10_I,
        Np=asifn10_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn10_out
    ) -> asifn10_time
#    uncomment for debugging
#    list(logLik=logLik(asifn10_out),time=asifn10_time["elapsed"],asifn10_out)
    list(logLik=logLik(asifn10_out),time=asifn10_time["elapsed"])
  } -> asifn10_list
})
asifn10_jobs <- asifn_jobs
asifn10_jobs$logLik <- vapply(asifn10_list,function(x)x$logLik,numeric(1))
asifn10_jobs$time <- vapply(asifn10_list,function(x) x$time,numeric(1))
asifn10_jobs$Method <- asifn10_name
@

<<asifn11,cache=F,echo=F>>=
asifn11 <- stew(file=paste0(asifn_files_dir,"asifn11.rda"),seed=186711,{
  foreach(job=iter(asifn_jobs,"row")) %do% {
    system.time(
      asif(asifn_measles,
        islands = asifn11_I,
        Np=asifn11_Np,
        nbhd = asifn_nbhd, tol=asifn_tol) -> asifn11_out
    ) -> asifn11_time
#    uncomment for debugging
#    list(logLik=logLik(asifn11_out),time=asifn11_time["elapsed"],asifn11_out)
    list(logLik=logLik(asifn11_out),time=asifn11_time["elapsed"])
  } -> asifn11_list
})
asifn11_jobs <- asifn_jobs
asifn11_jobs$logLik <- vapply(asifn11_list,function(x)x$logLik,numeric(1))
asifn11_jobs$time <- vapply(asifn11_list,function(x) x$time,numeric(1))
asifn11_jobs$Method <- asifn11_name
@

<<asifn_loglik_plot, echo=F, fig.height=5, fig.width=4, out.width="3.5in", fig.cap = paste('log likelihood estimates for simulated data from the measles model using {\\ABF}, with varying algorithmic parameters.')>>=

asifn_results <- rbind(asifn1_jobs,asifn2_jobs,asifn3_jobs,asifn4_jobs,asifn5_jobs,asifn6_jobs,asifn7_jobs,asifn8_jobs,asifn9_jobs,asifn10_jobs,asifn11_jobs)
#asifn_results <- rbind(asifn1_jobs,asifn2_jobs,asifn3_jobs)
asifn_results$logLik_per_unit <- asifn_results$logLik/asifn_results$U
asifn_results$logLik_per_obs <- asifn_results$logLik_per_unit/asifn_N

save(file=paste0(asifn_files_dir,"asifn_results.rda"),asifn_results)

asifn_max <- max(asifn_results$logLik_per_obs)

#ggplot(asifn_results,mapping = aes(x = Method, y = logLik_per_obs)) +
#  geom_point() +
#  coord_cartesian(ylim=c(asifn_max-2,asifn_max))+
#  ylab("log likelihood per observation")
# qplot(Method,logLik,data=asifn_results,geom="boxplot")
qplot(Method,logLik_per_obs,data=asifn_results,log="x",
  ylab="log likelihood per unit per time",
  xlab="Number of particles per island")+stat_summary(fun=mean, geom="line")

#ggplot(bm_results,mapping = aes(x = U, y = logLik_per_obs, group=Method,color=Method)) +
#  geom_point() +
#  stat_summary(fun=mean, geom="line") +
#  coord_cartesian(ylim=c(bm_max-0.35,bm_max))+
#  ylab("log likelihood per unit per time")

@

We investigated the tradeoff between the number of islands, $\Island$, and the number of particles per island, $\Np$, for the measles model.
We filtered simulated data for $U=\Sexpr{asifn_U}$ and $N=\Sexpr{asifn_N}$, with $\Sexpr{asifn_replicates}$ replications.
The results are shown in Figure~\ref{fig:asifn_loglik_plot}.
The total time taken was $\Sexpr{myround(sum(asifn1_jobs$time)/60,1)}$ mins for {\ABF}1, $\Sexpr{myround(sum(asifn2_jobs$time)/60,1)}$ mins for {\ABF}2,  $\Sexpr{myround(sum(asifn3_jobs$time)/60,1)}$ mins for {\ABF}3, $\Sexpr{myround(sum(asifn4_jobs$time)/60,1)}$ mins for {\ABF}4, $\Sexpr{myround(sum(asifn5_jobs$time)/60,1)}$ mins for {\ABF}5, $\Sexpr{myround(sum(asifn6_jobs$time)/60,1)}$ mins for {\ABF}6.
The algorithmic settings were as follows:

\begin{tabular}{ccc}
  $\Np$ & $\Island$ & time\\
\hline
\Sexpr{asifn1_Np} & \Sexpr{asifn1_I}  & \Sexpr{myround(sum(asifn1_jobs$time)/60,1)} \\
\Sexpr{asifn2_Np} & \Sexpr{asifn2_I}  & \Sexpr{myround(sum(asifn2_jobs$time)/60,1)} \\
\Sexpr{asifn3_Np} & \Sexpr{asifn3_I}  & \Sexpr{myround(sum(asifn3_jobs$time)/60,1)} \\
\Sexpr{asifn4_Np} & \Sexpr{asifn4_I}  & \Sexpr{myround(sum(asifn4_jobs$time)/60,1)} \\
\Sexpr{asifn5_Np} & \Sexpr{asifn5_I}  & \Sexpr{myround(sum(asifn5_jobs$time)/60,1)} \\
\Sexpr{asifn6_Np} & \Sexpr{asifn6_I}  & \Sexpr{myround(sum(asifn6_jobs$time)/60,1)} \\
\Sexpr{asifn7_Np} & \Sexpr{asifn7_I}  & \Sexpr{myround(sum(asifn7_jobs$time)/60,1)} \\
\Sexpr{asifn8_Np} & \Sexpr{asifn8_I}  & \Sexpr{myround(sum(asifn8_jobs$time)/60,1)} \\
\Sexpr{asifn9_Np} & \Sexpr{asifn9_I}  & \Sexpr{myround(sum(asifn9_jobs$time)/60,1)} \\
\Sexpr{asifn11_Np} & \Sexpr{asifn11_I}  & \Sexpr{myround(sum(asifn11_jobs$time)/60,1)} \\
\Sexpr{asifn10_Np} & \Sexpr{asifn10_I}  & \Sexpr{myround(sum(asifn10_jobs$time)/60,1)} 
\end{tabular}

\clearpage

%%%%%%%%%%%%%%   lz2lz2 %%%%%%%%%%%%%%%%%%%%%5

\section{Varying parameters for the Lorenz example}
\label{sec:vary_params}

This is to assess the hypothesis that scalable particle filter methods may have larger advantages over EnKF when the nonlinear dynamics are not enveloped in relatively large amounts of noise.
First, we tried reducing the process noise to $\sigma_p=0.25$.
Reducing process noise while maintaining measurement noise appears to  makes it harder for guiding and adapted simulation to operate.
However, the relatively large amount of Gaussian measurement noise fits the assumptions of EnKF.

The Jan 2020 version of the manuscript used $\sigma_p=0.25$ but we later reverted to $\sigma_p=1$.

The simulated data are plotted in Figure~\ref{fig:lz2_image_plot}, and the algorithmic parameters with corresponding run times are listed in Table~\ref{tab:lz2}.

The numerical results here were not kept updated, and have been deleted.

%%%%%%%%%%%%%%   lz4lz4 %%%%%%%%%%%%%%%%%%%%%5

\section{Choosing the observation interval}

This is to assess the hypothesis that scalable particle filter methods may have larger advantages over EnKF when the time between observations is larger. the nonlinear dynamics are not enveloped in relatively large amounts of noise.
We set the process noise to $\sigma_p=1$, the measurement noise to $\tau=1$ and observation times $t_n=n$ instead of $t_n=n/2$ as was the case previously.

The numerical results here were not kept updated with spatPomp changes, and have been deleted.
The hypothesis seemed to be refuted.
The system is already highly nonliner on the scale $t_n=n/2$ and quite challenging for particle methods.
Making the problem harder can favor EnKF since its failure mode is much softer.

\clearpage

%%%%%%%%%%%%%%   lz3lz3 %%%%%%%%%%%%%%%%%%%%%5

\section{Varying intermediate steps for {\ABFIR} on Lorenz}

This is to investigate the effect of changing the number of intermediate steps, $\Ninter$, in ABF-IR for the Lorenz '96 model.

The numerical results here were not kept updated with spatPomp changes, and have been deleted.

Parameter values were $\sigma_p = 0.25$ and $\tau=0.25$.
We found improvement as $S$ increased, with diminishing returns. It looked like $S=U/2$ was sufficient - for example, $S=5$ was as good as $S=10$ when $U=10$.

Changing to $\tau = 1$, keeping $\sigma_p=0.25$, gave dramatically different results. The results favored EnKF, and the improvement with increasing $S$ was small, in which case ABF might be preferred to ABF-IR.
An interpretation is that the reduced measurement noise is both problematic for an unguided method and helpful for the guided method.
The combination of small $\sigma_p$ and high $\tau$ is problematic for particle methods in a highly nonlinear, chaotic system: particles that find themselves in the wrong place cannot catch up by favorable choice of process noise.

\clearpage

%%%% mnmnmnmnmnmnmnmnmn

\section{\secTitleSpace
  Measles scaling with reduced noise}

The simpler {\ABF} method performed better than the more complex {\ABFIR} and GIRF algorithms on the measles example of Section~\ref{sec:mscale}.
This may be because the coupling is weak, or because the guide functions used by {\ABFIR} and GIRF are not guiding well.
To investigate further, we repeated the experiment with reduced process noise, setting $\sigma_{SE}=0.02$.
We hypothesized that reducing the process noise would improve the predictive capability of the deterministic skeleton $\mu^P_{\unit,\time,\ninter,\ell,\np}$ and therefore improve the guide function.

The numerical results here were not kept updated with spatPomp changes, and have been deleted.
Due to changes in the measles model, it is not clear if the previous results have much value.
We found that, in this example, {\ABF} and {\ABFIR} both performed apparently successfully and equally.

\clearpage

\section{\secTitleSpace
Bagged local filters}

In this section, we present a method that is related to, but different from, {\UBF} or {\ABF}.
The method runs a collection of independent particle filters, each of which is based on a spatially local sequence of observations.
Each of the independent islands, $i\in \mathcal I$, is a particle filter using the joint propagation kernel,
\[
f_{\mathbf X_n | \mathbf X_{n-1}},
\]
and the local observation density for a certain spatial unit $\mathfrak u(i)$ in $1:U$,
\[
f_{Y_{\mathfrak u(i), n}| X_{\mathfrak u(i),n}} (y_{\mathfrak u(i), n} | \cdot).
\]
The propagated and filter particles at time $n$ and location $u$ in the $i$-th island will be denoted by $X_{u,n,i,j}^P$ and $X_{u,n,i,j}^F$.
We write an estimate of $f_{Y_{\mathfrak u(i), n}|Y_{\mathfrak u(i), 1:n-1}}$ by the $i$-th island as
\[
\hat \ell_{n,i} := \frac{1}{J} \sum_{j=1}^J f_{Y_{\mathfrak u(i), n} | X_{\mathfrak u(i), n}} (y_{\mathfrak u(i), n} | X^P_{\mathfrak u(i), n, i, j}).
\]
We use the particles in $\mathcal I$ islands to estimate the likelihood $f_{Y_{1:U,1:N}}$ as follows.
The approach is similar to that of {\ABF} or {\ABFIR}, in that each particle is weighted by the prediction weight, and a weighted sum of measurement weights gives an estimate of $f_{Y_{u,n} | Y_{A_{u,n}}}$.
The prediction weight $w_{u,n,i,j}^P$ is given by
\[
w_{u,n,i,j}^P = \prod_{(\tilde u, \tilde n) \in B(u,n)} w^{u,n}_{\tilde u, \tilde n, i, j}
\]
where
\begin{equation}
w^{u,n}_{\tilde u, \tilde n, i, j} = \left\{
\begin{array}{ll}
  \hat \ell_{\tilde n, i} & \text{ if } \mathfrak u(i) = \tilde u\\
  f_{Y_{\tilde u,\tilde n}|X_{\tilde u, \tilde n}}(y_{\tilde u, \tilde n}| X^P_{\tilde u, \tilde n, i, a^{u,n}_{\tilde n} (j;i)}) & \text{ if } \mathfrak u(i) \neq \tilde u. 
\end{array}
\right.
\label{eqn:wun}
\end{equation}
Here, $a^{u,n}_{\tilde n}(j;i)$ is defined such that
\begin{equation}
X^P_{\tilde n, i, a^{u,n}_{\tilde n}(j;i)} \text{ is the ancestor particle of }
\left\{
\begin{array}{ll}
  X^F_{n,i,j} & \text{ if }\mathfrak u(i)<u\\
  X^P_{n,i,j} & \text{ if }\mathfrak u(i) \geq u,
\end{array}
\right.
\label{eqn:aun}
\end{equation}
where a particle is considered its own ancestor when $n=\tilde n$.
We define $a^{u,n}_{\tilde n}(j;i)$ as in \eqref{eqn:aun} because the prediction weights when estimating $f_{Y_{u,n}|Y_{A_{u,n}}}$ cannot use particles that already contain the information about $Y_{\mathfrak u(i),n}$ with $\mathfrak u(i) \geq u$.
According to a standard notation for ancestor particles, the ancestor particle of $X^F_{n,i,j}$ in the $i$-th island can be written as
\[
X^F_{n',i,a^n_{n'}(j;i)}.% \text{ is the ancestor particle of } X^F_{n,i,j}.
\]
If $n=n'$, we let $a^n_{n'}(j;i) = j$.
Using this notation, we can express $a^{u,n}_{\tilde n}(j;i)$ as
\[
a^{u,n}_{\tilde n}(j;i) = \left\{
\begin{array}{ll}
  a^n_{\tilde n-1}(j; i) & \text{ if } \mathfrak u(i) < u \\
  a^{n-1}_{\tilde n-1}(j; i) & \text{ if } \mathfrak u(i) \geq u.
\end{array}
\right.
\]
The measurement weights are given by
\[
w^M_{u,n,i,j} = \left\{
\begin{array}{ll}
  f_{Y_{u,n} | X_{u,n}} (y_{u,n} | X^F_{u,n,i,j}) & \text{ if } \mathfrak u(i) < u\\
  f_{Y_{u,n} | X_{u,n}} (y_{u,n} | X^P_{u,n,i,j}) & \text{ if } \mathfrak u(i) \geq u.
\end{array}
\right.
\]
The Monte Carlo estimate of $f_{Y_{u,n}| Y_{A_{u,n}}}$ is given by, as in {\ABF}(-IR),
\[
\MC{\ell_{u,n}} = \log \left( \frac{\sum_{i=1}^\mathcal I \sum_{j=1}^\mathcal J w_{u,n,i,j}^M w_{u,n,i,j}^P}{\sum_{i=1}^\mathcal I \sum_{j=1}^\mathcal J w_{u,n,i,j}^P} \right).
\]
The pseudocode for this method (which is tentatively called a bagged local filter) is given below.

\begin{center}
\noindent\begin{tabular}{l}
\hline
{\bf {BLF}. Bagged Local Filter}
\vspace{0.4mm} \\
\hline
\firstLineSpace
For each $i\in 1:\mathcal I$, run the bootstrap particle filter with the joint propagation kernel $f_{\mathbf X_n| \mathbf X_{n-1}}$\\ and local measurement density $f_{Y_{\mathfrak u(i),n} | X_{\mathfrak u(i),n}}(y_{\mathfrak u(i), n} | \cdot)$ for some spatial unit $\mathfrak u(i)$.\\
For $n$ in $1:N$\\
\asp $w_{u,n,i,j}^P = \prod_{(\tilde u, \tilde n) \in B(u,n)} w^{u,n}_{\tilde u, \tilde n, i, j}$ where $w^{u,n}_{\tilde u, \tilde n, i,j}$ is defined in \eqref{eqn:wun} \\
\asp Measurement weights: $w^M_{u,n,i,j} = \left\{
\begin{array}{ll}
  f_{Y_{u,n} | X_{u,n}} (y_{u,n} | X^F_{u,n,i,j}) & \text{ if } \mathfrak u(i) < u\\
  f_{Y_{u,n} | X_{u,n}} (y_{u,n} | X^P_{u,n,i,j}) & \text{ if } \mathfrak u(i) \geq u.
\end{array}
\right.
$\\
End For\\
$\MC{\ell_{u,n}} = \log \left( \frac{\sum_{i=1}^\mathcal I \sum_{j=1}^\mathcal J w_{u,n,i,j}^M w_{u,n,i,j}^P}{\sum_{i=1}^\mathcal I \sum_{j=1}^\mathcal J w_{u,n,i,j}^P} \right)$
\vspace{1mm}
\\
\hline
\end{tabular}
\end{center}

\clearpage

\bibliography{../bib-iif}

\end{document}

### 2020-11-28 ms.Rnw
setting up files

\documentclass[12pt]{article}

%%% JASA. Attention Mac Users. The latest version of the pdfTex compiler for Mac (pdfTeX-1.40.11) outputs files not currently supported by ScholarOne Manuscripts. To resolve this and have a PDF file that will convert properly, the PDF file will need to be distilled to Adobe version 1.4 or earlier. To create a PDF file that uploads correctly, place the command \pdfminorversion=4 in the preamble (before the begin{document}) of your LaTeX file and run pdfLaTeX again.

\pdfminorversion=4

<<debug,echo=F>>=
run_level <- 2
#run_level <- "jasa-sept-2020"
#run_level <- 2
bm_run_level <- run_level
#mscale_run_level <- run_level
mscale_run_level <- 3
slice_run_level <- run_level

if(run_level=="jasa-sept-2020"){  
#### values used for jasa submission, sept 2020  
  bm_run_level <- 4
  mscale_run_level <- 5
  slice_run_level <- 3
}

@

%%% IN THE FINAL VERSION, THE INPUT HEADER FILE SHOULD BE PASTED IN.
\input{theorems.tex}
\input{header-ms.tex}

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

\newcommand\rewrite[1]{{\textcolor{red}{#1}}}

\usepackage{fullpage}
\addtolength{\textheight}{-.3in}%

\bibliographystyle{apalike}

% to run code from R: knitr::purl("ms.Rnw") ; source("ms.R")

<<packages,include=F,cache=F>>=
library("ggplot2")
library("spatPomp")
library(doParallel)
library(doRNG)

# 40 cores for doob, 8 for ito
doob_cores <- 40
gl_cores <- 36

cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()  
registerDoParallel(cores)

<<set-opts,include=F,cache=F>>=

options(
        scipen=2,
        help_type="html",
        stringsAsFactors=FALSE,
        prompt="R> ",
        continue="+  ",
        width=70,
        useFancyQuotes=FALSE,
        reindent.spaces=2,
        xtable.comment=FALSE
        )
@

<<knitr-opts,include=F,cache=F,purl=F>>=
library("knitr")
opts_knit$set(concordance=TRUE)
opts_chunk$set(
    progress=TRUE,prompt=TRUE,highlight=FALSE,
    tidy=TRUE,
    tidy.opts=list(
        keep.blank.line=FALSE
    ),
    comment="",
    warning=FALSE,
    message=FALSE,
    error=TRUE,
    echo=TRUE,
    cache=FALSE,
    strip.white=TRUE,
    results="markup",
    background="#FFFFFF00",
    size="normalsize",
    fig.path="figure/",
    fig.lp="fig:",
    fig.align="left",
    fig.show="asis",
#    dpi=300,
    dev="pdf",
    dev.args=list(
        bg="transparent",
        pointsize=9
    )
)

myround<- function (x, digits = 1) {
  # taken from the broman package
  if (digits < 1) 
    stop("This is intended for the case digits >= 1.")
  if (length(digits) > 1) {
    digits <- digits[1]
    warning("Using only digits[1]")
  }
  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
  tmp[tmp == paste0("-", zero)] <- zero
  tmp
}

@

% \date{This manuscript was compiled on \today}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\date{}

\if1\blind
{
  \title{\vspace{-10mm}\bf \mytitle}
  \author{Edward L. Ionides\thanks{This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108, and National Institutes of Health grants 1-U54-GM111274 and 1-U01-GM110712.}\hspace{.2cm}\\
    Department of Statistics, University of Michigan\\
    Kidus Asfaw\\
    Department of Statistics, University of Michigan\\
    %and \\
    Joonha Park\\
    Department of Mathematics \& Statistics, Boston University\\
    %and\\
    Aaron A. King\\
    Department of Ecology \& Evolutionary Biology, University of Michigan
}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf \mytitle}
\end{center}
  \medskip
} \fi

\vspace{-10mm}

\begin{abstract}
Bagging (i.e., bootstrap aggregating) involves combining an ensemble of bootstrap estimators.
The particle filter, also known as sequential Monte Carlo or the bootstrap filter, is a widely used algorithm for estimating the latent states or likelihood of a partially observed Markov process.
Our bagged filter (BF) methodology combines an ensemble of particle filters, using spatiotemporally localized weights to select successful filters at each point of space and time.
In some situations, BF methodology can theoretically beat a curse of dimensionality affecting standard particle filters;
in others, BF fails in theory to beat the curse, but nevertheless evinces practical advantages.
Our focus is on evaluation of the likelihood function, though BF theory and methodology are also pertinent to latent state estimation.
Applications suited to BF methodology include spatiotemporal analysis of epidemiological and ecological systems.
We show that BF methodology can out-perform an ensemble Kalman filter on a coupled population dynamics model arising in the epidemiology of infectious disease.
We also find that a block particle filter performs well on this task, though the bagged filter respects smoothness and conservation laws that a block particle filter can violate.

\end{abstract}

\noindent%
{\it Keywords:}  Particle filter; Sequential Monte Carlo; Markov process; Population dynamics

%\vfill
\pagebreak

\spacingset{1.45} % DON'T change the spacing!

\section{Introduction}

Particle filter (PF) algorithms are a widespread tool for inference on nonlinear partially observed dynamic systems \citep{doucet01,doucet11,breto09,ionides15}.
Modern particle filtering traces back to \citet{gordon93} where the algorithm was called the {\it bootstrap filter}. 
Standard PF methods suffer from a curse of dimensionality (COD), defined as an exponential increase in computational demands as the problem size grows, which limits its applicability to spatiotemporal systems \citep{bengtsson08,snyder15,rebeschini15}.
The COD presents empirically as numerical instability of the Monte Carlo algorithm for attainable numbers of particles.
Bootstrap aggregating (i.e., bagging) is a methodology to improve numerically unstable estimators by combining an ensemble of replicated bootstrap calculations \citep{breiman96}.
We consider algorithms combining an ensemble of replicated particle filters, and we use the name {\it bagged filter}.
The task of each bootstrap replicate is to provide some predictive skill at some points in space and time.
The bagged filter then proceeds to combine these replicates, assessing which replicates are reliable predictors at a given spatiotemporal location.

Much previous research has investigated scalable approaches to filtering and inference with applications to spatiotemporal systems.
Our bagged filters are in the class of {\it plug-and-play} algorithms, meaning that they require as input a simulator for the latent dynamic process but not an evaluator of transition probabilities \citep{breto09,he10}.
This simulation-based approach, also known as {\it likelihood-free} \citep{brehmer20} or {\it equation-free} \citep{kevrekidis09}, facilitates application to a wide class of models.
The ensemble Kalman filter \citep{evensen96,katzfuss19} is a widely used plug-and-play method which uses simulations to parameterize a Gaussian-inspired filtering rule.
Another plug-and-play approach to combat the COD is the block particle filter \citep{rebeschini15,ng02}.
Both ensemble Kalman filter and block particle filter methods construct trajectories that can violate smoothness and conservation properties of the dynamic model.
By contrast, our bagged filters are built using valid trajectories of the dynamic model, but make localization approximations when comparing these trajectories to data.
It is an empirical question how these strengths and weaknesses play out for a particular data analysis.
The plug-and-play property of an algorithm makes a general numerical implementation possible: we provide a software environment within which alternative methods can be applied.

The replicated stochastic trajectories in a bagged filter form an ensemble of representations of the dynamic system.
Unlike ensemble Kalman filters \citep{lei10} the bagged replicates are independent in a Monte Carlo sense.
Bagged filters therefore bear some resemblance to \textit{poor man's ensemble} forecasting methodology in which a collection of independently constructed forecasts is generated using different models and methods \citep{ebert01}. 
Poor man's ensembles have sometimes been found to have greater forecasting skill than any one forecast \citep{leutbecher08,palmer02,chandler13}.
One explanation for this phenomenon is that even a hypothetically perfect model cannot filter and forecast well using methodology afflicted by the COD.  
We show that bagged filter methodology can relieve this limitation.
From this perspective, the independence of the forecasts in the poor man's ensemble, rather than the diversity of model structures, may be the key to its success.

In a basic bagged filter, each replicate simply simulates a realization of the latent process model, corresponding to a degenerate instance of PF with only a single particle.
We call this the unadapted bagged filter ({\UBF}) since each replicate in the ensemble does not track the data.
While formally beating the COD under a weak mixing assumption, {\UBF} can have poor numerical behavior if a very large number of replicates are needed to reach this happy asymptotic limit.
Our empirical results show that {\UBF} may indeed be a useful algorithm in some situations.
More importantly, however, it provides a route to two other algorithms, which we call the adapted bagged filter (\ABF) and adapted bagged filter with intermediate resampling (\ABFIR).
{\ABF} and {\ABFIR} carry out some resampling on each replicate and therefore enable each replicate to track the data, in a weak sense, while resisting the COD.
This comes at a price: {\ABF} no longer asymptotically beats COD, and {\ABFIR} does so only under restrictive conditions.
However, we demonstrate useful behavior for these algorithms on finite samples.

To set up notation, we suppose the latent process is comprised of a collection of units, labeled $\{1,2,\dots,\Unit\}$ which we write as $1\mycolon\Unit$. 
An unobserved, latent Markov process at a discrete sequence of observation times is written as $\{\myvec{X}_{\time},\time\in 0\mycolon\Time\}$, with $\myvec{X}_{\time}=X_{1:\Unit,\time}$ taking values in $\Xspace^\Unit$.
The initial value $\myvec{X}_{0}$ may be stochastic or deterministic.
Observations are made on each unit, modeled by an observable process $\{\myvec{Y}_{\time}=Y_{1:\Unit,\time},\time\in 1\mycolon\Time\}$ which takes values in $\Yspace^\Unit$.
Observations are modeled as being conditionally independent given the latent process.
The conditional independence of measurements applies over both space and time, so $Y_{\unit\comma\time}$ is conditionally independent of
$\big\{
  X_{\altUnit\comma\altTime},
  Y_{\altUnit\comma\altTime},
  (\altUnit\comma\altTime)\neq(\unit\comma\time)
\big\}$
given $X_{\unit\comma\time}$.
This unit structure for the observation process is not necessary for all that follows (see Sec.~\SuppSecGeneralization).
We suppose the existence of a joint density $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$  of $\{X_{1:\Unit,\time}\}$ and $\{Y_{1:\Unit,\time}\}$ with respect to some appropriate measure.
We use the same subscript notation to denote other marginal and conditional densities.
The data are ${y}_{\unit\comma\time}$ for spatial unit $\unit$ at time $\time$. 
For a fixed value of $\Unit$, this model fits into the partially observed Markov process (POMP) framework \citep{breto09}, also known as a state space model or hidden Markov model.
Spatiotemporal POMP models are characterized by a latent Markov process with dimension proportional to $\Unit$.
We use the term {SpatPOMP} to describe POMP models indexed by $\unit$ as well as $\time$.
In accordance with the spatial interpretation of the units $\unit\in 1\mycolon\Unit$, we may suppose that units sufficiently distant in some sense approach independence, a phenomenon called {\it weak coupling}. 
It is worth pointing out that inference techniques designed for SpatPOMPs are also relevant in high-dimensional stochastic dynamic systems where $\unit$ does not index spatial locations.
In the following, we use the ordering on the set of observations corresponding to their labels $1\mycolon\Unit$ to define the set of observations preceding unit $\unit$ at time $\time$ as
\begin{equation}\label{eq:setA}
A_{\unit\comma\time}=\big\{(\tilde\unit,\tilde\time): 1 \leq \tilde\time<\time \mbox{ or } (\tilde\time=\time \mbox{ and } \tilde\unit <\unit)\big\}.
\end{equation}
The ordering of the spatial locations in \myeqref{eq:setA} might seem artificial, and indeed densities such as
$f_{X_{\unit\comma\time}|X_{A_{\unit\comma\time}}}$ will frequently be hard to compute or simulate from.
However, the algorithms we study do not carry out such computations, but only evaluate the measurement model on these neighborhoods.
To represent weak coupling, we suppose there is a neighborhood $B_{\unit\comma\time}\subset A_{\unit\comma\time}$ such that the latent process on 
$A_{\unit\comma\time} \setminus B_{\unit\comma\time}$ 
is approximately conditionally independent of $X_{\unit\comma\time}$ given data on $B_{\unit\comma\time}$.

The inputs for the {\UBF}, {\ABF} and {\ABFIR} are listed in Table~1.
The plug-and-play property is evident since all three algorithms require as input a simulator for the latent dynamic process but not an evaluator of transition probabilities.
Our primary interest is in the estimation of the log likelihood for the data given the model, $\loglik=\log f_{\myvec{Y}_{1:\Time}}(\data{\myvec{y}}_{1:\Time})$, which is of fundamental importance in both Bayesian and non-Bayesian statistical inference.
Monte Carlo estimates of the log likelihood underpin modern methods for full-information statistical inference on partially observed dynamic systems \citep{andrieu10,ionides15,ionides17}.
Filtering theory ties together the three tasks of likelihood evaluation, forecasting, and inferring latent dynamic variables \citep{crisan11}.
Successful likelihood evaluation is therefore closely related to success at forecasting and latent process reconstruction.

The pseudocodes for the algorithms that follow will adopt a convention that implicit loops are carried out over all free indices.
For example, the construction of $w^P_{\unit,\time,\rep}$ in {\UBF} has an implicit loop over $\unit$, $\time$ and $\rep$. 
The construction of $w^P_{\unit,\time,\rep,\np}$ in {\ABF} has an implicit loop over $\unit$, $\rep$ and $\np$ but not over $\time$ since $\time$ is already defined in an explicit loop and is therefore not a free index.
Table~1 lists the ranges of the implicit loops for all following algorithms.

\begin{center}
\noindent\begin{tabular}{l}
\hline
\inputSpace {\bf Table 1. Bagged filter inputs, outputs and implicit loops.}\\
\hline
\inputSpace {\bf input:}
\\
simulator for $f_{\myvec{X}_0}(\myvec{x}_0)$ and $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1})$\\
    evaluator for $f_{{Y}_{\unit\comma\time}|{X}_{\unit\comma\time}}({y}_{\unit\comma\time}\given {x}_{\unit\comma\time})$\\
    number of replicates, ${\Rep}$\\
    neighborhood structure, $B_{\unit\comma\time}$\\
    data, $\data{\myvec{y}}_{1:\Time}$\\
{\ABF} and {\ABFIR}:    particles per replicate,  $\Np$\\
{\ABFIR}: number of intermediate timesteps, $\Ninter$ \\
{\ABFIR}: measurement variance parameterizations, ${\VtoTheta}_{\unit\comma\time}$ and ${\thetaToV}_{\unit\comma\time}$\\
{\ABFIR}: approximate process and observation mean functions, $\myvec{\mu}$ and $h_{\unit\comma\time}$\\
\inputSpace {\bf output:}\\
  Log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit\comma\time}$\\
\inputSpace {\bf implicit loops:}\\
$\; \unit \mbox{ in } \seq{1}{\Unit}$, 
$\; \time \mbox{ in } \seq{1}{\Time}$, 
$\; \rep \mbox{ in } \seq{1}{\Rep}$, 
$\; \np \mbox{ in } \seq{1}{\Np}$
%$\; \npgir \mbox{ in } \seq{1}{\Npgir}$
\lastLineSpace \\
\hline
\end{tabular}
\end{center}

The~{\UBF} algorithm that follows amounts to a bagged filter with a single particle per replicate, with no weighting or resampling within each replicate.

%%%%%%%%%%%%% ttttttttttttttttttttttttt

\begin{center}
\noindent\begin{tabular}{l}
\hline
\inputSpace {\bf {\UBF}. Unadapted bagged filter.}\\
\hline
%\inputSpace {\bf input:} From Table~1, with $\Np=1$ \\
%\hline
\firstLineSpace 
Simulate $\myvec{X}^{\tif}_{0:\Time,\rep}\sim f_{\myvec{X}_{0:\Time}}(\myvec{x}_{0:\Time})$\\ 
Measurement weights,
$w^M_{\unit,\time,\rep}=f_{Y_{\unit,\time}|X_{\unit,\time}}(\data{y}_{\unit,\time}\given X^{\tif}_{\unit,\time,\rep})$
\\
Prediction weights, 
$w^P_{\unit,\time,\rep}=\prod_{(\tilde \unit,\tilde n)\in B_{\unit,\time}}
w^M_{\tilde\unit,\tilde n,\rep}$\\
$\MC{\loglik}_{\unit,\time}= 
\log\left(
  \sum_{\rep=1}^\Rep w^M_{\unit,\time,\rep}w^P_{\unit,\time,\rep}
\right)
-\log\left(
  \sum_{\rep=1}^\Rep w^P_{\unit,\time,\rep}
\right)
$
\lastLineSpace
\\
\hline
\end{tabular}
\end{center}

Theorem~\ref{thm:tif} shows that {\UBF} can beat COD. 
However, an algorithm such as {\UBF}, which includes no resampling, will fail on long time series absent rapid spatiotemporal mixing, which allows simulated particles to remain relevant over the course of a long time series.
It may in practice be necessary to select simulations consistent with the data, much as standard PF algorithms do.
We look for approaches that build on the basic insight of {\UBF} while having superior practical performance.

Whereas the full global filtering problem may be intractable via importance sampling methods, a version of this problem local in space and time may nevertheless be feasible.
Drawing a sample from the conditional density, $f_{\myvec{X}_{\time}|\myvec{Y}_{\time},\myvec{X}_{\time-1}}$, is distinct from the filtering problem: we call this operation {\it adapted simulation}.
For models where $\myvec{X}_{\time}$ is close to $\myvec{X}_{\time-1}$, adapted simulation requires a local calculation that may be much easier than the full filtering calculation.
The following adapted bagged filter ({\ABF}) is constructed under a hypothesis that the adapted simulation problem is tractable.
The adapted simulations are then reweighted in a neighborhood of each point in space and time to construct a local approximation to the filtering problem which leads to an estimate of the likelihood.

%%%%%%%%%% aaaaaaaaaaaaaaaaa

\begin{center}
\noindent\begin{tabular}{l}
\hline
{\bf 
{\ABF}. Adapted bagged filter.}\inputSpace\\
\hline
\firstLineSpace
Initialize adapted simulation: $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\myvec{x}_0)$
\\
For $\time\ \mathrm{in}\ \seq{1}{\Time}$
\\
\asp  Proposals:
    $\myvec{X}_{\time,\rep,\np}^{\IP} \sim 
    f_{\myvec{X}_{\time}|X_{1:\Unit,\time-1}} 
    \big( \myvec{x}_{\time}\given \myvec{X}^{\IF}_{\time-1,\rep}\big)$
\\
\asp Measurement weights:
  $w^M_{\unit,\time,\rep,\np} = 
    f_{Y_{\unit,\time}|X_{\unit\comma\time}} 
    \big (\data{y}_{\unit\comma\time}\given X^{\IP}_{\unit\comma\time,\rep,\np}\big)$
\\
\asp  Adapted resampling weights:
  $w^{\IF}_{\time,\rep,\np} = 
    \prod_{\unit=1}^{\Unit} w^M_{\unit,\time,\rep,\np}$
\\
\asp
      Resampling:
        $\prob\big[\resampleIndex({\rep})=a \big] = w^{\IF}_{\time,\rep,a}
  \Big( 
  \sum_{\altNp=1}^{\Np} w^{\IF}_{\time,\rep,\altNp}
  \Big)^{-1}$
\\
\asp 
$\myvec{X}^{\IF}_{\time,\rep} = \myvec{X}^{\IP}_{\time,\rep,r(\rep)}$ 
\\
\asp % Prediction weights:
  $w^{\LCP}_{\unit,\time,\rep,\np}= \displaystyle
  \prod_{\altTime=1}^{\time-1}
  \Big[
    \frac{1}{\Np}\sum_{k=1}^{\Np}
    \hspace{1mm}
       \prod_{\altUnit:(\altUnit,\altTime)\in B_{\unit,\time}} 
    \hspace{-1mm}
        w^M_{\altUnit,\altTime,\rep,k}
  \Big] \prod_{\altUnit:(\altUnit,\time)\in B_{\unit,\time}} 
    \hspace{-1mm}
        w^M_{\altUnit,\time,\rep,\np}$
\\
End for
\\
$\displaystyle \MC{\loglik}_{\unit,\time}= 
\log\Bigg(
\frac{
\sum_{\rep=1}^\Rep \sum_{\np=1}^{\Np} w^M_{\unit,\time,\rep,\np}w^P_{\unit,\time,\rep,\np}
}{
\sum_{\rep=1}^\Rep \sum_{\np=1}^{\Np} w^P_{\unit,\time,\rep,\np}
}
\Bigg)
$
\vspace{1mm}
\\
\hline
\end{tabular}
\end{center}

In addition, we consider an algorithm which adds intermediate resampling to the adapted simulation replicates, and is therefore termed {\ABFIR}.
Intermediate resampling involves assessing the satisfactory progress of particles toward the subsequent observation at a collection of times between observations.
This is well defined when the latent system is a continuous time process $\{\myvec{X}(t)\}$ with observation times $t_{1:\Time}$ such that $\myvec{X}_\time=\myvec{X}(t_{\time})$.
For a continuous time latent process model, intermediate resampling can have favorable scaling properties when the intermediate times 
\begin{equation}
\nonumber
t_{\time-1}=t_{\time,0}<t_{\time,1}<\dots<t_{\time,\Ninter}=t_{\time} 
\end{equation}
scale with $\Ninter\approx \Unit$ \citep{park20}.
In the case $\Ninter=1$, {\ABFIR} reduces to {\ABF}.
Intermediate resampling was developed in the context of sequential Monte Carlo \citep{delmoral15,park20}; however, the same theory and methodology can be applied to the simpler and easier problem of adapted simulation. 
{\ABFIR} employs a guide function to gauge the compatibility of each particle with future data.
This is a generalization of the popular auxiliary particle filter \citep{pitt99}.
The implementation in this pseudocode constructs the guide $g_{\time,\ninter,\rep,\np}$ using a simulated moment method proposed by \citet{park20}.
The quantities  $\myvec{X}_{\time,\rep,\npgir}^{G}$, $V_{\unit,\time,\rep}$, $\myvec{\mu}^{\GP}_{\time,\ninter,\rep,\np}$, $V^{\mathrm{meas}}_{\unit,\time,\ninter,\rep,\np}$, $V^{\mathrm{proc}}_{\unit,\time,\ninter,\rep}$ and $\theta_{\unit,\time,\ninter,\rep,\np}$ constructed in {\ABFIR} are used only to construct $g_{\time,\ninter,\rep,\np}$.
Heuristically, we use guide simulations to approximate the variance of the increment in each particle between time points, and we augment the measurement variance to account for both dynamic variability and measurement error.
The guide function affects numerical performance of the algorithm but not its correctness: it enables a computationally convenient approximation to improve performance on the intractable target problem.
Our guide function supposes the availability of a deterministic function approximating evolution of the mean of the latent process, written as
\begin{equation}
\nonumber
\myvec{\mu}(\myvec{x},s,t)\approx \E\big[\myvec{X}(t) \given \myvec{X}(s)=\myvec{x}\big]. 
\end{equation}
Further, the guide requires that the measurement model has known conditional mean and variance as a function of the model parameter vector $\theta$, written as
\begin{eqnarray}
\nonumber
h_{\unit\comma\time}(x_{\unit\comma\time})
  &=& \E\big[Y_{\unit\comma\time}\given X_{\unit\comma\time}=x_{\unit\comma\time}\big]
\\
\nonumber
{\thetaToV}_{\unit\comma\time}(x_{\unit\comma\time},\theta)
  &=& \var\big(Y_{\unit\comma\time}\given X_{\unit\comma\time}=x_{\unit\comma\time} \giventh \theta\big)
\end{eqnarray}
Also required for {\ABFIR} is an inverse function ${\VtoTheta}_{\unit\comma\time}$ such that
\begin{equation}
\nonumber
{\thetaToV}_{\unit\comma\time} \big( x_{\unit\comma\time},\VtoTheta_{\unit\comma\time}(V,x_{\unit\comma\time},\theta) \big) = V.
\end{equation}

%%%%%%%%%%%%%%%%% iiiiiiiiiiiiiiiiiii

\begin{center}
\noindent\begin{tabular}{l}
\hline
{\bf {\ABFIR}. Adapted bagged filter with intermediate resampling.} 
\firstLineSpace \\  
\vspace{0.4mm} \\
\hline
\firstLineSpace
Initialize adapted simulation: $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\myvec{x}_0)$
\\
For $\time\ \mathrm{in}\ \seq{1}{\Time}$
\\
\asp Guide simulations:
    $\myvec{X}_{\time,\rep,\npgir}^{G} \sim 
    f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}} 
    \big( \myvec{x}_{\time}\given \myvec{X}^{\IF}_{\time-1,\rep} \big)$
\\
\asp Guide sample variance: $V_{\unit,\time,\rep}=
      \var \big\{
        h_{\unit\comma\time}\big( {X}_{\unit,\time,\rep,\npgir}^{G}\big), \npgir \mbox{ in } \seq{1}{\Npgir}
      \big\}$ 
\\
\asp $\guideFunc^{\resample}_{\time,0,\rep,\np}=1 \; \; $ and
$\; \myvec{X}_{\time,0,\rep,\np}^{\GR}=\myvec{X}^{\IF}_{\time-1,\rep}$
\\
\asp For $\ninter  \,\, \mathrm{in} \,\, \seq{1}{\Ninter}$
\\
\asp\asp Intermediate proposals:
        ${\myvec{X}}_{\time,\ninter,\rep,\np}^{\GP}
          \sim {f}_{{\myvec{X}}_{\time,\ninter}|{\myvec{X}}_{\time,\ninter-1}}
          \big(\mydot|{\myvec{X}}_{\time,\ninter-1,\rep,\np}^{\GR}\big)$ 
\\
\asp\asp 
        $\myvec{\mu}^{\GP}_{\time,\ninter,\rep,\np} 
           = \myvec{\mu}\big( \myvec{X}^{\GP}_{\time,\ninter,\rep,\np},t_{\time,\ninter},t_{\time} \big)$
\\
\asp\asp      %Measurement variance at skeleton: 
        $V^{\mathrm{meas}}_{\unit,\time,\ninter,\rep,\np}
           = \thetaToV_{\unit}(\theta,\mu^{\GP}_{\unit,\time,\ninter,\rep,\np})$
\\
\asp\asp  %Process variance:
        $V^{\mathrm{proc}}_{\unit,\time,\ninter,\rep}
          = V_{\unit,\time,\rep} \,
          \big(t_{\time}-t_{\time,\ninter}\big) \Big/
          \big(t_{\time}-t_{\time,0}\big)$ 
\\
\asp\asp
%      Moment matching:
        $\theta_{\unit,\time,\ninter,\rep,\np}= 
          \VtoTheta_{\unit}\big(
            V^{\mathrm{meas}}_{\unit,\time,\ninter,\rep,\np} + V^{\mathrm{proc}}_{\unit,\time,\ninter,\rep}, 
            \, \mu^{\GP}_{\unit,\time,\ninter,\rep,\np}
          \big)$
\\
\asp\asp  % Guide function: 
        $
\guideFunc_{\time,\ninter,\rep,\np}=
          \prod_{\unit=1}^{\Unit}
          f_{Y_{\unit,\time}|X_{\unit,\time}}
          \big(
            \data{y}_{\unit,\time}\given \mu^{\GP}_{\unit,\time,\ninter,\rep,\np} \giventh \theta_{\unit,\time,\ninter,\rep,\np} 
          \big)$
\\
\asp\asp Guide weights:
$w^G_{\time,\ninter,\rep,\np}= \guideFunc^{}_{\time,\ninter,\rep,\np}
         \big/ \guideFunc^{\resample}_{\time,\ninter-1,\rep,\np}$
\\
\asp\asp
      Resampling:
        $\prob\big[\resampleIndex({\rep,\np})=a \big] = w^G_{\time,\ninter,\rep,a}
\Big( \sum_{\altNp=1}^{\Np}w^G_{\time,\ninter,\rep,\altNp}\Big)^{-1}$
\\
\asp\asp
        $\myvec{X}_{\time,\ninter,\rep,\np}^{\GR}=\myvec{X}_{\time,\ninter,\rep,\resampleIndex({\rep,\np})}^{\GP}\; \; $ and
        $\; \guideFunc^{\resample}_{\time,\ninter,\rep,\np}= \guideFunc^{}_{\time,\ninter,\rep,\resampleIndex({\rep,\np})}\,$
\\
\asp
End For
\\
\asp
  Set $\myvec{X}^{\IF}_{\time,\rep}=\myvec{X}^{\GR}_{\time,\Ninter,\rep,1}$ 
\\ 
\asp Measurement weights:
  $w^M_{\unit,\time,\rep,\npgir} = 
    f_{Y_{\unit,\time}|X_{\unit,\time}} 
    \big (\data{y}_{\unit,\time}\given X^{G}_{\unit,\time,\rep,\npgir} \big)$
\\
\asp % Prediction weights:
  $w^{\LCP}_{\unit,\time,\rep,\npgir}= \displaystyle
  \prod_{\altTime=1}^{\time-1}
  \Big[
    \frac{1}{\Npgir}\sum_{a=1}^{\Npgir}
    \hspace{1mm}
       \prod_{\altUnit:(\altUnit,\altTime)\in B_{\unit,\time}} 
    \hspace{-1mm}
        w^M_{\altUnit,\altTime,\rep,a}
  \Big] \prod_{\altUnit:(\altUnit,\time)\in B_{\unit,\time}} 
    \hspace{-1mm}
        w^M_{\altUnit,\time,\rep,\npgir}$
\\
End for
\\
$\displaystyle \MC{\loglik}_{\unit,\time}= 
\log\Bigg(
\frac{
\sum_{\rep=1}^\Rep \sum_{\npgir=1}^{\Npgir} w^M_{\unit,\time,\rep,\npgir}w^P_{\unit,\time,\rep,\npgir}
}{
\sum_{\rep=1}^\Rep \sum_{\npgir=1}^{\Npgir} w^P_{\unit,\time,\rep,\npgir}
}
\Bigg)
$
\vspace{1mm}
\\
\hline
\end{tabular}
\end{center}

\clearpage

This guide function is applicable to spatiotemporal versions of a broad range of population and compartment models used to model dynamic systems in ecology, epidemiology, and elsewhere. 
Other guide functions could be developed and inserted into the {\ABFIR} algorithm, including other constructions considered by \citet{park20}.
The number of particles for the guide simulations in {\ABFIR} does not necessarily have to be set equal to the number of intermediate proposals. 
We make that simplification here since there is no compelling reason against it.

One might wonder why it is appropriate to keep many particle representations at intermediate timesteps while resampling down to a single representative at each observation time.
Part of the answer is that adaptive simulation can fail when one resamples down to a single particle too often (Sec.~\SuppSecAdaptedSimulation).

%% llllllllllllllllllllllllllllllllll

\section{Likelihood factorizations and their approximations}

The approximation error for the algorithms we study here can be divided into two sources: a localization bias due to conditioning on a finite neighborhood, and Monte Carlo error. 
The localization bias does not disappear in the limit as Monte Carlo effort increases.
It does become small as the conditioning neighborhood increases, but the Monte Carlo effort grows exponentially in the size of this neighborhood.
The class of problems for which these algorithms are useful are ones where a relatively small neighborhood is adequate. 
Although the filtering inference is carried out using localization, the simulation of the process is carried out globally which avoids the introduction of additional boundary effects and ensures that the simulations comply with any constraint properties of the global process simulator.

In this subsection, we consider the deterministic limit of each algorithm for infinite Monte Carlo effort.
We explain why the {\UBF}, {\ABF} and {\ABFIR} algorithms approximately target the likelihood function, subject to suitable mixing behavior.
Subsequently, we consider the scaling properties as Monte Carlo effort increases.
We adopt a convention that densities involving $Y_{\unit\comma\time}$ are implicitly evaluated at the data, $\data{y}_{\unit\comma\time}$, and densities involving $X_{\unit\comma\time}$ are implicitly evaluated at $x_{\unit\comma\time}$ unless otherwise specified.
We write $A^{+}_{\unit\comma\time}=A_{\unit\comma\time}\cup (\unit,\time)$ and $B^{+}_{\unit\comma\time}=B_{\unit\comma\time}\cup (\unit,\time)$.
The essential ingredient in all the algorithms is a localization of the likelihood, which may be factorized sequentially as
\begin{equation}
%\label{eq:ordering}
\nonumber
f_{Y_{1:\Unit,1:\Time}}
=
\prod_{\time=1}^\Time\prod_{\unit=1}^\Unit 
f_{Y_{\unit\comma\time}|Y_{A_{\unit\comma\time}}}
= \prod_{\time=1}^\Time\prod_{\unit=1}^\Unit 
\frac{f_{Y^{}_{A^+_{\unit\comma\time}}}}{f_{Y^{}_{A^{}_{\unit\comma\time}}}}.
\end{equation}
In particular, the approximations assume that the full history $A_{\unit\comma\time}$ can be well approximated by a neighborhood $B_{\unit\comma\time}\subset A_{\unit\comma\time}$.
{\UBF} approximates $f_{Y_{\unit\comma\time}|Y_{A_{\unit\comma\time}}}$ by
\begin{equation}
\nonumber
f_{Y_{\unit\comma\time}|Y_{B_{\unit\comma\time}}} 
=
\frac{f^{}_{Y^{}_{B^+_{\unit\comma\time}}}}{f^{}_{Y^{}_{B^{}_{\unit\comma\time}}}}
=
\frac{\int f_{Y_{B^+_{\unit\comma\time}}|X_{B^+_{\unit\comma\time}}} f_{X_{B^+_{\unit\comma\time}}}\, dx_{B^+_{\unit\comma\time}} }
{\int f_{Y_{B_{\unit\comma\time}}|X_{B_{\unit\comma\time}}} f_{X_{B_{\unit\comma\time}}}\, dx_{B_{\unit\comma\time}} }.
\end{equation}
For $\unitTimeSubset \subset \seq{1}{\Unit}\times\seq{1}{\Time}$, define $\unitTimeSubset^{[m]}=\unitTimeSubset \cap \big(\UnitSet\times \{m\}\big)$.
{\ABF} and {\ABFIR} build on the following identity, 
\begin{equation}
\nonumber
%\label{eq:adapted}
\hspace{-1mm}
f_{Y_{A_{\unit\comma\time}}}{=} \int 
\! \!
f_{\myvec{X}_0} 
\! \!
\left[
\prod_{\altAltTime=1}^{\time}
f_{\myvec{X}_{\altAltTime}|\myvec{X}_{\altAltTime-1},\myvec{Y_{\altAltTime}}}
f_{Y_{A^{[\altAltTime]}_{\unit\comma\time}}|\myvec{X}_{\altAltTime-1}}
\right]
\! d\myvec{x}_{0:\time},
\end{equation}
where
$f_{\myvec{X}_{\altAltTime}|\myvec{X}_{\altAltTime-1},\myvec{Y}_{\altAltTime}}$ is called the adapted transition density.
The adapted process (i.e., a stochastic process following the adapted transition density) can be interpreted as a one-step greedy procedure using the data to guide the latent process.
Let 
$g^{}_{\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time}}(\myvec{x}_{0:\Time},\myvec{x}^P_{1:\Time})$ 
be the joint density of the adapted process and the proposal process, 
\begin{eqnarray}
\nonumber
\hspace{-3mm} g^{}_{\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time}}(\myvec{x}_{0:\Time},\myvec{x}^P_{1:\Time})
&=& f_{\myvec{X}_0}(\myvec{x}_0) \times 
\\
\label{eq:g}
%\nonumber
&& \hspace{-42mm}
\prod_{\time=1}^{\Time} 
f_{\myvec{X}_{\time}|\myvec{X}_{\time-1},\myvec{Y}_{\time}} 
  \big( 
    \myvec{x}_{\time} \given \myvec{x}_{\time-1},\data{\myvec{y}}_{\time} 
  \big)
\,\,
f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}
  \big(
    \myvec{x}^P_{\time} \given \myvec{x}_{\time-1}
  \big).
\end{eqnarray}
For 
$\unitTimeSubset \subset \UnitSet\times 1{\mycolon}\Time$, 
define 
$\unitTimeSubset^{[m]}=\unitTimeSubset \cap \big(\UnitSet\times \{m\}\big)$ 
and set
\begin{equation}
%\label{eq:h_S}
\nonumber
\adapted^{}_{\unitTimeSubset}
%(\myvec{x}_{0:\Time})
= 
\prod_{m=1}^{\Time} f_{Y_{\unitTimeSubset^{[m]}}|\myvec{X}^{}_{m-1}} \big( \data{y}_{\unitTimeSubset^{[m]}} \given \myvec{X}^{}_{m-1} \big),
\end{equation}
using the convention that an empty density $f_{Y_{\emptyset}}$ evaluates to 1.
Denoting $\E_{g}$ for expectation for $(\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time})$ having density $g_{\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time}}$, we have
\begin{equation}
%\label{eq:AB:ratio}
\nonumber
f_{Y_{\unit\comma\time}|Y_{A_{\unit\comma\time}}}
= \frac{\E_{g}\big[\adapted^{}_{A^+_{\unit\comma\time}}\big] }{\E_{g}\big[\adapted^{}_{A^{}_{\unit\comma\time}}\big] }.
\end{equation}
Estimating this ratio by Monte Carlo sampling from $g$ is problematic due to the growing size of $A_{\unit\comma\time}$.
Thus, {\ABF} and {\ABFIR} make a localized approximation,
\begin{equation}
\label{eq:AB:ratio2}
 \frac{\E_{g}\big[\adapted^{}_{A^+_{\unit\comma\time}}\big] }{\E_{g}\big[\adapted^{}_{A^{}_{\unit\comma\time}}\big] }
\approx \frac{\E_{g}\big[\adapted^{}_{B^+_{\unit\comma\time}}\big] }{\E_{g}\big[\adapted^{}_{B^{}_{\unit\comma\time}}\big] }.
\end{equation}
The conditional log likelihood estimate $\MC{\loglik}_{\unit\comma\time}$ in {\ABF} and {\ABFIR} come from replacing the expectations on the right hand side of \myeqref{eq:AB:ratio2} with averages over 
Monte Carlo replicates of simulations from the adapted process. 
To see that we expect the approximation in \myeqref{eq:AB:ratio2} to hold when dependence decays across spatiotemporal distance, we can write 
%\begin{equation}
\begin{eqnarray}
\nonumber
\adapted^{}_{A^{}_{\unit\comma\time}} 
&=&
 \adapted^{}_{B^{}_{\unit\comma\time}}  \hspace{1mm} \adapted^{}_{B^{c}_{\unit\comma\time}} 
\\
%\mbox{ and }
\nonumber
\adapted^{}_{A^+_{\unit\comma\time}}
&=&
 \adapted^{}_{B^{+}_{\unit\comma\time}}  \hspace{1mm}  \adapted^{}_{B^{c}_{\unit\comma\time}} ,
%\end{equation}
\end{eqnarray}
where $B^c_{\unit\comma\time}$ is the complement of $B_{\unit\comma\time}$ in $A_{\unit\comma\time}$.
As long as $\adapted_{B^c_{\unit\comma\time}}$ is approximately independent of $X_{\unit\comma\time}$ under $g$, this term approximately cancels in the numerator and denominator of the right hand side of \myeqref{eq:AB:ratio2}.

\section{{\UBF} theory}

For each pair $(\Unit,\Time)$ we suppose there is a dataset $\data{\myvec{y}}_{1:\Time}$ and a model $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$.
We are interested in results that hold for all $(\Unit,\Time)$, but we do not require that the models for $(\Unit_1,\Time_1)\neq(\Unit_2,\Time_2)$ are nested in any way. 
The following conditions impose requirements that distant regions of space-time behave similarly and have only weak dependence.
The conditions are written non-asymptotically in terms of $\epsilon>0$ and $Q>1$ which are used to bound the asymptotic bias and variance in Theorem~\ref{thm:tif}.
Stronger bounds are obtained when the conditions hold for small $\eone$, $\etwo$ and $Q$.

\Ai  %% Assumption 1  Assumption~\ref{A1}

\Aii  %% Assumption 2  Assumption~\ref{A1b}

\Aiii   %% Assumption 3  Assumption~\ref{A2}

\Aiv    %% Assumption 4  Assumption~\ref{A:unconditional:mix}

The two mixing conditions in Assumptions~\ref{A1} and~\ref{A:unconditional:mix} are subtly different. 
Assumption~\ref{A1} is a conditional mixing property, dependent on the data, whereas~\ref{A:unconditional:mix} asserts a form of unconditional mixing.
Although both capture a similar concept of weak coupling, conditional and unconditional mixing properties do not readily imply one another.
Assumption~\ref{A2} is a compactness condition of a type that has proved useful in the theory of particle filters despite the rarity of its holding exactly.
Theorem~\ref{thm:tif} shows that these conditions let {\UBF} compute the likelihood with a Monte Carlo variance of order $\Unit\Time\Rep^{-1}$ with a bias of order $\Unit\Time\epsilon$.

\TheoremI %% Theorem~\ref{thm:tif}

\begin{proof}
A complete proof is given in Sec.~\SuppSecThmI. 
Briefly, the assumptions imply a multivariate central limit theorem for $\{\MC{\loglik}_{\unit\comma\time}, (\unit,\time)\in \UnitSet{\times}\ObsTimeSet\}$ as $\Rep\to\infty$.
The limiting variances and covariances are uniformly bounded, using Assumptions~\ref{A1b} and~\ref{A2}. 
Assumption~\ref{A1} provides a uniform bound on the discrepancy between ${\loglik}_{\unit\comma\time}$ and mean of the Gaussian limit.
This is enough to derive \myeqref{th1:lik:bound}.
Assumption~\ref{A:unconditional:mix} gives a stronger bound on covariances between sufficiently distant units, leading to \myeqref{th1:lik:bound2}.
\end{proof}

\section{{\ABFIR} theory}

Since {\ABF} is {\ABFIR} with $\Ninter=1$, we focus attention on {\ABFIR}.
At a conceptual level, the localized likelihood estimate in {\ABFIR} has the same structure as its {\UBF} counterpart.
However, {\ABFIR} additionally requires the capability to satisfactorily implement adapted simulation.
Adapted simulation is a local calculation, making it an easier task than the global operation of filtering.
Nevertheless, adapted simulation via importance sampling is vulnerable to COD for sufficiently large values of $\Unit$.
For a continuous time model, the use of $\Ninter>1$ is motivated by a result that guided intermediate resampling can reduce, or even remove, the COD in the context of a particle filtering algorithm \cite{park20}.
Assumptions~\ref{B1}--\ref{B:unconditional:mix} are analogous to ~\ref{A1}--\ref{A:unconditional:mix} and are non-asymptotic assumptions involving $\ethree>0$, $\efourA>0$ and $Q>1$ which are required to hold uniformly over space and time.
Assumption~\ref{B:temporal:mix}--\ref{B:XG_XA_ind} control the Monte~Carlo error arising from adapted simulation.
~\ref{B:temporal:mix} is a stability property which asserts that the effect of the latent process on the future of the adapted process decays over time. 
Assumption~\ref{B:girf} is a non-asymptotic bound on Monte~Carlo error for a single step of adapted simulation.
The scaling of the constant $\BvConstant$ with  $\Unit$,  $\Time$ and $\Ninter$ in  Assumption~\ref{B:girf} has been studied by \citet{park20}, where it was established that setting $\Ninter=\Unit$ can lead to $\BvConstant$ being constant, when using an ideal guide function, or slowly growing with $\Unit$ otherwise.
The $\efive^{-3}$ error rate in Assumption~\ref{B:girf} follows from balancing the two sources of error defined in the statement of Theorem~2 of \citet{park20}.
Assumption~\ref{B:XG_XA_ind} can be guaranteed by the construction of the algorithm, if independently generated Monte~Carlo random variables are used for building the guide function and the one-step prediction particles.
The asymptotic limit in Theorem~\ref{thm:abf} arises as the number of replicates increases.

\Bi %% Assumption~\ref{B1}

\Bii %% Assumption~\ref{B1b}

\Biii %% Assumption~\ref{B2}

\BivA %% Assumption~\ref{B:unconditional:mix}

\BivB %% Assumption~\ref{B:temporal:mix}

\Bv  %% Assumption~\ref{B:girf}

\Bvi    %% Assumption~\ref{B:XG_XA_ind}

\TheoremII %% Theorem~\ref{thm:abf}

\begin{proof}
 A full proof is provided in Sec.~\SuppSecThmII. The extra work to prove Theorem~\ref{thm:abf} beyond the argument for Theorem~\ref{thm:tif} is to bound the error arising from the importance sampling approximation to a draw from the adapted transition density. 
This bound is constructed using Assumptions~\ref{B:temporal:mix}, \ref{B:girf} and~\ref{B:XG_XA_ind}.
The remainder of the proof follows the same approach as Theorem~\ref{thm:tif}, with the adapted process replacing the unconditional latent process.
\end{proof}

The theoretical results foreshadow our empirical observations that the relative performance of {\UBF}, {\ABF} and {\ABFIR} is situation-dependent.
Assumption~\ref{A:unconditional:mix} is a mixing assumption for the unconditional latent process, whereas Assumption~\ref{B:unconditional:mix} replaces this with a mixing assumption for the adapted process conditional on the data. 
For a non-stationary process, Assumption~\ref{A:unconditional:mix} may fail to hold uniformly in $\Unit$ whereas the adapted process may provide stable tracking of the latent process (Sec.~\SuppSecAdaptedSimulation).
When Assumption~\ref{A:unconditional:mix} holds, {\UBF} can benefit from not requiring Assumptions~\ref{B:temporal:mix}, \ref{B:girf} and~\ref{B:XG_XA_ind}. 
Adapted simulation is an easier problem than filtering, but nevertheless can become difficult in high dimensions, with the consequence that Assumption~\ref{B:girf} could require large $\BvConstant$.
The tradeoff between {\ABF} and {\ABFIR} depends on the effectiveness of the guide function for the problem at hand.
Intermediate resampling and guide function calculation require additional computational resources, which will necessitate smaller values of $\Rep$ and $\Np$.
In some situations, the improved scaling properties of {\ABFIR} compared to {\ABF}, corresponding to a lower value of $\BvConstant$, will outweigh this cost.

\medskip

\section{Examples}

Likelihood evaluation via filtering does not by itself enable parameter estimation for POMP models, however it provides a foundation for Bayesian \citep{andrieu10} and likelihood-based \citep{ionides15} inference.
We therefore investigate likelihood evaluation with the expectation that improved methods for likelihood evaluation will lead to improved statistical methodology. 
Monte Carlo methods for computing the log likelihood generally suffer from both bias and variance, both of which can be considerable for large datasets and complex models.
Appropriate inference methodology, such as Monte Carlo adjusted profile (MCAP) confidence intervals, can accommodate substantial Monte Carlo variance so long as the bias is slowly varying across the statistically plausible region of the parameter space \citep{ionides17}.

We compare the performance of the three bagged filters ({\UBF}, {\ABF} and {\ABFIR}) against each other and against alternative approaches.
Ensemble Kalman filter (EnKF) methods propagate the ensemble members by simulation from the dynamic model and then update the ensemble to assimilate observations using a Gaussian-inspired rule \citep{evensen96,lei10}.
The block particle filter (BPF) \citep{rebeschini15,ng02} partitions the latent space and combines independently drawn components from each partition.
BPF overcomes COD under weak coupling assumptions \citep{rebeschini15}.
Unlike these two methods, our bagged filters modify particles only according to the latent dynamics.
Thus, our methods satisfy any conservation laws, continuity or smoothness that arise when simulating from the dynamic model.

We first consider a spatiotemporal Gaussian process for which the exact likelihood is available via a Kalman filter. 
We see in Fig.~\ref{fig:bm_alt_plot} that {\ABFIR} can have a considerable advantages over {\UBF} and {\ABF} for problems with an intermediate level of coupling.
We then develop a model for measles transmission within and between cities.
The model is weakly coupled, leading to successful performance for all three bagged filters.
This class of metapopulation models was the primary motivation for the development of these methodologies.
However, for a highly coupled system of moderate dimensions, a global particle filter such as GIRF may outperform our bagged methods;
an example is the Lorenz-96 model demonstrated in Sec.~\SuppSecLorenz.

\subsection{Correlated Brownian motion}

Suppose $\myvec{X}(t) = \Omega \myvec{W}(t)$ where $\myvec{W}(t)=W_{1:\Unit}(t)$ comprises $\Unit$ independent standard Brownian motions, and $\Omega_{\unit,\altUnit}=\rho^{\dist(\unit,\altUnit)}$ with $\dist(\unit,\altUnit)$ being the circle distance,
\[
\dist(\unit,\altUnit) 
= \min\big(|\unit-\altUnit|, |\unit-\altUnit+\Unit|, |\unit-\altUnit-\Unit|\big).
\]
Set $t_\time=\time$ for $\time=0,1,\dots,\Time$ with initial value $\myvec{X}(0)=\myvec{0}$ and suppose measurement errors are independent and normally distributed, $Y_{\unit\comma\time}= X_{\unit\comma\time}+ \eta_{\unit\comma\time}$ with
$\eta_{\unit\comma\time}\sim \normal(0,\tau^2)$.
The parameter $\rho$ determines the strength of the spatial coupling.

%%%%% bbbbbbbbbbbbbbbbbb

<<bm-settings,cache=FALSE,echo=F>>=

bm_files_dir <- paste0("bm_",bm_run_level,"/")
if(!dir.exists(bm_files_dir)) dir.create(bm_files_dir)

stew(file=paste0(bm_files_dir,"bm_settings.rda"),{

  # copy variables that should be included in the stew
  bm_run_level <- bm_run_level 
  bm_cores <- cores

  bm_tol <- 1e-300

  # run_level 1 for debugging; 2 for quick run; 3 for long run; 4 for production(?)

  if(bm_run_level==1){
    bm_U <- c(6, 4)
    bm_N <- 2
    bm_replicates <- 2 # number of Monte Carlo replicates
    bm_girf_Np <- 20
    bm_girf_lookahead <- 2
    bm_girf_nguide <- 10
    bm_pfilter_Np <- 100
    bm_abf_Nrep <- 3
    bm_abf_Np_per_replicate <- 10
    bm_ubf_Nrep <- 3
    bm_abfir_Nrep <- bm_abf_Nrep
    bm_abfir_Np_per_replicate <- bm_abf_Np_per_replicate
    bm_bpf_Np <- 50
    bm_bpf_units_per_block <- 2
    bm_enkf_Np <- 20
  } else if(bm_run_level==2){
    bm_U <- c(32,16,8,4)
    bm_N <- 20
    bm_replicates <- 5 # number of Monte Carlo replicates
    bm_girf_Np <- 1000
    bm_girf_lookahead <- 2
    bm_girf_nguide <- 50
    bm_pfilter_Np <- 10000
    bm_abf_Nrep <- 500
    bm_abf_Np_per_replicate <- 100
    bm_ubf_Nrep <- 5000
    bm_abfir_Nrep <- 200
    bm_abfir_Np_per_replicate <- 50
    bm_bpf_Np <- 5000
    bm_bpf_units_per_block <- 2
    bm_enkf_Np <- 5000
  } else if(bm_run_level==3){
    bm_replicates <- 5 # number of Monte Carlo replicates
    bm_U <- c(40,30,20,10,5)
    bm_N <- 50
    bm_girf_Np <- 1000
    bm_girf_lookahead <- 2
    bm_girf_nguide <- 50
    bm_pfilter_Np <- 50000
    bm_abf_Nrep <- 400
    bm_abf_Np_per_replicate <- 400
    bm_ubf_Nrep <- 10000
    bm_abfir_Nrep <- 200
    bm_abfir_Np_per_replicate <- 200
    bm_bpf_Np <- 10000
    bm_bpf_units_per_block <- 5
    bm_enkf_Np <- 5000    
  } else if(bm_run_level==4){
    bm_replicates <- 5 # number of Monte Carlo replicates
    bm_U <- c(100,80,60,40,30,20,10,5)
    bm_N <- 50
    bm_girf_Np <- 1000
    bm_girf_lookahead <- 2
    bm_girf_nguide <- 50
    bm_pfilter_Np <- 100000
    bm_abf_Nrep <- 400
    bm_abf_Np_per_replicate <- 400
    bm_ubf_Nrep <- 40000
    bm_abfir_Nrep <- 200
    bm_abfir_Np_per_replicate <- 200
    bm_bpf_Np <- 20000
    bm_bpf_units_per_block <- 3
    bm_enkf_Np <- 10000
  }

  set.seed(4512)
  bm_list <- foreach(u=bm_U) %do% bm(U=u, N=bm_N)

})

bm_jobs <- expand.grid(U=bm_U,reps=1:bm_replicates)
bm_jobs$U_id <- rep(seq_along(bm_U),times=bm_replicates)

@

<<bm_nbhd,echo=F>>=
bm_nbhd <- function(object, time, unit) {
  nbhd_list <- list()
  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
  if(unit>1) nbhd_list <- c(nbhd_list, list(c(unit-1, time)))
  if(unit>2) nbhd_list <- c(nbhd_list, list(c(unit-2, time)))
  return(nbhd_list)
}
@

<<bm_kf,cache=F,echo=F>>=
bm_kf <- stew(file=paste0(bm_files_dir,"bm_kf.rda"),seed=25487,{
  foreach(i=seq_along(bm_U),.combine=c)%do%{
    bm <- bm_list[[i]]
    bm_units <- bm_U[i]
    bm_dist <- function(u,v,U) min(abs(u-v),abs(u-v+U),abs(u-v-U))
    bm_dmat <- matrix(0,bm_units,bm_units)
    for(u in 1:bm_units) {
      for(v in 1:bm_units) {
        bm_dmat[u,v] <- bm_dist(u,v,bm_units)
      }
    }
    rootQ <- coef(bm)["rho"]^bm_dmat * coef(bm)["sigma"]
    kf <- pomp:::kalmanFilter(
      t=1:bm_N,
      y=obs(bm),
      X0=rinit(bm),
      A= diag(bm_units),
      Q= rootQ %*% rootQ,
      C=diag(bm_units),
      R=diag(coef(bm)["tau"]^2, nrow=bm_units)
    )$loglik
  } -> bm_kf_results
})

bm_jobs$logLik <- rep(bm_kf_results,times=bm_replicates)
@

<<bm_girf,cache=F,echo=F>>=
bm_girf <- stew(file=paste0(bm_files_dir,"bm_girf.rda"),seed=5981724,{
  foreach(job=iter(bm_jobs,"row")) %dopar% {
    system.time(
      girf(bm_list[[job$U_id]], method='adams',
        Np=bm_girf_Np,
        Ninter = job$U,
        lookahead = bm_girf_lookahead,
        Nguide = bm_girf_nguide,
        tol = bm_tol) -> bm_girf_out 
    ) -> bm_girf_time
#    uncomment for debugging 
#    list(logLik=logLik(bm_girf_out),time=bm_girf_time["elapsed"],bm_girf_out)
    list(logLik=logLik(bm_girf_out),time=bm_girf_time["elapsed"])
  } -> bm_girf_list
})
bm_jobs$girf_logLik <- vapply(bm_girf_list,function(x)x$logLik,numeric(1))
bm_jobs$girf_time <- vapply(bm_girf_list,function(x) x$time,numeric(1))
@

<<bm_apf,cache=F,echo=F,eval=F>>=
bm_apf <- stew(file=paste0(bm_files_dir,"bm_apf.rda"),seed=5981724,{
  foreach(job=iter(bm_jobs,"row")) %dopar% {
    system.time(
      girf(bm_list[[job$U_id]], method='adams',
        Np=bm_apf_Np,
        Ninter = 1,
        lookahead = 2,
        Nguide = bm_apf_nguide,
        tol = bm_tol 
      )-> bm_apf_out 
    ) -> bm_apf_time
#    uncomment for debugging 
#    list(logLik=logLik(bm_apf_out),time=bm_apf_time["elapsed"],bm_apf_out)
    list(logLik=logLik(bm_apf_out),time=bm_apf_time["elapsed"])
  } -> bm_apf_list
})

bm_jobs$apf_logLik <- vapply(bm_apf_list,function(x)x$logLik,numeric(1))
bm_jobs$apf_time <- vapply(bm_apf_list,function(x) x$time,numeric(1))

@

<<bm_abf,cache=F,echo=F>>=
bm_abf <- stew(file=paste0(bm_files_dir,"bm_abf.rda"),seed=844424,{
  foreach(job=iter(bm_jobs,"row")) %do% {
    system.time(
      abf(bm_list[[job$U_id]], 
        Nrep = bm_abf_Nrep,
        Np=bm_abf_Np_per_replicate,
        nbhd = bm_nbhd, tol=bm_tol) -> bm_abf_out 
    ) -> bm_abf_time
#    uncomment for debugging 
#    list(logLik=logLik(bm_abf_out),time=bm_abf_time["elapsed"],bm_abf_out)
    list(logLik=logLik(bm_abf_out),time=bm_abf_time["elapsed"])
  } -> bm_abf_list
})
bm_jobs$abf_logLik <- vapply(bm_abf_list,function(x)x$logLik,numeric(1))
bm_jobs$abf_time <- vapply(bm_abf_list,function(x) x$time,numeric(1))

@

<<bm_ubf,cache=F,echo=F>>=
bm_ubf <- stew(file=paste0(bm_files_dir,"bm_ubf.rda"),seed=844424,{
  foreach(job=iter(bm_jobs,"row")) %do% {
    system.time(
      abf(bm_list[[job$U_id]], 
        Nrep = bm_ubf_Nrep,
        Np=1,
        nbhd = bm_nbhd, tol=bm_tol) -> bm_ubf_out 
    ) -> bm_ubf_time
#    uncomment for debugging 
#    list(logLik=logLik(bm_ubf_out),time=bm_ubf_time["elapsed"],bm_ubf_out)
    list(logLik=logLik(bm_ubf_out),time=bm_ubf_time["elapsed"])
  } -> bm_ubf_list
})
bm_jobs$ubf_logLik <- vapply(bm_ubf_list,function(x)x$logLik,numeric(1))
bm_jobs$ubf_time <- vapply(bm_ubf_list,function(x) x$time,numeric(1))

@

<<bm_abfir,cache=F,echo=F>>=
bm_abfir <- stew(file=paste0(bm_files_dir,"bm_abfir.rda"),seed=53398,{
  foreach(job=iter(bm_jobs,"row")) %do% {
    system.time(
      abfir(bm_list[[job$U_id]], method='adams',
        Nrep = as.integer(bm_abfir_Nrep),
        Np=bm_abfir_Np_per_replicate,
        Ninter = as.integer(job$U/2),
        nbhd = bm_nbhd, tol=bm_tol) -> bm_abfir_out 
    ) -> bm_abfir_time
#    uncomment for debugging 
#    list(logLik=logLik(bm_abfir_out),time=bm_abfir_time["elapsed"],bm_abfir_out)
    list(logLik=logLik(bm_abfir_out),time=bm_abfir_time["elapsed"])
  } -> bm_abfir_list
})
bm_jobs$abfir_logLik <- vapply(bm_abfir_list,function(x)x$logLik,numeric(1))
bm_jobs$abfir_time <- vapply(bm_abfir_list,function(x) x$time,numeric(1))

@

<<bm_pfilter,cache=F,echo=F>>=
bm_pfilter <- stew(file=paste0(bm_files_dir,"bm_pfilter.rda"),seed=53285,{
  foreach(job=iter(bm_jobs,"row")) %dopar% {
    system.time(
      pfilter(bm_list[[job$U_id]],Np=bm_pfilter_Np, tol=bm_tol) -> bm_pfilter_out
    ) -> bm_pfilter_time
#    uncomment for debugging 
#    list(logLik=logLik(bm_pfilter_out),time=bm_pfilter_time["elapsed"],bm_pfilter_out)
    list(logLik=logLik(bm_pfilter_out),time=bm_pfilter_time["elapsed"])
  } -> bm_pfilter_list
})
bm_jobs$pfilter_logLik <- vapply(bm_pfilter_list,function(x)x$logLik,numeric(1))
bm_jobs$pfilter_time <- vapply(bm_pfilter_list,function(x) x$time,numeric(1))
@

<<bm_bpf,cache=F,echo=F>>=
#
#  to test:
#  bpfilter(bm_list[[1]],50, block_size=1) -> tmp
#
bm_bpf <- stew(file=paste0(bm_files_dir,"bm_bpf.rda"),seed=53285,{
  foreach(job=iter(bm_jobs,"row")) %dopar% {
    system.time(
      bpfilter(bm_list[[job$U_id]],
        Np=bm_bpf_Np,
	block_size=bm_bpf_units_per_block) -> bm_bpf_out
    ) -> bm_bpf_time
#    uncomment for debugging 
#    list(logLik=logLik(bm_bpf_out),time=bm_bpf_time["elapsed"],bm_bpf_out)
#
# replace using logLik(bm_bpf_out) when that method exists
#
    list(logLik=bm_bpf_out@loglik,time=bm_bpf_time["elapsed"])
  } -> bm_bpf_list
})
bm_jobs$bpf_logLik <- vapply(bm_bpf_list,function(x)x$logLik,numeric(1))
bm_jobs$bpf_time <- vapply(bm_bpf_list,function(x) x$time,numeric(1))
@

<<bm_enkf,cache=F,echo=F>>=
bm_enkf <- stew(file=paste0(bm_files_dir,"bm_enkf.rda"),seed=53285,{
  foreach(job=iter(bm_jobs,"row")) %dopar% {
    system.time(
      enkf(bm_list[[job$U_id]],
        Np=bm_enkf_Np) -> bm_enkf_out
    ) -> bm_enkf_time
    list(logLik=bm_enkf_out@loglik,time=bm_enkf_time["elapsed"])
  } -> bm_enkf_list
})
bm_jobs$enkf_logLik <- vapply(bm_enkf_list,function(x)x$logLik,numeric(1))
bm_jobs$enkf_time <- vapply(bm_enkf_list,function(x) x$time,numeric(1))
@

<<bm_alt_plot, echo=F, fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('log likelihood estimates for a correlated Brownian motion model of various dimensions. {\\UBF}, {\\ABF} and {\\ABFIR} are compared with a guided intermediate resampling filter (GIRF), standard particle filter (PF), block particle filter (BPF) and ensemble Kalman filter (EnKF). The exact likelihood was computed via a Kalman filter (KF).')>>=

# put output in tall format for plotting
#bm_methods <- c("APF","ABF", "ABFIR","GIRF","KF","PF")
bm_methods <- c("ABF", "ABF-IR","UBF","GIRF","KF","PF","BPF","EnKF")
bm_results <- data.frame(
  Method=rep(bm_methods,each=nrow(bm_jobs)),
  logLik=c(
#    bm_jobs$apf_logLik,
    bm_jobs$abf_logLik,
    bm_jobs$abfir_logLik,
    bm_jobs$ubf_logLik,
    bm_jobs$girf_logLik,
    bm_jobs$logLik,
    bm_jobs$pfilter_logLik,
    bm_jobs$bpf_logLik,
    bm_jobs$enkf_logLik
  ),
  Units=rep(bm_jobs$U,reps=length(bm_methods))
)
bm_point_perturbation <- 1.5
bm_results$U <- bm_results$Units+
  rep( bm_point_perturbation*seq(from=-1,to=1,length=length(bm_methods)),
    each=nrow(bm_jobs))
bm_results$logLik_per_unit <- bm_results$logLik/bm_results$Units
bm_results$logLik_per_obs <- bm_results$logLik_per_unit/bm_N

save(file=paste0(bm_files_dir,"bm_results.rda"),bm_results,bm_jobs)

if(0){
  ggplot(bm_results,mapping = aes(x = U, y = logLik_per_unit)) +
  geom_point(aes(color = Method)) +
  scale_y_continuous(limits=c(-108,NA)) +
# scale_y_continuous() +
  ylab("log likelihood per unit")
}

bm_max <- max(bm_results$logLik_per_obs)

cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

ggplot(bm_results,mapping = aes(x = U, y = logLik_per_obs, group=Method,color=Method,shape=Method,linetype=Method)
) +
#  geom_point(aes(shape=Method),size=1) +
#  geom_line(aes(linetype=Method), size=1) +
  scale_linetype_manual(values=c(1,1,1,1,2,2,2,2)) +
  scale_shape_manual(values=c(1,2,4,5,1,2,4,5)) +
  scale_color_manual(values=cbPalette[c(2,3,4,5,6,1,7,8)])+
  geom_point() +
  stat_summary(fun=mean, geom="line") +
  coord_cartesian(ylim=c(bm_max-0.35,bm_max))+
  theme(legend.key.width = unit(1,"cm"))+
  ylab("log likelihood per unit per time")

@

Fig.~\ref{fig:bm_alt_plot} shows how the bagged filters scale on this Gaussian model, compared to a standard particle filter (PF), a guided intermediate resampling filter (GIRF), a block particle filter (BPF), and an ensemble Kalman filter.
For our numerical results, we use $\tau=\Sexpr{coef(bm)['tau']}$, $\rho=\Sexpr{coef(bm)['rho']}$ and $\Time=\Sexpr{length(time(bm))}$.
The algorithmic parameters and run times are listed in Sec.~\SuppSecBM, together with a plot of the simulated data and supplementary discussion.
In this case, the exact likelihood is computable via the Kalman filter (KF). 
Since EnKF is based on a Gaussian approximation, it is also exact in this case, up to a small Monte Carlo error. 
The GIRF framework encompasses lookahead particle filter techniques, such as the auxiliary particle filter \citep{pitt99}, and intermediate resampling techniques \citep{delmoral17}. 
GIRF methods combining these techniques were found to perform better than either of these component techniques alone \citep{park20}.
Thus, GIRF here represents a state-of-the-art auxiliary particle filter that targets the complete joint filter density for all units.
We use the general-purpose, plug-and-play implementation of GIRF provided by the \texttt{spatPomp} R package \citep{asfaw20github};
 this implementation of GIRF is thus not tailored in any way to exploit special features of this toy model.
PF works well for small values of $\Unit$ in Fig.~\ref{fig:bm_alt_plot}, but rapidly starts struggling as $\Unit$ increases.
GIRF behaves comparably to PF for small $\Unit$ but its performance is maintained for larger $\Unit$.
{\ABF} and {\ABFIR} have some efficiency loss, for small $\Unit$, relative to PF and GIRF due to the localization involved in the filter weighting, but for large $\Unit$ this cost is paid back by the benefit of the reduced Monte Carlo variability.
{\UBF} has a larger efficiency loss for small $\Unit$, but its favorable scaling properties lead it to overtake {\ABF} for larger $\Unit$.
BPF shows stable scaling and modest efficiency loss.
This linear Gaussian SpatPOMP model provides a simple scenario to demonstrate scaling behavior.
For filters that cannot take direct advantage of the Gaussian property of the model, we see that there is a tradeoff between efficiency at low $\Unit$ and scalability.
This is unavoidable, since there is no known algorithm that is simultaneously fully efficient (up to Monte Carlo error), scalable, and applicable to general SpatPOMP models.
We now explore this tradeoff empirically on to a more complex SpatPOMP exemplifying the nonlinear non-Gaussian models motivating our new filtering approach.

%%% mmmmmmmmmmmmmmmmmmmmmm

\subsection{Spatiotemporal measles epidemics}

Data analysis for spatiotemporal systems featuring nonlinear, nonstationary mechanisms and partial observability has been a longstanding open challenge for ecological and epidemiological analysis \citep{bjornstad01}.
A compartment modeling framework for spatiotemporal population dynamics divides the population at each spatial location into categories, called compartments, which are modeled as homogeneous. 
Spatiotemporal compartment models can be called patch models or metapopulation models in an ecological context.
Ensemble Kalman filter (EnKF) methods provide a state-of-the-art approach to inference for metapopulation models \citep{li20} despite concerns that the approximations inherent in the EnKF can be problematic for models that are highly nonlinear or non-Gaussian \citep{ades15}.
Our bagged filter methodologies have theoretical guarantees for arbitrarily nonlinear and non-Gaussian models, while having improved scaling properties compared to particle filters.

We consider a spatiotemporal model for disease transmission dynamics of measles within and between multiple cities, based on the model of \citet{park20} which adds spatial interaction to the compartment model presented by \cite{he10}.
The model compartmentalizes the population of each city into susceptible ($S$), exposed ($E$), infectious ($I$), and recovered/removed ($R$) categories.
The number of individuals in each compartment city $\unit$ at time $t$ are denoted by integer-valued random variables $S_\unit(t)$, $E_\unit(t)$, $I_\unit(t)$, and $R_\unit(t)$.
The population dynamics are written in terms of counting processes $N_{\bullet\bullet,\unit}(t)$ enumerating cumulative transitions in city $\unit$, up to time $t$, between compartments identified by the subscripts.
Our model is described by the following system of stochastic differential equations, for $\unit=1,\dots, \Unit$,
\begin{equation}
\nonumber
%\label{eq:measles:system}
\begin{array}{lllllll}
\displaystyle dS_\unit(t) &=& dN_{BS,\unit}(t) &-& dN_{SE,\unit}(t) &-& dN_{SD,\unit}(t) \\
\displaystyle dE_\unit(t) &=& dN_{SE,\unit}(t) &-& dN_{EI,\unit}(t) &-& dN_{ED,\unit}(t) \\
\displaystyle dI_\unit(t) &=& dN_{EI,\unit}(t) &-& dN_{IR,\unit}(t) &-& dN_{ID,\unit}(t) 
\end{array}
\end{equation}
Here,  $N_{BS,\unit}(t)$ models recruitment into the susceptible population, and $N_{\bullet D,\unit}(t)$ models emigration and death. 
The total population $P_\unit(t)=S_\unit(t)+E_\unit(t)+I_\unit(t)+R_\unit(t)$ is calculated by smoothing census data and is treated as known.
The number of recovered individuals $R_\unit(t)$ in city $\unit$ is therefore defined implicitly.
$N_{SE,\unit}(t)$ is modeled as negative binomial death processes \citep{breto09,breto11}
with over-dispersion parameter $\sigma_{SE}$, and rate given by
\begin{eqnarray}
\nonumber
\mathbb{E} \big[ N_{SE,\unit}(t+dt) - N_{SE,\unit}(t) \big] 
&=& 
\beta(t) \, S_\unit(t) 
\Big[ 
  \left( \frac{I_\unit+\iota}{P_\unit} \right)^\alpha
\\
\label{eq:dEdt}
&& \hspace{-3.8cm}
 + \sum_{\altUnit \neq \unit} \frac{v_{\unit\altUnit}}{P_\unit} 
  \left\{ 
    \left(
      \frac{ I_{\altUnit} }{ P_{\altUnit} }
    \right)^\alpha - 
    \left(
      \frac{I_\unit}{P_\unit}  
    \right)^\alpha
  \right\}
\Big] dt + o(dt),\label{eqn:transmissionrate}
\end{eqnarray}
where $\beta(t)$ models seasonality driven by high contact rates between children at school, described by
\begin{equation}
\nonumber
  \beta(t)=\begin{cases}
\big(1+\amplitude(1-p)p^{-1} \big)\, \meanBeta & \mbox{ during school term},\\
\big( 1-\amplitude\big) \, \meanBeta& \mbox{ during vacation}
  \end{cases} \label{eq:term}
\end{equation}
with $p = 0.759$ being the proportion of the year taken up by the school terms, $\meanBeta$ is the mean transmission rate, and $\amplitude$ measures the reduction of transmission during school holidays.
In \myeqref{eq:dEdt}, $\alpha$ is a mixing exponent modeling inhomogeneous contact rates within a city, and $\iota$ models immigration of infected individuals which is appropriate when analyzing a subset of cities that cannot be treated as a closed system.
The number of travelers from city $\unit$ to $\altUnit$ is denoted by $v_{\unit\altUnit}$. 
Here, $v_{\unit\altUnit}$ is constructed using the gravity model of \cite{xia04}, 
\[
v_{\unit\altUnit} = \gravity \cdot \frac{\;\overline{\dist}\;}{\bar{P}^2} \cdot \frac{P_\unit \cdot P_{\altUnit}}{\dist(\unit,\altUnit)},
\]
where $\dist(\unit,\altUnit)$ denotes the distance between city $\unit$ and city $\altUnit$, $P_\unit$ is the average population for city $\unit$ across time, $\bar{P}$ is the average population across cities, and $\overline{\dist}$ is the average distance between a randomly chosen pair of cities.
Here, we model $v_{\unit\altUnit}$ as fixed through time and symmetric between any two arbitrary cities, though a natural extension would allow for temporal variation and asymmetric movement between two cities.
The transition processes $N_{EI,\unit}(t)$, $N_{EI,\unit}(t)$ and $N_{\bullet D,\unit}(t)$ are modeled as conditional Poisson processes with per-capita rates $\mu_{EI}$, $\mu_{IR}$ and $\mu_{\bullet D}$ respectively, and we fix $\mu_{\bullet D}=50 \mbox{ year}^{-1}$.
The birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson processes with rate $\mu_{BS,\unit}(t)$, given by interpolated census data.

To complete the model specification, we must describe the measurement process.
Let $Z_{\unit\comma\time}=N_{IR\comma\unit}(t_\time)-N_{IR\comma\unit}(t_{\time-1})$ be the number of removed infected individuals in the $n$th reporting interval.
Suppose that cases are quarantined once they are identified, so that reported cases comprise a fraction $\rho$ of these removal events.
The case report $\data{y}_{\unit\comma\time}$ is modeled as a realization of a discretized conditionally Gaussian random variable $Y_{\unit\comma\time}$, defined for $y>0$ via
\begin{eqnarray}
\nonumber
\prob\big[Y_{\unit\comma\time}{=}y\mid Z_{\unit\comma\time}{=}z\big] &=& \Phi\big(y+0.5; \rho z,\rho(1-\rho)z+\psi^2\rho^2z^2\big)
\\
&&\hspace{-20mm}
- \Phi\big(y-0.5; \rho z,\rho(1-\rho)z+\psi^2\rho^2z^2\big)
\label{eq:obs}
\end{eqnarray}
where $\Phi(\cdot;\mu,\sigma^2)$ is the $\normal(\mu,\sigma^2)$ cumulative distribution function, and $\psi$ models overdispersion relative to the binomial distribution.
For $y=0$, we replace $y-0.5$ by $-\infty$ in \myeqref{eq:obs}.

%%%%%%  msmsmsmsmsmsmsmsmsmsms %%%%%%%%%%%%

<<measles_spatPomp,eval=T,echo=F>>=
library(spatPomp)
measles_model_dir <- "measlesModel/"
if(!dir.exists(measles_model_dir)) dir.create(measles_model_dir)

measles_spatPomp <- stew(file=paste0(measles_model_dir,"measles_spatPomp.rda"),{
  measles_U <- 40    # number of cities (units) for simulation that can subsequently be subsetted
  measles_N <- 15*26 # number of 2-week observation intervals for simulation that can subsequently be subsetted
  measles_unit_statenames <- c('S','E','I','R', 'C','W')
  measles_S_0 <- 0.032; measles_E_0 <- 0.00005; measles_I_0 <- 0.00004
  measles_unit_params <- c(
    alpha=1,
    iota=0,  # set to zero for a closed population
    R0=30,
    cohort=0,
    amplitude=0.5, gamma=52, sigma=52,mu=0.02,
    sigmaSE=0.15, rho=0.5,
    psi=0.15,
    g=400,
    S_0=measles_S_0, E_0=measles_E_0, I_0=measles_I_0,
    R_0=1-measles_S_0-measles_E_0-measles_I_0
  )
  measles_uk <- measles(measles_U)
  measles_statenames <- paste0(rep(measles_unit_statenames,
    each=measles_U),1:measles_U)
  measles_IVPnames <- paste0(measles_statenames[1:(4*measles_U)],"_0")
  measles_RPnames <- c("alpha","iota","R0","cohort","amplitude",
    "gamma","sigma","mu","sigmaSE","rho","psi","g")
  measles_paramnames <- c(measles_RPnames,measles_IVPnames)

  measles_params <- rep(NA,length(measles_paramnames))
  names(measles_params) <- measles_paramnames

  measles_params[measles_RPnames] <- measles_unit_params[measles_RPnames]
  measles_params[paste0("S",1:measles_U,"_0")] <-measles_unit_params["S_0"]
  measles_params[paste0("E",1:measles_U,"_0")] <-measles_unit_params["E_0"]
  measles_params[paste0("I",1:measles_U,"_0")] <-measles_unit_params["I_0"]
  measles_params[paste0("R",1:measles_U,"_0")] <-measles_unit_params["R_0"]

  set.seed(34)
  measles_sim <- simulate(measles_uk,params=measles_params,
    times=time(measles_uk)[1:measles_N])

  measles_subset <- function(m_U,m_N){
    m <- measles(U=m_U)
    m@data <- measles_sim@data[1:m_U,1:m_N]
    time(m) <- measles_sim@times[1:m_N]
    m_statenames <- paste0(rep(measles_unit_statenames,each=m_U),1:m_U)
    m_IVPnames <- paste0(m_statenames[1:(4*m_U)],"_0")
    m_paramnames <- c(measles_RPnames,m_IVPnames)
    m_params <- measles_params[names(measles_params)%in%m_paramnames]
    coef(m) <- m_params
    return(m)
  }
})
@

<<measles_image_plot, echo=F, fig.height=6, fig.width=7, out.width="6.5in", fig.cap = paste('Log(reported cases $+$ 1) for (A) the measles simulation used for the likelihood slice; (B) the corresponding UK measles data. The simulation shares the biennial pattern, with most but not all cities locked in phase most of the time.')>>=
library(spatPomp)
library(fields)
par(mai=c(0.5,0.7,0.1,0.1))
par(mfrow=c(2,1))
image.plot(y=1:dim(obs(measles_sim))[1],x=time(measles_sim),z=log(t(obs(measles_sim)+1)),ylab="unit",xlab="",col=gray.colors(33))
mtext("A",side=2,line=2.5,las=1,padj=-5.5,cex=1.8)
image.plot(y=1:dim(obs(measles_uk))[1],x=time(measles_uk),z=log(t(obs(measles_uk)+1)),ylab="unit",xlab="",col=gray.colors(33))
mtext("B",side=2,line=2.5,las=1,padj=-5.5,cex=1.8)
mtext("time",side=1,line=2.5)
@

<<mscale_settings,cache=FALSE,echo=F>>=

  # run_level 1 for debugging; 2 for quick run; 3 for long run;
  # 4 even longer; 5 for production(?)
  # 6 for a small-U test figure

mscale_files_dir <- paste0("mscale_",mscale_run_level,"/")
if(!dir.exists(mscale_files_dir)) dir.create(mscale_files_dir)

stew(file=paste0(mscale_files_dir,"mscale_settings.rda"),{

  # copy variables that should be included in the stew
  mscale_run_level <- mscale_run_level 
  mscale_cores <- cores

  mscale_tol <- 1e-300

  if(mscale_run_level==1){
    mscale_U <- c(4, 2)
    mscale_N <- 5
    mscale_replicates <- 2 # number of Monte Carlo replicates
    mscale_girf_Np <- 50
    mscale_girf_lookahead <- 2
    mscale_girf_nguide <- 10
    mscale_pfilter_Np <- 100
    mscale_abf_Nrep <- 3
    mscale_abf_Np_per_replicate <- 10
    mscale_ubf_Nrep <- 20
    mscale_abfir_Nrep <- mscale_abf_Nrep
    mscale_abfir_Np_per_replicate <- mscale_abf_Np_per_replicate
    mscale_enkf_Np <- 100
    mscale_bpf_units_per_block <- 1
    mscale_bpf_Np <- 50
    mscale_bootgirf_Np <- 50
    mscale_bootgirf_nguide <- 10
    mscale_bootgirf_lookahead <- 2
  } else if(mscale_run_level==2){
    mscale_U <- c(16,8,4,2)
    mscale_N <- 4*26
    mscale_replicates <- 4 # number of Monte Carlo replicates
    mscale_girf_Np <- 1000
    mscale_girf_lookahead <- 1
    mscale_girf_nguide <- 50
    mscale_pfilter_Np <- 10000
    mscale_abf_Nrep <- 500
    mscale_abf_Np_per_replicate <- 100
    mscale_ubf_Nrep <- 5000
    mscale_abfir_Nrep <- 200
    mscale_abfir_Np_per_replicate <- 50
    mscale_enkf_Np <- 5000
    mscale_bpf_units_per_block <- 1
    mscale_bpf_Np <- 5000
    mscale_bootgirf_Np <- 1000
    mscale_bootgirf_nguide <- 20
    mscale_bootgirf_lookahead <- 2
  } else if(mscale_run_level==3){
    mscale_replicates <- 5 # number of Monte Carlo replicates
    mscale_U <- c(32, 16, 8, 4, 2)
    mscale_N <- 5*26
    mscale_girf_Np <- 2000
    mscale_girf_lookahead <- 2
    mscale_girf_nguide <- 20
    mscale_pfilter_Np <- 50000
    mscale_abf_Nrep <- 500
    mscale_abf_Np_per_replicate <- 100
    mscale_ubf_Nrep <- 5000
    mscale_abfir_Nrep <- 200
    mscale_abfir_Np_per_replicate <- 100
    mscale_enkf_Np <- 10000
    mscale_bpf_units_per_block <- 1
    mscale_bpf_Np <- 10000
    mscale_bootgirf_Np <- 2000
    mscale_bootgirf_nguide <- 20
    mscale_bootgirf_lookahead <- 2
  } else if(mscale_run_level==4){
    mscale_replicates <- 5 # number of Monte Carlo replicates
    mscale_U <- c(20,16,12,10,8,6,4,2)
    mscale_N <- 52
    mscale_girf_Np <- 2000
    mscale_girf_lookahead <- 1
    mscale_girf_nguide <- 50
    mscale_pfilter_Np <- 100000
    mscale_abf_Nrep <- 500
    mscale_abf_Np_per_replicate <- 500
    mscale_ubf_Nrep <- 10000
    mscale_abfir_Nrep <- 200
    mscale_abfir_Np_per_replicate <- 200
    mscale_enkf_Np <- 10000
    mscale_bpf_units_per_block <- 1
    mscale_bpf_Np <- 50
  } else if(mscale_run_level==5){
    mscale_replicates <- 5 # number of Monte Carlo replicates
    mscale_U <- c(40,36,32,28,24,20,16,14,12,10,8,6,4,2)
    mscale_N <- 5*26
    mscale_girf_Np <- 2000
    mscale_girf_lookahead <- 1
    mscale_girf_nguide <- 40
    mscale_pfilter_Np <- 100000
    mscale_abf_Nrep <- 500
    mscale_abf_Np_per_replicate <- 500
    mscale_ubf_Nrep <- 20000
    mscale_abfir_Nrep <- 200
    mscale_abfir_Np_per_replicate <- 200
    mscale_enkf_Np <- 10000
    mscale_bpf_units_per_block <- 4
    mscale_bpf_Np <- 20000
  }  else if(mscale_run_level==6){
    mscale_replicates <- 5 # number of Monte Carlo replicates
    mscale_U <- c(8,6,4,3,2)
    mscale_N <- 52
    mscale_girf_Np <- 2000
    mscale_girf_lookahead <- 1
    mscale_girf_nguide <- 50
    mscale_pfilter_Np <- 100000
    mscale_abf_Nrep <- 500
    mscale_abf_Np_per_replicate <- 200
    mscale_ubf_Nrep <- 40000
#    mscale_abfir_Nrep <- 200
#    mscale_abfir_Np_per_replicate <- 200
    mscale_abfir_Nrep <- 10
    mscale_abfir_Np_per_replicate <- 10
    mscale_enkf_Np <- 1000
    mscale_bpf_units_per_block <- 1
    mscale_bpf_Np <- 2000
  }

})

mscale_jobs <- expand.grid(U=mscale_U,reps=1:mscale_replicates)
mscale_jobs$U_id <- rep(seq_along(mscale_U),times=mscale_replicates)

@

<<mscale_spatPomp,eval=T,echo=F>>=
mscale_spatPomp <- stew(file=paste0(mscale_files_dir,"mscale_spatPomp.rda"),{

  mscale_model_dir <-  measles_model_dir
  # note: care is required if mscale_model_dir is set to something other than measles_model_dir
  # however, it may be useful to do that while testing model variations.

  load(file=paste0(mscale_model_dir,"measles_spatPomp.rda"))  
  mscale_list <- vector("list", length = length(mscale_U))  
  for(i in 1:length(mscale_U)) mscale_list[[i]] <- measles_subset(m_U=mscale_U[i], m_N=mscale_N)
})
@

<<mscale_nbhd,echo=F>>=
mscale_nbhd <- function(object, time, unit) {
  nbhd_list <- list()
  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
  return(nbhd_list)
}
@

<<mscale_girf,cache=F,echo=F>>=
mscale_girf <- stew(file=paste0(mscale_files_dir,"mscale_girf.rda"),seed=5981724,{
  foreach(job=iter(mscale_jobs,"row")) %dopar% {
    system.time(
      girf(mscale_list[[job$U_id]], method='adams',
        Np=mscale_girf_Np,
        Ninter = job$U,
        lookahead = mscale_girf_lookahead,
        Nguide = mscale_girf_nguide,
        tol = mscale_tol) -> mscale_girf_out 
    ) -> mscale_girf_time
#    uncomment for debugging 
#    list(logLik=logLik(mscale_girf_out),time=mscale_girf_time["elapsed"],mscale_girf_out)
    list(logLik=logLik(mscale_girf_out),time=mscale_girf_time["elapsed"])
  } -> mscale_girf_list
})
mscale_jobs$girf_logLik <- vapply(mscale_girf_list,function(x)x$logLik,numeric(1))
mscale_jobs$girf_time <- vapply(mscale_girf_list,function(x) x$time,numeric(1))
@

<<mscale_bootgirf,cache=F,echo=F>>=
mscale_bootgirf <- stew(file=paste0(mscale_files_dir,"mscale_bootgirf.rda"),seed=5981724,{
### begin test
#mt <- mscale_list[[2]]
#bootgirf2(mt, method='adams',
#        Np=100,
#        Ninter = 4,
#        lookahead = 2,
#        Nguide = 10) -> bt
### end test
  foreach(job=iter(mscale_jobs,"row")) %dopar% {
    system.time(
# job <- mscale_jobs[1,] # for debugging
      bootgirf2(mscale_list[[job$U_id]], method='adams',
        Np=mscale_bootgirf_Np,
        Ninter = job$U,
        lookahead = mscale_bootgirf_lookahead,
        Nguide = mscale_bootgirf_nguide,
        tol = mscale_tol) -> mscale_bootgirf_out
   ) -> mscale_bootgirf_time   
#    uncomment for debugging 
#    list(logLik=logLik(mscale_bootgirf_out),time=mscale_bootgirf_time["elapsed"],mscale_bootgirf_out)
    list(logLik=logLik(mscale_bootgirf_out),time=mscale_bootgirf_time["elapsed"])
  } -> mscale_bootgirf_list
})
mscale_jobs$bootgirf_logLik <- vapply(mscale_bootgirf_list,function(x)x$logLik,numeric(1))
mscale_jobs$bootgirf_time <- vapply(mscale_bootgirf_list,function(x) x$time,numeric(1))
@

<<mscale_abf,cache=F,echo=F>>=
mscale_abf <- stew(file=paste0(mscale_files_dir,"mscale_abf.rda"),seed=844424,{
  foreach(job=iter(mscale_jobs,"row")) %do% {
    system.time(
      abf(mscale_list[[job$U_id]], 
        Nrep = mscale_abf_Nrep,
        Np=mscale_abf_Np_per_replicate,
        nbhd = mscale_nbhd, tol=mscale_tol) -> mscale_abf_out 
    ) -> mscale_abf_time
#    uncomment for debugging 
#    list(logLik=logLik(mscale_abf_out),time=mscale_abf_time["elapsed"],mscale_abf_out)
    list(logLik=logLik(mscale_abf_out),time=mscale_abf_time["elapsed"])
  } -> mscale_abf_list
})
mscale_jobs$abf_logLik <- vapply(mscale_abf_list,function(x)x$logLik,numeric(1))
mscale_jobs$abf_time <- vapply(mscale_abf_list,function(x) x$time,numeric(1))

@

<<mscale_ubf,cache=F,echo=F>>=
mscale_ubf <- stew(file=paste0(mscale_files_dir,"mscale_ubf.rda"),seed=844424,{
  foreach(job=iter(mscale_jobs,"row")) %do% {
    system.time(
      abf(mscale_list[[job$U_id]], 
        Nrep = mscale_ubf_Nrep,
        Np=1,
        nbhd = mscale_nbhd, tol=mscale_tol) -> mscale_ubf_out 
    ) -> mscale_ubf_time
#    uncomment for debugging 
#    list(logLik=logLik(mscale_ubf_out),time=mscale_ubf_time["elapsed"],mscale_ubf_out)
    list(logLik=logLik(mscale_ubf_out),time=mscale_ubf_time["elapsed"])
  } -> mscale_ubf_list
})
mscale_jobs$ubf_logLik <- vapply(mscale_ubf_list,function(x)x$logLik,numeric(1))
mscale_jobs$ubf_time <- vapply(mscale_ubf_list,function(x) x$time,numeric(1))

@

<<mscale_abfir,cache=F,echo=F>>=
mscale_abfir <- stew(file=paste0(mscale_files_dir,"mscale_abfir.rda"),seed=53398,{
  foreach(job=iter(mscale_jobs,"row")) %do% {
    system.time(
      abfir(mscale_list[[job$U_id]], method='adams',
        Nrep = as.integer(mscale_abfir_Nrep),
        Np=mscale_abfir_Np_per_replicate,
        Ninter = as.integer(job$U/2),
        nbhd = mscale_nbhd, tol=mscale_tol) -> mscale_abfir_out 
    ) -> mscale_abfir_time
#    uncomment for debugging 
#    list(logLik=logLik(mscale_abfir_out),time=mscale_abfir_time["elapsed"],mscale_abfir_out)
    list(logLik=logLik(mscale_abfir_out),time=mscale_abfir_time["elapsed"])
  } -> mscale_abfir_list
})
mscale_jobs$abfir_logLik <- vapply(mscale_abfir_list,function(x)x$logLik,numeric(1))
mscale_jobs$abfir_time <- vapply(mscale_abfir_list,function(x) x$time,numeric(1))

@

<<mscale_pfilter,cache=F,echo=F>>=
mscale_pfilter <- stew(file=paste0(mscale_files_dir,"mscale_pfilter.rda"),seed=53285,{
  foreach(job=iter(mscale_jobs,"row")) %dopar% {
    system.time(
      pfilter(mscale_list[[job$U_id]],Np=mscale_pfilter_Np) -> mscale_pfilter_out
    ) -> mscale_pfilter_time
#    uncomment for debugging 
#    list(logLik=logLik(mscale_pfilter_out),time=mscale_pfilter_time["elapsed"],mscale_pfilter_out)
    list(logLik=logLik(mscale_pfilter_out),time=mscale_pfilter_time["elapsed"])
  } -> mscale_pfilter_list
})
mscale_jobs$pfilter_logLik <- vapply(mscale_pfilter_list,function(x)x$logLik,numeric(1))
mscale_jobs$pfilter_time <- vapply(mscale_pfilter_list,function(x) x$time,numeric(1))
@

<<mscale_bpf,cache=F,echo=F>>=
mscale_bpf <- stew(file=paste0(mscale_files_dir,"mscale_bpf.rda"),seed=53285,{
  foreach(job=iter(mscale_jobs,"row")) %dopar% {
    system.time(
      bpfilter(mscale_list[[job$U_id]],
        Np=mscale_bpf_Np,
	block_size=mscale_bpf_units_per_block) -> mscale_bpf_out
    ) -> mscale_bpf_time
#    uncomment for debugging 
#    list(logLik=logLik(mscale_bpf_out),time=mscale_bpf_time["elapsed"],mscale_bpf_out)
    list(logLik=mscale_bpf_out@loglik,time=mscale_bpf_time["elapsed"])
  } -> mscale_bpf_list
})
mscale_jobs$bpf_logLik <- vapply(mscale_bpf_list,function(x)x$logLik,numeric(1))
mscale_jobs$bpf_time <- vapply(mscale_bpf_list,function(x) x$time,numeric(1))
@

<<mscale_enkf,cache=F,echo=F>>=
###  genkf(mscale_list[[1]],Np=mscale_enkf_Np) -> mscale_enkf_out
mscale_enkf <- stew(file=paste0(mscale_files_dir,"mscale_enkf.rda"),seed=53285,{
  foreach(job=iter(mscale_jobs,"row")) %dopar% {
    system.time(
      enkf(mscale_list[[job$U_id]],Np=mscale_enkf_Np) -> mscale_enkf_out
    ) -> mscale_enkf_time
#    uncomment for debugging 
#    list(logLik=logLik(mscale_enkf_out),time=mscale_enkf_time["elapsed"],mscale_enkf_out)
    list(logLik=logLik(mscale_enkf_out),time=mscale_enkf_time["elapsed"])
  } -> mscale_enkf_list
})
mscale_jobs$enkf_logLik <- vapply(mscale_enkf_list,function(x)x$logLik,numeric(1))
mscale_jobs$enkf_time <- vapply(mscale_enkf_list,function(x) x$time,numeric(1))
@

<<mscale_loglik_plot, echo=F,  fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('log likelihood estimates for simulated data from the measles model of various dimensions. {\\UBF}, {\\ABF} and {\\ABFIR} are compared with a guided intermediate resampling filter (GIRF), a standard particle filter (PF), a block particle filter (BPF) and an ensemble Kalman filter (EnKF).')>>=

# put output in tall format for plotting

#mscale_methods <- c("ABF", "ABF-IR","UBF","GIRF","PF","BPF","EnKF")
mscale_methods <- c("ABF", "ABF-IR","UBF","GIRF","PF","BPF","EnKF","bootgirf")
mscale_results <- data.frame(
  Method=rep(mscale_methods,each=nrow(mscale_jobs)),
  logLik=c(
    mscale_jobs$abf_logLik,
    mscale_jobs$abfir_logLik,
    mscale_jobs$ubf_logLik,
    mscale_jobs$girf_logLik,
    mscale_jobs$pfilter_logLik,
    mscale_jobs$bpf_logLik,
    mscale_jobs$enkf_logLik,
    mscale_jobs$bootgirf_logLik
  ),
  Units=rep(mscale_jobs$U,reps=length(mscale_methods))
)
mscale_point_perturbation <- 0.25
mscale_results$U <- mscale_results$Units+
  rep( mscale_point_perturbation*seq(from=-1,to=1,length=length(mscale_methods)),
    each=nrow(mscale_jobs))
mscale_results$logLik_per_unit <- mscale_results$logLik/mscale_results$Units
mscale_results$logLik_per_obs <- mscale_results$logLik_per_unit/mscale_N

save(file=paste0(mscale_files_dir,"mscale_results.rda"),mscale_results,mscale_jobs)

mscale_max <- max(mscale_results$logLik_per_obs)

ggplot(mscale_results,mapping = aes(x = U, y = logLik_per_obs, group=Method,color=Method,linetype=Method,shape=Method)) +
  scale_linetype_manual(values=c(1,1,1,1,2,2,2,3)) +
  scale_shape_manual(values=c(1,2,4,5,1,2,4,3)) +
#  scale_color_manual(values=c(1,2,3,4,5,6,8))+
  scale_color_manual(values=cbPalette[c(2:8,1)])+
  geom_point() +
  stat_summary(fun=mean, geom="line") +
  coord_cartesian(ylim=c(mscale_max-2,mscale_max))+
  theme(legend.key.width = unit(1,"cm"))+
  ylab("log likelihood per observation")

@

This model includes many features that have been proposed to be relevant for understanding measles transmission dynamics \citep{he10}.
Our plug-and-play methodology permits consideration of all these features, and readily extends to the investigation of further variations. 
Likelihood-based inference via plug-and-play methodology therefore provides a framework for evaluating which features of a dynamical model are critical for explaining the data \citep{king08}.
By contrast, \cite{xia04} developed a linearization for a specific spatiotemporal measles model which is numerically convenient but not readily adaptable to assess alternative model choices.
Fig.~\ref{fig:measles_image_plot} shows a simulation from our model, showing that trajectories from this model can capture some features of the system that have been hard to understand: how can it be that disease transmission dynamics between locations have important levels of interaction yet are not locked in synchrony \citep{becker20}?
Here, we are developing statistical tools rather than engaging directly in the scientific debate.

We first assess the scaling properties of the filters on the measles model by evaluating the likelihood over varying numbers of units, $\Unit$, for fixed parameters.
The results are given in Fig.~\ref{fig:mscale_loglik_plot}, with additional information about timing, algorithmic choices, parameter values and a plot of the data provided in Sec.~\SuppSecMeasles.
In Fig.~\ref{fig:mscale_loglik_plot}, the log likelihood per unit per time increases with $\Unit$ because city size decreases with $\Unit$. 
Smaller cities have fewer measles cases, resulting in a narrower and taller probability density function.
Fig.~\ref{fig:mscale_loglik_plot} shows a rapid decline in the performance of the particle filter (PF) beyond $\Unit=4$.
This is a challenging filtering problem, with dynamics including local fadeouts and high stochasticity in each city stabilized at the metapopulation level by the coupling.
In this example, GIRF performs poorly suggesting that the simulated moment guide function is less than successful.
We used the general-purpose implementation of GIRF in the \code{spatPomp} package, and there might be room for improvement by developing a model-specific guide function.
{\ABFIR} uses the same guide function, and this may explain why {\ABFIR} performs worse than {\ABF} here, though {\ABFIR} is much less sensitive than GIRF to the quality of the guide.
{\ABF} and {\UBF} are competing with BPF as winners on this challenge.
The bagged filters and BPF have substantial advantages compared to EnKF, amounting to more than 0.2 log likelihood units per observation.
We suspect that the limitations of EnKF on this problem are due to the nonlinearity, non-Gaussianity, and discreteness of fadeout and reintroduction dynamics.
All the algorithms have various tuning parameters that could influence the results. 
Some investigations of alternatives are presented in the Sec.~\SuppSecMeasles.
Generalizable conclusions are hard to infer from numerical comparisons of complex algorithms on complex models.
Experimentation with different methods, and their tuning parameters, is recommended when investigating a new model.

<<slice_settings,echo=F,eval=T>>=

slice_files_dir <- paste0("slice_",slice_run_level,"/")
if(!dir.exists(slice_files_dir)) dir.create(slice_files_dir)

stew(file=paste0(slice_files_dir,"slice_settings.rda"),{

  # copy variables that should be included in the stew
  slice_run_level <- slice_run_level 
  slice_cores <- cores

  g_lo <- 200
  g_hi <- 600

  g_bpf_lo <- g_lo
  g_bpf_hi <- g_hi

#  g_enkf_lo <- 50
#  g_enkf_hi <- 750

   g_enkf_lo <- g_lo
   g_enkf_hi <- g_hi

  if(slice_run_level==1){
    slice_Ntheta <- 5
    slice_U <- 4
    slice_N <- 3
    slice_replicates <- 2
    slice_abf_Nrep <- 3
    slice_abf_Np_per_replicate <- 10
    slice_bpf_units_per_block <- 1
    slice_bpf_Np <- 100
    slice_ubf_Nrep <- 5
    slice_enkf_Np <- 20
}

  if(slice_run_level==2){
    slice_Ntheta <- 10
    slice_U <- 20
    slice_N <- 10*26
    slice_replicates <- 2 # number of Monte Carlo replicates
    slice_abf_Nrep <- 200
    slice_abf_Np_per_replicate <- 50
    slice_bpf_units_per_block <- 1
    slice_bpf_Np <- 5000
    slice_ubf_Nrep <- 5000
    slice_enkf_Np <- 5000
  }

  if(slice_run_level==3){
    slice_Ntheta <- 10
    slice_replicates <- 4 # number of Monte Carlo replicates
    slice_U <- 40
    slice_N <- 15*26
    slice_abf_Nrep <- 1000
    slice_abf_Np_per_replicate <- 100
    slice_bpf_units_per_block <- 2
    slice_bpf_Np <- 40000
    slice_ubf_Nrep <- 40000
    slice_enkf_Np <- 25000
  }

  if(slice_run_level==4){
    slice_Ntheta <- 12
    slice_replicates <- 5 # number of Monte Carlo replicates
    slice_U <- 40
    slice_N <- 15*26
    slice_abf_Nrep <- 5000
    slice_abf_Np_per_replicate <- 100
    slice_bpf_units_per_block <- 2
    slice_bpf_Np <- 40000
    slice_ubf_Nrep <- 40000
    slice_enkf_Np <- 5000
  }

})
@

<<slice_spatPomp,eval=T,echo=F>>=
slice_spatPomp <- stew(file=paste0(slice_files_dir,"slice_spatPomp.rda"),{

  slice_model_dir <-  measles_model_dir
  # note: care is required if slice_model_dir is set to something other than measles_model_dir
  # however, it may be useful to do that while testing model variations.

  load(file=paste0(slice_model_dir,"measles_spatPomp.rda"))
  slice_sim <- measles_subset(m_U=slice_U, m_N=slice_N)

  slice_sliceJobs <- slice_design(
    center=coef(slice_sim),
    g=rep(seq(from=g_lo,to=g_hi,length=slice_Ntheta),each=slice_replicates)
  )

  slice_bpf_sliceJobs <- slice_design(
    center=coef(slice_sim),
    g=rep(seq(from=g_bpf_lo,to=g_bpf_hi,length=slice_Ntheta),each=slice_replicates)
  )

  slice_enkf_sliceJobs <- slice_design(
    center=coef(slice_sim),
    g=rep(seq(from=g_enkf_lo,to=g_enkf_hi,length=slice_Ntheta),each=slice_replicates)
  )

})
@

<<slice_nbhd,echo=F>>=
slice_nbhd <- function(object, time, unit) {
  nbhd_list <- list()
  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
  return(nbhd_list)
}
@

<<slice_abf, eval=T, cache=F,echo=F>>=
slice_abf_files <- paste0(slice_files_dir,"slice_abf",1:nrow(slice_sliceJobs),".rda")
# parallelization is carried out within abf so we use a serial foreach
foreach (slice_row=1:nrow(slice_sliceJobs), .combine=rbind) %do% {
  slice_abf <- stew(file=slice_abf_files[slice_row],seed=4000817+slice_row,{
    theta <- unlist(slice_sliceJobs[slice_row,names(coef(slice_sim))])
    system.time(
      abf(slice_sim,params=theta, 
        Nrep = slice_abf_Nrep,
        Np=slice_abf_Np_per_replicate,
        nbhd = slice_nbhd,tol=1e-300
      ) -> abf1
    ) -> slice_abf_time
    abf_logLik <- logLik(abf1)
    rm(abf1)
  })
  c(logLik=abf_logLik,theta, Np=slice_abf_Np_per_replicate,
    Nrep = slice_abf_Nrep, U=length(unit_names(slice_sim)), N=length(time(slice_sim)),
    time=slice_abf_time["elapsed"]
  )
} -> slice_abf_results

@

<<slice_ubf, eval=T, cache=F,echo=F>>=
slice_ubf_files <- paste0(slice_files_dir,"slice_ubf",1:nrow(slice_sliceJobs),".rda")
# parallelization is carried out within ubf so we use a serial foreach
foreach (slice_row=1:nrow(slice_sliceJobs), .combine=rbind) %do% {
  slice_ubf <- stew(file=slice_ubf_files[slice_row],seed=4000817+slice_row,{
    theta <- unlist(slice_sliceJobs[slice_row,names(coef(slice_sim))])
    system.time(
      abf(slice_sim,params=theta, 
        Nrep = slice_ubf_Nrep,
        Np=1,
        nbhd = slice_nbhd,tol=1e-300
      ) -> ubf1
    ) -> slice_ubf_time
    ubf_logLik <- logLik(ubf1)
    rm(ubf1)
  })
  c(logLik=ubf_logLik,theta,
    Nrep = slice_ubf_Nrep, U=length(unit_names(slice_sim)), N=length(time(slice_sim)),
    time=slice_ubf_time["elapsed"]
  )
} -> slice_ubf_results

@

<<slice_abfir,eval=F,cache=F,echo=F>>=
slice_abfir_files <- paste0(slice_files_dir,"slice_abfir",1:nrow(slice_sliceJobs),".rda")
# parallelization is carried out within abfir so we use a serial foreach
foreach (slice_row=1:nrow(slice_sliceJobs), .combine=rbind) %do% {  
  slice_abfir <- stew(file=slice_abfir_files[slice_row],seed=4000817+slice_row,{
    theta <- unlist(slice_sliceJobs[slice_row,names(coef(slice_sim))])
    system.time(
      abfir(slice_sim,params=theta,  method='adams',
        Nrep = slice_abfir_Nrep,
        Np=slice_abfir_Np_per_replicate,
        nbhd = slice_nbhd,
        Ninter = as.integer(slice_U/2),
        tol = 1e-300
      ) -> abfir1
    ) -> slice_abfir_time
    abfir_logLik <- logLik(abfir1)
    rm(abfir1)
  })    
  c(logLik=abfir_logLik,theta, Np=slice_abfir_Np_per_replicate,
      Nrep = slice_abfir_Nrep, U=length(unit_names(slice_sim)), N=length(time(slice_sim)),
      time=slice_abfir_time["elapsed"]
  )
} -> slice_abfir_results 
@

<<slice_bpf,cache=F,echo=F,eval=T>>=
slice_bpf_files <- paste0(slice_files_dir,"slice_bpf",1:nrow(slice_bpf_sliceJobs),".rda")
foreach (slice_row=1:nrow(slice_bpf_sliceJobs), .combine=rbind) %dopar% {
  theta_eval <<- unlist(slice_bpf_sliceJobs[slice_row,names(coef(slice_sim))])
  stew(file=slice_bpf_files[slice_row],seed=4000817+slice_row,{
    theta <- theta_eval
    system.time(
      bpfilter(slice_sim,params=theta,
        Np=slice_bpf_Np,
	block_size=slice_bpf_units_per_block
      ) -> slice_bpf_out
    ) -> slice_bpf_time
    bpf_logLik <- slice_bpf_out@loglik
    rm(slice_bpf_out)
  })
  c(logLik=bpf_logLik,theta, Np=slice_bpf_Np,
    units_per_block = slice_bpf_units_per_block,
    U=length(unit_names(slice_sim)), N=length(time(slice_sim)),
    time=slice_bpf_time["elapsed"]
  )
} -> slice_bpf_results    
@

<<slice_enkf,cache=F,echo=F,eval=T>>=

slice_enkf_files <- paste0(slice_files_dir,"slice_enkf",1:nrow(slice_enkf_sliceJobs),".rda")
foreach (slice_row=1:nrow(slice_enkf_sliceJobs), .combine=rbind) %dopar% {
  theta_eval <<- unlist(slice_enkf_sliceJobs[slice_row,names(coef(slice_sim))])
  stew(file=slice_enkf_files[slice_row],seed=4000817+slice_row,{
    theta <- theta_eval
    # an extra step due to bug identified in enkf()
    sim_tmp <- slice_sim
    coef(sim_tmp) <- theta
    system.time(
      enkf(sim_tmp,
        Np=slice_enkf_Np
      ) -> slice_enkf_out
    ) -> slice_enkf_time
    enkf_logLik <- slice_enkf_out@loglik
    rm(slice_enkf_out)
    rm(sim_tmp)
  })
  c(logLik=enkf_logLik,theta, Np=slice_enkf_Np,
    U=length(unit_names(slice_sim)), N=length(time(slice_sim)),
    time=slice_enkf_time["elapsed"]
  )
} -> slice_enkf_results    
@

<<slice_mcap,eval=T,echo=F>>=

slice_abf_mcap <- panelPomp::mcap(
  lp=slice_abf_results[,"logLik"],
  parameter=slice_abf_results[,"g"],
  lambda = 1
)

slice_ubf_mcap <- panelPomp::mcap(
  lp=slice_ubf_results[,"logLik"],
  parameter=slice_ubf_results[,"g"],
  lambda = 0.8
)

slice_bpf_mcap <- panelPomp::mcap(
  lp=slice_bpf_results[,"logLik"],
  parameter=slice_bpf_results[,"g"],
  lambda = 1
)

slice_enkf_mcap <- panelPomp::mcap(
  lp=slice_enkf_results[,"logLik"],
  parameter=slice_enkf_results[,"g"],
  lambda = 1
)

plot.profile2 <- function(mcap1,ylab,xline=2.5,yline=2.5,xlab="",quadratic=FALSE,...){
  if(missing(ylab)) ylab <- "log likelihood"
  # par(mai=c(0.7,0.7,0.3,0.3))
  ggplot() + geom_point(aes(x = mcap1$parameter, y = mcap1$lp)) + 
    geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$smoothed),
              color = "red") +
    {if(quadratic) geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$quadratic),
                             color = "blue",
                             lwd = 1.25)} +
    labs(x = xlab, y = ylab) +
    geom_vline(xintercept = mcap1$ci, color = "red") + 
    geom_hline(yintercept = max(mcap1$fit$smoothed, na.rm = T) - mcap1$delta, color = "red") +
    theme(panel.border = element_rect(colour = "black", fill=NA))
  }

@

<<slice_abf_plot, eval = F, echo=F,  fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Likelihood slice varying the coupling parameter, computed via {\\ABF} with $U=40$ cities. The solid lines construct a 95\\% Monte Carlo adjusted confidence interval \\citep{ionides17}. For this model realization, this interval includes the true parameter value identified by a blue dashed line.')>>=

if(0){
slice_results <- as.data.frame(rbind(
  slice_abf_results[,c("logLik","g")],
  slice_bpf_results[,c("logLik","g")]
))
slice_results$Method <- rep( c("ABF", "BPF"),each=nrow(slice_sliceJobs))
} else {
  slice_results <-  as.data.frame(slice_abf_results[,c("logLik","g")])
  slice_results$Method <- rep("ABF",each=nrow(slice_sliceJobs))
}

plot.profile2(slice_abf_mcap,xlab="G", quadratic = FALSE) +
  geom_vline(xintercept=measles_params["g"],color="blue",linetype="dashed")

@

<<slice_bpf_plot, eval = F, echo=F,  fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Likelihood slice varying the coupling parameter, computed via {BPF} with $U=40$ cities. The solid lines construct a 95\\% Monte Carlo adjusted confidence interval \\citep{ionides17}. For this model realization, this interval includes the true parameter value identified by a blue dashed line.')>>=

plot.profile2(slice_bpf_mcap,xlab="G", quadratic = FALSE) +
  geom_vline(xintercept=measles_params["g"],color="blue",linetype="dashed")

@

<<slice_enkf_plot, eval = F, echo=F,  fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Likelihood slice varying the coupling parameter, computed via {EnKF} with $U=40$ cities. The solid lines construct a 95\\% Monte Carlo adjusted confidence interval \\citep{ionides17}. For this model realization, this interval includes the true parameter value identified by a blue dashed line.')>>=

plot.profile2(slice_enkf_mcap,xlab="G", quadratic = FALSE) +
  geom_vline(xintercept=measles_params["g"],color="blue",linetype="dashed")

@

<<slice_ubf_plot, eval = F, echo=F,  fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Likelihood slice varying the coupling parameter, computed via {\\UBF} with $U=40$ cities. The solid lines construct a 95\\% Monte Carlo adjusted confidence interval \\citep{ionides17}. For this model realization, this interval includes the true parameter value identified by a blue dashed line.')>>=

plot.profile2(slice_ubf_mcap,xlab="G", quadratic = FALSE) +
  geom_vline(xintercept=measles_params["g"],color="blue",linetype="dashed")

@

<<slice_combined_plot, echo=F, fig.height=6, fig.width=4, out.height="7in", fig.cap = paste('Likelihood slices varying the coupling parameter, for the measles model with $U=40$ cities, computed via (A) {ABF}; (B) BPF; (C) EnKF. The solid perpendicular lines construct 95\\% Monte Carlo adjusted confidence intervals \\citep{ionides17}. The true parameter value is identified by a blue dashed line.')>>=

plot.profile <- function(mcap1,ylab,xline=2.5,yline=2.5,xlab="",...){
  if(missing(ylab)) ylab <- "log likelihood"
  plot(mcap1$lp ~ mcap1$parameter, xlab="",ylab="",xaxt="n",...)
  mtext(side=2,text=ylab,line=yline)
  mtext(side=1,text=xlab,line=yline)
  lines(mcap1$fit$parameter,mcap1$fit$smoothed, col = "red", lwd = 1.5)
  abline(v=mcap1$ci,col="red")
  abline(h=max(mcap1$fit$smoothed,na.rm=T)-mcap1$delta,col="red")
}

#pdf(file="tmp.pdf",height=7,width=3.5)

par(omi=c(0.5,0.1,0.1,0.1))
par(mai=c(0.1,0.7,0.1,0.1))
par(mfrow=c(3,1))

plot.profile(slice_abf_mcap,xlab=""
# ,ylab="log likelihood (ABF)"
)
axis(side=1,labels=F)
abline(v=measles_params["g"],col="blue",lty="dashed")

yloc <- 0.05; xloc <- 570
mtext("A",side=2,line=2.5,las=1,padj=-4,cex=1.8)
text(xloc,
  (1-yloc)*min(slice_abf_mcap$lp) + yloc*max(slice_abf_mcap$lp),
  cex=1.4,"ABF")

plot.profile(slice_bpf_mcap,xlab="")
axis(side=1,labels=F)
abline(v=measles_params["g"],col="blue",lty="dashed")

mtext("B",side=2,line=2.5,las=1,padj=-4,cex=1.8)
text(xloc,
  (1-yloc)*min(slice_bpf_mcap$lp) + yloc*max(slice_bpf_mcap$lp),
  cex=1.4,"BPF")

plot.profile(slice_enkf_mcap,xlab="")
axis(side=1,labels=T)
abline(v=measles_params["g"],col="blue",lty="dashed")

mtext("C",side=2,line=2.5,las=1,padj=-4,cex=1.8)
mtext("G",side=1,line=2.5)
text(xloc,
  (1-yloc)*min(slice_enkf_mcap$lp) + yloc*max(slice_enkf_mcap$lp),
  cex=1.4,"EnKF")

@

Fig.~\ref{fig:slice_combined_plot}(A) demonstrates an application of {\ABF} to the task of computing a slice of the likelihood function over the coupling parameter, $\gravity$, for simulated data.
This slice varies $\gravity$ while fixing the other parameters at the values used for the simulation.
Scientifically, this gives us an upper bound on the identifiability of $\gravity$ from such data, since the likelihood slice provides statistically efficient inference when all other parameters are known.
In this case, the Monte Carlo error is comparable to the change in the log likelihood over a wide range of values of $\gravity$. 
Fig.~\ref{fig:slice_combined_plot}(B) shows a similar plot calculated using BPF with comparable computational effort.
Both ABF and BPF are successful here, though BPF is more computationally efficient.
By contrast, Fig.~\ref{fig:slice_combined_plot}(C) shows that EnKF has substantial bias in estimating $\gravity$, as well as considerably lower likelihood.

%Fig~\ref{fig:slice_abf_plot} took $\Sexpr{myround(mean(slice_abf_results[,"time.elapsed"]*gl_cores/60),1)}$ core minutes to evaluate each point using a parallelized implementation on a computing node with $\Sexpr{gl_cores}$ cores, corresponding to a run time of  $\Sexpr{myround(mean(slice_abf_results[,"time.elapsed"]/60),1)}$ per likelihood evaluation.
%By contrast, Fig~\ref{fig:slice_bpf_plot} took $\Sexpr{myround(mean(slice_bpf_results[,"time.elapsed"]/60),1)}$ minutes per likelihood evaluation on a single core.
%
%, equivalent in effort to $\Sexpr{myround(mean(slice_bpf_results[,"time.elapsed"]/60/gl_cores),1)}$ minutes for a parallel implementation.
%
%and Fig~\ref{fig:slice_ubf_plot} took $\Sexpr{myround(mean(slice_ubf_results[,"time.elapsed"]/60),1)}$ minutes,

%%% dddddddddddddd

\section{Discussion}

The pseudocode presented for the bagged filters describes how the outputs are calculated given the inputs, but does not prescribe details of how these quantities are calculated. 
There is scope for implementations to trade off memory, computation and communication by making differing decisions on how the various loops defined in the pseudocode are coded, such as decisions on memory over-writing and parallelization.
This article focuses on  the logical structure of the algorithms, leaving room for future research on implementation-specific considerations, though some supplementary discussion of memory-efficient implementation is given in Sec.~\SuppSecMemoryEfficient.

This article has not exhausted the possibilities for bagging different PF algorithms.
Bagging a standard PF is one possibility, but the algorithm becomes rapidly degenerate due to COD.
Our approach has been to bag PF variants that are already degenerate, so there is less scope for COD to damage them.
Alternatively, one could bag PF variants that avoid the COD by using local weights themselves.

We have focused on likelihood evaluation, a critical quantity for testing scientific hypotheses about model structure and the values of unknown parameters.
However, our bagged filters also provide localized solutions to the filtering problem, using the local weights to combine estimates across replicates.
A solution to the filtering problem, in turn, provides a foundation for forecasting.
Effective filtering and likelihood evaluation can enable likelihood maximization via iterated filtering algorithms \citep{ionides11,ionides15}.

Plug-and-play inference based on sequential Monte Carlo likelihood evaluation has proved successful for investigating highly nonlinear partially observed dynamic systems of low dimension arising in analysis of epidemiological and ecological time series data \citep{breto18statSci,pons-salort18,decelles18,marino18}.
A benefit of the plug-and-play property is that it facilitates development of broadly applicable software, which in turn promotes creative scientific model development \citep{breto09,he10}.

Geophysical data assimilation has similarities to inference on spatiotemporal biological systems.
Relative to biological systems, geophysical applications are characterized by a greater number of spatial locations, better mathematical understanding of the underlying processes, and lower stochasticity.
From this literature, the locally weighted particle filter of \citep{poterjoy16} is perhaps closest to our approach, but those local weights are used to construct a localized Kalman gain which is motivated by a Gaussian approximation.

The EnKF algorithm arose originally via geophysical research \citep{evensen96} and has since become used more widely for inference on SpatPOMP models \citep{katzfuss19,lei10}.
However, EnKF can fail entirely even on simple POMP models if the structure is sufficiently non-Gaussian.
For example, let $X_n$ be a one-dimensional Gaussian random walk, and let $Y_n$ given $X_n=x_n$ be normally distributed with mean $0$ and variance $x_n^2$.
The linear filter rule used by EnKF to update the estimate of $X_n$ given $Y_n$ has mean zero for any value of $X_n$, since $X_n$ and $Y_n$ are uncorrelated.
Therefore, the EnKF filter estimate of the latent process remains essentially constant regardless of the data.
Models of this form are used in finance to describe stochastic volatility.
EnKF could be applied more successfully by modifying model, such as replacing $Y_n$ by $|Y_n|$, but for complex models it may be unclear whether and where such problems are arising.
Our results show that there is room for improvement over EnKF on a spatiotemporal epidemiology model, though in our example there is no clear advantage for BF methods over BPF.

Latent state trajectories constructed in our BF algorithms are all generated from the model simulator, appropriately reweighted and resampled, and so are necessarily valid sample paths of the model.
For example, spatial smoothness properties of the model through space, or conservation properties where some function of the system remains unchanged through time, are maintained in the BF trajectories.
This is not generally true for the block particle filter (since resampling blocks can lead to violations at block boundaries) or for EnKF (since the filter procedure perturbs particles using a linear update rule that cannot respect nonlinear relationships).
The practical importance of smoothness and conservation considerations will vary with the system under investigation, but this property of BF gives the scientific investigator one less thing to worry about.

The algorithms {\UBF}, {\ABF}, {\ABFIR}, GIRF, PF, BPF, and EnKF compared in this article all enjoy the plug-and-play property, facilitating their implementations in general-purpose software.
The numerical results for this paper use the \code{abf}, \code{abfir}, \code{girf}, \code{pfilter}, \code{bpfilter} and \code{enkf} functions via the open-source R package \code{spatPomp} \citep{asfaw20github} that provides a spatiotemporal extension of the R package \code{pomp} \citep{king16}.
{\UBF} was implemented using \code{abf} with $\Np=1$ particles per replicate.
The source code for this paper will be contributed to an open-source scientific archive upon acceptance for publication.

\bibliography{bib-iif}

\end{document}

### 2020-11-28 si.Rnw
setting up files

\documentclass[11pt]{article}

%%% JASA. Attention Mac Users. The latest version of the pdfTex compiler for Mac (pdfTeX-1.40.11) outputs files not currently supported by ScholarOne Manuscripts. To resolve this and have a PDF file that will convert properly, the PDF file will need to be distilled to Adobe version 1.4 or earlier. To create a PDF file that uploads correctly, place the command \pdfminorversion=4 in the preamble (before the begin{document}) of your LaTeX file and run pdfLaTeX again.

\pdfminorversion=4

<<debug,echo=F>>=
#run_level <- 1
#run_level <- "jasa-sept-2020"
run_level <- 2
abfNbhd_run_level <- run_level
lz_run_level <- run_level
bm_run_level <- run_level
mscale_run_level <- run_level
slice_run_level <- run_level

if(run_level=="jasa-sept-2020"){  
#### values used for jasa submission, sept 2020  
  abfNbhd_run_level <- 3
  lz_run_level <- 5
  bm_run_level <- 4
  mscale_run_level <- 5
  slice_run_level <- 3
}

@

\newcommand\secTitleSpace{\hspace{3mm}}

<<packages,include=F,cache=F>>=
library("ggplot2")
#library("xtable")
library("spatPomp")
library(doParallel)
library(doRNG)

cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()  
registerDoParallel(cores)

gl_cores <- 36
doob_cores <- 40
ito_cores <- 8

<<set-opts,include=F,cache=F>>=

options(
        scipen=2,
        help_type="html",
        stringsAsFactors=FALSE,
        prompt="R> ",
        continue="+  ",
        width=70,
        useFancyQuotes=FALSE,
        reindent.spaces=2,
        xtable.comment=FALSE
        )
@

<<knitr-opts,include=F,cache=F,purl=F>>=
library("knitr")
opts_knit$set(concordance=TRUE)
opts_chunk$set(
    progress=TRUE,prompt=TRUE,highlight=FALSE,
    tidy=TRUE,
    tidy.opts=list(
        keep.blank.line=FALSE
    ),
    comment="",
    warning=FALSE,
    message=FALSE,
    error=TRUE,
    echo=TRUE,
    cache=FALSE,
    strip.white=TRUE,
    results="markup",
    background="#FFFFFF00",
    size="normalsize",
    fig.path="figure/",
    fig.lp="fig:",
    fig.align="left",
    fig.show="asis",
#    dpi=300,
    dev="pdf",
    dev.args=list(
        bg="transparent",
        pointsize=9
    )
)

myround<- function (x, digits = 1) {
  # taken from the broman package
  if (digits < 1) 
    stop("This is intended for the case digits >= 1.")
  if (length(digits) > 1) {
    digits <- digits[1]
    warning("Using only digits[1]")
  }
  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
  tmp[tmp == paste0("-", zero)] <- zero
  tmp
}

@

%% copied from pnas-new.cls
\usepackage{amsmath,amsfonts,amssymb,graphicx}

\input{header-ms.tex}
\input{theorems.tex}

\renewcommand\siOnly[1]{#1}
\renewcommand\msOnly[1]{}

\bibliographystyle{apalike}
%% added or modified from pnas-new.cls
\usepackage{natbib} 
\usepackage{fullpage}
\renewcommand\myeqref[1]{(\ref{#1})}

%\usepackage{relsize}

%\templatetype{pnassupportinginfo}

% \readytosubmit %% Uncomment this line before submitting, so that the instruction page is removed.

%\author{Author1, Author2 and Author3 (Complete author list)}
%\correspondingauthor{Corresponding Author Name.\\E-mail: author.two@email.com}

\renewcommand{\contentsname}{Supplementary Content}
\renewcommand{\refname}{Supplementary References}
\renewcommand\thefigure{S-\arabic{figure}}
\renewcommand\thetable{S-\arabic{table}}
%\renewcommand\thepage{S-\arabic{page}}
\renewcommand\thesection{S\arabic{section}}
\renewcommand\theequation{S\arabic{equation}}
\renewcommand\theprop{S\arabic{prop}}
\renewcommand\thelemma{S\arabic{lemma}}

\setcounter{tocdepth}{1}

\begin{document}

%% Comment/remove this line before generating final copy for submission 
%\instructionspage

\date{\today}
\title{Supplement to ``{\mytitle}''}
\author{E. L. Ionides, K. Asfaw, J. Park and A. A. King}

\newcommand{\blind}{1}

\if1\blind
{
\maketitle
}
\fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf \mytitle}
\end{center}
  \bigskip
  \bigskip
}\fi

%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.

%\SItext

%Draft: \today

\tableofcontents

\newpage

\section{\secTitleSpace A generalization to models without latent unit structure}

Variations of the algorithms in the main text apply when there is no latent unit structure. 
In this case, the observation vector $\myvec{Y}_{\time}=(Y_{1,\time},\dots,Y_{\Unit,\time})$ consists of a collection of measurements on a general latent vector ${X}_{\time}$.
We may have the structure that $Y_{1,\time},\dots,Y_{\Unit,\time}$ are conditionally independent given ${X}_{\time}$, but even this is not essential to the approach.
This is most readily seen in the context of the unadapted bagged filter, giving rise to the generalized unadapted bagged filter (G-{\UBF}) algorithm defined as follows.

\begin{center}
\noindent\begin{tabular}{l}
\hline
{\bf Algorithm~G-{\UBF} (Generalized unadapted bagged filter).}\rule[-1.5mm]{0mm}{6mm}\\
\hline
{\bf input:}\rule[-1.5mm]{0mm}{6mm} \\
Simulator for $f_{{X}_{\time}|{X}_{\time-1}}({x}_{\time}|{x}_{\time-1})$\\
Evaluator for $f_{Y_{\unit,\time}|{X}_{\time}}(\data{y}_{\unit,\time}\given {x}_{\time})$\\
Number of bootstrap filters, $\Rep$\\
Neighborhood structure, $B_{\unit,\time}$, for $\unit\in 1\mycolon\Unit$ and $\time\in 1\mycolon\Time$\\
Data, $\data{y}_{\unit,\time}$ for $\unit\in 1\mycolon\Unit$ and $\time\in 1\mycolon\Time$ \\
{\bf output:}\rule[-1.5mm]{0mm}{5mm} \\
Log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit,\time}$  \rule[-2mm]{0mm}{3mm}
\\
\hline
For $\rep$ in $1\mycolon\Rep$ \rule[-1.5mm]{0mm}{6mm}
\\
\asp simulate ${X}_{\time,\rep}^{\simulation}$ from the dynamic model, for $\time\in 1{\mycolon}\Time$\\
End For\\
Prediction weights, 
$w^P_{\unit,\time,\rep}= f_{Y_{B_{\unit,\time}}|{X}_{1:\time}}(\data{y}_{B_{\unit,\time}}\given {X}^{\simulation}_{1:\time,\rep})$\\
Measurement weights,
$w^M_{\unit,\time,\rep}=f_{Y_{\unit,\time}|X_{\time},Y_{B_{\unit,\time}}}(\data{y}_{\unit,\time}\given X^{\simulation}_{\time,\rep},\data{y}_{B_{\unit,\time}})$
\\
Conditional log likelihood estimate,
$\MC{\loglik}_{\unit,\time}= 
\log \left(
  \sum_{\rep=1}^\Rep w^M_{\unit,\time,\rep}\,   w^P_{\unit,\time,\rep}
\right)
-
\log \left(
    \sum_{\tilde \rep=1}^\Rep w^P_{\unit,\time,\tilde \rep}
\right)
$
\\
\hline
\end{tabular}
\end{center}

The algorithm G-{\UBF} operates on an arbitrary POMP model. 
G-{\UBF} therefore provides a potential approach to extending methodologies from SpatPOMP models to models that have some similarity to a SpatPOMP without formally meeting the definition.
For example, there may be collections of interacting processes at different spatial scales in a spatiotemporal system. 
Alternatively, the potential outcomes of the latent process may vary between spatial units, such as when modeling interactions between terrestrial and aquatic ecosystems.
We do not further explore G-{\UBF} here.

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  asasasas #####################

\section{\secTitleSpace Adapted simulation for an Euler approximation}

\newcommand\nextn{n+1}
\newcommand\thisn{n}

We investigate the adapted simulation process by considering a continuous-time limit where it becomes a diffusion process.
We find that adapted simulation can effectively track the latent process when the measurement error is on an appropriate scale.
However, when the measurement error is large compared to the latent process noise, adapted simulation can fail in situations where filtering succeeds.
We work with a one-dimensional POMP model having a latent process constructed as an Euler approximation,
\begin{eqnarray}\label{x:euler}
X_{\nextn}&=& X_{\thisn} + \mu(X_{\thisn})\delta + \sigma \sqrt{\delta} \epsilon_{\nextn}, 
\end{eqnarray}
which provides a numerical solution to a one-dimensional stochastic differential equation,
\begin{eqnarray}
\nonumber
dX(t)&=& \mu\big(X(t)\big)\, dt + \sigma\, dU(t) ,
\end{eqnarray}
where $\{U(t)\}$ is a standard Brownian motion. 
We will consider several different measurement processes.

%\vspace{3mm}

%\noindent{\bf \large M1. 
\subsection{Measurement error on the same scale as the process noise}
\label{subsec:m1}

%\vspace{2mm}

\noindent Here, we consider the measurement model
\begin{equation} \label{m1}
Y_{\nextn}= \mu(X_{\thisn})\delta + \sigma \sqrt{\delta} \epsilon_{\nextn} + \tau\sqrt{\delta}\,\eta_{\nextn}.
\end{equation}
This is an approximation to the increment $Y(t+\delta)-Y(t)$ of a continuous time measurement model
\begin{equation}
\label{m1:cts}
dY(t) = dX(t) + \tau\, dV(t),
\end{equation}
where $\{V(t)\}$ is a standard Brownian motion independent of $\{U(t)\}$.
The measurement model (\ref{m1:cts}) makes inference on $X(t)$ given $Y(t)$ a continuous time version of the filtering problem.
A feature of this model is that $Y(t)$ does not directly track the level of the state, since the solution with initial conditions $Y(t_0)=X(t_0)$ and $V(t_0)=0$ is
\begin{equation}
\nonumber
Y(t) = X(t) + \tau V(t).
\end{equation}
The measurement error, $\tau V(t)$, has variance $\tau^2 t$ that increases with $t$. 
However, under appropriate conditions, information on changes in $\{X(t)\}$ obtained via $\{Y(t)\}$ are enough to track $X(t)$ indirectly via the filtering equations.
For the POMP given by (\ref{x:euler}) and (\ref{m1}), we can calculate exactly the adapted simulation distribution $f_{X_{\nextn}|Y_{\nextn},X_{\thisn}}$.
It is convenient to work conditionally on $X_{\thisn}$, allowing us to treat $X_{\thisn}$ and $\mu(X_{\thisn})$ as constants, with $X_{\nextn}$ and $Y_{\nextn}$ therefore being jointly normally distributed.
A Gaussian distribution calculation then gives the conditional moments.
First, we find
\begin{eqnarray*}
%\label{m1.1}
\E\big[ X_{\nextn}|Y_{\nextn},X_{\thisn} \big]
&=& 
X_{\thisn} + \mu(X_{\thisn})\delta + \E\big[{\sigma \sqrt{\delta} \epsilon_{\nextn} \, \big| \, \sigma \sqrt{\delta}\epsilon_{\nextn}+\tau{\sqrt{\delta}}\eta_{\nextn}}\big]
\\
&=&
X_{\thisn} + \mu(X_{\thisn})\delta + \frac{\sigma^2}{\sigma^2 +\tau^2}
\big(
  \sigma \sqrt{\delta}\epsilon_{\nextn}+\tau{\sqrt{\delta}}\eta_{\nextn}
\big)
%\label{m1.2}
\\
&=& X_{\thisn} + \mu(X_{\thisn})\delta + \frac{\sigma^2}{\sigma^2 +\tau^2} 
  \big( Y_{\nextn}- \mu(X_{\thisn})\delta \big).
\end{eqnarray*} 
Then,
\begin{eqnarray*}
\var\big[{X_{\nextn} \, \big| \, Y_{\nextn},X_{\thisn}}\big]
&=& 
\var\big[
   \sigma \sqrt{\delta} \epsilon_{\nextn} \, \big| \,
   \sigma \sqrt{\delta}\epsilon_{\nextn}+\tau{\sqrt{\delta}}\eta_{\nextn}
\big]
\\
&=& \sigma^2\delta - \frac{\sigma^4\delta^2}{\sigma^2\delta+\tau^2\delta}
\\
&=& \delta\frac{\sigma^2\tau^2}{\sigma^2+\tau^2}.
\end{eqnarray*}
Call the adapted simulation process $\{A_n\}$, defined conditionally on $\{Y_n\}$. 
We see from the above calculation that 
\begin{eqnarray*}
%\label{m1:a}
A_{\nextn} &=& A_{\thisn} + \mu(A_{\thisn})\delta +
  \frac{\sigma^2}{\sigma^2 +\tau^2} 
  \Big( 
    \mu(X_{\thisn})\delta + \sigma \sqrt{\delta} \, \epsilon_{\nextn}
    + \tau\sqrt{\delta} \, \eta_{\nextn}  -\mu(A_{\thisn})\, \delta
  \Big) 
\\
&& 
\hspace{4cm} + 
  \frac{\sigma\tau}{\sqrt{\sigma^2+\tau^2}} \sqrt{\delta} \, \zeta_{\nextn}
\end{eqnarray*}
where $\{\zeta_n\}$ is an iid standard normal sequence independent of $\{\epsilon_n,\eta_n\}$.
To study how well the adapted simulation tracks $\{X_n\}$, we subtract $X_{\nextn}$ from both sides to get
\begin{eqnarray*}
%\label{m1:b}
\hspace{-8mm} [A_{\nextn}-X_{\nextn}] &=& [A_{\thisn} -X_{\thisn}] + [\mu(A_{\thisn}) - \mu(X_{\thisn})]\delta - \sigma \sqrt{\delta} \, \epsilon_{\nextn} 
\\
&& \hspace{1mm}
+ 
  \frac{\sigma^2}{\sigma^2 +\tau^2} 
  \Big( 
     [\mu(X_{\thisn})-\mu(A_{\thisn})]\delta + \sigma \sqrt{\delta} \, \epsilon_{\nextn} + \tau\sqrt{\delta} \, \eta_{\nextn}  
   \Big) 
   + \frac{\sigma\tau}{\sqrt{\sigma^2+\tau^2}} \sqrt{\delta} \, \zeta_{\nextn}
\\
&=&   [A_{\thisn} -X_{\thisn}] +  \frac{2\sigma^2+\tau^2}{\sigma^2 +\tau^2}[\mu(X_{\thisn})-\mu(A_{\thisn})]\delta
\\
&& \hspace{3mm}
+   \frac{\sigma^2\tau\sqrt{\delta}\eta_{\nextn} - \sigma\tau^2\sqrt{\delta}\epsilon_{\nextn}}{\sigma^2 +\tau^2} 
   + \frac{\sigma\tau}{\sqrt{\sigma^2+\tau^2}} \sqrt{\delta} \, \zeta_{\nextn}.
\end{eqnarray*}
$\{A_n\}$ tracks $\{X_n\}$ when the process $\{A_n-X_n\}$ is stable.
This happens when $\mu(x)-\mu(y)$ is negative when $x$ is sufficiently larger than $y$. 
For example, a stable autoregressive process with $\mu(x)=-ax$ gives a stable adapted filter process.

\subsection{Independent measurement error on a scale that gives a finite limiting amount of information about $X(t)$ from measurements on a unit time interval}
\label{subsec:m2}

\noindent We now consider the measurement model
\begin{eqnarray} 
%\label{m2}
\nonumber
Y_{\nextn}&=& X_{\nextn}+ \frac{\tau}{\sqrt{\delta}}\, \eta_{\nextn}
\\
  &=& X_{\thisn} + \mu(X_{\thisn}) \, \delta + \sigma \sqrt{\delta}  \, \epsilon_{\nextn}+\frac{\tau}{\sqrt{\delta}}\eta_{\nextn},
\label{e0}
\end{eqnarray}
where $\{\epsilon_n,\eta_n\}$ is a collection of independent standard normal random variables.
The conditional mean is now
\begin{eqnarray}
%\label{m2:e1}
\nonumber
\E\big[{X_{\nextn}|Y_{\nextn},X_{\thisn}}\big].
&=&
X_{\thisn} + \mu(X_{\thisn})\delta +
\E\Big[
  \sigma \sqrt{\delta} \epsilon_{\nextn}\,\Big| \, \sigma \sqrt{\delta}\epsilon_{\nextn}+\frac{\tau}{\sqrt{\delta}} \eta_{\nextn}\Big]
\\
&=&
X_{\thisn} + \mu(X_{\thisn})\delta + \frac{\sigma^2\delta}{\sigma^2\delta +\tau^2/\delta}
\big(
  \sigma \sqrt{\delta}\epsilon_{\nextn}+\tau{\sqrt{\delta}} \, \eta_{\nextn}
\big)
\label{m2.2}
\end{eqnarray} 
Using (\ref{e0}) and (\ref{m2.2}) gives
\begin{eqnarray*}
\E\big[{X_{\nextn}|Y_{\nextn},X_{\thisn}}\big] &=& 
X_{\thisn} + \mu(X_{\thisn})\delta + \frac{\sigma^2\delta^2}{\sigma^2\delta^2+\tau^2} 
\Big(
  Y_{\nextn}- X_{\thisn} - \mu(X_{\thisn})\, \delta
\Big).
\end{eqnarray*} 
In the limit as $\delta\to 0$, the contribution from the measurement is order $\delta^2$ and is therefore negligible.
Although the observation process is meaningfully informative about the latent process, the adapted simulation fails to track the latent process in this limit. 
Intuitively, this is because the adapted simulation is trying to track differences in the latent process, but for this model the signal to noise ratio for the difference in each interval of length $\delta$ tends to zero.

\subsection{Independent measurements of the latent process with measurement error on a scale that gives a useful adapted process as $\delta\to 0$}
\label{subsec:m3}

\noindent We now consider the measurement model
\begin{eqnarray} 
%\label{m3}
\nonumber
Y_{\nextn}&=& X_{\nextn}+ \tau\eta_{\nextn}
\\
  &=& X_{\thisn} + \mu(X_{\thisn})\delta + \sigma \sqrt{\delta} \epsilon_{\nextn}+\tau\eta_{\nextn}.
\label{m3:e0}
\end{eqnarray}
The conditional mean is now
\begin{eqnarray}
%\label{m3:e1}
\nonumber
\E\big[{X_{\nextn}|Y_{\nextn},X_{\thisn}}\big].
&=&
X_{\thisn} + \mu(X_{\thisn})\, \delta +
\E\big[
  \sigma \sqrt{\delta} \epsilon_{\nextn}\,\big| \,
  \sigma \sqrt{\delta}\epsilon_{\nextn}+\tau \eta_{\nextn}
\big]
\\
&=&
X_{\thisn} + \mu(X_{\thisn})\delta + \frac{\sigma^2\delta}{\sigma^2\delta +\tau^2}
\big(
  \sigma \sqrt{\delta}\epsilon_{\nextn}+\tau \eta_{\nextn}
\big)
\label{m3.2}
\end{eqnarray} 
Using (\ref{m3:e0}) and (\ref{m3.2}) gives
\begin{eqnarray*}
\E\big[ X_{\nextn}|Y_{\nextn},X_{\thisn}\big] &=& 
X_{\thisn} + \mu(X_{\thisn})\delta + \frac{\sigma^2\delta}{\sigma^2\delta+\tau^2} 
\Big(
  Y_{\nextn}- X_{\thisn} - \mu(X_{\thisn})\delta
  \Big)
\\
&=& X_n + \mu(X_n)\delta + \frac{\sigma^2}{\tau^2} \delta
\Big(
  Y_{\nextn}- X_{\thisn} - \mu(X_{\thisn}) \, \delta
  \Big)
+ o(\delta)
\end{eqnarray*} 
In the limit as $\delta\to 0$, the adapted simulation has a diffusive drift toward the value of the latent process.

For disease models, incidence data can arguably be considered as noisy measurements of the change of a state variable (number of susceptibles) that is not directly measured. 
This could correspond to a situation where the measurement error is on the same scale as the process noise (Subsection~\ref{subsec:m1}). 
Alternatively, we could think of weekly aggregated incidence as a noisy measurement of the infected class, in which case the measurement error could match the scaling in Subsection~\ref{subsec:m3}.

The model in Subsection~\ref{subsec:m2} is a cautionary tale, warning us against carrying out adapted simulation on short time intervals.
An interpretation is that one should not carry out adapted simulation unless a reasonable amount of information has accrued.
When each observation has low information, a particle filter may enable solution to the filtering problem without particle depletion.
It is when the data are highly informative that the curse of dimensionality makes basic particle filters ineffective, opening up demand for alternative methods.

We are now in a better position to understand why it may be appropriate to keep many particle representations at intermediate timesteps while resampling down to a single representative at each observation time, as {\ABF} and {\ABFIR} do.
We have seen that adaptive simulation can fail when observations occur frequently.
Resampling down to a single particle too often can lose the ability for the adapted process to track the latent process.
This implies that adapted simulation should not be relied upon more than necessary to ameliorate the curse of dimensionality: once proper importance sampling for filtering problem becomes tractable in a sufficiently small spatiotemporal neighborhood, one should maintain weighted particles on this spatiotemporal scale rather than resorting to adapted simulation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 1111111111111111
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 2222222222222222

\section{\secTitleSpace {\UBF} convergence: Proof of Theorem~1}

We consider a collection of models $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$ and data $\data{\myvec{y}}_{1:\Time}$ defined for each $(\Unit,\Time)$.
These models and datasets are not required to have any nesting relationship, so we do not insist that $X_{1,1}$ or $\data{y}_{1,1}$ should be the same for $(\Unit,\Time)=(10,10)$ as for $(\Unit,\Time)=(100,100)$.
Formally, we define probability and expectation on a product space of the stochastic model and Monte Carlo outcomes. 
Monte Carlo quantities such as the output of the {\UBF}, {\ABF} and {\ABFIR} algorithms depend on the data but not on any random variables constructed under the model.
As a consequence, we can use $\E$ to correspond both to expectations over Monte Carlo stochasticity (for Monte Carlo quantities) and model stochasticity (for random variables constructed under the model).
We suppress discussion of measurability by assuming that all functions considered have appropriate measurability properties.
We restate the assumptions and statement for Theorem~1.

\Ai  %% Assumption 1  Assumption~\ref{A1}

We use the total variation bound in Assumption~~\ref{A1} via the following Proposition~\ref{weak_coupling:lemma}, which replaces conditioning on $X_{B^c_{\unit,\time}}$ with conditioning on $Y_{B^c_{\unit,\time}}$. The bound in \myeqref{eq:implication:of:A1} could be used in place of Assumption~\ref{A1}.

\begin{prop}\label{weak_coupling:lemma}
Under the conditions of Assumption~\ref{A1}, \myeqref{eq:weak_coupling2} implies
\begin{equation}\label{eq:implication:of:A1}
\bigg| \int h(x_{\unit,\time}) f_{X_{\unit,\time}|Y_{A_{\unit,\time}}}(x_{\unit,\time}\given \data{y}_{A_{\unit,\time}}) \, dx_{\unit,\time}
- \int h(x_{\unit,\time})f_{X_{\unit,\time}|Y_{B_{\unit,\time}}}(x_{\unit,\time}\given \data{y}_{B_{\unit,\time}})
\, dx_{\unit,\time} \, 
\bigg| < \eone
\end{equation}
\end{prop}
\begin{proof} For notational compactness, we suppress the arguments $x_{\unit,\time}$, $x_{B^c_{\unit,\time}}$, $\data{y}_{A_{\unit,\time}}$, $\data{y}_{B_{\unit,\time}}$ matching the subscripts of conditional densities.
Using the conditional independence of the measurements given the latent process, we calculate
\begin{eqnarray}\nonumber
&& \hspace{-15mm}\bigg| \int h(x_{\unit,\time}) f_{X_{\unit,\time}|Y_{A_{\unit,\time}}} \, dx_{\unit,\time}
- \int h(x_{\unit,\time})f_{X_{\unit,\time}|Y_{B_{\unit,\time}}} \, dx_{\unit,\time}
\bigg|
\\
\nonumber
&=& \bigg| \int  \bigg\{
{\mediumint}  h(x_{\unit,\time}) f_{X^{}_{\unit,\time}|Y^{}_{B_{\unit,\time}},X_{B^c_{\unit,\time}
}} \, dx_{\unit,\time} -  {\mediumint}
 h(x_{\unit,\time})f_{X^{}_{\unit,\time}|Y^{}_{B_{\unit,\time}}}\, dx_{\unit,\time} \bigg\} \, 
 f_{X_{B^c_{\unit,\time}} \given Y_{A^{}_{\unit,\time}}}\, dx^{}_{B^c_{\unit,\time}} \bigg| 
%\label{eq:weak_coupling:lemma}
\nonumber
\\
\nonumber
&\le& \int  \bigg| \,
{\mediumint}  h(x_{\unit,\time}) f_{X_{\unit,\time}|Y_{B_{\unit,\time}},X_{B^c_{\unit,\time}
}} \, dx_{\unit,\time} -  {\mediumint}
 h(x_{\unit,\time})f_{X_{\unit,\time}|Y_{B_{\unit,\time}}}\, dx_{\unit,\time} \, \bigg| \, 
 f_{X_{B^c_{\unit,\time}} \given Y_{A^{}_{\unit,\time}}}\, dx^{}_{B^c_{\unit,\time}} 
\\
\nonumber
& < & \int \eone \, f_{X_{B^c_{\unit,\time}} \given Y^{}_{A_{\unit,\time}}}\, dx^{}_{B^c_{\unit,\time}} \hspace{3mm} = \hspace{3mm} \eone.
\end{eqnarray}
\end{proof}

\Aii  %% Assumption 2  Assumption~\ref{A1b}

\Aiii   %% Assumption 3   Assumption~\ref{A2}

\Aiv    %% Assumption 4 \ref{A:unconditional:mix}

%Note that requiring the inequality \myeqref{eq:A:unconditional:mix} to hold over $\altB^{}_{\unit,\time}$ implies that it holds over all subsets of  ${\altB^{}_{\unit,\time}}$. 
%This is demonstrated by the following proposition.

%\begin{prop}
%Let $X$, $Y$ and $Z$ be random variables with joint density $f_{XYZ}(x,y,z)$.
%If $\big|f_{XY|Z}-f_{XY}\big| < \epsilon \, f_{XY}$ then $\big| f_{X|Z}-f_{X} \big| < \epsilon \, f_{X}$.
%\end{prop}
\newcommand\extraSpace{\hspace{1mm}}
%\begin{proof}
%$\big|f_{X|Z}-f_{X}\big| 
%\extraSpace = \extraSpace \left| \int \big( f_{XY|Z}-f_{XY} \big) \, dy \right|
%\extraSpace \le \extraSpace \int  \big| f_{XY|Z}-f_{XY} \big| \, dy
%\extraSpace < \extraSpace \epsilon \int f_{XY}\, dy  
%\extraSpace = \extraSpace \epsilon \, f_X$.
%\end{proof}

Assumption~\ref{A:unconditional:mix} is needed only to ensure that the variance bound in Theorem~\ref{thm:tif} is essentially $O(\Unit \Time)$ rather than $O(\Unit^2 \Time^2)$. 
Both these rates avoid the exponentially increasing variance characterizing the curse of dimensionality.
Lower variance than $O(\Unit \Time)$ cannot be anticipated for any sequential Monte Carlo method since the log likelihood estimate can be written as a sum of $\Unit\Time$ terms each of which involves its own sequential Monte Carlo calculation.

\TheoremI %% Theorem~\ref{thm:tif}

\begin{proof}
Suppose the quantities $w^M_{\unit,\time,\rep}$ and $w^P_{\unit,\time,\rep}$ constructed in Algorithm~{\UBF} are considered i.i.d.\ replicates of jointly defined random variables $w^M_{\unit,\time}$ and $w^P_{\unit,\time}$, for each $(\unit,\time)\in \spaceTime$.
Also, write 
\begin{equation}
\nonumber
\centersum^{MP}_{\unit,\time}= \frac{1}{\sqrt{\Rep}}\sum_{\rep=1}^\Rep\big(w^M_{\unit,\time,\rep}w^P_{\unit,\time,\rep} - \EMC[w^M_{\unit,\time}w^P_{\unit,\time}]\big),
\hspace{2cm}
\centersum^{P}_{\unit,\time}= \frac{1}{\sqrt{\Rep}}\sum_{\rep=1}^\Rep\big(w^P_{\unit,\time,\rep} - \EMC[w^P_{\unit,\time}]\big),
\end{equation}
Then, using the delta method (e.g., Section 2.5.3 in \cite{liu01}) we find
\begin{eqnarray}
\nonumber
\MC{\loglik}_{\unit,\time} &=& \log\left( \frac{\sum_{\rep=1}^\Rep w^M_{\unit,\time,\rep}w^P_{\unit,\time,\rep}}{\sum_{\rep=1}^\Rep w^P_{\unit,\time,\rep}}\right)
\\
\nonumber
&=&  \log\Big(\EMC[w^M_{\unit,\time}w^P_{\unit,\time}] + \Rep^{-1/2}\centersum^{MP}_{\unit,\time}\Big) - \log\Big({\EMC[w^P_{\unit,\time}] + \Rep^{-1/2} \centersum^P_{\unit,\time}}\Big)
\\
\label{eq:likelihood:at:point}
&=& \log\left(\frac{\EMC[w^M_{\unit,\time}w^P_{\unit,\time}]}{\EMC[w^P_{\unit,\time}]}\right) +  \Rep^{-1/2}
\left(\frac{\centersum^{MP}_{\unit,\time}}{\EMC[w^M_{\unit,\time}w^P_{\unit,\time}]} - \frac{\centersum^{P}_{\unit,\time}}{\EMC[w^P_{\unit,\time}]} \right) + o_P\big(\Rep^{-1/2}\big)
\end{eqnarray}
The joint distribution of $\big\{ (\centersum^{MP}_{\unit,\time},\centersum^{P}_{\unit,\time}), (\unit,\time)\in \spaceTime \big\}$ follows a standard central limit theorem as $\Rep\to\infty$.
Each term has mean zero, with covariances uniformly bounded over $(\unit,\time,\altUnit,\altTime)$ due to Assumption~\ref{A2}.
Specifically,
\begin{equation}
\nonumber
\hspace{-3mm}
\var
\hspace{-1mm}
\left(
\hspace{-2mm}
\begin{array}{c}
\centersum^{MP}_{\unit,\time} \\
\centersum^{P}_{\unit,\time} \\
\centersum^{MP}_{\altUnit,\altTime} \\
\centersum^{P}_{\altUnit,\altTime} \\
\end{array} 
\hspace{-2mm}
\right)
\hspace{-1mm}
{=}
\hspace{-1mm}
\left(
\hspace{-2mm}
\begin{array}{cccc}
\var(w^{M}_{\unit,\time}w^{P}_{\unit,\time}) 
& \cov(w^{M}_{\unit,\time}w^{P}_{\unit,\time},w^{P}_{\unit,\time}) 
\hspace{-2mm}
& \cov(w^{M}_{\unit,\time}w^{P}_{\unit,\time},w^{M}_{\altUnit,\altTime}w^{P}_{\altUnit,\altTime}) 
\hspace{-2mm}
& \cov(w^{M}_{\unit,\time}w^{P}_{\unit,\time},w^{P}_{\altUnit,\altTime}) 
\\
\cov(w^{M}_{\unit,\time}w^{P}_{\unit,\time},w^{P}_{\unit,\time}) 
& \var(w^{P}_{\unit,\time}) 
& \cov(w^{P}_{\unit,\time},w^{M}_{\altUnit,\altTime}w^{P}_{\altUnit,\altTime}) 
& \cov(w^{P}_{\unit,\time},w^{P}_{\altUnit,\altTime}) 
\\
\cov(w^{M}_{\unit,\time}w^{P}_{\unit,\time},w^{M}_{\altUnit,\altTime}w^{P}_{\altUnit,\altTime}) 
\hspace{-2mm}
& \cov(w^{P}_{\unit,\time},w^{M}_{\altUnit,\altTime}w^{P}_{\altUnit,\altTime}) 
& \var(w^{M}_{\altUnit,\altTime}w^{P}_{\altUnit,\altTime}) 
& \cov(w^{M}_{\altUnit,\altTime}w^{P}_{\altUnit,\altTime},w^{P}_{\altUnit,\altTime}) 
\\
\cov(w^{M}_{\unit,\time}w^{P}_{\unit,\time},w^{P}_{\altUnit,\altTime}) 
& \cov(w^{P}_{\unit,\time},w^{P}_{\altUnit,\altTime}) 
& \cov(w^{M}_{\altUnit,\altTime}w^{P}_{\altUnit,\altTime},w^{P}_{\altUnit,\altTime}) 
& \var(w^{P}_{\altUnit,\altTime})
\end{array}
\hspace{-2mm}
\right)
\end{equation}
Note that
\begin{eqnarray*}
\nonumber
\log\left[ 
  \frac{\EMC[w^{M}_{\unit,\time}w^{P}_{\unit,\time}]}{\EMC[w^{P}_{\unit,\time}]} 
\right]
&=&\log \left[
\frac{
\int f_{Y_{\unit,\time}|X_{\unit,\time}}(\data{y}_{\unit,\time}\given x_{\unit,\time,\rep})\, f_{Y_{B_{\unit,\time}}|X_{B_{\unit,\time}}}(\data{y}_{B_{\unit,\time}}\given x_{B_{\unit,\time}})\,
f_{X_{B^+_{\unit,\time}}}(x_{B^+_{\unit,\time}}) \, dx_{B^+_{\unit,\time}}
}{
\int f_{Y_{B_{\unit,\time}}|X_{B_{\unit,\time}}}(\data{y}_{B_{\unit,\time}}\given x_{B_{\unit,\time}})\,
f_{X_{B_{\unit,\time}}}(x_{B_{\unit,\time}}) \, dx_{B_{\unit,\time}}
}
\right]
\\
\nonumber
&=& \log \big[ 
  f_{Y_{\unit,\time}|Y_{B_{\unit,\time}}}(\data{y}_{\unit,\time}\given \data{y}_{B_{\unit,\time}})
\big],
\end{eqnarray*}
where $B^+_{\unit,\time}=B_{\unit,\time}\cup (\unit,\time)$. Now, define
\begin{equation}
\nonumber
\centersum^{\loglik}_{\unit,\time}=
\left(\frac{\centersum^{MP}_{\unit,\time}}{\EMC[w^M_{\unit,\time}w^P_{\unit,\time}]} - \frac{\centersum^{P}_{\unit,\time}}{\EMC[w^P_{\unit,\time}]} \right)
\end{equation}
Summing over all $({\unit,\time})\in \spaceTime$, we get
\begin{equation}
\label{thm1:loglik:linearization}
\sqrt{\Rep}\left( \MC{\loglik} - \sum_{({\unit,\time})\in\, \spaceTime}
\log f_{Y_{\unit,\time}|Y_{B_{\unit,\time}}}(\data{y}_{\unit,\time}\given \data{y}_{B_{\unit,\time}})
\right)
= \sum_{({\unit,\time})\in \,  \spaceTime} \centersum^{\loglik}_{\unit,\time} + o(1).
\end{equation}
Now,
\begin{equation}
\nonumber
\cov\big(\centersum^{\loglik}_{\unit,\time},\centersum^{\loglik}_{\altUnit,\altTime}\big)
=\cov\left(
  \frac{w^M_{\unit,\time}w^P_{\unit,\time}}{\EMC[w^M_{\unit,\time}w^P_{\unit,\time}]}
  - \frac{w^P_{\unit,\time}}{\EMC[w^P_{\unit,\time}]},
  \frac{w^M_{\altUnit,\altTime}w^P_{\altUnit,\altTime}}{\EMC[w^M_{\altUnit,\altTime}w^P_{\altUnit,\altTime}]}
  - \frac{w^P_{\altUnit,\altTime}}{\EMC[w^P_{\altUnit,\altTime}]}
\right).
\end{equation}
Since 
\begin{equation}
\nonumber
\left|  \frac{w^M_{\unit,\time}w^P_{\unit,\time}}{\EMC[w^M_{\unit,\time}w^P_{\unit,\time}]}
  - \frac{w^P_{\unit,\time}}{\EMC[w^P_{\unit,\time}]} \right| <Q^{2\Bsize},
\end{equation}
we have 
\begin{equation}
\nonumber
\big| \cov\big(\centersum^{\loglik}_{\unit,\time},\centersum^{\loglik}_{\altUnit,\altTime}\big)
\big| < Q^{4\Bsize},
\end{equation}
implying that
\begin{equation}
\label{var:bound:without:assumption}
\var\left(\sum_{({\unit,\time})\in \spaceTime} \centersum^{\loglik}_{\unit,\time}\right)
< Q^{4\Bsize}\Unit^2\Time^2.
\end{equation}
If, in addition, $(\unit,\time)$ and $(\altUnit,\altTime)\in A_{\unit,\time}$ are sufficiently separated in the sense of Assumption~\ref{A:unconditional:mix}, then Lemma~\ref{lemma:covariance-bound} shows that Assumption~\ref{A:unconditional:mix} implies
\begin{equation}
\nonumber
\big| \cov\big(\centersum^{\loglik}_{\unit,\time},\centersum^{\loglik}_{\altUnit,\altTime}\big)
\big| < \etwo Q^{4\Bsize}.
\end{equation}
The number of insufficiently separated neighbors to $(\unit,\time)$ is bounded by $\altb$, and so we obtain
\begin{equation}
\label{var:bound:with:assumption}
\var\left(\sum_{({\unit,\time})\in S} \centersum^{\loglik}_{\unit,\time}\right)
< \ThmOneVarBound.
%Q^{4\Bsize}\big[ \Unit\Time \big( \altb + \etwo\Unit\Time \big) \big].
\end{equation}
Now we proceed to bound the bias in the Monte Carlo central limit estimator of $\loglik$.
Putting $h(x_{\unit,\time})=f_{Y_{\unit,\time}|X_{\unit,\time}}(\data{y}_{\unit,\time}\given x_{\unit,\time})$ into Assumption~\ref{A1}, using Assumption~\ref{A2}, gives
\begin{equation}
\nonumber
\big|
f_{Y_{\unit,\time}|Y_{B_{\unit,\time}}}(\data{y}_{\unit,\time}\given \data{y}_{B_{\unit,\time}})
- f_{Y_{\unit,\time}|Y_{A_{\unit,\time}}}(\data{y}_{\unit,\time}\given \data{y}_{A_{\unit,\time}})
\big|
< \eone Q.
\end{equation}
Noting that 
\begin{equation}
\label{LogInequality}
|a-b|<\delta, \hspace{3mm} a>Q^{-1} \mbox{ and } b>Q^{-1} \mbox{ implies }|\log(a)-\log(b)| < \delta Q, 
\end{equation}
we find
\begin{equation}
\label{eq:log:diff:identity}
\big|
\log f_{Y_{\unit,\time}|Y_{B_{\unit,\time}}}(\data{y}_{\unit,\time}\given \data{y}_{B_{\unit,\time}})
- \log f_{Y_{\unit,\time}|Y_{A_{\unit,\time}}}(\data{y}_{\unit,\time}\given \data{y}_{A_{\unit,\time}})
\big|
< \eone Q^2.
\end{equation}
Summing over $(\unit,\time)$, we get
\begin{equation}
\label{thm1:bias:bound}
\left| \loglik - 
\sum_{({\unit,\time})\in S}
\log f_{Y_{\unit,\time}|Y_{B_{\unit,\time}}}(\data{y}_{\unit,\time}\given \data{y}_{B_{\unit,\time}})
\right| 
< \eone Q^2 \Unit\Time.
\end{equation}
Together, the results in [\ref{thm1:loglik:linearization}], [\ref{var:bound:without:assumption}], [\ref{var:bound:with:assumption}] and [\ref{thm1:bias:bound}] confirm the assertions of the theorem.
\end{proof}

%%%%%%%%%%%%%%%%% lemma2 ##########################

%\newcommand\Xlemma{\Phi}
%\newcommand\Ylemma{\Psi}
%\newcommand\xlemma{\phi}
%\newcommand\ylemma{\psi}
\newcommand\Xlemma{U}
\newcommand\Ylemma{V}
\newcommand\xlemma{u}
\newcommand\ylemma{v}

\begin{lemma}\label{lemma:covariance-bound}
Suppose $\Xlemma$ and $\Ylemma$ are random variables with joint density satisfying 
\begin{equation}
\big| f_{\Ylemma|\Xlemma}(\ylemma\given \xlemma)-f_{\Ylemma}(\ylemma) \big| < \epsilon f_{\Ylemma}(\ylemma).
\end{equation}
Suppose $|g(\Xlemma)| < a$ and $|h(\Ylemma)| < b$ for some real-valued function $g$ and $h$.
Then, $\cov\big(g(\Xlemma),h(\Ylemma)\big) <  ab\epsilon$.
\end{lemma}

\begin{proof}
The result is obtained by direct calculation, as follows.
\begin{eqnarray}
\nonumber
\cov\big(g(\Xlemma),h(\Ylemma)\big) &=& \E\bigg[ g(\Xlemma) \, \E\Big[ h(\Ylemma)-\E[h(\Ylemma)] \, \Big| \,  \Xlemma\Big] \bigg]
\\
\nonumber
&=& \int \left\{ \int g(\xlemma) h(\ylemma) \big( f_{\Ylemma|\Xlemma}(\ylemma \given \xlemma)-f_{\Ylemma}(\ylemma)\big)\, d{\ylemma} \right\} f_{\Xlemma}(\xlemma)\, d{\xlemma}
\\
\nonumber
&<& \int \left\{ \int a b \epsilon f_{\Ylemma}(\ylemma) \, d{\ylemma} \right\} f_{\Xlemma}(\xlemma)\, d{\xlemma}
\\
\nonumber
&=& ab\epsilon.
\end{eqnarray}
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 22222222222222222

\section{\secTitleSpace {\ABF} and {\ABFIR} convergence: Proof of Theorem~2}

Let 
$g^{}_{\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time}}(\myvec{x}_{0:\Time},\myvec{x}^P_{1:\Time})$ 
be the joint density of the adapted process and the proposal process, 
\begin{equation}
\label{eq:g}
g^{}_{\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time}}(\myvec{x}_{0:\Time},\myvec{x}^P_{1:\Time})
= f_{\myvec{X}_0}(\myvec{x}_0)
\prod_{\time=1}^{\Time} 
f_{\myvec{X}_{\time}|\myvec{X}_{\time-1},\myvec{Y}_{\time}} 
  \big( 
    \myvec{x}_{\time} \given \myvec{x}_{\time-1},\data{\myvec{y}}_{\time} 
  \big)
\,\,
f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}
  \big(
    \myvec{x}^P_{\time} \given \myvec{x}_{\time-1}
  \big).
\end{equation}
For 
$\unitTimeSubset \subset \UnitSet\times 1:\Time$, 
define 
$\unitTimeSubset^{[m]}=\unitTimeSubset \cap \big(\UnitSet\times \{m\}\big)$ 
and set
\begin{equation}
\label{eq:h_S}
\adapted^{}_{\unitTimeSubset}
%(\myvec{x}_{0:\Time})
= 
\prod_{m=1}^{\Time} f_{Y_{\unitTimeSubset^{[m]}}|\myvec{X}^{}_{m-1}} \big( \data{y}_{\unitTimeSubset^{[m]}} \given \myvec{X}^{}_{m-1} \big),
\end{equation}
using the convention that an empty density $f_{Y_{\emptyset}}$ evaluates to 1.
If we denoting $\E_{g}$ for expectation for $(\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time})$ having density $g_{\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time}}$, \eqref{eq:h_S} can be written as
\begin{equation}
\label{eq:h_S2}
\nonumber
\adapted^{}_{\unitTimeSubset}
%(\myvec{x}_{0:\Time})
= 
\E_g\left[
f_{Y_{\unitTimeSubset}|{X}^{}_{\unitTimeSubset}} 
\big( 
  \data{y}_{\unitTimeSubset} \given {X}^{P}_{\unitTimeSubset}
\big)
\, \Big| \, 
\myvec{X}^{}_{0:\Time}
%=\myvec{x}^{}_{0:\Time}
\right],
\end{equation}
so we have
\[
\E_g \big[ \gamma_{B} \big] = \E_g \left[ f_{Y_{B}|X_{B}} \big(\data y_{B}| X_{B}^P \big) \right].
\]
Two useful identities are
\begin{eqnarray}
\nonumber
%% \label{eq:AB:ratio}
f_{X^{}_{\unit,\time}|Y_{A_{\unit,\time}}} 
\big( 
  x^{}_{\unit,\time}|\data{y}_{A_{\unit,\time}} 
\big) 
&=&
\frac{
  \E_{g}\Big[
    f_{Y^{}_{A_{\unit,\time}}|X^{}_{A_{\unit,\time}}} \big( \data y_{A_{\unit,\time}} \given X^P_{A^{}_{\unit,\time}} \big) \, 
    f^{}_{X_{\unit,\time}|X^{[\time]}_{A_{\unit,\time}},\myvec{X}_{\time-1}}\big( x^{}_{\unit,\time}|X^P_{A^{[\time]}_{\unit,\time}},\myvec{X}_{\time-1} \big)
  \Big]
}{
  \E_{g}\Big[
    f^{}_{Y_{A_{\unit,\time}}|X_{A_{\unit,\time}}}\big( \data{y}_{A_{\unit,\time}}|X^P_{A_{\unit,\time}} \big)
  \Big]
},
\\
\nonumber
%% \label{eq:AB:ratio2}
f_{Y^{}_{\unit,\time}|Y_{A_{\unit,\time}}} \big( \data{y}_{\unit,\time}|\data{y}_{A_{\unit,\time}} \big) &=&
\frac{\E_{g}\big[\adapted^{}_{A^+_{\unit,\time}} 
%\big( \myvec{X}_{0:\Time} \big) 
\big]
}{\E_{g}\big[\adapted^{}_{A_{\unit,\time}} 
%\big( \myvec{X}_{0:\Time} \big) 
\big]}.
\end{eqnarray}
%Thus, $\adapted_{\unitTimeSubset}$ is the proper weight for $f$ with respect to $g$ on $\unitTimeSubset$.

\Bi %% Assumption~\ref{B1}

\Bii %% Assumption~\ref{B1b}

\Biii %% Assumption~\ref{B2}

\begin{prop}\label{prop:ABF_AB}
Setting $h(x)= f_{Y_{\unit,\time}|X_{\unit,\time}}(\data{y}_{\unit,\time}|x)$, assumptions~\ref{B1} and~\ref{B2} imply
\begin{equation}
\Bigg| \,
\frac{\E_{g}\big[\adapted^{}_{A^+_{\unit,\time}}\big]}{\E_{g}\big[\adapted^{}_{A_{\unit,\time}}]}
- 
\frac{\E_{g}\big[\adapted^{}_{B^+_{\unit,\time}}\big]}{\E_{g}\big[\adapted^{}_{B_{\unit,\time}}]}
\,
\Bigg|
< Q \epsilon.
\end{equation}
\end{prop}
\begin{proof}
Using the non-negativity of all terms to justify interchange of integral and expectation,
\begin{equation}
\begin{split}
 & \int h(x) \E_g \Big[ f_{Y_{B_{u,n}}|X_{B_{u,n}}}(\data y_{B_{u,n}} | X_{B_{u,n}}^P) f_{X_{u,n}|X_{B_{u,n}^{[n]}},\myvec X_{n-1}}(x | X_{B_{u,n}^{[n]}}^P, \myvec X_{n-1}) \Big] dx \\
 & = \E_g \Big[ \int f_{Y_{u,n}|X_{u,n}}(\data y_{u,n}|x) f_{X_{u,n}|X_{B_{u,n}^{[n]}},\myvec X_{n-1}}(x | X_{B_{u,n}^{[n]}}^P, \myvec X_{n-1}) dx \cdot f_{Y_{B_{u,n}}|X_{B_{u,n}}}(\data y_{B_{u,n}}|X_{B_{u,n}}^P) \Big]
\end{split}
\label{eq:int_h_fubini}
\end{equation}
But by the construction of $g$,
\[
\begin{split}
f_{X_{u,n}|X_{B_{u,n}^{[n]}}, \myvec X_{n-1}} (x| X_{B_{u,n}^{[n]}}^P, \myvec X_{n-1})
&= g_{X_{u,n}^P|X_{B_{u,n}^{[n]}}^P, \myvec X_{n-1}}(x | X_{B_{u,n}^{[n]}}^P, \myvec X_{n-1}) \\
&= g_{X_{u,n}^P|X_{B_{u,n}}^P, \myvec X_{n-1}} (x|X_{B_{u,n}}^P, \myvec X_{n-1}).
\end{split}
\]
Thus \eqref{eq:int_h_fubini} becomes
\[
\begin{split}
  &\E_g \Big[ \int f_{Y_{u,n}|X_{u,n}}(\data y_{u,n}|x) g_{X_{u,n}^P|X_{B_{u,n}^P}, \myvec X_{n-1}} (x|X_{B_{u,n}}^P, \myvec X_{n-1}) dx \cdot f_{Y_{B_{u,n}}|X_{B_{u,n}}}(\data y_{B_{u,n}}|X_{B_{u,n}}^P) \Big]\\
  &= \E_g \left[ \E_g \Big[ f_{Y_{u,n}|X_{u,n}}(\data y_{u,n} | X_{u,n}^P) | X_{B_{u,n}}^P, \myvec X_{n-1} \Big] \cdot f_{Y_{B_{u,n}}|X_{B_{u,n}}}(\data y_{B_{u,n}} | X_{B_{u,n}}^P) \right]\\
  &= \E_g \Big[ f_{Y_{B_{u,n}^+} | X_{B_{u,n}^+}} (\data y_{B_{u,n}^+} | X_{B_{u,n}^+}^P) \Big]\\
  &= \E_g \gamma_{B_{u,n}^+}.
\end{split}
\]
Applying the same argument for the special case of $B_{u,n}=A_{u,n}$, we substitute into Assumption~\ref{B1} to complete the proof with the fact that $h<Q$.
\end{proof}

\BivA %% Assumption~\ref{B:unconditional:mix}

The mixing of the adapted process in Assumption~\ref{B:unconditional:mix} replaces the mixing of the unconditional process in  Assumption~\ref{A:unconditional:mix}.
Though mixing of the adapted process may be hard to check, one may suspect that the adapted process typically mixes more rapidly than the unconditional process.
%As for the TIF analysis, we note that requiring the inequality in Assumption~\ref{B:unconditional:mix} to hold over ${B^{+2c}_{\unit,\time}}$ implies that it holds over all subsets of  ${B^{+2c}_{\unit,\time}}$. 
Assumption~\ref{B:unconditional:mix} is needed only to ensure that the variance bound in Theorem~\ref{thm:abf} is essentially $O(\Unit \Time)$ rather than $O(\Unit^2 \Time^2)$. 
Either of these rates avoids the exponentially increasing variance characterizing the curse of dimensionality.
Lower variance than $O(\Unit \Time)$ cannot be anticipated for any sequential Monte Carlo method since the log likelihood estimate can be written as a sum of $\Unit\Time$ terms each of which involves its own sequential Monte Carlo calculation.
The following Proposition~\ref{prop:cov:mix} gives an implication of Assumption~\ref{B:unconditional:mix}

%%%% pppppppppp

\begin{prop} \label{prop:cov:mix}
  Assumption~\ref{B:unconditional:mix} implies that, if $(\altUnit, \altTime) \notin \altB_{\unit\comma\time}$,
\begin{equation}
\label{supp:eq:cov:g:bound}
\cov_{g_{}}\big(\adapted^{}_{B^{}_{\unit,\time}},\adapted^{}_{B^{}_{\altUnit\comma\altTime}} \big) < \efourA Q^{|B^{}_{\unit,\time}|+|B^{}_{\altUnit\comma\altTime}|}.
\end{equation}
\end{prop}
\begin{proof}
Write 
$\gamma=\adapted^{}_{B^{}_{\unit,\time}}
%(\myvec{X}_{0:\time})
$ and 
$\tilde\gamma=\adapted^{}_{B^{}_{\altUnit\comma\altTime}}
%(\myvec{X}_{0:\time})
$. 
Also, write $B=B^{}_{\unit,\time}$, $\tilde B=B^{}_{\altUnit,\altTime}$ and $f^{}_{B}(x^P_B)=f^{}_{Y_B|X_B}(\data{y}_B\given x_B^P)$.
Then,
\begin{eqnarray}
\nonumber
\hspace{-5mm} \E[\gamma\tilde\gamma]&=&
\int
  \left[
    \int f^{}_{B}(x^P_B)
         g^{}_{X^P_B|\myvec{X}_{0:\Time}}(x^P_B\given \myvec{x}_{0:\Time})\, dx^P_B
  \right]
\\
\nonumber
&& \hspace{15mm} 
\times
  \left[
    \int f^{}_{{\tilde B}}(x^P_{\tilde B})
         g^{}_{X^P_{\tilde B}|\myvec{X}_{0:\Time}}(x^P_{\tilde B}\given \myvec{x}_{0:\Time})\, dx^P_{\tilde B}
  \right]
  g^{}_{\myvec{X}_{0:\Time}}(\myvec{x}_{0:\Time})\, d\myvec{x}_{0:\Time}
\\
\nonumber
&=& \int \int f^{}_{B}(x^P_B)\, f^{}_{{\tilde B}}(x^P_{\tilde B})\,  \bigg\{ \int g^{}_{X^P_B|\myvec{X}_{0:\Time}}(x^P_B\given \myvec{x}_{0:\Time}) 
\\
\label{eq:prop:long:integral}
&& \hspace{15mm} 
\times \, 
g^{}_{X^P_{\tilde B}|\myvec{X}_{0:\Time}}(x^P_{\tilde B}\given \myvec{x}_{0:\Time}) \, g^{}_{\myvec{X}_{0:\Time}}(\myvec{x}_{0:\Time}) \, d\myvec{x}_{0:\Time} \bigg\}
\, dx^P_B \, dx^P_{\tilde B} 
\end{eqnarray}
Putting the approximate conditional independence requirement of Assumption~\ref{B:unconditional:mix} into \myeqref{eq:prop:long:integral}, we have
\begin{eqnarray}
\nonumber
&& \hspace{-15mm}
\Big|  
  \E[\gamma\tilde\gamma] -
  \int f^{}_{B}(x^P_B)\, f^{}_{{\tilde B}}(x^P_{\tilde B}) \, 
         g^{}_{X^P_{B}X^P_{\tilde B}|\myvec{X}_{0:\Time}}(x^{P}_{B},x^P_{\tilde B}\given \myvec{x}_{0:\Time})
\, g^{}_{\myvec{X}_{0:\Time}}(\myvec{x}_{0:\Time})
\, dx^P_B \, dx^P_{\tilde B} \, d\myvec{x}_{0:\Time}
\Big| 
\\
\nonumber
&& \hspace{40mm} < (1/2)\, \efourA \, Q^{|B|+|\tilde B|}.
\end{eqnarray}
This gives
\begin{equation}
\label{eq:prop:medium:integral}
\Big|  
  \E[\gamma\tilde\gamma] -
  \int f^{}_{B}(x^P_B)\, f^{}_{{\tilde B}}(x^P_{\tilde B}) \, 
         g^{}_{X^P_{B}X^P_{\tilde B}}(x^{P}_{B},x^P_{\tilde B})\, dx^P_B \, dx^P_{\tilde B} 
\Big| < (1/2) \, \efourA \, Q^{|B|+|\tilde B|}.
\end{equation}
Then, using the approximate unconditional independence requirement of Assumption~\ref{B:unconditional:mix}  combined with the triangle inequality, \myeqref{eq:prop:medium:integral} implies
\begin{equation}
\label{eq:prop:short:integral}
\Big| 
  \E[\gamma\tilde\gamma] -
  \int f^{}_{B}(x^P_B)\, f^{}_{{\tilde B}}(x^P_{\tilde B}) \, 
         g^{}_{X^P_{B}}(x^{P}_{B}) \,  g^{}_{X^P_{\tilde B}}(x^P_{\tilde B})\, dx^P_B \, dx^P_{\tilde B} 
\Big| < \efourA \, Q^{|B|+|\tilde B|}.
\end{equation}
We can rewrite \myeqref{eq:prop:short:integral} as
\begin{equation}
\label{eq:prop:without:integral}
\big|  \, 
  \E[\gamma\tilde\gamma] - \E[\gamma] \, \E[\tilde\gamma] 
\, \big|
< \efourA \, Q^{|B|+|\tilde B|},
\end{equation}
proving the proposition.
\end{proof}

\BivB %% Assumption~\ref{B:temporal:mix}

Assumption~\ref{B:temporal:mix} is needed to ensure the stability of the Monte Carlo approximation to the adapted process. 
It ensures that any error due to finite Monte Carlo sample size has limited consequences at sufficiently remote time points.
One could instead propose a bound that decreases exponentially with $K$, but that is not needed for the current purposes.
The following Proposition~\ref{lemma:E1E2} is useful for taking advantage of Assumption~\ref{B:temporal:mix}.

\begin{prop}\label{lemma:E1E2}
  Suppose that $f$ is a non-negative function and that for some $\epsilon>0$,
  \[|f(x) - f(x')| < \epsilon f(x')\]
  holds for all $x,x'$.
  Then for any two probability distributions where the expectations are denoted by $\E_1$ and $\E_2$ and for any random variable $X$, we have
  \[
  |\E_1 f(X) - \E_2 f(X) | \leq 
%\frac{\epsilon}{1+\epsilon} 
\epsilon \, \E_2 f(X).
  \]
%  and thus
%  \[
%  \frac{1}{1+\epsilon} \leq \frac{\E_1 f(X)}{\E_2 f(X)} \leq 1+\epsilon,
%  \]
%  provided that $\E_2 f(X) > 0$.
\end{prop}
\begin{proof}
  Let the two probability laws be denoted by $P_1$ and $P_2$. We have
  \[\begin{split}
  |\E_1 f(X) - \E_2 f(X)| &= \left| \int f(x) P_1(dx) - \int f(x') P_2(dx') \right|\\
  &\leq \left| \int \int f(x) P_1(dx) P_2(dx') - \int\int f(x') P_1(dx) P_2(dx') \right|\\
  &\leq \int\int | f(x) - f(x') | P_1(dx) P_2(dx')\\
  &\leq  \int 
%\frac{\epsilon}{1+\epsilon} 
\epsilon
f(x') P_2(dx') = 
%\frac{\epsilon}{1+\epsilon} 
\epsilon \, 
\E_2 f(X).
  \end{split}.\]
%  Thus,
%  \[
%  \frac{1}{1+\epsilon} = 1- \frac{\epsilon}{1+\epsilon} \leq \frac{\E_1 f(X)}{\E_2 f(X)} \leq 1 + \frac{\epsilon}{1+\epsilon} < 1+\epsilon.
%  \]
\end{proof}

\Bv  %% Assumption~\ref{B:girf}

Assumption~\ref{B:girf} controls the Monte Carlo error for a single time interval on a single bootstrap replicate. 
In the case $\Ninter=1$, {\ABFIR} becomes {\ABF} and this assumption is one of many alternatives for bounding error from importance sampling.  
The purpose behind the selection of Assumption~\ref{B:girf} is to draw on the results of \citet{park20} for intermediate resampling, and our assumption is a restatement of their Theorem~2.
When $\Ninter=1$, the curse of dimensionality for importance sampling has the consequence that $C_0$ grows exponentially with $\Unit$.
However, \citet{park20} showed that setting $\Ninter=\Unit$ can lead to situations where $C_0(\Unit,\Time,\Ninter)$ in Assumption~\ref{B:girf} grows polynomially with $\Unit$.
Here, we do not place requirements concerning the dependence of $C_0$ on $\Unit$, $\Time$ and $\Ninter$ since our immediate concern is a limit where $\Rep$ and $\Np$ increase. 
Nevertheless, the numerical results are consistent with the theoretical and empirical results obtained for intermediate resampling in the context of particle filtering by \cite{park20}.

\Bvi    %% Assumption~\ref{B:XG_XA_ind}

The Monte Carlo conditional independence required by Assumption~\ref{B:XG_XA_ind} would hold for {\ABFIR} if the guide variance $V_{\unit,\time,i}$ were calculated using an independent set of guide simulations to those used for evaluating the measurement weights $w^M_{\unit,\time,i,j}$.
For numerical efficiency, the {\ABFIR} algorithm implemented here constructs a shared pool of simulations for both purposes rather than splitting the pool up between them, in the expectation that the resulting minor violation of Assumption~\ref{B:XG_XA_ind} has negligible impact.

\TheoremII %% Theorem~\ref{thm:abf}

%%%%%%%%%%%%%%%%%%%%%%%% p2p2p2p2

\begin{proof}   
First, we set up some notation. For $\unitTimeSubset_{\unit\comma\time}$ and $w^M_{\unit,\time,\rep,\np}$ constructed by {\ABFIR}, define
\begin{equation}\label{eq:gamma:def}
\adapted^{MC,\rep}_{\unitTimeSubset_{\unit\comma\time}}
=
  \prod_{m=1}^{\time}
  \left[
    \frac{1}{\Np}\sum_{\np=1}^{\Np}
    \hspace{1mm}
       \prod_{(\altUnit,m)\in \unitTimeSubset^{[m]}_{\unit\comma\time}} 
    \hspace{-1mm}
        w^M_{\altUnit,m,\rep,\np}
  \right]
\hspace{5mm}
\mbox{and}
\hspace{5mm}
\bar\adapted^{MC}_{\unitTimeSubset_{\unit\comma\time}}
=\frac{1}{\Rep} \sum_{\rep=1}^{\Rep} \adapted^{MC,\rep}_{\unitTimeSubset_{\unit\comma\time}}.
\end{equation}
The Monte Carlo conditional likelihoods output by {\ABFIR} can be written as
\begin{equation}
\MC{\loglik}_{\unit,\time} = \log\MC{\bar\adapted}_{B^+_{\unit,\time}} - \log\MC{\bar\adapted}_{B^{\plusStrut}_{\unit,\time}}.
\end{equation}
We proceed with a similar argument to the proof of Theorem~\ref{thm:tif}. 
Since $\adapted^{MC,\rep}_{\unitTimeSubset_{\unit\comma\time}}$ are i.i.d. for $\rep \in \seq{1}{\Rep}$, we can suppose they are replicates of a Monte Carlo random variable $\adapted^{MC}_{\unitTimeSubset_{\unit\comma\time}}$.
We define
\begin{equation}
\nonumber
\centersum^{+}_{\unit,\time}= \frac{1}{\sqrt{\Rep}}\sum_{\rep=1}^\Rep\big(
\adapted^{MC,\rep}_{B^+_{\unit,\time}} - \E\big[\adapted^{MC}_{B^+_{\unit,\time}}\big]\big),
\hspace{2cm}
\centersum^{}_{\unit,\time}= \frac{1}{\sqrt{\Rep}}\sum_{\rep=1}^\Rep\big(
\adapted^{MC,\rep}_{B^{}_{\unit,\time}} - \E\big[\adapted^{MC}_{B^{}_{\unit,\time}}\big]\big).
\end{equation}
The same calculation as \myeqref{eq:likelihood:at:point} gives
\begin{eqnarray}
\label{eq:likelihood:at:point:B}
\MC{\loglik}_{\unit,\time} 
&=& 
\log\left(\frac{\E\big[\adapted^{MC}_{B^+_{\unit,\time}}\big]}{\E\big[\adapted^{MC}_{B^{\plusStrut}_{\unit,\time}}\big]}\right) +  \Rep^{-1/2}
\left(\frac{\centersum^{+}_{\unit,\time}}{\E\big[\adapted^{MC}_{B^+_{\unit,\time}}\big]} - \frac{\centersum^{}_{\unit,\time}}{\E\big[\adapted^{MC}_{B^{}_{\unit,\time}}\big]} \right) + o_P\big(\Rep^{-1/2}\big)
\end{eqnarray}
The joint distribution of $\big\{ (\centersum^{+}_{\unit,\time},\centersum^{}_{\unit,\time}), (\unit,\time)\in \spaceTime \big\}$ follows a standard central limit theorem as $\Rep\to\infty$.
Each term has mean zero, with variances and covariances uniformly bounded over $(\unit,\time,\altUnit,\altTime)$ due to Assumption~\ref{B2}.
From Proposition~\ref{prop:ABF_AB}, using the same reasoning as \myeqref{eq:log:diff:identity},
\begin{equation}
\label{thm2:eq3}
\left|  
  \log \left(
    \frac{\E_{g}\big[\adapted^{}_{A^+_{\unit,\time}}
\big]}{\E_{g}\big[\adapted^{}_{A^{}_{\unit,\time}}
\big]} 
  \right)
  - 
  \log \left(
     \frac{\E_{g}\big[\adapted^{}_{B^+_{\unit,\time}}
\big]}{\E_{g}\big[\adapted^{}_{B^{}_{\unit,\time}}
\big]}
  \right)
\right|
< \ethree Q^2.
\end{equation}
Now we use  Lemma~\ref{lemma:AS_bias} and \eqref{LogInequality} to obtain
\begin{eqnarray}
\nonumber
&&\left|  
  \log \left(
     \frac{\E_{g}\big[\adapted^{}_{B^+_{\unit,\time}}
\big]}{\E_{g}\big[\adapted^{}_{B^{}_{\unit,\time}}
\big]}
  \right)
  -
  \log \left(
     \frac{\EMC\big[\adapted^{MC}_{B^+_{\unit,\time}}\big]}{\EMC\big[\adapted^{}_{B^{}_{\unit,\time}}\big]}
  \right)
\right|
\\
\nonumber
&& \le
  \left|  
    \log \E_{g}\big[\adapted^{}_{B^+_{\unit,\time}}
    \big]
    -
    \log \EMC\big[\adapted^{}_{B^{+}_{\unit,\time}}\big]
  \right|
  +
  \left|
    \log \E_{g}\big[\adapted^{}_{B^{}_{\unit,\time}}
    \big]
    -
    \log \EMC\big[\adapted^{}_{B^{}_{\unit,\time}}\big]
  \right|
\\
\label{thm2:eq3b}
&& < 2 Q^{2\Bsize}\big(\boundLemmaUniform\big).
\end{eqnarray}
The proof of the central limit result in \myeqref{th:abf:lik:bound} is completed by combining \myeqref{eq:likelihood:at:point:B}, \myeqref{thm2:eq3} and \myeqref{thm2:eq3b}.
To show \myeqref{th:abf:lik:bound2} we check that $\centersum^{}_{\unit,\time}$ and $\centersum^{}_{\altUnit,\altTime}$ are weakly correlated when $(\unit,\time)$ and $(\altUnit,\altTime)$ are sufficiently separated. 
By the same reasoning as the proof of Theorem~\ref{thm:tif}, it is sufficient to show that $\adapted^{MC}_{B^{}_{\unit\comma\time}}$ and $\adapted^{MC}_{B^{}_{\altUnit\comma\altTime}}$ are weakly correlated.
These Monte Carlo quantities approximate $\adapted^{}_{B^{}_{\unit,\time}}(\myvec{X}_{0:\time-1})$ and $\adapted^{}_{B^{}_{\altUnit\comma\altTime}}(\myvec{X}_{0:\altTime-1})$ with $\myvec{X}$ drawn from $g$.
Let us suppose $\time\ge\altTime$, and write $d_{\unit,\time}=\time-\inf_{(v,m)\in B_{\unit,\time}}m$.
First, we consider the situation $\time-\altTime>K+d_{\unit,\time}$, in which case we can use the Markov property to give
\begin{equation}
\label{supp:eq:cov:MC:bound}
\cov\big(\adapted^{MC}_{B^{}_{\unit,\time}},\adapted^{MC}_{B^{}_{\altUnit\comma\altTime}}\big)
< \EMC \big[ \adapted^{MC}_{B^{}_{\altUnit\comma\altTime}} \big]
\sup_{\myvec{x}}\left\{
\EMC \big[ \adapted^{MC}_{B^{}_{\unit\comma\time}} \big| \myvec{X}^A_{\time-d_{\unit,\time}-K,1}=\myvec{x} \big]
- \EMC \big[ \adapted^{MC}_{B^{}_{\unit\comma\time}} \big] \right\}
\end{equation}
Then, the triangle inequality followed by applications of Assumption~\ref{B:temporal:mix} and Lemma~\ref{lemma:AS_bias} gives
\newcommand\boundA{2\efourB + 2 \big( K+d_{\unit,\time}\big)\efive^{} }
\begin{eqnarray}
\nonumber
&&\hspace{-15mm} \left|
  \EMC \big[ \adapted^{MC}_{B^{}_{\unit\comma\time}} \big| \myvec{X}^A_{\time-d_{\unit,\time}-K,1}=\myvec{x} \big]
  - \EMC \big[ \adapted^{MC}_{B^{}_{\unit\comma\time}} \big] 
\right|
\\
\nonumber
&& %\hspace{5mm}
\le 
  \Big|
    \E_g\big[ \adapted^{}_{B^{}_{\unit\comma\time}}
    \big| \myvec{X}_{\time-d_{\unit,\time}-K}=\myvec{x} \big]
    - \E_g \big[ \adapted^{}_{B^{}_{\unit\comma\time}}
    \big] 
  \Big|
\\
\nonumber
&& \hspace{15mm}
  +
  \Big|
    \EMC \big[ \adapted^{MC}_{B^{}_{\unit\comma\time}} \big| \myvec{X}^A_{\time-d_{\unit,\time}-K,1}=\myvec{x} \big]
    -    \E_g\big[ \adapted^{}_{B^{}_{\unit\comma\time}}
    \big| \myvec{X}_{\time-d_{\unit,\time}-K}=\myvec{x} \big]
  \Big|
\\
\nonumber
&& \hspace{15mm}
  +
  \Big| 
    \EMC \big[ \adapted^{MC}_{B^{}_{\unit\comma\time}} \big] 
   - \E_g\big[ \adapted^{}_{B^{}_{\unit\comma\time}}
    \big] 
  \Big|
\\
\label{supp:eq:cov:MC:triangle}
&& 
\le
Q^b 
\Big(
  \boundA
\Big) 
\end{eqnarray}
Putting \eqref{supp:eq:cov:MC:triangle} into \eqref{supp:eq:cov:MC:bound}, we get
\begin{equation}
\cov
\big(
  \adapted^{MC}_{B^{}_{\unit,\time}},\adapted^{MC}_{B^{}_{\altUnit\comma\altTime}}
\big) 
< 
Q^{2\Bsize} 
\Big(
  \boundA
\Big).
\end{equation}
Now we address the situation $\time-\altTime\le K+d_{\unit,\time}$. 
We apply Lemma~\ref{lemma:AS_bias} on the union $B^{}_{\unit\comma\time} \cup B^{}_{\altUnit\comma\altTime}$ for which the temporal depth is bounded by 
%$d_{B^{}_{\unit\comma\time} \cup B^{}_{\altUnit\comma\altTime}}\le K+d_{\unit,\time}+d_{\altUnit,\altTime}$.
$d \le K+d_{\unit,\time}+d_{\altUnit,\altTime}$.
This gives
\newcommand\boundB{(2K+d_{\unit,\time}+d_{\altUnit,\altTime})\efive^{} +\efourB}
\begin{equation}
\label{application:of:thm2:eq2}
\Big| \,
  \EMC \big[\adapted^{MC}_{\unitTimeSubset_{\unit\comma\time}}\adapted^{MC}_{\unitTimeSubset_{\altUnit\comma\altTime}}\big] - \E_{g_{}}\big[\adapted^{}_{\unitTimeSubset_{\unit\comma\time}}
   \adapted^{}_{\unitTimeSubset_{\altUnit\comma\altTime}}
   \big]
\, \Big|
< Q^{2\Bsize}
\Big( 
\boundB
\Big).
\end{equation}
From
Proposition~\ref{prop:cov:mix}, if $(\altUnit\comma\altTime) \notin \altB_{\unit\comma\time}$,
\begin{equation}
\label{supp:eq:cov:g:bound}
\cov_{g_{}}\big(\adapted^{}_{B^{}_{\unit,\time}}
  ,\adapted^{}_{B^{}_{\altUnit\comma\altTime}}
  \big) < \efourA Q^{2\Bsize}.
\end{equation}
Now, we establish that 
$\cov\big(\adapted^{MC}_{\unitTimeSubset_{\unit\comma\time}},
    \adapted^{MC}_{\unitTimeSubset_{\altUnit\comma\altTime}}\big)$ 
is close to  
$\cov_g\big(\adapted^{}_{\unitTimeSubset_{\unit\comma\time}},
     \adapted^{}_{\unitTimeSubset_{\altUnit\comma\altTime}}\big)$.
\begin{eqnarray}
\nonumber
&& \hspace{-15mm} 
\Big| 
  \cov\big(\adapted^{MC}_{\unitTimeSubset_{\unit\comma\time}},
    \adapted^{MC}_{\unitTimeSubset_{\altUnit\comma\altTime}}\big)
  - 
  \cov_g\big(\adapted^{}_{\unitTimeSubset_{\unit\comma\time}},
     \adapted^{}_{\unitTimeSubset_{\altUnit\comma\altTime}}\big) 
\Big| 
\\
\nonumber
&&\le   
\Big| \,
  \EMC \big[\adapted^{MC}_{\unitTimeSubset_{\unit\comma\time}}\adapted^{MC}_{\unitTimeSubset_{\altUnit\comma\altTime}}\big] - \E_{g}\big[\adapted^{}_{\unitTimeSubset_{\unit\comma\time}}
  \adapted^{}_{\unitTimeSubset_{\altUnit\comma\altTime}}
  \big]
\, \Big|
\\
\nonumber
&& \hspace{10mm}
+ 
\Big|
  \EMC \big[\adapted^{MC}_{\unitTimeSubset_{\unit\comma\time}}\big]
  \big(
    \EMC \big[\adapted^{MC}_{\unitTimeSubset_{\altUnit\comma\altTime}}\big]
    -  \E_g\big[\adapted^{}_{\unitTimeSubset_{\altUnit\comma\altTime}}
  \big]
  \big)
\Big|
\\
\nonumber
&& \hspace{10mm}
+ 
\Big|
  \EMC \big[\adapted^{}_{\unitTimeSubset_{\altUnit\comma\altTime}}
  \big]
  \big(
    \EMC \big[\adapted^{MC}_{\unitTimeSubset_{\unit\comma\time}}\big]
    -  \E_g\big[\adapted^{}_{\unitTimeSubset_{\unit\comma\time}}
    \big]
  \big)
\Big|
\\
\nonumber
&&
\label{supp:eq:lots_of_K}
< 
Q^{2\Bsize}
\Big(
  \boundB + 2(\boundLemmaUniform)
\Big).
\\
&&
<
Q^{2\Bsize}
\Big(
\PartOfBoundOffDiagonal
\Big)
\end{eqnarray}
Using \myeqref{supp:eq:lots_of_K} together with \myeqref{supp:eq:cov:g:bound} to bound the $\Unit\Time(\Unit\Time-\altb)$ off-diagonal covariance terms completes the derivation of \myeqref{th:abf:lik:bound2}.

\end{proof}

%% lllllllllllllllllllllllll

\begin{lemma} \label{lemma:AS_bias}
Suppose Assumptions~\ref{B2}, \ref{B:temporal:mix}, \ref{B:girf} and \ref{B:XG_XA_ind}. 
Suppose the number of particles $\Np$ exceeds the requirement for~\ref{B:girf}.
If we write $d_B=\max_{(u_1,n_1), (u_2,n_2) \in B} |n_1-n_2|$ for $B \subset 1\mycolon U \times 1\mycolon N$,
then for any $B$,
\begin{equation*}
\Big| \,
  \EMC\big[\adapted^{MC}_{\unitTimeSubset} \big| \myvec X_{n-d_B-K,1}^A = \myvec x \big] - \E_{g}\big[\adapted^{}_{\unitTimeSubset}
 \big| \myvec X_{n-d_B-K}=\myvec x \big]
\, \Big|
< Q^{|B|}(K+d_B) \epsilon_{\mathrm{B6}}^{}, \quad \forall \myvec x \in \mathbb X^U,
\end{equation*}
and
\begin{equation}
\label{thm2:eq2}
\Big| \,
  \EMC\big[\adapted^{MC}_{\unitTimeSubset} \big] - \E_{g}\big[\adapted^{}_{\unitTimeSubset}
  \big]
\, \Big|
< Q^{|B|}(\epsilon_{\mathrm{B5}}^{} + (K+d_B) \epsilon_{\mathrm{B6}}^{}).
\end{equation}
\end{lemma}
\begin{proof}
Suppose that $\max_{(u',n')\in B} n' = n$.
Define $\eta_\time(\myvec{x}_{\time})=1$ and, for $0\le m \le \time-1$,
\begin{equation}
\label{eq:eta}
\eta_{m}(\myvec{x}_{m})
= \E_{g} \! \left[ \, \prod_{k=m+1}^{\time} \adapted^{}_{B^{[k]}} 
  \, \Big| \myvec{X}_{m}=\myvec{x}_{m} \right].
\end{equation}
We have a recursive identity
\begin{equation}
\label{eq:thm2:recursion}
\eta_m(\myvec{X}_m)=\E_{g} \! \left[ \adapted^{}_{B^{[m+1]}}
  \, \eta_{m+1}(\myvec{X}_{m+1}) \Big| \myvec{X}_m
\right].
\end{equation}
By taking the expectation of \eqref{eq:eta}, we have
\begin{equation}
\E_{g}\big[\eta_0(\myvec{X}_0)\big]=\E_{g}\big[\adapted^{}_{B}
  \big].
\end{equation}
Note that $g$ has marginal density $f_{\myvec{X}_{0}}$ for $\myvec{X}_0$.
We analyze an {\ABFIR} approximation to \myeqref{eq:thm2:recursion}. 
The function $\eta_{m+1}(\myvec{x})$ is not in practice computationally available for evaluation via {\ABFIR}, but the recursion nevertheless leads to a useful bound.
Let $\myvec{X}^{\IF}_{m+1}[\np](\myvec{x}_{m})$ correspond to the variable $\myvec{X}^{IR}_{m+1,\Ninter,1,\np}$ constructed by {\ABFIR} conditional on $\myvec{X}^{\IF}_{m,1}=\myvec{x}_{m}$.
Equivalently,  $\myvec{X}^{\IF}_{m+1}[\np](\myvec{x}_{m})$ matches the variable $\myvec{X}^{\IF}_{m+1,1}$ in {\ABFIR} if the assignment $\myvec{X}^{\IF}_{m+1,1}=\myvec{X}^{IR}_{m+1,\Ninter,1,1}$ is replaced by $\myvec{X}^{\IF}_{m+1,1}=\myvec{X}^{IR}_{m+1,\Ninter,1,\np}$ conditional on $\myvec{X}^{\IF}_{m,1}=\myvec{x}_{m}$.
We define an approximation error $e_m(\myvec{x}_m)$ by
\begin{equation}
\label{thm2:eq7}
\eta_m(\myvec{x}_m) =  
\frac{1}{\Np}\sum_{\np=1}^{\Np}
f_{Y_{B^{[m+1]}}|\myvec{X}_m}(\data{y}_{B^{[m+1]}}\given \myvec{x}_m)
 \, \eta_{m+1}\big( \myvec{X}_{m+1}^{\IF}[\np](\myvec{x}_{m})\big) + e_m(\myvec{x}_{m}).
\end{equation}
From Assumptions~\ref{B2} and~\ref{B:girf}, $\EMC\big|e_m(\myvec{x}_m)\big|<\efive^{} \, Q^{\big|B^{[m+1:n]}\big|}$ uniformly over $\myvec{x}_m$, 
Thus, setting $r_m = \EMC | e_m(\myvec{X}^A_{m,1}) |$, we have
\begin{equation}
\label{eq:lemma:r}
r_m < \efive^{} \, Q^{\big|B^{[m+1:n]}\big|}.
\end{equation} 
Now, setting $\K=K+d_{\unit,\time}$, we commence to prove inductively that, for $\time-\K\le m \le \time$,
\begin{equation}
\label{lemma:inductive:hypothesis}
\Bigg| \,
  \eta_{\time-\K}(\myvec{x}) - \EMC 
  \Big[ \eta_m(\myvec{X}^A_{m,1})
    \prod_{k=\time-\K+1}^m 
f_{Y_{B^{[k]}}|\myvec{X}_{k-1}}(\data{y}_{B^{[k]}}\given \myvec{X}^A_{k-1,1})
    \Big| \myvec{X}^A_{n-\K,1}=\myvec{x}
  \Big]
\, \Bigg|
< (m-n+\K)\efive \, Q^{|B|}.
\end{equation}
First, suppose that \myeqref{lemma:inductive:hypothesis} holds for $m$. 
From \myeqref{thm2:eq7} and \myeqref{eq:lemma:r},
\begin{equation}
\left| 
 \eta_m(\myvec x_m)
 - 
 \EMC\left[
  \frac{1}{\Np}\sum_{\np=1}^{\Np}
 f_{Y_{B^{[m+1]}}|\myvec{X}_m}(\data{y}_{B^{[m+1]}}\given \myvec{x}_m)
 \, \eta_{m+1}\big( \myvec{X}_{m+1}^{\IF}[\np](\myvec{x}_{m})\big)
 \right]
\right| 
< \efive \, Q^{\big|B^{[m+1:n]}\big|}.
\end{equation}
Since the particles are exchangeable, the expectation of the mean of $\Np$ particles can be replaced with the expectation of the first particle.
Plugging in $\myvec x_m = \myvec X_{m,1}^A$ gives us
\begin{equation}
\label{eq:lemma:eta:bound}
\left| 
 \eta_m(\myvec X_{m,1}^A)
 - 
  f_{Y_{B^{[m+1]}}|\myvec{X}_m}\big(\data{y}_{B^{[m+1]}}\given  \myvec{X}_{m,1}^{A}\big)
 \EMC\left[
   \eta_{m+1}\big(  \myvec{X}_{m+1,1}^{A} \big) \middle| \myvec X_{m,1}^A
 \right]
\right| 
< \efive \, Q^{\big| B^{[m+1:n]} \big| }
\end{equation}
Putting \myeqref{eq:lemma:eta:bound} into \myeqref{lemma:inductive:hypothesis}, for $m\le \time$, and taking an iterated expectation with respect to $\myvec{X}^A_{m,1}$, we find that \myeqref{lemma:inductive:hypothesis} holds also for $m+1$.
Since \myeqref{lemma:inductive:hypothesis} holds trivially for $m=\time-\K$, it holds for $\time-\K\le m\le \time$ by induction.
Then, noting $\eta_{\time}(\myvec{x})=1$, we have from \myeqref{lemma:inductive:hypothesis} that
\[
\left|
\eta_{n-\K}(x)
- \EMC \left[ \prod_{k=n-\K+1}^n 
  f_{Y_{B^{[k]}}|\myvec{X}_{k-1}}\big(\data{y}_{B^{[k]}}\given  \myvec{X}_{k-1,1}^{A}\big)
\middle \vert \myvec X_{n-\K, 1}^A = x \right]
\right|
< \K \efive \, Q^{|B|}.
\]
Integrating the above inequality over $\myvec x$ with respect to the law of $\myvec X_{n-\K,1}^A$, we obtain
\begin{equation}\label{eq:AS_bias_induction_result}
\left|
\EMC [\eta_{n-\K}(\myvec X_{n-\K,1}^A)]
- \EMC \left[ \prod_{k=n-\K+1}^n 
  f_{Y_{B^{[k]}}|\myvec{X}_{k-1}}\big(\data{y}_{B^{[k]}}\given  \myvec{X}_{k-1,1}^{A}\big)
\right]
\right|
< \K\efive \, Q^{| B |}.
\end{equation}
But under Assumption~\ref{B:XG_XA_ind}, we have
\begin{equation}\label{eq:gammaMC_underB6}
\EMC \left[ \prod_{k=n-\K+1}^n 
  f_{Y_{B^{[k]}}|\myvec{X}_{k-1}}\big(\data{y}_{B^{[k]}}\given  \myvec{X}_{k-1,1}^{A}\big)
\right]
= \EMC \big[ \gamma^{MC}_{B}\big].
\end{equation}
Assumption~\ref{B:temporal:mix} says
\begin{equation}
\label{eq:eta:bound}
| \eta_{n-\K}(\myvec x_{n-\K}^{(1)}) - \eta_{n-\K}(\myvec x_{n-\K}^{(2)}) |
< \efourB \, \eta_{n-\K}(\myvec x_{n-\K}^{(2)}).
\end{equation}
Application of Proposition~\ref{lemma:E1E2} to \myeqref{eq:eta:bound} gives
\begin{equation}\label{eq:Eg_EMC_eta}
\Big| \E_g \big[ \eta_{n-\K}(\myvec X_{n-\K})\big] - \EMC \big[\eta_{n-\K}(\myvec X_{n-\K,1}^A) \big] \Big|
< \efourB \, \E_g \big[ \eta_{n-\K}(\myvec X_{n-\K}) \big]
< \efourB \, Q^{| B |}.
\end{equation}
Combining \eqref{eq:AS_bias_induction_result}, \eqref{eq:gammaMC_underB6}, and \eqref{eq:Eg_EMC_eta} completes the proof of Lemma~\ref{lemma:AS_bias}.
\end{proof}

\clearpage

\clearpage

\section{\secTitleSpace The correlated Brownian motion example}

%bbbbbbbbbbbbbbbbbbbbb

<<bm_load, echo=F>>=

bm_files_dir <- paste0("bm_",bm_run_level,"/")
if(!dir.exists(bm_files_dir)) dir.create(bm_files_dir)

bm_files_list <- list.files(path=bm_files_dir,full.names=TRUE)
for(filename in bm_files_list) load(file=filename)

bm_girf_times <- sapply(split(bm_jobs$girf_time,bm_jobs$U),mean)
bm_ubf_times <- sapply(split(bm_jobs$ubf_time,bm_jobs$U),mean)
bm_abf_times <- sapply(split(bm_jobs$abf_time,bm_jobs$U),mean)
bm_abfir_times <- sapply(split(bm_jobs$abfir_time,bm_jobs$U),mean)
bm_pfilter_times <- sapply(split(bm_jobs$pfilter_time,bm_jobs$U),mean)
bm_bpf_times <- sapply(split(bm_jobs$bpf_time,bm_jobs$U),mean)
bm_enkf_times <- sapply(split(bm_jobs$enkf_time,bm_jobs$U),mean)

@

<<bm_image_plot, echo=F, fig.height=3.5, fig.width=7, out.width="6.5in", fig.cap = paste('correlated Brownian motion simulation used in the main text')>>=
library(spatPomp)
library(fields)
par(mai=c(0.7,0.7,0.1,0.5))
bm_big <- bm_list[[1]]
image.plot(y=1:dim(states(bm_big))[1],x=time(bm_big),z=t(states(bm_big)),ylab="unit",xlab="time",col=gray.colors(33))
@

To help visualize the correlated Brownian motion model, Fig.~\ref{fig:bm_image_plot} shows one of the simulations used for the results in Figure~{\MainFigureMeaslesSlice} of the main text.
Table~\ref{tab:bm} gives the algorithmic settings used for the filters and corresponding computational resource requirements.
Broadly speaking, $\Rep\Np$ for {\ABF} and {AIRSIF} should be compared with $\Rep$ for {\UBF}, $\Np$ for PF, and $\Np\Nguide$ for GIRF.

The computational effort allocated to each algorithm in Table~\ref{tab:bm} is given in core minutes.
{\UBF}, {\ABF} and {\ABFIR} parallelize readily, which is less true for PF and GIRF.
Therefore, the {\UBF}, {\ABF} and {\ABFIR} implementations run on all available cores (\Sexpr{bm_cores} for this experiment)  whereas the PF and GIRF implementations run on a single core.
If sufficient replications are being carried out to utilize all available cores, comparison of core minute utilization is equivalent to comparison of total computation time.
However, a single replication of {\UBF}, {\ABF} or {\ABFIR} proceeds more quickly due to the parallelization.

{\ABFIR} and GIRF have computational time scaling quadratically with $\Unit$ in this example, whereas the other methods scale linearly.
This is because the number of intermediate steps used, $\Ninter$, grows linearly with $\Unit$.

The main purpose of this example is not to provide a comparison between the functional capabilities of the methods on interesting scientific problems.
It is a toy example without the complexities that the methods are intended to address.
This simple example does show clearly the quick decline of PF and the slower declines of GIRF and {\ABF} as dimension increases.

\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline &
  {\UBF}  &
  {\ABF} &
  {\ABFIR}  &
  GIRF  &
  PF &
  BPF &
  EnKF
\\ 
\hline 
\rule{0mm}{4.5mm}particles,
$\Np$ &
  --- &
  \Sexpr{bm_abf_Np_per_replicate} &
  \Sexpr{bm_abfir_Np_per_replicate} &
  \Sexpr{bm_girf_Np} &
  \Sexpr{bm_pfilter_Np} &
  \Sexpr{bm_bpf_Np} &
  \Sexpr{bm_enkf_Np}
\\
bootstrap replications, $\Rep$ &
  \Sexpr{bm_ubf_Nrep} &
  \Sexpr{bm_abf_Nrep} &
  \Sexpr{bm_abfir_Nrep} &
  --- &
  --- &
  --- &
  ---
\\
guide simulations, $\Nguide$ &
  ---&
  ---&
  ---&
  \Sexpr{bm_girf_nguide}&
  --- &
  --- &
  ---
\\
lookahead lag, $L$ &
  ---&
  ---&
  ---&
  \Sexpr{bm_girf_lookahead} &
  --- &
  --- &
  ---
\\
intermediate steps, $S$ &
  ---&
  --- &
  $\Unit/2$ &
  $\Unit$ &
  --- &
  --- &
  ---
\\
\cline{2-4}
\hspace{-3mm} \begin{tabular}{l}
neighborhood, $B_{\unit\comma\time}$ \\
or block size
\end{tabular} &
\multicolumn{3}{c|}{
$ \rule[-4mm]{0mm}{11mm}
\begin{array}{c} 
$\big\{(\unit-1,\time),(\unit-2,\time),$ \\  $(\unit,\time-1),(\unit,\time-2)\big\}$
\end{array}$
} &
  --- &
  --- &
  \Sexpr{bm_bpf_units_per_block} &
  ---
\\
\cline{2-5}
forecast mean, $\myvec{\mu}(\myvec{x},s,t)$ &
  --- &
  --- &
  \multicolumn{2}{c|}{$\myvec{x}$} &
  --- &
  --- &
  ---
\\
measurement mean, $h_{\unit\comma\time}(x)$ &
  --- &
  --- &
  \multicolumn{2}{c|}{$x$} &
  --- &
  --- &
  $x$
\\
\cline{4-5}
$\tau={\VtoTheta}_{\unit\comma\time}(V,x)$ &
  --- &
  --- &
  \multicolumn{2}{c|}{\rule{0mm}{5mm}$\sqrt{V}$}  &
  --- &
  --- &
  ---
\\
$V={\thetaToV}_{\unit\comma\time}(\tau,x)$ &
  --- &
  --- &
  \multicolumn{2}{c|}{$\tau^2$}  &
  --- &
  --- &
  $\tau^2$
\\
\hline
\rule{0mm}{4.5mm}effort (core mins, $U=\Sexpr{bm_U[1]}$) & 
  \Sexpr{myround(bm_ubf_times[1]/60,1)} & 
  \Sexpr{myround(bm_abf_times[1]*bm_cores/60,1)} & 
  \Sexpr{myround(bm_abfir_times[1]*bm_cores/60,1)} & 
  \Sexpr{myround(bm_girf_times[1]/60,1)}  & 
  \Sexpr{myround(bm_pfilter_times[1]/60,1)} &
  \Sexpr{myround(bm_bpf_times[1]/60,1)} &
  \Sexpr{myround(bm_enkf_times[1]/60,1)} 
\\
effort  (core mins,  $U=\Sexpr{bm_U[2]}$) & 
  \Sexpr{myround(bm_ubf_times[2]/60,1)} & 
  \Sexpr{myround(bm_abf_times[2]*bm_cores/60,1)} & 
  \Sexpr{myround(bm_abfir_times[2]*bm_cores/60,1)} & 
  \Sexpr{myround(bm_girf_times[2]/60,1)} & 
  \Sexpr{myround(bm_pfilter_times[2]/60,1)} &
  \Sexpr{myround(bm_bpf_times[2]/60,1)} &
  \Sexpr{myround(bm_enkf_times[2]/60,1)} 
\\
effort  (core mins, $U=\Sexpr{bm_U[3]}$) & 
  \Sexpr{myround(bm_ubf_times[3]/60,1)} & 
  \Sexpr{myround(bm_abf_times[3]*bm_cores/60,1)} & 
  \Sexpr{myround(bm_abfir_times[3]*bm_cores/60,1)} & 
  \Sexpr{myround(bm_girf_times[3]/60,1)} & 
  \Sexpr{myround(bm_pfilter_times[3]/60,1)} &
  \Sexpr{myround(bm_bpf_times[3]/60,1)} &
  \Sexpr{myround(bm_enkf_times[3]/60,1)} 
\\
effort  (core mins,  $U=\Sexpr{bm_U[5]}$) & 
  \Sexpr{myround(bm_ubf_times[5]/60,1)} & 
  \Sexpr{myround(bm_abf_times[5]*bm_cores/60,1)} & 
  \Sexpr{myround(bm_abfir_times[5]*bm_cores/60,1)} & 
  \Sexpr{myround(bm_girf_times[5]/60,1)} & 
  \Sexpr{myround(bm_pfilter_times[5]/60,1)} &
  \Sexpr{myround(bm_bpf_times[5]/60,1)} &
  \Sexpr{myround(bm_enkf_times[5]/60,1)}   
\\
effort  (core mins,  $U=\Sexpr{bm_U[7]}$) & 
  \Sexpr{myround(bm_ubf_times[7]/60,1)} & 
  \Sexpr{myround(bm_abf_times[7]*bm_cores/60,1)} & 
  \Sexpr{myround(bm_abfir_times[7]*bm_cores/60,1)} & 
  \Sexpr{myround(bm_girf_times[7]/60,1)} & 
  \Sexpr{myround(bm_pfilter_times[7]/60,1)} &
  \Sexpr{myround(bm_bpf_times[7]/60,1)} &
  \Sexpr{myround(bm_enkf_times[7]/60,1)} 
\\
\hline

\end{tabular}
\caption{Algorithmic settings for the correlated Brownian motion numerical example. 
Computational effort is measured in core minutes for running one filter, corresponding to a point on Figure~{\MainFigureBmScaling} in the main text. 
The time taken for computing a single point using the parallel {\UBF}, {\ABF} and {\ABFIR} implementations is the effort divided by the number of cores, here $\Sexpr{doob_cores}$.
The time taken for computing a single point using the single-core GIRF, PF, BPF and EnKF implementations is equal to the effort in core minutes.
}\label{tab:bm}
\end{center}
\end{table}

\clearpage

%%%% sssssss

%%%% mmmmmmmmmm

\section{\secTitleSpace The measles example}

<<measlesModel_load, echo=F>>=

measles_model_dir <- "measlesModel/"
load(file=paste0(measles_model_dir,"measles_spatPomp.rda"))

@

<<measles_image_plot, echo=F, eval=F, fig.height=6, fig.width=7, out.width="6.5in", fig.cap = paste('Log(reported cases $+$ 1) for (A) the measles simulation used for the likelihood slice; (B) the corresponding UK measles data.')>>=
library(spatPomp)
library(fields)
par(mai=c(0.5,0.7,0.1,0.1))
par(mfrow=c(2,1))
image.plot(y=1:dim(obs(measles_sim))[1],x=time(measles_sim),z=log(t(obs(measles_sim)+1)),ylab="unit",xlab="",col=gray.colors(33))
mtext("A",side=2,line=2.5,las=1,padj=-5.5,cex=1.8)
image.plot(y=1:dim(obs(measles_uk))[1],x=time(measles_uk),z=log(t(obs(measles_uk)+1)),ylab="unit",xlab="",col=gray.colors(33))
mtext("B",side=2,line=2.5,las=1,padj=-5.5,cex=1.8)
mtext("time",side=1,line=2.5)
@

%%%% msmsmsmsmsmsmsmsmsmsmsmsmsmsms

<<mscale_load,cache=FALSE,echo=F>>=

mscale_files_dir <- paste0("mscale_",mscale_run_level,"/")

load(file=paste0(mscale_files_dir,"mscale_settings.rda"))
load(file=paste0(mscale_files_dir,"mscale_results.rda"))

if(0){
mscale_jobs <- expand.grid(U=mscale_U,reps=1:mscale_replicates)
mscale_jobs$U_id <- rep(seq_along(mscale_U),times=mscale_replicates)

load(file=paste0(mscale_files_dir,"mscale_girf.rda"))

mscale_jobs$girf_logLik <- vapply(mscale_girf_list,function(x)x$logLik,numeric(1))
mscale_jobs$girf_time <- vapply(mscale_girf_list,function(x) x$time,numeric(1))

load(file=paste0(mscale_files_dir,"mscale_abf.rda"))

mscale_jobs$abf_logLik <- vapply(mscale_abf_list,function(x)x$logLik,numeric(1))
mscale_jobs$abf_time <- vapply(mscale_abf_list,function(x) x$time,numeric(1))

load(file=paste0(mscale_files_dir,"mscale_ubf.rda"))

mscale_jobs$ubf_logLik <- vapply(mscale_ubf_list,function(x)x$logLik,numeric(1))
mscale_jobs$ubf_time <- vapply(mscale_ubf_list,function(x) x$time,numeric(1))

load(file=paste0(mscale_files_dir,"mscale_abfir.rda"))

mscale_jobs$abfir_logLik <- vapply(mscale_abfir_list,function(x)x$logLik,numeric(1))
mscale_jobs$abfir_time <- vapply(mscale_abfir_list,function(x) x$time,numeric(1))

load(file=paste0(mscale_files_dir,"mscale_pfilter.rda"))

mscale_jobs$pfilter_logLik <- vapply(mscale_pfilter_list,function(x)x$logLik,numeric(1))
mscale_jobs$pfilter_time <- vapply(mscale_pfilter_list,function(x) x$time,numeric(1))

mscale_results <- data.frame(
  Method=rep(mscale_methods,each=nrow(mscale_jobs)),
  logLik=c(
    mscale_jobs$ubf_logLik,
    mscale_jobs$abf_logLik,
    mscale_jobs$abfir_logLik,
    mscale_jobs$enkf_logLik,
    mscale_jobs$girf_logLik,
    mscale_jobs$pfilter_logLik
  ),
  Units=rep(mscale_jobs$U,reps=length(mscale_methods))
)
mscale_point_perturbation <- 0.25
mscale_results$U <- mscale_results$Units+
  rep( mscale_point_perturbation*seq(from=-1,to=1,length=length(mscale_methods)),
    each=nrow(mscale_jobs))
mscale_results$logLik_per_unit <- mscale_results$logLik/mscale_results$Units
mscale_results$logLik_per_obs <- mscale_results$logLik_per_unit/mscale_N
}
@

\begin{table}
\renewcommand{\arraystretch}{1.2}
  \centering
  \begin{tabular}{|crcl|}
    \hline
    parameter & value & unit & description  \\  \hline
    $\meanBeta$ & \Sexpr{myround(measles_params["R0"]*(measles_params["gamma"]+measles_params["mu"]),1)} & year$^{-1}$& mean contact rate\\
    $\mu_{\bullet D}^{-1}$ & \Sexpr{myround(1/measles_params["mu"],1)} &year& mean duration in the population \\
    $\mu_{EI}^{-1}$ & \Sexpr{myround(365/measles_params["sigma"],1)} &day& latent period  \\
    $\mu_{IR}^{-1}$ & \Sexpr{myround(365/measles_params["gamma"],1)} &day& infectious period \\
    $\sigma_{SE}$ & \Sexpr{myround(measles_params["sigmaSE"],3)} &year$^{1/2}$& process noise\\
    $\amplitude$ & \Sexpr{myround(measles_params["amplitude"],3)} &--- & amplitude of seasonality  \\
    $\alpha$ & \Sexpr{measles_params["alpha"]} &---& mixing exponent \\
    $\tau$ & 4 & year & delay from birth to entry into susceptibles\\    
%%    $\cohort$ & \Sexpr{measles_params["amplitude"]} &---& cohort fraction entering $S$ on September 6 \\
    $\rho$ & \Sexpr{measles_params["rho"]} &---& reporting probability \\
    $\psi$ & \Sexpr{measles_params["psi"]} &---& reporting overdispersion \\
    $G$ & \Sexpr{measles_params["g"]} &---& gravitation constant \\  \hline
  \end{tabular}
  \caption{\label{tab:measles_parameters} Table of parameters for the spatiotemporal measles transmission model}
\end{table}

% Fig.~\ref{fig:slice_image_plot} shows the simulation used for the measles scaling experiment in  Figure~{\MainFigureMeaslesScaling} and the likelihood slice shown in Figure~{\MainFigureMeaslesSlice} of the main text.
% Unit $\unit$ models the $u$th largest city in UK.

Table~\ref{tab:measles_parameters} gives the model parameter values and Table~\ref{tab:mscale} gives the algorithmic settings used for the filters.
The times in Table~\ref{tab:mscale} give the total time required by each algorithm to calculate all its results for Figure~{\MainFigureMeaslesScaling} in the main text, using \Sexpr{mscale_cores} cores.
The expected forecast function $\mu(x,s,t)$ needed for {\ABFIR} and GIRF was computed using a numerical solution to the deterministic skeleton of the stochastic model, i.e, a system of ODEs with derivative matching the infinitesimal mean function of the stochastic dynamic model.
In the specifications of $h_\unit\big(x)$, $\VtoTheta_\unit(V,x,\theta)$ and $\thetaToV_\unit(\theta,x)$, the latent process value $x$ contains a variable $C$ giving the cumulative removed infections in the current observation interval.

In Table~\ref{tab:mscale}, we see that the effort allocated to {\UBF}, {\ABF} and PF scales linearly with $\Unit$, since the number of bootstrap replications and particles is fixed in this experiment. 
GIRF computational effort scales fairly linearly in $\Unit$, since its effort is dominated by the guide simulations (which are linear in $\Unit$) rather than by the intermediate timestep calculations (which are quadratic in $\Unit$ since we carry out $\Unit$ intermediate calculations each of size linear in $\Unit$).
The effort allocated to {\ABFIR} scales with $\Unit^2$, since {\ABFIR} is more parsimonious with guide simulations (all particles in one bootstrap replication share the same guide simulations) and so the intermediate timestep calculations dominate the effort.
To obtain stable variance in the log likelihood estimate, the number of particles and bootstrap replications would have to grow with $\Unit$.
However, given a constraint on total computational resources, the number of particles and bootstrap replications would have to shrink as $\Unit$ increases.
The limit studied in this experiment is a balance between the two: the assumption is that one is prepared to invest a growing amount of computational effort as the data grow, but this should not grow too fast.
{\ABFIR} was permitted the greatest computational effort, but two considerations balance this.
\begin{enumerate}
\item Parallelization. {\UBF}, {\ABF} and {\ABFIR} are trivially parallelizable. 
The value of parallelization depends, among other things, on how many replications are being computed simultaneously and on how many cores are available.
Nevertheless, it is helpful that the core minute effort requirement for {\ABF} and {\ABFIR} can be divided by the number of available cores to give the computational time. 
Parallelizations of GIRF and PF can be constructed \citep{park20} but these involve non-trivial interaction between processors leading to additional algorithmic complexity and computational overhead. 
\item Memory. 
The intermediate timestep calculations in {\ABFIR} and GIRF do not add to the memory requirement, and the memory demands of {\UBF}, {\ABF} and {\ABFIR} are distributed across the parallel computations. 
A basic PF implementation for a large model can become constrained by its memory requirement (linear in the number of particles) before it can match the processor effort employed by the other algorithms.
\end{enumerate}

<<mscale_times, echo=F>>=

mscale_girf_times <- sapply(split(mscale_jobs$girf_time,mscale_jobs$U),mean)
mscale_abf_times <- sapply(split(mscale_jobs$abf_time,mscale_jobs$U),mean)
mscale_ubf_times <- sapply(split(mscale_jobs$ubf_time,mscale_jobs$U),mean)
mscale_abfir_times <- sapply(split(mscale_jobs$abfir_time,mscale_jobs$U),mean)
mscale_pfilter_times <- sapply(split(mscale_jobs$pfilter_time,mscale_jobs$U),mean)
mscale_bpf_times <- sapply(split(mscale_jobs$bpf_time,mscale_jobs$U),mean)
mscale_enkf_times <- sapply(split(mscale_jobs$enkf_time,mscale_jobs$U),mean)

K <- mscale_bpf_units_per_block
if(K==1) mscale_blocks_text <- "{\\Unit}" else mscale_blocks_text <- paste0("{\\Unit/",mscale_bpf_units_per_block,"}")

@

\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
  \hline & 
  {\UBF}  & 
  {\ABF} & 
  {\ABFIR}  & 
  GIRF &
  EnKF &
  PF &
  BPF 
\\ 
  \hline 
  particles, $\Np$    \rule{0mm}{4.5mm} & 
  1 &
  \Sexpr{mscale_abf_Np_per_replicate} & 
  \Sexpr{mscale_abfir_Np_per_replicate} & 
  \Sexpr{mscale_girf_Np}  & 
  \Sexpr{mscale_enkf_Np} &
  \Sexpr{mscale_pfilter_Np} &
  \Sexpr{mscale_bpf_Np} 
\\
  replicates, $\Rep$ & 
  \Sexpr{mscale_ubf_Nrep} & 
  \Sexpr{mscale_abf_Nrep} & 
  \Sexpr{mscale_abfir_Nrep} &
  --- & 
  --- &
  --- &
  ---
\\
  guide simulations, $\Nguide$ &
  --- & 
  --- &
  --- & 
  \Sexpr{mscale_girf_nguide} & 
  --- &
  --- &
  ---
\\
  lookahead lag, $L$ &
  --- &
  --- &
  --- & \Sexpr{mscale_girf_lookahead} & 
  --- &
  --- &
  ---
\\ 
  intermediate steps, $S$ & 
  --- & 
  --- &
  $\Unit/2$ & 
  $\Unit$ & 
  --- &
  --- &
  ---
\\
  \cline{2-4}
\hspace{-3mm} \begin{tabular}{l}
neighborhood, $B_{\unit\comma\time}$ \\
or block size
\end{tabular}
&
  \multicolumn{3}{c|}{
    \rule[-3mm]{0mm}{8mm}$\big\{(\unit,\time-1),(\unit,\time-2)\big\}$
  } &
  --- &
  --- &
  --- &
  \Sexpr{mscale_bpf_units_per_block} 
\\
  \cline{2-6}

  measurement mean, $h_{\unit\comma\time}\big(x)$ &
  --- &
  --- &
  \multicolumn{3}{c|}{\rule[-3mm]{0mm}{8mm}$\rho C$} &
  --- &
  --- 
\\  
  $V=\thetaToV_{\unit\comma\time}(\psi,\rho,x)$ &
  --- &
  --- &
  \multicolumn{3}{c|}{\rule[-3mm]{0mm}{8mm}$\rho(1-\rho)C + \rho^2C^2\psi^2$} &
  --- &
  ---
\\
  \cline{2-6}

  forecast mean, $\myvec{\mu}(\myvec{x},s,t)$ &
  --- &
  --- &
  \multicolumn{2}{c|}{\rule{0mm}{5mm}ODE model} &
  --- &
  --- &
  ---
\\
  $\psi=\VtoTheta_{\unit\comma\time}(V,x)$ &
  --- &
  --- &
  \multicolumn{2}{c|}{$\frac{\sqrt{V - \rho(1-\rho)C}}{\rho C}$} &
  --- &
  --- &
  ---
\\
\hline
  effort (core mins, $U=\Sexpr{mscale_U[which(mscale_U==2)]}$)   \rule{0mm}{4.5mm}  & 
  \Sexpr{myround(mscale_ubf_times[names(mscale_girf_times)==2]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_abf_times[names(mscale_girf_times)==2]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_abfir_times[names(mscale_girf_times)==2]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_girf_times[names(mscale_girf_times)==2]/60,1)}  & 
  \Sexpr{myround(mscale_enkf_times[names(mscale_girf_times)==2]/60,1)}  & 
  \Sexpr{myround(mscale_pfilter_times[names(mscale_girf_times)==2]/60,1)} &
  \Sexpr{myround(mscale_bpf_times[names(mscale_girf_times)==2]/60,1)} 
\\
  effort (core mins, $U=\Sexpr{mscale_U[which(mscale_U==4)]}$)  & 
  \Sexpr{myround(mscale_ubf_times[names(mscale_girf_times)==4]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_abf_times[names(mscale_girf_times)==4]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_abfir_times[names(mscale_girf_times)==4]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_girf_times[names(mscale_girf_times)==4]/60,1)}  & 
  \Sexpr{myround(mscale_enkf_times[names(mscale_girf_times)==4]/60,1)}  & 
  \Sexpr{myround(mscale_pfilter_times[names(mscale_girf_times)==4]/60,1)} &
  \Sexpr{myround(mscale_bpf_times[names(mscale_girf_times)==4]/60,1)} 
\\
  effort (core mins, $U=\Sexpr{mscale_U[which(mscale_U==8)]}$)  & 
  \Sexpr{myround(mscale_ubf_times[names(mscale_girf_times)==8]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_abf_times[names(mscale_girf_times)==8]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_abfir_times[names(mscale_girf_times)==8]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_girf_times[names(mscale_girf_times)==8]/60,1)}  &
  \Sexpr{myround(mscale_enkf_times[names(mscale_girf_times)==8]/60,1)}  &   
  \Sexpr{myround(mscale_pfilter_times[names(mscale_girf_times)==8]/60,1)} &
  \Sexpr{myround(mscale_bpf_times[names(mscale_girf_times)==8]/60,1)} 
\\
  effort (core mins, $U=\Sexpr{mscale_U[which(mscale_U==16)]}$)   & 
  \Sexpr{myround(mscale_ubf_times[names(mscale_girf_times)==16]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_abf_times[names(mscale_girf_times)==16]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_abfir_times[names(mscale_girf_times)==16]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_girf_times[names(mscale_girf_times)==16]/60,1)}  & 
  \Sexpr{myround(mscale_enkf_times[names(mscale_girf_times)==16]/60,1)}  & 
  \Sexpr{myround(mscale_pfilter_times[names(mscale_girf_times)==16]/60,1)} &
  \Sexpr{myround(mscale_bpf_times[names(mscale_girf_times)==16]/60,1)} 
\\
  effort (core mins, $U=\Sexpr{mscale_U[which(mscale_U==32)]}$)   & 
  \Sexpr{myround(mscale_ubf_times[names(mscale_girf_times)==32]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_abf_times[names(mscale_girf_times)==32]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_abfir_times[names(mscale_girf_times)==32]*mscale_cores/60,1)} & 
  \Sexpr{myround(mscale_girf_times[names(mscale_girf_times)==32]/60,1)}  & 
  \Sexpr{myround(mscale_enkf_times[names(mscale_girf_times)==32]/60,1)}  & 
  \Sexpr{myround(mscale_pfilter_times[names(mscale_girf_times)==32]/60,1)} &
  \Sexpr{myround(mscale_bpf_times[names(mscale_girf_times)==32]/60,1)}  
\\
\hline

\end{tabular}
\caption{Algorithmic settings for the measles example calculations in Figures~{\MainFigureMeaslesScaling} and~{\MainFigureMeaslesSlice}. 
Computational effort is measured in core minutes for running one filter, corresponding to a point on Figure~\MainFigureMeaslesScaling.
The time taken for computing a single point using the parallel {\UBF}, {\ABF} and {\ABFIR} implementations is the effort divided by the number of cores, here $\Sexpr{mscale_cores}$.
The time taken for computing a single point using the single core GIRF, EnKF, PF and BPF implementations is equal to the effort in core minutes.
}\label{tab:mscale}
\end{center}
\end{table}

\clearpage

\clearpage

\section{\secTitleSpace Varying the neighborhood for measles}

%%% nnnnnnnnnnnnnnnnnnnnn

<<abfNbhd_settings,cache=FALSE,echo=F>>=

abfNbhd_files_dir <- paste0("abfNbhd_",abfNbhd_run_level,"/")
if(!dir.exists(abfNbhd_files_dir)) dir.create(abfNbhd_files_dir)

stew(file=paste0(abfNbhd_files_dir,"abfNbhd_settings.rda"),{

  # copy variables that should be included in the stew
  abfNbhd_run_level <- abfNbhd_run_level 
  abfNbhd_cores <- cores

  abfNbhd_tol <- 1e-300

    abfNbhd1_name <- "NBHD2"
    abfNbhd1_nbhd <- function(object, time, unit) {
      nbhd_list <- list()
      if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
      if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
      return(nbhd_list)
    }

   abfNbhd2_name <- "NBHD4"
    abfNbhd2_nbhd <- function(object, time, unit) {
      nbhd_list <- list()
      if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
      if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
      if(unit>1) nbhd_list <- c(nbhd_list, list(c(unit-1, time)))
      return(nbhd_list)
    }

    abfNbhd3_name <- "NBHD5"
    abfNbhd3_nbhd <- function(object, time, unit) {
      nbhd_list <- list()
      if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
      if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
      if(unit>1) nbhd_list <- c(nbhd_list, list(c(1, time)))
      return(nbhd_list)
    }

    abfNbhd4_name <- "NBHD1"
    abfNbhd4_nbhd <- function(object, time, unit) {
      nbhd_list <- list()
      if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
      return(nbhd_list)
    }

    abfNbhd5_name <- "NBHD3"
    abfNbhd5_nbhd <- function(object, time, unit) {
      nbhd_list <- list()
      if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
      if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
      if(time>3) nbhd_list <- c(nbhd_list, list(c(unit, time-3)))
      return(nbhd_list)
    }

  if(abfNbhd_run_level==1){ #######
    abfNbhd_Np <- 10
    abfNbhd_I <- 10
    abfNbhd_U <- 2
    abfNbhd_N <- 5
    abfNbhd_replicates <- 2 
    abfNbhd1_Np <- abfNbhd_Np
    abfNbhd2_Np <- abfNbhd_Np
    abfNbhd3_Np <- abfNbhd_Np
    abfNbhd4_Np <- abfNbhd_Np
    abfNbhd5_Np <- abfNbhd_Np
    abfNbhd1_I <- abfNbhd_I
    abfNbhd2_I <- abfNbhd_I
    abfNbhd3_I <- abfNbhd_I
    abfNbhd4_I <- abfNbhd_I
    abfNbhd5_I <- abfNbhd_I    
  } else if(abfNbhd_run_level==2){ #######
    abfNbhd_Np <- 100
    abfNbhd_I <- 500
    abfNbhd_U <- 40
    abfNbhd_N <- 4*26
    abfNbhd_replicates <- 5
    abfNbhd1_Np <- abfNbhd_Np
    abfNbhd2_Np <- abfNbhd_Np
    abfNbhd3_Np <- abfNbhd_Np
    abfNbhd4_Np <- abfNbhd_Np
    abfNbhd5_Np <- abfNbhd_Np
    abfNbhd1_I <- abfNbhd_I
    abfNbhd2_I <- abfNbhd_I
    abfNbhd3_I <- abfNbhd_I
    abfNbhd4_I <- abfNbhd_I
    abfNbhd5_I <- abfNbhd_I    
  } else if(abfNbhd_run_level==3){
    abfNbhd_Np <- 200
    abfNbhd_I <- 1000
    abfNbhd_U <- 40
    abfNbhd_N <- 5*26
    abfNbhd_replicates <- 10
    abfNbhd1_Np <- abfNbhd_Np
    abfNbhd2_Np <- abfNbhd_Np
    abfNbhd3_Np <- abfNbhd_Np
    abfNbhd4_Np <- abfNbhd_Np
    abfNbhd5_Np <- abfNbhd_Np
    abfNbhd1_I <- abfNbhd_I
    abfNbhd2_I <- abfNbhd_I
    abfNbhd3_I <- abfNbhd_I
    abfNbhd4_I <- abfNbhd_I
    abfNbhd5_I <- abfNbhd_I    
  } else if(abfNbhd_run_level==4){
  } else if(abfNbhd_run_level==5){
  }

})

abfNbhd_jobs <- data.frame(reps=1:abfNbhd_replicates,U=abfNbhd_U,N=abfNbhd_N)

@

<<abfNbhd_spatPomp,eval=T,echo=F>>=
abfNbhd_spatPomp <- stew(file=paste0(abfNbhd_files_dir,"abfNbhd_spatPomp.rda"),{

  abfNbhd_model_dir <- "measlesModel/"
  # note: care is required if abfNbhd_model_dir is set to something other than the default measlesModel
  # however, it may be useful to do that while testing model variations.

  load(file=paste0(abfNbhd_model_dir,"measles_spatPomp.rda"))
  abfNbhd_sim <- measles_subset(m_U=abfNbhd_U, m_N=abfNbhd_N)

})
@

<<abfNbhd_nbhd,echo=F>>=
abfNbhd_nbhd <- function(object, time, unit) {
  nbhd_list <- list()
  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
  return(nbhd_list)
}
@

<<abfNbhd1,cache=F,echo=F>>=
abfNbhd1 <- stew(file=paste0(abfNbhd_files_dir,"abfNbhd1.rda"),seed=844424,{
  foreach(job=iter(abfNbhd_jobs,"row")) %do% {
    system.time(
      abf(abfNbhd_sim, 
        Nrep = abfNbhd1_I,
        Np=abfNbhd1_Np,
        nbhd = abfNbhd1_nbhd, tol=abfNbhd_tol) -> abfNbhd1_out 
    ) -> abfNbhd1_time
#    uncomment for debugging 
#    list(logLik=logLik(abfNbhd1_out),time=abfNbhd1_time["elapsed"],abfNbhd1_out)
    list(logLik=logLik(abfNbhd1_out),time=abfNbhd1_time["elapsed"])
  } -> abfNbhd1_list
  rm(abfNbhd1_out)
})
abfNbhd1_jobs <- abfNbhd_jobs
abfNbhd1_jobs$logLik <- vapply(abfNbhd1_list,function(x)x$logLik,numeric(1))
abfNbhd1_jobs$time <- vapply(abfNbhd1_list,function(x) x$time,numeric(1))
abfNbhd1_jobs$Method <- abfNbhd1_name
@

<<abfNbhd2,cache=F,echo=F>>=
abfNbhd2 <- stew(file=paste0(abfNbhd_files_dir,"abfNbhd2.rda"),seed=389855,{
  foreach(job=iter(abfNbhd_jobs,"row")) %do% {
    system.time(
      abf(abfNbhd_sim, 
        Nrep = abfNbhd2_I,
        Np=abfNbhd2_Np,
        nbhd = abfNbhd2_nbhd, tol=abfNbhd_tol) -> abfNbhd2_out 
    ) -> abfNbhd2_time
#    uncomment for debugging 
#    list(logLik=logLik(abfNbhd2_out),time=abfNbhd2_time["elapsed"],abfNbhd2_out)
    list(logLik=logLik(abfNbhd2_out),time=abfNbhd2_time["elapsed"])
  } -> abfNbhd2_list
  rm(abfNbhd2_out)
})
abfNbhd2_jobs <- abfNbhd_jobs
abfNbhd2_jobs$logLik <- vapply(abfNbhd2_list,function(x)x$logLik,numeric(1))
abfNbhd2_jobs$time <- vapply(abfNbhd2_list,function(x) x$time,numeric(1))
abfNbhd2_jobs$Method <-  abfNbhd2_name
@

<<abfNbhd3,cache=F,echo=F>>=
abfNbhd3 <- stew(file=paste0(abfNbhd_files_dir,"abfNbhd3.rda"),seed=186711,{
  foreach(job=iter(abfNbhd_jobs,"row")) %do% {
    system.time(
      abf(abfNbhd_sim,
        Nrep = abfNbhd3_I,
        Np=abfNbhd3_Np,
        nbhd = abfNbhd3_nbhd, tol=abfNbhd_tol) -> abfNbhd3_out
    ) -> abfNbhd3_time
#    uncomment for debugging
#    list(logLik=logLik(abfNbhd3_out),time=abfNbhd3_time["elapsed"],abfNbhd3_out)
    list(logLik=logLik(abfNbhd3_out),time=abfNbhd3_time["elapsed"])
  } -> abfNbhd3_list
  rm(abfNbhd3_out)
})
abfNbhd3_jobs <- abfNbhd_jobs
abfNbhd3_jobs$logLik <- vapply(abfNbhd3_list,function(x)x$logLik,numeric(1))
abfNbhd3_jobs$time <- vapply(abfNbhd3_list,function(x) x$time,numeric(1))
abfNbhd3_jobs$Method <-  abfNbhd3_name
@

<<abfNbhd4,cache=F,echo=F>>=
abfNbhd4 <- stew(file=paste0(abfNbhd_files_dir,"abfNbhd4.rda"),seed=186711,{
  foreach(job=iter(abfNbhd_jobs,"row")) %do% {
    system.time(
      abf(abfNbhd_sim,
        Nrep = abfNbhd4_I,
        Np=abfNbhd4_Np,
        nbhd = abfNbhd4_nbhd, tol=abfNbhd_tol) -> abfNbhd4_out
    ) -> abfNbhd4_time
#    uncomment for debugging
#    list(logLik=logLik(abfNbhd4_out),time=abfNbhd4_time["elapsed"],abfNbhd4_out)
    list(logLik=logLik(abfNbhd4_out),time=abfNbhd4_time["elapsed"])
  } -> abfNbhd4_list
  rm(abfNbhd4_out)
})
abfNbhd4_jobs <- abfNbhd_jobs
abfNbhd4_jobs$logLik <- vapply(abfNbhd4_list,function(x)x$logLik,numeric(1))
abfNbhd4_jobs$time <- vapply(abfNbhd4_list,function(x) x$time,numeric(1))
abfNbhd4_jobs$Method <-  abfNbhd4_name
@

<<abfNbhd5,cache=F,echo=F,eval=T>>=
abfNbhd5 <- stew(file=paste0(abfNbhd_files_dir,"abfNbhd5.rda"),seed=186711,{
  foreach(job=iter(abfNbhd_jobs,"row")) %do% {
    system.time(
      abf(abfNbhd_sim,
        Nrep = abfNbhd5_I,
        Np=abfNbhd5_Np,
        nbhd = abfNbhd5_nbhd, tol=abfNbhd_tol) -> abfNbhd5_out
    ) -> abfNbhd5_time
#    uncomment for debugging
#    list(logLik=logLik(abfNbhd5_out),time=abfNbhd5_time["elapsed"],abfNbhd5_out)
    list(logLik=logLik(abfNbhd5_out),time=abfNbhd5_time["elapsed"])
  } -> abfNbhd5_list
  rm(abfNbhd5_out)
})
abfNbhd5_jobs <- abfNbhd_jobs
abfNbhd5_jobs$logLik <- vapply(abfNbhd5_list,function(x)x$logLik,numeric(1))
abfNbhd5_jobs$time <- vapply(abfNbhd5_list,function(x) x$time,numeric(1))
abfNbhd5_jobs$Method <-  abfNbhd5_name
@

<<abfNbhd_loglik_plot, echo=F, fig.align='center', fig.height=4, fig.width=4, out.width="3in", fig.cap = paste('log likelihood estimates for simulated data from the measles model using {\\ABF}, with varying neighborhoods.')>>=

abfNbhd_results <- rbind(abfNbhd1_jobs,abfNbhd2_jobs,abfNbhd3_jobs,abfNbhd4_jobs,abfNbhd5_jobs)
#abfNbhd_results <- rbind(abfNbhd1_jobs,abfNbhd2_jobs,abfNbhd3_jobs,abfNbhd4_jobs)
#abfNbhd_results <- rbind(abfNbhd1_jobs,abfNbhd2_jobs,abfNbhd3_jobs)
abfNbhd_results$logLik_per_unit <- abfNbhd_results$logLik/abfNbhd_results$U
abfNbhd_results$logLik_per_obs <- abfNbhd_results$logLik_per_unit/abfNbhd_N

save(file=paste0(abfNbhd_files_dir,"abfNbhd_results.rda"),abfNbhd_results)

abfNbhd_max <- max(abfNbhd_results$logLik_per_obs)

#ggplot(abfNbhd_results,mapping = aes(x = Unit, y = logLik_per_obs, group=Method,color=Method)) +
#  geom_point() +
#  coord_cartesian(ylim=c(abfNbhd_max-2,abfNbhd_max))+
#  ylab("log likelihood per observation")
# qplot(Method,logLik,data=abfNbhd_results,geom="boxplot")
qplot(Method,logLik_per_obs,data=abfNbhd_results,
  ylab="log likelihood per unit per time",
  xlab="Neighborhood")
@

We compared five different neighborhoods for the measles model:
\begin{center}
\begin{tabular}{lll}
\hline
NBHD1 & One co-located lag & $\{(\unit,\time-1)\}$ 
\\
NBHD2 & Two co-located lags & $\{(\unit,\time-1),(\unit,\time-2)\}$ 
\\
NBHD3 & Three co-located lags & $\{(\unit,\time-1),(\unit,\time-2),(\unit,\time-3)\}$ 
\\
NBHD4 & Two co-located lags and the previous city& $\{(\unit,\time-1),(\unit,\time-2),(\unit-1,\time)\}$ 
\\
NBHD5 & Two co-located lags and London& $\{(\unit,\time-1),(\unit,\time-2),(1,\time)\}$ 
\\
\hline
\end{tabular}
\end{center}
We filtered simulated data for $U=\Sexpr{abfNbhd_U}$ and $N=\Sexpr{abfNbhd_N}$, with $\Sexpr{abfNbhd_replicates}$ replications.
We used {\ABF} with $\Sexpr{abfNbhd1_Np}$ particles on each of $\Sexpr{abfNbhd1_I}$ bootstrap replications.
The results are shown in Fig.~\ref{fig:abfNbhd_loglik_plot}.
Larger neighborhoods should increase the expected likelihood, but their increased Monte Carlo variability can decrease the expected log likelihood due to Jensen's inequality.
In this case, we see that a neighborhood of two co-located lags provides a reasonable bias-variance tradeoff.
The time taken for the above calculation was insensitive to the size of the neighborhood.
The total run time for each neighborhood in Fig.~\ref{fig:abfNbhd_loglik_plot} was 
\Sexpr{myround(sum(abfNbhd1_jobs$time)/60,1)} mins for NBHD1,
\Sexpr{myround(sum(abfNbhd2_jobs$time)/60,1)} mins for NBHD2,
\Sexpr{myround(sum(abfNbhd3_jobs$time)/60,1)} mins for NBHD3,
\Sexpr{myround(sum(abfNbhd4_jobs$time)/60,1)} mins for NBHD4,
\Sexpr{myround(sum(abfNbhd5_jobs$time)/60,1)} mins for NBHD5.

\clearpage

%%%%% lzlzlzlzlzlz

\clearpage

\section{\secTitleSpace A Lorenz-96 example}

<<lz_settings,cache=FALSE,echo=F>>=

lz_files_dir <- paste0("lz_",lz_run_level,"/")
if(!dir.exists(lz_files_dir)) dir.create(lz_files_dir)

# lz_files_dir <- "lz_5_nbhd2/"

lz_params <- c(F=8, sigma=1, tau=1)

stew(file=paste0(lz_files_dir,"lz_settings.rda"),{

  # copy variables that should be included in the stew
  lz_run_level <- lz_run_level 
  lz_cores <- cores

  lz_tol <- 1e-300

  if(lz_run_level==1){
    lz_U <- c(8, 4)
    lz_N <- 2
    lz_replicates <- 2 # number of Monte Carlo replicates
    lz_girf_Np <- 20
    lz_girf_lookahead <- 2
    lz_girf_nguide <- 10
    lz_pfilter_Np <- 100
    lz_abf_Nrep <- 3
    lz_abf_Np_per_replicate <- 10
    lz_ubf_Nrep <- 20
    lz_abfir_Nrep <- lz_abf_Nrep
    lz_abfir_Np_per_replicate <- lz_abf_Np_per_replicate
    lz_enkf_Np <- 20
    lz_bpf_Np <- 20
    lz_bpf_units_per_block <- 2
  } else if(lz_run_level==2){ 
    lz_U <- c(32,16,8,4)
    lz_N <- 20
    lz_replicates <- 4 # number of Monte Carlo replicates
    lz_girf_Np <- 1000
    lz_girf_lookahead <- 2
    lz_girf_nguide <- 50
    lz_pfilter_Np <- 50000
    lz_abf_Nrep <- 500
    lz_abf_Np_per_replicate <- 100
    lz_ubf_Nrep <- 10000
    lz_abfir_Nrep <- 50
    lz_abfir_Np_per_replicate <- 50
    lz_enkf_Np <- 10000
    lz_bpf_Np <- 10000
    lz_bpf_units_per_block <- 2
  } else if(lz_run_level==3){
    lz_replicates <- 5 # number of Monte Carlo replicates
    lz_U <- c(40,30,20,10,5)
    lz_N <- 10
    lz_girf_Np <- 1000
    lz_girf_lookahead <- 2
    lz_girf_nguide <- 50
    lz_pfilter_Np <- 50000
    lz_abf_Nrep <- 400
    lz_abf_Np_per_replicate <- 400
    lz_ubf_Nrep <- 5000
    lz_abfir_Nrep <- 200
    lz_abfir_Np_per_replicate <- 200
    lz_enkf_Np <- 2000
    lz_bpf_Np <- 10000
    lz_bpf_units_per_block <- 3
  } else if(lz_run_level==4){
    lz_replicates <- 5 # number of Monte Carlo replicates
    lz_U <- c(60, 50, 40, 30, 20, 10, 5)
    lz_N <- 50
    lz_girf_Np <- 2000
    lz_girf_lookahead <- 1
    lz_girf_nguide <- 50
    lz_pfilter_Np <- 100000
    lz_abf_Nrep <- 400
    lz_abf_Np_per_replicate <- 400
    lz_ubf_Nrep <- 40000
    lz_abfir_Nrep <- 200
    lz_abfir_Np_per_replicate <- 200
    lz_enkf_Np <- 5000
    lz_bpf_Np <- 20000
    lz_bpf_units_per_block <- 4
  } else if(lz_run_level==5){
    lz_replicates <- 5 # number of Monte Carlo replicates
    lz_U <- c(50, 40, 30, 20, 16, 12, 10, 8, 6, 4)
    lz_N <- 50
    lz_girf_Np <- 1000
    lz_girf_lookahead <- 2
    lz_girf_nguide <- 50
    lz_pfilter_Np <- 100000
    lz_abf_Nrep <- 400
    lz_abf_Np_per_replicate <- 400
    lz_ubf_Nrep <- 40000
    lz_abfir_Nrep <- 200
    lz_abfir_Np_per_replicate <- 200
    lz_enkf_Np <- 10000
    lz_bpf_Np <- 10000
    lz_bpf_units_per_block <- 4
  }

  lz_dt <- 0.005
  lz_dt_obs <- 0.5
  lz_ivp_mean <- 5
  lz_ivp_sd <- 2

  set.seed(4512)
  lz_list <- foreach(u=lz_U) %do% {
    lz1 <- lorenz(
      U=u, N=lz_N,
      dt=lz_dt,dt_obs=lz_dt_obs,
      regular_params=lz_params)
    lz_ivps <- rnorm(u,mean=lz_ivp_mean,sd=lz_ivp_sd)
    names(lz_ivps) <- paste0("X",1:u,"_0")
    lz1_params <- coef(lz1)
    lz1_params[names(lz_ivps)] <- lz_ivps
    simulate(lz1,params=lz1_params)
  }
})

lz_jobs <- expand.grid(U=lz_U,reps=1:lz_replicates)
lz_jobs$U_id <- rep(seq_along(lz_U),times=lz_replicates)

@

Our primary motivation for {\ABF} and {\ABFIR} is application to population dynamics arising in ecological and epidemiological models.
Geophysical models provide an alternative situation involving spatiotemporal data analysis.
We compare methods on the Lorenz-96 model, a nonlinear chaotic system providing a toy model for global atmospheric circulation \citep{lorenz96,vankekem18}.
We consider a stochastic Lorenz-96 model with added Gaussian process noise \citep{park20} defined
as the solution to the following system of stochastic differential equations,
\begin{equation}
  dX_{\unit}(t) =
\big \{
\big(X_{\unit+1}(t) - X_{\unit-2}(t)\big)
\cdot X_{\unit-1}(t) - X_{\unit}(t) + F
\big\} dt
+ \sigma_p dB_{\unit}(t), \qquad \unit \in \seq{1}{\Unit}.
  \label{eqn:stoLorenz}\end{equation}
We define $X_{0} = X_{\Unit}$, $X_{-1} = X_{\Unit-1}$, and $X_{\Unit+1} = X_{1}$ so that the $\Unit$ spatial locations are placed on a circle.
The terms $\{B_{\unit}(t), \unit \in \seq{1}{\Unit} \}$ denote $\Unit$ independent standard Brownian motions.
$F$ is a forcing constant, and we use the value $F{\,=\,}8$ which was demonstrated by \citet{lorenz96} to induce chaotic behavior.
The process noise parameter is set to $\sigma_p=\Sexpr{coef(lz_list[[1]])["sigma"]}$.
The system is started with initial state $X_{\unit}(0)$ drawn as an independent normal random variable with mean $\Sexpr{lz_ivp_mean}$ and standard deviation $\Sexpr{lz_ivp_sd}$ for $\unit\in \seq{1}{\Unit}$.
This initialization leads to short transient behavior.
Observations are independently made for each dimension at 
%%%%%%%%%%
%%%%%%%%%% $t_\time = \Sexpr{lz_dt_obs} \time$ 
%%%%%%%%%%
$t_\time = \time$ 
for $\time \in \seq{1}{\Time}$ with Gaussian measurement noise of mean zero and standard deviation $\tau=\Sexpr{coef(lz_list[[1]])["tau"]}$,
\begin{equation}
Y_{\unit, \time} = X_{\unit}(t_{\time}) + \eta_{\unit, \time}\, \hspace{20mm} \eta_{\unit, \time} \sim N(0, \tau^2).
\end{equation}
We used an Euler-Maruyama method for numerical approximation of the sample paths of $\{\myvec{X}(t)\}$, with timestep of $\Sexpr{lz_dt}$. 
A simulation from this model is shown in Fig.~\ref{fig:lz_image_plot}.

The ensemble Kalman filter (EnKF) is a widely used filtering method in weather forecasting for high dimensional systems \citep{evensen96}.  
EnKF involves a local Gaussian approximation which is problematic in highly nonlinear systems \citep{ades15}.
Methods that make local Gaussian assumptions like EnKF are necessary to scale up to the dimensions of the problems in weather forecasting. 
Figure~\ref{fig:lz_loglik_plot} shows that for a small number of units, the basic particle filter (PF) and GIRF out-perform EnKF.
Then, as the number of spatial units increases, the performance of PF rapidly deteriorates whereas GIRF continues to perform well up to a moderate number of units.
{\UBF}, {\ABF}, and particularly {\ABFIR}, scales well despite under-performing EnKF on this example.
The additive Gaussian observation and process noise in the Lorenz-96 model is well suited to the approximations involved in EnKF.
By contrast, it is less clear how to apply EnKF to discrete population non-Gaussian models such as the measles example, and how effective the resulting approximations might be.

<<lz_nbhd,echo=F>>=
lz_nbhd <- function(object, time, unit) {
  nunits <- length(unit_names(object))
  nbhd_list <- list()
  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
  if(unit>1) nbhd_list <- c(nbhd_list, list(c(unit-1, time)))
  if(unit>2) nbhd_list <- c(nbhd_list, list(c(unit-2, time)))
  return(nbhd_list)
}

lz_nbhd2 <- function(object, time, unit) {
  nunits <- length(unit_names(object))
  nbhd_list <- list()
  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
  return(nbhd_list)
}

@

<<lz_girf,cache=F,echo=F>>=
lz_girf <- stew(file=paste0(lz_files_dir,"lz_girf.rda"),seed=5981724,{
  foreach(job=iter(lz_jobs,"row")) %dopar% {
    system.time(
      girf(lz_list[[job$U_id]], method='adams',
        Np=lz_girf_Np,
        Ninter = job$U,
        lookahead = lz_girf_lookahead,
        Nguide = lz_girf_nguide,
        tol = lz_tol) -> lz_girf_out 
    ) -> lz_girf_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_girf_out),time=lz_girf_time["elapsed"],lz_girf_out)
    list(logLik=logLik(lz_girf_out),time=lz_girf_time["elapsed"])
  } -> lz_girf_list
})
lz_jobs$girf_logLik <- vapply(lz_girf_list,function(x)x$logLik,numeric(1))
lz_jobs$girf_time <- vapply(lz_girf_list,function(x) x$time,numeric(1))
@

<<lz_abf,cache=F,echo=F>>=
lz_abf <- stew(file=paste0(lz_files_dir,"lz_abf.rda"),seed=844424,{
  foreach(job=iter(lz_jobs,"row")) %do% {
    system.time(
      abf(lz_list[[job$U_id]], 
        Nrep = lz_abf_Nrep,
        Np=lz_abf_Np_per_replicate,
        nbhd = lz_nbhd, tol=lz_tol) -> lz_abf_out 
    ) -> lz_abf_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_abf_out),time=lz_abf_time["elapsed"],lz_abf_out)
    list(logLik=logLik(lz_abf_out),time=lz_abf_time["elapsed"])
  } -> lz_abf_list
})
lz_jobs$abf_logLik <- vapply(lz_abf_list,function(x)x$logLik,numeric(1))
lz_jobs$abf_time <- vapply(lz_abf_list,function(x) x$time,numeric(1))

@

<<lz_ubf,cache=F,echo=F>>=
lz_ubf <- stew(file=paste0(lz_files_dir,"lz_ubf.rda"),seed=844424,{
  foreach(job=iter(lz_jobs,"row")) %do% {
    system.time(
      abf(lz_list[[job$U_id]], 
        Nrep = lz_ubf_Nrep,
        Np=1,
        nbhd = lz_nbhd, tol=lz_tol) -> lz_ubf_out 
    ) -> lz_ubf_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_ubf_out),time=lz_ubf_time["elapsed"],lz_ubf_out)
    list(logLik=logLik(lz_ubf_out),time=lz_ubf_time["elapsed"])
  } -> lz_ubf_list
})
lz_jobs$ubf_logLik <- vapply(lz_ubf_list,function(x)x$logLik,numeric(1))
lz_jobs$ubf_time <- vapply(lz_ubf_list,function(x) x$time,numeric(1))

@

<<lz_ubf2,cache=F,echo=F,eval=F>>=
lz_ubf2 <- stew(file=paste0(lz_files_dir,"lz_ubf2.rda"),seed=844424,{
  foreach(job=iter(lz_jobs,"row")) %do% {
    system.time(
      abf(lz_list[[job$U_id]],
        Nrep = lz_ubf_Nrep,
        Np=1,
        nbhd = lz_nbhd2, tol=lz_tol) -> lz_ubf2_out
    ) -> lz_ubf2_time
#    uncomment for debugging
#    list(logLik=logLik(lz_ubf2_out),time=lz_ubf2_time["elapsed"],lz_ubf2_out)
    list(logLik=logLik(lz_ubf2_out),time=lz_ubf2_time["elapsed"])
  } -> lz_ubf2_list
})
lz_jobs$ubf2_logLik <- vapply(lz_ubf2_list,function(x)x$logLik,numeric(1))
lz_jobs$ubf2_time <- vapply(lz_ubf2_list,function(x) x$time,numeric(1))

@

<<lz_abfir,cache=F,echo=F,eval=T>>=
lz_abfir <- stew(file=paste0(lz_files_dir,"lz_abfir.rda"),seed=53398,{
  foreach(job=iter(lz_jobs,"row")) %do% {
    system.time(
      abfir(lz_list[[job$U_id]], method='adams',
        Nrep = as.integer(lz_abfir_Nrep),
        Np=lz_abfir_Np_per_replicate,
        Ninter = as.integer(job$U/2),
        nbhd = lz_nbhd, tol=lz_tol) -> lz_abfir_out 
    ) -> lz_abfir_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_abfir_out),time=lz_abfir_time["elapsed"],lz_abfir_out)
    list(logLik=logLik(lz_abfir_out),time=lz_abfir_time["elapsed"])
  } -> lz_abfir_list
})
lz_jobs$abfir_logLik <- vapply(lz_abfir_list,function(x)x$logLik,numeric(1))
lz_jobs$abfir_time <- vapply(lz_abfir_list,function(x) x$time,numeric(1))
@

<<lz_enkf,cache=F,echo=F>>=

lz_h <- function(state.vec, param.vec){
  ix<-grep('X',names(state.vec))
  state.vec[ix]
}

lz_enkf <- stew(file=paste0(lz_files_dir,"lz_enkf.rda"),seed=837509,{
  foreach(job=iter(lz_jobs,"row")) %dopar% {
    system.time(
      enkf(lz_list[[job$U_id]], 
        Np=lz_enkf_Np,
        h = lz_h,
        R = diag((coef(lz_list[[job$U_id]])["tau"])^2, nrow = job$U)  
      ) -> lz_enkf_out 
    ) -> lz_enkf_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_enkf_out),time=lz_enkf_time["elapsed"],lz_enkf_out)
    list(logLik=logLik(lz_enkf_out),time=lz_enkf_time["elapsed"])
  } -> lz_enkf_list
})
lz_jobs$enkf_logLik <- vapply(lz_enkf_list,function(x)x$logLik,numeric(1))
lz_jobs$enkf_time <- vapply(lz_enkf_list,function(x) x$time,numeric(1))

@

<<lz_pfilter,cache=F,echo=F>>=
lz_pfilter <- stew(file=paste0(lz_files_dir,"lz_pfilter.rda"),seed=53285,{
  foreach(job=iter(lz_jobs,"row")) %dopar% {
    system.time(
      pfilter(lz_list[[job$U_id]],Np=lz_pfilter_Np) -> lz_pfilter_out
    ) -> lz_pfilter_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_pfilter_out),time=lz_pfilter_time["elapsed"],lz_pfilter_out)
    list(logLik=logLik(lz_pfilter_out),time=lz_pfilter_time["elapsed"])
  } -> lz_pfilter_list
})
lz_jobs$pfilter_logLik <- vapply(lz_pfilter_list,function(x)x$logLik,numeric(1))
lz_jobs$pfilter_time <- vapply(lz_pfilter_list,function(x) x$time,numeric(1))
@

<<lz_bpf,cache=F,echo=F>>=
lz_bpf <- stew(file=paste0(lz_files_dir,"lz_bpf.rda"),seed=53285,{
  foreach(job=iter(lz_jobs,"row")) %dopar% {
    system.time(
      bpfilter(lz_list[[job$U_id]],
        Np=lz_bpf_Np,
	block_size=lz_bpf_units_per_block) -> lz_bpf_out
    ) -> lz_bpf_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_bpf_out),time=lz_bpf_time["elapsed"],lz_bpf_out)
    list(logLik=lz_bpf_out@loglik,time=lz_bpf_time["elapsed"])
  } -> lz_bpf_list
})
lz_jobs$bpf_logLik <- vapply(lz_bpf_list,function(x)x$logLik,numeric(1))
lz_jobs$bpf_time <- vapply(lz_bpf_list,function(x) x$time,numeric(1))
@

<<lz_loglik_plot, echo=F, fig.align='center', fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('Log likelihood estimates for a Lorenz-96 model of various dimensions. {\\UBF}, {\\ABF} and {\\ABFIR} are compared with a guided intermediate resampling filter (GIRF), a standard particle filter (PF), a block particle filter (BPF) and an ensemble Kalman filter (EnKF).')>>=

# put output in tall format for plotting
#lz_methods <- c("ABF", "ABF-IR","ENKF","GIRF","PF")
lz_methods <- c("ABF", "ABF-IR","UBF","EnKF","GIRF","PF","BPF")
#lz_methods <- c("ABF","UBF","ENKF","GIRF","PF")
lz_results <- data.frame(
  Method=rep(lz_methods,each=nrow(lz_jobs)),
  logLik=c(
    lz_jobs$abf_logLik,
    lz_jobs$abfir_logLik,
    lz_jobs$ubf_logLik,
    lz_jobs$enkf_logLik,
    lz_jobs$girf_logLik,
    lz_jobs$pfilter_logLik,
    lz_jobs$bpf_logLik
  ),
  Units=rep(lz_jobs$U,reps=length(lz_methods))
)
lz_point_perturbation <- 1
lz_results$U <- lz_results$Units+
  rep( lz_point_perturbation*seq(from=-1,to=1,length=length(lz_methods)),
    each=nrow(lz_jobs))
lz_results$logLik_per_unit <- lz_results$logLik/lz_results$Units
lz_results$logLik_per_obs <- lz_results$logLik_per_unit/lz_N

save(file=paste0(lz_files_dir,"lz_results.rda"),lz_results,lz_jobs)

lz_max <- max(lz_results$logLik_per_obs,na.rm=T)

ggplot(lz_results,mapping = aes(x = U, y = logLik_per_obs, group=Method,color=Method,shape=Method,linetype=Method)) +
  scale_linetype_manual(values=c(1,1,1,1,2,2,2)) +
  scale_shape_manual(values=c(1,2,4,5,1,2,4)) +
  theme(legend.key.width = unit(1,"cm"))+
  geom_point() +
  stat_summary(fun=mean, geom="line") +
  coord_cartesian(ylim=c(lz_max-2,lz_max))+
  ylab("log likelihood per observation")

@

<<lz_load, echo=F>>=

# lz_files_list <- list.files(path=lz_files_dir,full.names=TRUE)
# for(filename in lz_files_list) load(file=filename)

lz_girf_times <- sapply(split(lz_jobs$girf_time,lz_jobs$U),mean)
lz_abf_times <- sapply(split(lz_jobs$abf_time,lz_jobs$U),mean)
lz_ubf_times <- sapply(split(lz_jobs$ubf_time,lz_jobs$U),mean)

lz_abfir_times <- sapply(split(lz_jobs$abfir_time,lz_jobs$U),mean)

lz_pfilter_times <- sapply(split(lz_jobs$pfilter_time,lz_jobs$U),mean)
lz_enkf_times <- sapply(split(lz_jobs$enkf_time,lz_jobs$U),mean)
lz_bpf_times <- sapply(split(lz_jobs$bpf_time,lz_jobs$U),mean)

lz_K <- lz_bpf_units_per_block
if(lz_K==1) lz_blocks_text <- "{\\Unit}" else lz_blocks_text <- paste0("{\\Unit/",lz_bpf_units_per_block,"}")

@

<<lz_image_plot, echo=F, fig.height=3.5, fig.width=7, out.width="6.5in", fig.cap = paste('Lorenz \'96 simulation used in the main text')>>=
library(fields)
lz_obj <- lz_list[[which.max(lz_U)]]
par(mai=c(0.7,0.7,0.1,0.5))
image.plot(
  y=1:dim(states(lz_obj))[1],
  x=time(lz_obj),z=t(states(lz_obj)),
  ylab="unit",xlab="time",col=gray.colors(33)
)
@

\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline &
  {\UBF} &
  {\ABF} &
  {\ABFIR} &
  GIRF &
  PF &
  EnKF &
  BPF
\\ 
\hline 
\rule{0mm}{4.5mm}particles, $\Np$ &
  1 &
  \Sexpr{lz_abf_Np_per_replicate} &
  \Sexpr{lz_abfir_Np_per_replicate} &
  \Sexpr{lz_girf_Np}  &
  \Sexpr{lz_pfilter_Np} &
  \Sexpr{lz_enkf_Np} &
  \Sexpr{lz_bpf_Np} 
\\
bootstrap replicates, $\Rep$ &
  \Sexpr{lz_ubf_Nrep} &
  \Sexpr{lz_abf_Nrep} &
  \Sexpr{lz_abfir_Nrep} &
  --- &
  --- &
  --- &
  ---
\\
guide simulations, $\Nguide$ &
  --- &
  --- &
  --- &
  \Sexpr{lz_girf_nguide} &
  --- &
  --- &
  ---
\\
lookahead lag, $L$ &
  --- &
  --- &
  --- &
  \Sexpr{lz_girf_lookahead} &
  --- &
  --- &
  ---
\\ 
intermediate steps, $S$ &
  --- &
  --- &
  $\Unit/2$ &
  $\Unit$ &
  --- &
  --- &
  ---
\\
\cline{2-4}

\hspace{-2mm}\begin{tabular}{l}
neighborhood, $B_{\unit\comma\time}$ \\
or block size
\end{tabular}
&
  \multicolumn{3}{c|}{
    $ \rule[-4mm]{0mm}{11mm}
    \begin{array}{c} 
    $\big\{(\unit,\time-1),(\unit,\time-2),$ \\
    $(\unit-1,\time),(\unit-2,\time)\big\}$
    \end{array}$
  } &
  --- &
  --- &
  --- &
  \Sexpr{lz_bpf_units_per_block}
\\
\cline{2-5}
forecast mean, $\myvec{\mu}(\myvec{x},s,t)$ &
  --- &
  --- &
  \multicolumn{2}{c|}{ODE model} &
  --- &
  --- &
  ---
\\
measurement mean, $h_{\unit\comma\time}(x)$ &
  --- &
  --- &
  \multicolumn{2}{c|}{$x$} &
  --- &
  $x$ &
  ---
\\
\cline{4-5}
$\tau={\VtoTheta}_{\unit\comma\time}(V,x)$ &
  --- &
  --- &
  \multicolumn{2}{c|}{\rule{0mm}{5mm}$\sqrt{V}$}  &
  --- &
  --- &
  ---
\\
$V={\thetaToV}_{\unit\comma\time}(\tau,x)$ &
  --- &
  --- &
  \multicolumn{2}{c|}{$\tau^2$} &
  --- &
  $\tau^2$ &
  ---
\\
%\cline{4-5}
%Units in each blocks&
%  --- &
%  --- &
%  --- &
%  --- &
%  --- &
%  --- &
%  $\Sexpr{lz_bpf_units_per_block}$
%\\
\hline
\rule{0mm}{4.5mm}effort (core mins, $U=\Sexpr{names(lz_abf_times)[1]}$) & 
  \Sexpr{myround(lz_ubf_times[1]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_abf_times[1]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_abfir_times[1]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_girf_times[1]/60,1)}  & 
  \Sexpr{myround(lz_pfilter_times[1]/60,1)} & 
  \Sexpr{myround(lz_enkf_times[1]/60,1)} &
  \Sexpr{myround(lz_bpf_times[1]/60,1)}
\\
effort  (core mins,  $U=\Sexpr{names(lz_abf_times)[2]}$) & 
  \Sexpr{myround(lz_ubf_times[2]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_abf_times[2]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_abfir_times[2]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_girf_times[2]/60,1)} &
  \Sexpr{myround(lz_pfilter_times[2]/60,1)} & 
  \Sexpr{myround(lz_enkf_times[2]/60,1)} &
  \Sexpr{myround(lz_bpf_times[2]/60,1)}  
\\
effort  (core mins,  $U=\Sexpr{names(lz_abf_times)[4]}$) & 
  \Sexpr{myround(lz_ubf_times[4]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_abf_times[4]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_abfir_times[4]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_girf_times[4]/60,1)} & 
  \Sexpr{myround(lz_pfilter_times[4]/60,1)} & 
  \Sexpr{myround(lz_enkf_times[4]/60,1)} &
  \Sexpr{myround(lz_bpf_times[4]/60,1)}
\\
effort  (core mins,  $U=\Sexpr{names(lz_abf_times)[6]}$) & 
  \Sexpr{myround(lz_ubf_times[6]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_abf_times[6]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_abfir_times[6]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_girf_times[6]/60,1)} & 
  \Sexpr{myround(lz_pfilter_times[6]/60,1)} & 
  \Sexpr{myround(lz_enkf_times[6]/60,1)} &
  \Sexpr{myround(lz_bpf_times[6]/60,1)}
\\
effort  (core mins,  $U=\Sexpr{names(lz_abf_times)[10]}$) & 
  \Sexpr{myround(lz_ubf_times[10]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_abf_times[10]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_abfir_times[10]*lz_cores/60,1)} & 
  \Sexpr{myround(lz_girf_times[10]/60,1)} & 
  \Sexpr{myround(lz_pfilter_times[10]/60,1)} & 
  \Sexpr{myround(lz_enkf_times[10]/60,1)} &
  \Sexpr{myround(lz_bpf_times[10]/60,1)}  
\\
\hline

\end{tabular}
\caption{Algorithmic settings for the Lorenz-96 numerical example. 
Computational effort is measured in core minutes for running one filter, corresponding to a point on Figure~ Figure~\ref{fig:lz_loglik_plot}. 
The time taken for computing a single point using the parallel {\UBF}, {\ABF} and {\ABFIR} implementations is the effort divided by the number of cores, here $\Sexpr{lz_cores}$.
The time taken for computing a single point using the single-core GIRF, PF, EnKF and BPF implementations is equal to the effort in core minutes.
}\label{tab:lz}
\end{center}
\end{table}

%\FloatBarrier

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% mmmmmmmmmmmmmmmmmmmm %%%%%%%%%%%%%%

\section{\secTitleSpace A memory-efficient representation of {\ABF}}

The {\ABF} pseudocode in the main text emphasizes the logical structure of the mathematical quantities computed, rather than describing a specific implementation. 
Arguably, an algorithm should specify not only what is to be computed, but also details of what variables should be created and saved to carry out these computations efficiently, and how computations and storage are shared across multiple locations when the algorithm is parallelized.
We use the term algorithm to denote a higher level description of the quantities to be calculated and we will say that alternative pseudocodes arriving at the same quantities are representations of the algorithm. 
Here, we present an alternative representations of {\ABF} which we call {\ABF}$_2$, and we refer to the representation in the main text as {\ABF}$_1$.
The most concrete form of an algorithm is the actual computer code implementing the algorithm in a programming language.
The implementation of {\ABF} in \texttt{spatPomp}, used for numerical results in this paper, uses an embarrassingly parallel approach based on the representation {\ABF}$_1$. 
This strategy facilitates robust and simple parallelization, and is appropriate when memory constraints are not limiting. 
To develop {\ABF}$_2$, we set up notation similar to that used for the mathematical theory.
Let 
\begin{equation}
\gamma^{}_{\unit,\time,\rep,k}=\frac{1}{\Np}\sum_{\np=1}^{\Np}\hspace{2mm}\prod_{\altUnit:(\altUnit,\time-k)\in B_{\unit,\time}} w^M_{\altUnit,\time-k,\rep,\np}
\end{equation}
Also, let 
\begin{equation}
\gamma^{+}_{\unit,\time,\rep,0}=\frac{1}{\Np}\sum_{\np=1}^{\Np}\prod_{(\altUnit,\time)\in B^{+}_{\unit,\time}} w^M_{\altUnit,\time,\rep,\np}
\end{equation}
Now set $K$ to be the largest value of $k$ for which $B^{[n-k]}_{\unit,\time}$ is nontrivial for some $(\unit,\time)$, i.e., $K$ is the largest temporal lag in any neighborhood. 
With this notation, we can write \myeqref{eq:gamma:def} as
\begin{eqnarray*}
\gamma^{MC,\rep}_{B_{\unit,\time}} &=& \prod_{k=0}^{K} \gamma^{}_{\unit,\time,\rep,k}
\\
\gamma^{MC,\rep}_{B^{+}_{\unit,\time}} &=& \gamma^{+}_{\unit,\time,\rep,0} \, \prod_{k=1}^{K} \gamma^{}_{\unit,\time,\rep,k}
\end{eqnarray*}
This motivates the following representation of {\ABF}.

\begin{center}
\noindent\begin{tabular}{l}
\hline
{\bf 
{{\ABF}$_2$}. Adapted bagged filter, representation $2$.}\inputSpace\\
\hline
%{\bf input:} From Table~1 \inputSpace \\
%\hline
\firstLineSpace
Initialize adapted simulation: $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\myvec{x}_0)$
\\
For $\time\ \mathrm{in}\ \seq{1}{\Time}$
\\
\asp  Proposals:
    $\myvec{X}_{\time,\rep,\np}^{\IP} \sim 
    f_{\myvec{X}_{\time}|X_{1:\Unit,\time-1}} 
    \big( \myvec{x}_{\time}\given \myvec{X}^{\IF}_{\time-1,\rep}\big)$
\\
\asp Measurement weights:
  $w^M_{\unit,\time,\rep,\np} = 
    f_{Y_{\unit,\time}|X_{\unit,\time}} 
    \big (\data{y}_{\unit,\time}\given X^{\IP}_{\unit,\time,\rep,\np}\big)$
\\
\asp  Adapted resampling weights:
  $w^{\IF}_{\time,\rep,\np} = 
    \prod_{\unit=1}^{\Unit} w^M_{\unit,\time,\rep,\np}$
\\
\asp
      Resampling:
        $\prob\big[\resampleIndex({\rep})=a \big] = w^{\IF}_{\time,\rep,a}
  \Big( 
  \sum_{\altNp=1}^{\Np} w^{\IF}_{\time,\rep,\altNp}
  \Big)^{-1}$
\\
\asp 
$\myvec{X}^{\IF}_{\time,\rep} = \myvec{X}^{\IP}_{\time,\rep,r(\rep)}$ 
\\
\asp Compute $\gamma^{}_{\unit,\time+k,\rep,k}$  for $k\ \mathrm{in}\ \seq{0}{K}$,
\\
\asp Compute $\gamma^{+}_{\unit,\time,\rep,0}$
\\
End for
\\
$\displaystyle \MC{\loglik}_{\unit,\time}= 
\log\Bigg(
\frac{
\sum_{\rep=1}^\Rep \gamma^{+}_{\unit,\time,\rep,0}\prod_{k=1}^K
\gamma^{}_{\unit,\time,\rep,k}
}{
\sum_{\rep=1}^\Rep \prod_{k=0}^K
\gamma^{}_{\unit,\time,\rep,k}
}
\Bigg)
$
\vspace{1mm}
\\
\hline
\end{tabular}
\end{center}

It is clearer from the {\ABF}$_2$ representation than from the {\ABF}$_1$ representation that we do not have to save every individual particle and its weight $w^M_{\unit,\time,\rep,\np}$ after the quantities $\myvec{X}^A_{\time,\rep}$, $\gamma^{}_{\unit,\time+k,\rep,k}$ and $\gamma^{+}_{\unit,\time,\rep,0}$ have been computed.
{\ABF}$_2$ has an embarrassingly parallel implementation, since the replicates do not need to interact until  they are combined to compute $\displaystyle \MC{\loglik}_{\unit,\time}$.
An embarrassingly parallel implementation therefore requires $O\big(\Unit(K\Time+\Np)\big)$ memory for each replicate, and $O\big(\Unit\Time \Rep K\big)$ memory when the results from each replicate are collected together.

By using additional communication, for example when the implementation is designed for a single core or multiple cores with shared memory, it is possible to further reduce the memory requirement.
At time $\time$, we need to save only $\myvec{X}^A_{\time,\rep}$ and $\gamma^{}_{\unit,\time-K:\time+K,\rep,k}$ to compute $\displaystyle \MC{\loglik}_{\unit,\time}$ and all subsequent quantities.
Computing  $\myvec{X}^A_{\time,\rep}$ requires $O(\Unit\Np)$ storage.
Therefore, {\ABF} can be implemented with memory requirement $O\big(\Unit (K\Rep+\Np) \big)$, independent of $\Time$.

\bibliography{bib-iif}

\end{document}

### 2020-11-28 theorems.tex
setting up files

\newcommand\PartOfBoundOffDiagonal{3\efourB+4(K + \dmax)\,\efive}
\newcommand\boundOffDiagonal{\efourA + \PartOfBoundOffDiagonal}
\newcommand\K{K^\prime}
\newcommand\ThmOneVarBound{Q^{4b} \, \Unit\Time  \big(\altb + \etwo \, (\Unit\Time-\altb) \big)}

\newcommand\BvConstant{\mathcal{C}_0}
\newcommand\dmax{d_{\max}}
\newcommand\boundLemma{\efourB+(K+d_{\unit,\time})\efive}
\newcommand\boundLemmaUniform{\efourB+(K+\dmax)\efive}

%\newcommand\altB{\tilde{B}}
\newcommand\altB{C}
%\newcommand\el{\epsilon_\ell}
\newcommand\el{\epsilon}   %%% epsilon for Thm 1
%\newcommand\eone{\epsilon_1}
%\newcommand\eone{\epsilon^{}_{\!\mathrm{\ref{A1}}}}
\newcommand\eone{\epsilon^{}_{\!\mathrm{A1}}}
%\newcommand\eone{\epsilon^{A}_{1}}
%\newcommand\etwo{\epsilon^{}_{\!\mathrm{\ref{A:unconditional:mix}}}}
\newcommand\etwo{\epsilon^{}_{\!\mathrm{A4}}}
%\newcommand\altb{\tilde{b}}
\newcommand\altb{c}
\newcommand\Ai{
\begin{assumption}\label{A1}
There is an $\eone>0$, independent of $\Unit$ and $\Time$, and a collection of neighborhoods $\{B_{\unit\comma\time}\subset A_{\unit\comma\time}, \unit\in 1\mycolon\Unit,\time\in 1\mycolon\Time\}$ such that, for all  $\unit$ and $\time$, any  bounded real-valued function $|h(x)|\le 1$, and any value of $x_{B^c_{\unit\comma\time}}$, 
%\begin{equation}
\begin{eqnarray}
\nonumber
&& \hspace{-12mm} \Bigg| \int h(x_{\unit\comma\time}) f_{X_{\unit\comma\time}|Y_{B_{\unit\comma\time}},X_{B^c_{\unit\comma\time}}}(x_{\unit\comma\time}\given  \data{y}_{B_{\unit\comma\time}}, x_{B^c_{\unit\comma\time}}) \, dx_{\unit\comma\time}
\\
\siOnly{\label{eq:weak_coupling2}}
\msOnly{\nonumber}
&& \hspace{-3mm} - \int h(x_{\unit\comma\time})f_{X_{\unit\comma\time}|Y_{B_{\unit\comma\time}}}(x_{\unit\comma\time}\given \data{y}_{B_{\unit\comma\time}})
\, dx_{\unit\comma\time}
\Bigg| \hspace{1mm} < \hspace{1mm} \eone.
%\end{equation}
\end{eqnarray}
\end{assumption}
}

\newcommand\Aii{
\begin{assumption}\label{A1b}
For the collection of neighborhoods in Assumption~\ref{A1}, with $B^+_{\unit\comma\time}=B^{}_{\unit\comma\time}\cup (\unit,\time)$, there is a constant $\Bsize$, depending on $\eone$ but not on $\Unit$ and $\Time$, such that
\begin{equation}
\nonumber
\sup_{\unit\in 1:\Unit, \, \time\in 1:\Time} \big| B^+_{\unit\comma\time}\big| \le \Bsize.
\end{equation}
\end{assumption}
}

\newcommand\Aiii{
\begin{assumption}\label{A2}
There is a constant $Q$, independent of $\Unit$ and $\Time$, such that, for all $\unit$ and $\time$,
\begin{equation}
\nonumber
Q^{-1} <f_{Y_{\unit\comma\time}|X_{\unit\comma\time}}(\data{y}_{\unit\comma\time}\given x_{\unit\comma\time}) < Q
\end{equation}
\end{assumption}
}

\newcommand\Aiv{
  \begin{assumption}\label{A:unconditional:mix}
    There exists $\etwo > 0$, independent of $\Unit$ and $\Time$, such that the following holds.
    For each $\unit, \time$, a set $\altB^{}_{\unit\comma\time} \subset (1\mycolon U)\times(0\mycolon N)$ exists such that $(\altUnit, \altTime) \notin \altB_{\unit\comma\time}$ implies $B_{\unit\comma\time}^+ \cap B_{\altUnit\comma\altTime}^+ = \emptyset$ and
      \begin{equation} 
\siOnly{\label{eq:A:unconditional:mix}}
\msOnly{\nonumber}
        \big| f_{X_{B^{+}_{\altUnit\comma\altTime}}|X^{}_{B^{+}_{\unit\comma\time}}} - f_{X_{B^{+}_{\altUnit\comma\altTime}}} \big|
        < \etwo \, f_{X_{B^{+}_{\altUnit\comma\altTime}}}
      \end{equation}
      Further, there is a uniform bound $|\altB_{\unit\comma\time}| \le \altb$.
\end{assumption}
}

\newcommand\TheoremI{
\begin{theorem}  \label{thm:tif}
Let $\MC{\loglik}$ denote the Monte Carlo likelihood approximation constructed by {\UBF}.
Consider a limit with a growing number of bootstrap replicates, $\Rep\to\infty$, and suppose assumptions~\ref{A1},~\ref{A1b} and~\ref{A2}.
There are quantities $\el(\Unit,\Time)$ and $V(\Unit,\Time)$, with bounds $|\el|<\eone Q^{2}$ and $V < Q^{4\Bsize} \, U^2 N^2$, such that
\begin{equation}
\label{th1:lik:bound}
\Rep^{1/2}\big[ \MC{\loglik}-\loglik-\el\Unit\Time \big]  \xrightarrow[\Rep \rightarrow \infty]{d} \normal\big[0,V\big],
\end{equation}
where $\xrightarrow[\Rep \rightarrow \infty]{d}$ denotes convergence in distribution and $\normal[\mu,\Sigma]$ is the normal distribution with mean $\mu$ and variance $\Sigma$.
If additionally Assumption~\ref{A:unconditional:mix} holds, we obtain an improved variance bound
\begin{equation}
\label{th1:lik:bound2}
V < \ThmOneVarBound.
\end{equation}
\end{theorem}
}

%%% 2222222222222222222222

%\newcommand\ethree{\epsilon^{}_{\mathrm{\ref{B1}}}}
\newcommand\ethree{\epsilon^{}_{\mathrm{B1}}}
%\newcommand\efourA{\epsilon^{}_{\mathrm{\ref{B:unconditional:mix}}}}
\newcommand\efourA{\epsilon^{}_{\mathrm{B4}}}
%\newcommand\efourB{\epsilon^{}_{\mathrm{\ref{B:temporal:mix}}}}
\newcommand\efourB{\epsilon^{}_{\mathrm{B5}}}
%\newcommand\efive{\epsilon^{}_{\mathrm{\ref{B:girf}}}}
\newcommand\efive{\epsilon_{\mathrm{B6}}}

\newcommand\Bi{

\begin{assumptionB}\label{B1}
  There is an $\ethree>0$, independent of $\Unit$ and $\Time$, and a collection of neighborhoods $\{B_{\unit\comma\time}\subset A_{\unit\comma\time}, \unit\in 1\mycolon\Unit,\time\in 1\mycolon\Time\}$ such that the following holds for all $\unit$ and $\time$, and any bounded real-valued function $|h(x)|\le 1$\emph{:} if we write $A=A_{u,n}$, $B=B_{u,n}$, $f_{A}(x_A) = f_{Y_{A}|X_{A}}(\data y_{A}|x_A)$, and $f_{B}(x_B) = f_{Y_{B}|X_{B}}(\data y_{B}|x_B)$,
%\begin{equation}
\begin{eqnarray}
\nonumber
%\label{asif:eq:weak_coupling2}
&&\hspace{-7mm}
\left| 
\int \hspace{-1mm} h(x)
\left\{
\frac{
  \E_{g} \! \big[ 
    f_{A}\big(X_{A}^P\big) f_{X_{u,n}|X_{A^{[n]}},\myvec X_{n-1}} \big(x | X_{A^{[n]}}^P, \myvec X_{n-1}\big) 
  \big]
} {
  \E_{g} \! \big[f_A\big(X_A^P\big)\big]
} - 
\right.
\right.
\\
\nonumber
&& 
\siOnly{\hspace{20mm}}
\hspace{-6.5mm} 
% - \hspace{-1mm} 
\siOnly{\hspace{2mm}}
\left.
\left.
\frac{
  \E_{g} \! \big[
    f_{B}\big(X_{B}^P\big) f_{X_{u,n}|X_{B^{[n]}},\myvec X_{n-1}}\big(x | X_{B^{[n]}}^P, \myvec X_{n-1}\big) \big] 
} {
  \E_{g} \! \big[f_B\big(X_B^P\big)\big]
} 
\!
\right\}
\!
dx
\right| \hspace{0mm} < \hspace{0mm} \ethree.
%\end{equation}
\end{eqnarray}
\end{assumptionB}
}

\newcommand\Bii{
\begin{assumptionB}\label{B1b}
The bound $\sup_{\unit\in 1:\Unit,\time\in 1:\Time} \big| B^+_{\unit\comma\time}\big| \le \Bsize$ in Assumption~\ref{A1b} applies for the neighborhoods defined in Assumption~\ref{B1}. 
This also implies there is a finite maximum temporal depth for the collection of neighborhoods, defined as
\begin{equation}
\nonumber
\dmax=\sup_{(\unit,\time)}\hspace{1mm} \sup_{(\altUnit,\altTime)\in B_{\unit,\time}}|\time-\altTime|.
\end{equation}
\end{assumptionB}
}

\newcommand\Biii{
\begin{assumptionB}\label{B2} Identically to Assumption~\ref{A2}, 
$Q^{-1} <f_{Y_{\unit\comma\time}|X_{\unit\comma\time}}(\data{y}_{\unit\comma\time}\given x_{\unit\comma\time}) < Q$.
\end{assumptionB}
}

\newcommand\BivA{
\begin{assumptionB}\label{B:unconditional:mix}
Define $g_{{X}_{A}|{X}_{B}}$ via \myeqref{eq:g}.
Suppose there is an $\efourA$, independent of $\Unit$ and $\Time$, such that the following holds.
For each $\unit$ and $\time$, a set $\altB^{}_{\unit\comma\time} \subset (1\mycolon U)\times(0\mycolon N)$ exists such that $(\altUnit, \altTime) \notin \altB_{\unit\comma\time}$ implies $B_{\unit\comma\time}^+ \cap B_{\altUnit\comma\altTime}^+ = \emptyset$ and
%      \begin{eqnarray} 
%\label{eq:B:unconditional:mix:1}
\begin{equation}
\nonumber
\big| 
   g^{}_{{X}^{P}_{B^{}_{\altUnit\comma\altTime}}{X}^{P}_{B^{}_{\unit\comma\time}}} 
   - g^{}_{{X}^{P}_{B^{}_{\altUnit\comma\altTime}}} \, g^{}_{{X}^{P}_{B^{}_{\unit\comma\time}}} 
\big|
<
%&\hspace{-3mm}<\hspace{-3mm} & 
%\\
%&&\hspace{-30mm}
(1/2) \, \efourA \, \,  g^{}_{{X}^{P}_{B^{}_{\altUnit\comma\altTime}}} \, g^{}_{{X}^{P}_{B^{}_{\unit\comma\time}}}
%\\
%\label{eq:B:unconditional:mix:2}
\end{equation}
\begin{eqnarray}
\nonumber
\big|
g^{}_{X^P_{B^{}_{\altUnit\comma\altTime}}|\myvec{X}_{0:\Time}} 
\, g^{}_{X^P_{B^{}_{\unit\comma\time}}|\myvec{X}_{0:\Time}} 
-
g^{}_{X^P_{B^{}_{\altUnit\comma\altTime}},X^P_{B^{}_{\unit\comma\time}}|\myvec{X}_{0:\Time}}
\big|
&&
%&\hspace{-3mm}<\hspace{-3mm}& 
\\
\nonumber
%&&\hspace{-30mm}(1/2) \, \efourA \, \, 
&&\hspace{-30mm} < (1/2) \, \efourA \, \, 
 g^{}_{X^P_{B^{}_{\altUnit\comma\altTime}},X^P_{B^{}_{\unit\comma\time}}|\myvec{X}_{0:\Time}}
      \end{eqnarray}
      Further, there is a uniform bound $|\altB_{\unit\comma\time}| \le \altb$.
\end{assumptionB}
}
\newcommand\BivB{
\begin{assumptionB}\label{B:temporal:mix}
There is a constant $K$, independent of $\Unit$ and $\Time$, such that, for any $0\leq d \leq \dmax$, any $\time\ge K+d$, and any set $D\subset (\UnitSet)\times(n{\mycolon}n-d)$, 
\begin{eqnarray}
\nonumber
&&\hspace{-6mm} \big|
g^{}_{{X}_{D}|\myvec{X}_{\time-d-K}}(x^{}_{D}\given\myvec{x}^{(1)}_{\time-d-K})
-
g^{}_{{X}_{D}|\myvec{X}_{\time-d-K}}(x^{}_{D}\given\myvec{x}^{(2)}_{\time-d-K})
\big|
\\
\nonumber
&& \siOnly{\hspace{10mm}} \hspace{4mm}
< 
\efourB  \,  \,
g^{}_{{X}_{D}|\myvec{X}_{\time-d-K}}(x^{}_{D}\given\myvec{x}^{(1)}_{\time-d-K})
\end{eqnarray}
holds for all $\myvec{x}^{(1)}_{\time-d-K}$, $\myvec{x}^{(2)}_{\time-d-K}$, and $x^{}_{D}$.
\end{assumptionB}
}

\newcommand\Bv{
\begin{assumptionB}\label{B:girf}
Let $h$ be a bounded function with $|h(x)|\le 1$. 
Let $\myvec{X}^{\GR}_{\time,\Ninter,\np,\rep}$ be the Monte Carlo quantity constructed in {\ABFIR}, conditional on $\myvec{X}^{\IF}_{\time-1,\Ninter,\rep}=\myvec{x}^{\IF}_{\time-1,\Ninter,\rep}$.
There is a constant $\BvConstant(\Unit,\Time,\Ninter)$ such that, for all $\efive^{}>0$ and $\myvec{x}^{\IF}_{\time-1,\Ninter,\rep}$, whenever the number of particles satisfies $J > \BvConstant(\Unit,\Time,\Ninter)/\efive^{3}$,
\begin{equation}
\nonumber
\hspace{-0.5mm}
\left|  
\, 
  \EMC\Big[ 
    \frac{1}{\Np} \sum_{\np=1}^{\Np} h(\myvec{X}^{\GR}_{\time,\Ninter,\np,\rep}) 
   \Big] \!
-  \E_g \! \big[h(\myvec{X}_\time)\given \myvec{X}_{\time-1}=\myvec{x}^{\IF}_{\time-1,\Ninter,\rep}\big] 
\right| \!
< \efive^{}.
\end{equation}
\end{assumptionB}
}

\newcommand\Bvi{
\begin{assumptionB}\label{B:XG_XA_ind}
For $1\leq \time \leq N$, the Monte Carlo random variable $X^A_{\time,i}$ is independent of $w^M_{\unit,\time,i,j}$ conditional on $X^A_{\time-1,i}$.
\end{assumptionB}
}

\newcommand\TheoremII{
\begin{theorem}  \label{thm:asif}
Let $\MC{\loglik}$ denote the Monte Carlo likelihood approximation constructed by {\ABFIR}, or by {\ABF} since this is the special case of {\ABFIR} with $\Ninter=1$.
Consider a limit with a growing number of bootstrap replicates, $\Rep\to\infty$, and suppose assumptions~\ref{B1},~\ref{B1b}, \ref{B2}, \ref{B:temporal:mix}, \ref{B:girf} and~\ref{B:XG_XA_ind}. 
Suppose the number of particles $\Np$ exceeds the requirement for~\ref{B:girf}.
There are quantities $\el(\Unit,\Time)$ and $V(\Unit,\Time)$ with $|\el|<
Q^2\ethree + 2 Q^{2\Bsize}\big(\boundLemmaUniform\big)$
and $V<Q^{4\Bsize}\Unit^2\Time^2$ 
such that
\begin{equation}
\siOnly{\label{th:asif:lik:bound}}
\msOnly{\nonumber}
\Rep^{1/2}
  \big[ 
    \MC{\loglik}-\loglik-\el\Unit\Time 
  \big]  
  \xrightarrow[\Rep \rightarrow \infty]{d} 
  \normal\big[ 0,V \big].
\end{equation}
If additionally Assumption~\ref{B:unconditional:mix} holds, we obtain an improved rate of
\begin{equation}
\siOnly{\label{th:asif:lik:bound2}}
\msOnly{\nonumber}
V < Q^{4\Bsize}\Time\Unit
\big\{ c + \big(\boundOffDiagonal \big) \big(\Time\Unit-c\big)
\big\}
\end{equation}
\end{theorem}
}

### 2020-11-29 ms.Rnw
a few more small edits

@@
 <<debug,echo=F>>=
-run_level <- 2
+run_level <- 1
+#run_level <- 2
 #run_level <- "jasa-sept-2020"
-#run_level <- 2
 bm_run_level <- run_level
-#mscale_run_level <- run_level
-mscale_run_level <- 3
+mscale_run_level <- run_level
 slice_run_level <- run_level

### 2021-09-02 ms.Rnw
code to make unformatted accepted version

@@
 <<debug,echo=F>>=
-run_level <- 1
-#run_level <- 2
-#run_level <- "jasa-sept-2020"
+# run_level <- "jasa-sept-2020"
+ run_level <- "manual"
+# run_level <- 2
+
+if(is.numeric(run_level)){
   bm_run_level <- run_level
@@
   slice_run_level <- run_level
+}
+
+if(run_level=="manual"){
+  bm_run_level <- 4
+  mscale_run_level <- 5
+  slice_run_level <- 3
+}
 
@@
 \usepackage{amsmath,amssymb}
+%\usepackage{graphicx,psfrag,epsf}
 \usepackage{graphicx}
@@
 \newcommand\rewrite[1]{{\textcolor{red}{#1}}}
+%\newcommand\jpc[1]{{\textcolor{orange}{#1}}}
+\newcommand\aak[1]{{\textcolor{blue}{#1}}}
 
 \usepackage{fullpage}
+% DON'T change margins - should be 1 inch all around.
+%\addtolength{\oddsidemargin}{-.5in}%
+%\addtolength{\evensidemargin}{-.5in}%
+%\addtolength{\textwidth}{1in}%
 \addtolength{\textheight}{-.3in}%
+%%\addtolength{\textheight}{.3in}%
+%\addtolength{\topmargin}{-.8in}%
 
@@
   \author{Edward L.
-Ionides\thanks{This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108, and National Institutes of Health grants 1-U54-GM111274 and 1-U01-GM110712.}\hspace{.2cm}\\
+Ionides\thanks{This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108, and National Institutes of Health grants 1-U54-GM111274, 1-U01-GM110712 and 1-R01-AI143852.
+We acknowledge constructive feedback from two anonymous referees and the associate editor.}\hspace{.2cm}\\
     Department of Statistics, University of Michigan\\
@@
     Joonha Park\\
-    Department of Mathematics \& Statistics, Boston University\\
+    Department of Mathematics, University of Kansas\\
     %and\\
     Aaron A. King\\
-    Department of Ecology \& Evolutionary Biology, University of Michigan
+    Department of Ecology and Evolutionary Biology \& \\ Center for the Study of Complex Systems, University of Michigan
 }
@@
 Bagging (i.e., bootstrap aggregating) involves combining an ensemble of bootstrap estimators.
-The particle filter, also known as sequential Monte Carlo or the bootstrap filter, is a widely used algorithm for estimating the latent states or likelihood of a partially observed Markov process.
-Our bagged filter (BF) methodology combines an ensemble of particle filters, using spatiotemporally localized weights to select successful filters at each point of space and time.
-In some situations, BF methodology can theoretically beat a curse of dimensionality affecting standard particle filters;
-in others, BF fails in theory to beat the curse, but nevertheless evinces practical advantages.
-Our focus is on evaluation of the likelihood function, though BF theory and methodology are also pertinent to latent state estimation.
-Applications suited to BF methodology include spatiotemporal analysis of epidemiological and ecological systems.
-We show that BF methodology can out-perform an ensemble Kalman filter on a coupled population dynamics model arising in the epidemiology of infectious disease.
-We also find that a block particle filter performs well on this task, though the bagged filter respects smoothness and conservation laws that a block particle filter can violate.
+We consider bagging for inference from noisy or incomplete measurements on a collection of interacting stochastic dynamic systems.
+Each system is called a unit, and each unit is associated with a spatial location.
+A motivating example arises in epidemiology, where each unit is a city: the majority of transmission occurs within a city, with smaller yet epidemiologically important interactions arising from disease transmission between cities.
+Monte~Carlo filtering methods used for inference on nonlinear non-Gaussian systems can suffer from a curse of dimensionality as the number of units increases.
+We introduce bagged filter (BF) methodology which combines an ensemble of Monte Carlo filters, using spatiotemporally localized weights to select successful filters at each unit and time.
+We obtain conditions under which likelihood evaluation using a BF algorithm can beat a curse of dimensionality, and we demonstrate applicability even when these conditions do not hold.
+BF can out-perform an ensemble Kalman filter on a coupled population dynamics model describing infectious disease transmission.
+A block particle filter also performs well on this task, though the bagged filter respects smoothness and conservation laws that a block particle filter can violate.
 
 \end{abstract}
+
+%%% JASA instructions
+%% Should be written with the following elements in the following order: title page; author footnote; abstract; keywords; article text (table(s); figures); acknowledgments; appendices; references
+%% Should be no more than 35 pages, inclusive of the abstract, tables, references, figure captions, footnotes, endnotes.
+%% Should contain an unstructured abstract of 200 words.
+%% Should contain between 3 and 5 keywords. Read making your article more discoverable, including information on choosing a title and search engine optimization.
 
@@
 
-Particle filter (PF) algorithms are a widespread tool for inference on nonlinear partially observed dynamic systems \citep{doucet01,doucet11,breto09,ionides15}.
-Modern particle filtering traces back to \citet{gordon93} where the algorithm was called the {\it bootstrap filter}. 
-Standard PF methods suffer from a curse of dimensionality (COD), defined as an exponential increase in computational demands as the problem size grows, which limits its applicability to spatiotemporal systems \citep{bengtsson08,snyder15,rebeschini15}.
-The COD presents empirically as numerical instability of the Monte Carlo algorithm for attainable numbers of particles.
-Bootstrap aggregating (i.e., bagging) is a methodology to improve numerically unstable estimators by combining an ensemble of replicated bootstrap calculations \citep{breiman96}.
-We consider algorithms combining an ensemble of replicated particle filters, and we use the name {\it bagged filter}.
-The task of each bootstrap replicate is to provide some predictive skill at some points in space and time.
-The bagged filter then proceeds to combine these replicates, assessing which replicates are reliable predictors at a given spatiotemporal location.
-
+Bagging is a technique to improve numerically unstable estimators by combining an ensemble of replicated bootstrap calculations \citep{breiman96}.
+In the context of nonlinear partially observed dynamic systems, the bootstrap filter of \citet{gordon93} has led to a variety of particle filter (PF) methodologies \citep{doucet01,doucet11};
+Here, we consider algorithms combining an ensemble of replicated particle filters, which we term {\it bagged filter} algorithms.
+Standard PF methods suffer from a curse of dimensionality (COD), defined as an exponential increase in computational requirement as the problem size grows, limiting its applicability to large systems \citep{bengtsson08,snyder15,rebeschini15}.
+The COD presents empirically as numerical instability of the Monte Carlo algorithm for affordable numbers of particles.
 Much previous research has investigated scalable approaches to filtering and inference with applications to spatiotemporal systems.
 Our bagged filters are in the class of {\it plug-and-play} algorithms, meaning that they require as input a simulator for the latent dynamic process but not an evaluator of transition probabilities \citep{breto09,he10}.
-This simulation-based approach, also known as {\it likelihood-free} \citep{brehmer20} or {\it equation-free} \citep{kevrekidis09}, facilitates application to a wide class of models.
-The ensemble Kalman filter \citep{evensen96,katzfuss19} is a widely used plug-and-play method which uses simulations to parameterize a Gaussian-inspired filtering rule.
+Similar properties to plug-and-play are {\it likelihood-free} \citep{brehmer20} and {\it equation-free} \citep{kevrekidis09}.
+The ensemble Kalman filter \citep{evensen09book,lei10,katzfuss19} is a widely used plug-and-play method which uses simulations to construct a nonlinear filter that is exact for a linear Gaussian model.
 Another plug-and-play approach to combat the COD is the block particle filter \citep{rebeschini15,ng02}.
 Both ensemble Kalman filter and block particle filter methods construct trajectories that can violate smoothness and conservation properties of the dynamic model.
-By contrast, our bagged filters are built using valid trajectories of the dynamic model, but make localization approximations when comparing these trajectories to data.
-It is an empirical question how these strengths and weaknesses play out for a particular data analysis.
-The plug-and-play property of an algorithm makes a general numerical implementation possible: we provide a software environment within which alternative methods can be applied.
+By contrast, our bagged filters are built using valid trajectories of the dynamic model, making localization approximations only when comparing these trajectories to data.
 
 The replicated stochastic trajectories in a bagged filter form an ensemble of representations of the dynamic system.
-Unlike ensemble Kalman filters \citep{lei10} the bagged replicates are independent in a Monte Carlo sense.
+Unlike the particles in a particle filter or ensemble Kalman filter, the bagged replicates are independent in a Monte Carlo sense.
 Bagged filters therefore bear some resemblance to \textit{poor man's ensemble} forecasting methodology in which a collection of independently constructed forecasts is generated using different models and methods \citep{ebert01}.
 Poor man's ensembles have sometimes been found to have greater forecasting skill than any one forecast \citep{leutbecher08,palmer02,chandler13}.
-One explanation for this phenomenon is that even a hypothetically perfect model cannot filter and forecast well using methodology afflicted by the COD.  
+One explanation for this phenomenon is that even a hypothetically perfect model cannot provide effective filtering using methodology afflicted by the COD.  
 We show that bagged filter methodology can relieve this limitation.
@@
 
-In a basic bagged filter, each replicate simply simulates a realization of the latent process model, corresponding to a degenerate instance of PF with only a single particle.
-We call this the unadapted bagged filter ({\UBF}) since each replicate in the ensemble does not track the data.
-While formally beating the COD under a weak mixing assumption, {\UBF} can have poor numerical behavior if a very large number of replicates are needed to reach this happy asymptotic limit.
-Our empirical results show that {\UBF} may indeed be a useful algorithm in some situations.
-More importantly, however, it provides a route to two other algorithms, which we call the adapted bagged filter (\ABF) and adapted bagged filter with intermediate resampling (\ABFIR).
-{\ABF} and {\ABFIR} carry out some resampling on each replicate and therefore enable each replicate to track the data, in a weak sense, while resisting the COD.
-This comes at a price: {\ABF} no longer asymptotically beats COD, and {\ABFIR} does so only under restrictive conditions.
-However, we demonstrate useful behavior for these algorithms on finite samples.
-
-To set up notation, we suppose the latent process is comprised of a collection of units, labeled $\{1,2,\dots,\Unit\}$ which we write as $1\mycolon\Unit$. 
-An unobserved, latent Markov process at a discrete sequence of observation times is written as $\{\myvec{X}_{\time},\time\in 0\mycolon\Time\}$, with $\myvec{X}_{\time}=X_{1:\Unit,\time}$ taking values in $\Xspace^\Unit$.
+We first consider a simple bagged filter where each replicate is an independent simulation of the latent process model.
+We call this the unadapted bagged filter ({\UBF}) since the replicates in the ensemble depend on the model but not on the data.
+{\UBF} is described in Sec.~\ref{sec:ubf}, with a theoretical analysis presented in Sec.~\ref{sec:ubf:theory}.
+Each {\UBF} replicate corresponds to a basic PF algorithm with a single particle.
+We show that {\UBF} formally beats the COD under a weak mixing assumption, though {\UBF} can have poor numerical behavior if a very large number of replicates are needed to reach this happy asymptotic limit.
+Subsequent empirical results show that {\UBF} may nevertheless be a useful algorithm in some situations.
+In Sec.~\ref{sec:abf}, we generalize {\UBF} to construct an adapted bagged filter (\ABF) where each replicate tracks the data.
+The price of adaptation is that {\ABF} no longer avoids the COD, a limitation that can be controlled in certain situations by supplementing {\ABF} with a technique called intermediate resampling, to obtain the {\ABFIR} algorithm.
+Theoretical results for {\ABF} and {\ABFIR} algorithms are developed in Sec.~\ref{sec:abf:theory}.
+The algorithms are demonstrated in action and compared with alternative approaches in Sec.~\ref{sec:examples}.
+
+\section{The unadapted bagged filter (UBF)}
+\label{sec:ubf}
+
+Suppose the collection of units is indexed by the set $\{1,2,\dots,\Unit\}$, which is written as $1\mycolon\Unit$.
+The latent Markov process is denoted by $\{\myvec{X}_{\time},\time\in 0\mycolon\Time\}$, with $\myvec{X}_{\time}=X_{1:\Unit,\time}$ taking values in a product space $\Xspace^\Unit$.
+This discrete time process may arise from a continuous time Markov process $\{\myvec{X}(t), t_0\le t\le t_{\Time} \}$ observed at times $t_{1:\Time}$, and in this case we set $\myvec{X}_\time=\myvec{X}(t_\time)$.
 The initial value $\myvec{X}_{0}$ may be stochastic or deterministic.
-Observations are made on each unit, modeled by an observable process $\{\myvec{Y}_{\time}=Y_{1:\Unit,\time},\time\in 1\mycolon\Time\}$ which takes values in $\Yspace^\Unit$.
+Observations are made on each unit, modeled by an observable process $\{\myvec{Y}_{\time}=Y_{1:\Unit,\time},\time\in 1\mycolon\Time\}$ which takes values in a product space $\Yspace^\Unit$.
 Observations are modeled as being conditionally independent given the latent process.
-The conditional independence of measurements applies over both space and time, so $Y_{\unit\comma\time}$ is conditionally independent of
-$\big\{
-  X_{\altUnit\comma\altTime},
-  Y_{\altUnit\comma\altTime},
-  (\altUnit\comma\altTime)\neq(\unit\comma\time)
-\big\}$
-given $X_{\unit\comma\time}$.
-This unit structure for the observation process is not necessary for all that follows (see Sec.~\SuppSecGeneralization).
-We suppose the existence of a joint density $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$  of $\{X_{1:\Unit,\time}\}$ and $\{Y_{1:\Unit,\time}\}$ with respect to some appropriate measure.
-We use the same subscript notation to denote other marginal and conditional densities.
-The data are ${y}_{\unit\comma\time}$ for spatial unit $\unit$ at time $\time$. 
-For a fixed value of $\Unit$, this model fits into the partially observed Markov process (POMP) framework \citep{breto09}, also known as a state space model or hidden Markov model.
-Spatiotemporal POMP models are characterized by a latent Markov process with dimension proportional to $\Unit$.
-We use the term {SpatPOMP} to describe POMP models indexed by $\unit$ as well as $\time$.
-In accordance with the spatial interpretation of the units $\unit\in 1\mycolon\Unit$, we may suppose that units sufficiently distant in some sense approach independence, a phenomenon called {\it weak coupling}.
-It is worth pointing out that inference techniques designed for SpatPOMPs are also relevant in high-dimensional stochastic dynamic systems where $\unit$ does not index spatial locations.
-In the following, we use the ordering on the set of observations corresponding to their labels $1\mycolon\Unit$ to define the set of observations preceding unit $\unit$ at time $\time$ as
+The conditional independence of measurements applies over both time and the unit structure, so the collection $\big\{Y_{\unit\comma\time},\unit\in\seq{1}{\Unit},\time\in\seq{1}{\Time}\big\}$ is conditionally independent given
+$\big\{X_{\unit\comma\time},\unit\in\seq{1}{\Unit},\time\in\seq{1}{\Time}\big\}$.
+The unit structure for the observation process is not necessary for all that follows (see Sec.~\SuppSecGeneralization).
+We suppose the existence of a joint density $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$  of $X_{1:\Unit,0:\Time}$ and $Y_{1:\Unit,1:\Time}$ with respect to some appropriate measure, following a notational convention that the subscripts of $f$ denote the joint or conditional density under consideration.
+The data are $\data{y}_{\unit\comma\time}$ for unit $\unit$ at time $\time$.
+This model is a special case of a  partially observed Markov process  \citep[POMP,][]{breto09}, also known as a state space model or hidden Markov model.
+The additional unit structure, not generally required for a POMP, is appropriate for modeling interactions between units characterized by a spatial location, and so we call the model a SpatPOMP.
+In the following, we use a lexicographical ordering on the set of observations;
+Specifically, we define the set of observations preceding unit $\unit$ at time $\time$ as
 \begin{equation}\label{eq:setA}
@@
 $f_{X_{\unit\comma\time}|X_{A_{\unit\comma\time}}}$ will frequently be hard to compute or simulate from.
-However, the algorithms we study do not carry out such computations, but only evaluate the measurement model on these neighborhoods.
-To represent weak coupling, we suppose there is a neighborhood $B_{\unit\comma\time}\subset A_{\unit\comma\time}$ such that the latent process on 
+The bagged filter algorithms we study do not evaluate or simulate such transition densities but only compute the measurement model on neighborhoods, unlike the filter of \citet{beskos17} built on a similar factorization.
+If sufficiently distant units are approximately independent, we say the system is {\it weakly coupled}.
+In this case, we suppose there is a neighborhood $B_{\unit\comma\time}\subset A_{\unit\comma\time}$ such that the latent process on 
 $A_{\unit\comma\time} \setminus B_{\unit\comma\time}$ 
@@
 
-The inputs for the {\UBF}, {\ABF} and {\ABFIR} are listed in Table~1.
-The plug-and-play property is evident since all three algorithms require as input a simulator for the latent dynamic process but not an evaluator of transition probabilities.
-Our primary interest is in the estimation of the log likelihood for the data given the model, $\loglik=\log f_{\myvec{Y}_{1:\Time}}(\data{\myvec{y}}_{1:\Time})$, which is of fundamental importance in both Bayesian and non-Bayesian statistical inference.
-Monte Carlo estimates of the log likelihood underpin modern methods for full-information statistical inference on partially observed dynamic systems \citep{andrieu10,ionides15,ionides17}.
-Filtering theory ties together the three tasks of likelihood evaluation, forecasting, and inferring latent dynamic variables \citep{crisan11}.
-Successful likelihood evaluation is therefore closely related to success at forecasting and latent process reconstruction.
-
-The pseudocodes for the algorithms that follow will adopt a convention that implicit loops are carried out over all free indices.
-For example, the construction of $w^P_{\unit,\time,\rep}$ in {\UBF} has an implicit loop over $\unit$, $\time$ and $\rep$. 
-The construction of $w^P_{\unit,\time,\rep,\np}$ in {\ABF} has an implicit loop over $\unit$, $\rep$ and $\np$ but not over $\time$ since $\time$ is already defined in an explicit loop and is therefore not a free index.
-Table~1 lists the ranges of the implicit loops for all following algorithms.
-
-\begin{center}
-\noindent\begin{tabular}{l}
-\hline
-\inputSpace {\bf Table 1. Bagged filter inputs, outputs and implicit loops.}\\
-\hline
-\inputSpace {\bf input:}
+Our primary interest is estimation of the log likelihood for the data given the model, $\loglik=\log f_{\myvec{Y}_{1:\Time}}(\data{\myvec{y}}_{1:\Time})$, which is of fundamental importance in both Bayesian and non-Bayesian statistical inference.
+A general filtering problem is to evaluate $\E\big[h(X_{\unit,\time}) \given Y_{A_{\unit,\time}}\equals \data{y}_{A_{\unit,\time}}\big]$ for some function $h:\Xspace\to\R$.
+Taking $h(x)= f_{Y_{\unit,\time}| X_{\unit,\time}}\big(\data{y}_{\unit,\time}\given x\big)$ gives a filtering representation of the likelihood evaluation problem.
+Further discussion on bagged filtering for other filtering problems is given in Sec.~{\SuppSecLatentStates}.
+For likelihood-based inference, maximization plays an important role in point estimation, confidence interval construction, hypothesis testing and model selection. 
+An extension of bagged filtering to likelihood maximization is demonstrated in Sec.~\ref{sec:profile} following the approach described in Sec.~{\SuppSecParameterEst}.
+
+Pseudocode for a {\UBF} algorithm for likelihood evaluation is given below.
+The prediction weight  $w^P_{u,n,i}$ gives an appropriate weighting for replicate $i$ for predicting $\data{y}_{u,n}$ based on the most relevant data, $\data{y}_{B_{u,n}}$.
+Conditional log likelihoods are estimated using an approximation
+\begin{eqnarray*}
+    \ell_{u,n} &=& \log f_{Y_{u,n}|Y_{A_{u,n}}}\big(\data{y}_{u,n}\given \data{y}_{A_{u,n}}\big)
+    = \log \left( \int f_{Y_{u,n}|X_{u,n}}(\data{y}_{u,n}\given x) \, f_{X_{u,n}|Y_{A_{u,n}}}(x \given \data{y}_{A_{u,n}}) \, dx \right)
 \\
-simulator for $f_{\myvec{X}_0}(\myvec{x}_0)$ and $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1})$\\
-    evaluator for $f_{{Y}_{\unit\comma\time}|{X}_{\unit\comma\time}}({y}_{\unit\comma\time}\given {x}_{\unit\comma\time})$\\
-    number of replicates, ${\Rep}$\\
-    neighborhood structure, $B_{\unit\comma\time}$\\
-    data, $\data{\myvec{y}}_{1:\Time}$\\
-{\ABF} and {\ABFIR}:    particles per replicate,  $\Np$\\
-{\ABFIR}: number of intermediate timesteps, $\Ninter$ \\
-{\ABFIR}: measurement variance parameterizations, ${\VtoTheta}_{\unit\comma\time}$ and ${\thetaToV}_{\unit\comma\time}$\\
-{\ABFIR}: approximate process and observation mean functions, $\myvec{\mu}$ and $h_{\unit\comma\time}$\\
-\inputSpace {\bf output:}\\
-  Log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit\comma\time}$\\
-\inputSpace {\bf implicit loops:}\\
-$\; \unit \mbox{ in } \seq{1}{\Unit}$, 
-$\; \time \mbox{ in } \seq{1}{\Time}$, 
-$\; \rep \mbox{ in } \seq{1}{\Rep}$, 
-$\; \np \mbox{ in } \seq{1}{\Np}$
-%$\; \npgir \mbox{ in } \seq{1}{\Npgir}$
-\lastLineSpace \\
-\hline
-\end{tabular}
-\end{center}
-
-The~{\UBF} algorithm that follows amounts to a bagged filter with a single particle per replicate, with no weighting or resampling within each replicate.
-
-%%%%%%%%%%%%% ttttttttttttttttttttttttt
-
-\begin{center}
-\noindent\begin{tabular}{l}
-\hline
-\inputSpace {\bf {\UBF}. Unadapted bagged filter.}\\
-\hline
-%\inputSpace {\bf input:} From Table~1, with $\Np=1$ \\
-%\hline
-\firstLineSpace 
-Simulate $\myvec{X}^{\tif}_{0:\Time,\rep}\sim f_{\myvec{X}_{0:\Time}}(\myvec{x}_{0:\Time})$\\ 
-Measurement weights,
-$w^M_{\unit,\time,\rep}=f_{Y_{\unit,\time}|X_{\unit,\time}}(\data{y}_{\unit,\time}\given X^{\tif}_{\unit,\time,\rep})$
-\\
-Prediction weights, 
-$w^P_{\unit,\time,\rep}=\prod_{(\tilde \unit,\tilde n)\in B_{\unit,\time}}
-w^M_{\tilde\unit,\tilde n,\rep}$\\
+  &\approx& \log \left( \int f_{Y_{u,n}|X_{u,n}}(\data{y}_{u,n}\given x) \, f_{X_{u,n}|Y_{B_{u,n}}}(x\given \data{y}_{B_{u,n}}) \, dx \right).
+\end{eqnarray*}
+The choice of $B_{u,n}$ is determined empirically, with a bias-variance trade-off used to compare small neighborhoods such as $B_{u,n} = \{ (u,n-1),(u-1,n)\}$ or $B_{u,n} = \{ (u,n-1),(u,n-2)\}$ against larger neighborhoods.
+The plug-and-play property is evident because {\UBF} requires as input a simulator for the latent coupled dynamic process but not an evaluator of transition probabilities.
+
+%%%%%%%%%%%%% uuuuuuuuu
+
+% UBF PSEUDOCODE uuuuuuuuuu
+\begin{algorithm}[H]
+  \caption{\bf Unadapted bagged filter (UBF).
+   }\label{alg:ubf}
+  \KwIn{
+    simulator for $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1})$ and $f_{\myvec{X}_0}(\myvec{x}_0)$;
+    evaluator for $f_{{Y}_{\unit,\time}|{X}_{\unit,\time}}({y}_{\unit,\time}\given {x}_{\unit,\time})$;
+    data, $\data{\myvec{y}}_{1:\Time}$;
+    number of replicates, $\Rep$;
+    neighborhood structure, $B_{\unit,\time}$
+  }
+\For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{
+initialize simulation, $\myvec{X}_{0,\rep} \sim f_{\myvec{X}_0}(\cdot)$
+  \;
+  \nllabel{alg:ubf:for:n}
+\For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{
+simulate,
+    $\myvec{X}_{\time,\rep} \sim
+      f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}
+      \big( \mydot \given \myvec{X}_{\time-1,\rep}
+    \big)$
+  \nllabel{alg:ubf:adapted:proposals}
+  \;
+measurement weights,
+  $w^M_{\unit,\tilde n,\rep}=
+    f_{Y_{\unit,\time}|X_{\unit,\time}}
+    \big (\data{y}_{\unit,\time}\given {X}_{\unit,\time,\rep}
+  \big)$
+    for $\unit$ in $\seq{1}{\Unit}$
+  \;
+prediction weights,
+  $w^P_{\unit,\time,\rep}=\prod_{(\tilde \unit,\tilde n)\in B_{\unit,\time}} w^M_{\tilde\unit,\tilde n,\rep}$
+    for $\unit$ in $\seq{1}{\Unit}$
+      \nllabel{alg:ubf:adapted:weights}
+  \;
+  }
+}
+%conditional log likelihood,
   $\MC{\loglik}_{\unit,\time}= 
@@
       \sum_{\rep=1}^\Rep w^P_{\unit,\time,\rep}
-\right)
-$
-\lastLineSpace
-\\
-\hline
-\end{tabular}
-\end{center}
+    \right)$
+  for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$
+  \;
+\KwOut{
+log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit\comma\time}$\\
+}
+\end{algorithm}
+
+\subsection{{\UBF} theory}
+\label{sec:ubf:theory}
+
+A dataset $\data{\myvec{y}}_{1:\Time}$ with $\Unit$ units is modeled via a joint density  $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$.
+We consider non-asymptotic bounds that apply for all values of $\Unit$ and $\Time$. 
+To impose a requirement that distant regions of space-time behave similarly and have only weak dependence, we assert the following conditions which define constants $\eone$, $\etwo$ and $Q$ used to bound the bias and variance in Theorem~\ref{thm:tif}.
+Stronger bounds are obtained when the conditions hold for small $\eone$, $\etwo$ and $Q$.
+
+\Ai  %% Assumption 1  Assumption~\ref{A1}
+
+\Aii  %% Assumption 2  Assumption~\ref{A1b}
+
+\Aiii   %% Assumption 3  Assumption~\ref{A2}
+
+\Aiv    %% Assumption 4  Assumption~\ref{A:unconditional:mix}
+
+The two mixing conditions in Assumptions~\ref{A1} and~\ref{A:unconditional:mix} are subtly different. 
+Assumption~\ref{A1} describes a conditional mixing property dependent on the data, whereas~\ref{A:unconditional:mix} asserts a form of unconditional mixing.
+Although both capture a similar concept of weak coupling, conditional and unconditional mixing properties do not readily imply one another.
+Assumption~\ref{A2} is a compactness condition of a type that has proved useful in the theory of particle filters despite the rarity of its holding exactly.
+Theorem~\ref{thm:tif} shows that these conditions let {\UBF} compute the likelihood with a Monte Carlo variance of order $\Unit\Time\Rep^{-1}$ with a bias of order $\Unit\Time\epsilon$.
+
+\TheoremI %% Theorem~\ref{thm:tif}
+
+\begin{proof}
+A complete proof is given in Sec.~\SuppSecThmI. 
+Briefly, the assumptions imply a multivariate central limit theorem for $\{\MC{\loglik}_{\unit\comma\time}, (\unit,\time)\in \UnitSet{\times}\ObsTimeSet\}$ as $\Rep\to\infty$.
+The limiting variances and covariances are uniformly bounded, using Assumptions~\ref{A1b} and~\ref{A2}. 
+Assumption~\ref{A1} provides a uniform bound on the discrepancy between ${\loglik}_{\unit\comma\time}$ and mean of the Gaussian limit.
+This is enough to derive \myeqref{th1:lik:bound}.
+Assumption~\ref{A:unconditional:mix} gives a stronger bound on covariances between sufficiently distant units, leading to \myeqref{th1:lik:bound2}.
+\end{proof}
+
+Theorem~\ref{thm:tif} does not guarantee uniformity over $\Unit$ and $\Time$ of the rate of convergence as $\Rep\to\infty$.
+However, it does guarantee that the polynomial bounds in (\ref{th1:lik:bound}) and (\ref{th1:lik:bound2}) hold for sufficiently large $\Rep$.
+The COD is characterized by exponential bounds, and so Theorem~\ref{thm:tif} shows a specific sense in which {\UBF} can avoid COD.
+%% This formalizes the heuristic that {\UBF} uses prediction weights $w^P_{u,n,i}$ which scale exponentially with the size of local neighborhoods $B_{u,n}$, avoiding the use of weights that scale exponentially with increasing $U$.
+Uniformity of the central limit convergence in Theorem~\ref{thm:tif} may be expected to hold via a Berry-Esseen theorem, but extension of existing Berry-Esseen results for dependent processes \citep{bentkus97,jirak16} is beyond the scope of this article.
+
+The approximation error for {\UBF} can be divided into two sources: a localization bias due to conditioning on a finite neighborhood, and Monte Carlo error. 
+The localization bias does not disappear in the limit as Monte Carlo effort increases.
+It does become small as the conditioning neighborhood increases, but the Monte Carlo effort grows exponentially in the size of this neighborhood.
+Although the filtering inference is carried out using localization, the simulation of the process is carried out globally which avoids the introduction of additional boundary effects and ensures that the simulations comply with any constraints satisfied by the model for the latent process.
+
+\section{Adaptation and intermediate resampling}
+\label{sec:abf}
 
 Theorem~\ref{thm:tif} shows that {\UBF} can beat COD.
-However, an algorithm such as {\UBF}, which includes no resampling, will fail on long time series absent rapid spatiotemporal mixing, which allows simulated particles to remain relevant over the course of a long time series.
-It may in practice be necessary to select simulations consistent with the data, much as standard PF algorithms do.
+However, {\UBF} can perform poorly on long time series unless weak temporal dependence allows simulated sample paths to remain relevant over the course of a long time series. 
+For example, we will find that {\UBF} performs well on an epidemiological model (Sec.~\ref{sec:discussion}) but less well on a geophysical model (Sec.~{\SuppSecLorenz}).
+It is sometimes necessary to select simulations consistent with the data, much as standard PF algorithms do.
 We look for approaches that build on the basic insight of {\UBF} while having superior practical performance.
 
-Whereas the full global filtering problem may be intractable via importance sampling methods, a version of this problem local in space and time may nevertheless be feasible.
-Drawing a sample from the conditional density, $f_{\myvec{X}_{\time}|\myvec{Y}_{\time},\myvec{X}_{\time-1}}$, is distinct from the filtering problem: we call this operation {\it adapted simulation}.
-For models where $\myvec{X}_{\time}$ is close to $\myvec{X}_{\time-1}$, adapted simulation requires a local calculation that may be much easier than the full filtering calculation.
-The following adapted bagged filter ({\ABF}) is constructed under a hypothesis that the adapted simulation problem is tractable.
-The adapted simulations are then reweighted in a neighborhood of each point in space and time to construct a local approximation to the filtering problem which leads to an estimate of the likelihood.
+Whereas the full global filtering problem of drawing from $f_{\myvec{X}_{\time}|\myvec{Y}_{1:\time}}$ may be intractable via importance sampling methods, a version of this problem localized in space and time may nevertheless be feasible.
+The conditional density, $f_{\myvec{X}_{\time}|\myvec{Y}_{\time},\myvec{X}_{\time-1}}$, is called the {\it adapted density}, and simulating from this density is called {\it adapted simulation}.
+For models where $\myvec{X}_{\time-1}$ is highly informative about $\myvec{X}_{\time}$, importance sampling for adapted simulation may be much easier than the full filter calculation.
+The following adapted bagged filter ({\ABF}) is constructed under a hypothesis that the adapted simulation problem is tractable, and it is applicable when the number of units is prohibitive for Monte Carlo sampling from the full filter distribution but not for sampling from the adapted distribution.
+In {\ABF}, the adapted simulations are reweighted in a neighborhood of each unit and time point to construct a local approximation to the filtering problem which leads to an estimate of the likelihood.
+The pseudocode for {\ABF}, below, reduces to {\UBF} when using a single particle per repicate, $\Np=1$.
 
@@
 
-\begin{center}
-\noindent\begin{tabular}{l}
-\hline
-{\bf 
-{\ABF}. Adapted bagged filter.}\inputSpace\\
-\hline
-\firstLineSpace
-Initialize adapted simulation: $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\myvec{x}_0)$
-\\
-For $\time\ \mathrm{in}\ \seq{1}{\Time}$
-\\
-\asp  Proposals:
+% ABF PSEUDOCODE AAAAAAAAAAAAAAAAAAAA
+\begin{algorithm}[H]
+  \caption{\bf Adapted bagged filter (ABF)
+  }\label{alg:abf}
+  \KwIn{
+    simulator for $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1})$ and $f_{\myvec{X}_0}(\myvec{x}_0)$;
+    evaluator for $f_{{Y}_{\unit,\time}|{X}_{\unit,\time}}({y}_{\unit,\time}\given {x}_{\unit,\time})$;
+    data, $\data{\myvec{y}}_{1:\Time}$;
+    number of particles per replicate, $\Np$;
+    number of replicates, $\Rep$;
+    neighborhood structure, $B_{\unit,\time}$
+  }
+\For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{
+initialize adapted simulation, $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\cdot)$
+  \;
+  \nllabel{alg:abf:for:n}
+\For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{
+proposals:
   $\myvec{X}_{\time,\rep,\np}^{\IP} \sim 
@@
     \big( \myvec{x}_{\time}\given \myvec{X}^{\IF}_{\time-1,\rep}\big)$
-\\
-\asp Measurement weights:
+  for $\np$ in $\seq{1}{\Np}$
+  \nllabel{alg:abf:adapted:proposals}
+  \;
+% measurement weights:
   $w^M_{\unit,\time,\rep,\np} = 
@@
     \big (\data{y}_{\unit\comma\time}\given X^{\IP}_{\unit\comma\time,\rep,\np}\big)$
-\\
-\asp  Adapted resampling weights:
+  for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$
+  \;
+adapted resampling weights,
     $w^{\IF}_{\time,\rep,\np} = 
     \prod_{\unit=1}^{\Unit} w^M_{\unit,\time,\rep,\np}$
-\\
-\asp
-      Resampling:
-        $\prob\big[\resampleIndex({\rep})=a \big] = w^{\IF}_{\time,\rep,a}
+    for $\unit$ in $\seq{1}{\Unit}$, 
+    $\np$ in $\seq{1}{\Np}$
+  \nllabel{alg:abf:adapted:weights}
+  \;
+%resampling:
+  $\myvec{X}^{\IF}_{\time,\rep} = \myvec{X}^{\IP}_{\time,\rep,r(\rep)}$ with $\prob\big[\resampleIndex({\rep})=a \big] = w^{\IF}_{\time,\rep,a}
     \Big( 
@@
     \Big)^{-1}$ 
-\\
-\asp 
-$\myvec{X}^{\IF}_{\time,\rep} = \myvec{X}^{\IP}_{\time,\rep,r(\rep)}$ 
-\\
-\asp % Prediction weights:
+  \;
+% Prediction weights:
   $w^{\LCP}_{\unit,\time,\rep,\np}= \displaystyle
@@
     \hspace{1mm}
-       \prod_{\altUnit:(\altUnit,\altTime)\in B_{\unit,\time}} 
-    \hspace{-1mm}
+       \prod_{(\altUnit,\altTime)\in B_{\unit,\time}} 
+    \hspace{-3mm}
         w^M_{\altUnit,\altTime,\rep,k}
-  \Big] \prod_{\altUnit:(\altUnit,\time)\in B_{\unit,\time}} 
-    \hspace{-1mm}
+  \Big]
+  \prod_{(\altUnit,\time)\in B_{\unit,\time}} 
+    \hspace{-3mm}
         w^M_{\altUnit,\time,\rep,\np}$
-\\
-End for
-\\
-$\displaystyle \MC{\loglik}_{\unit,\time}= 
+      for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$
+  \nllabel{alg:abf:lp:weights} \nllabel{alg:abf:end:n}
+  \;
+}
+}
+  $ \displaystyle
+    \MC{\loglik}_{\unit,\time}= 
     \log\Bigg(
@@
       }
-\Bigg)
-$
-\vspace{1mm}
-\\
-\hline
-\end{tabular}
-\end{center}
-
-In addition, we consider an algorithm which adds intermediate resampling to the adapted simulation replicates, and is therefore termed {\ABFIR}.
+   \Bigg)$
+   for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$
+   \;
+\KwOut{
+log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit\comma\time}$\\
+}
+\end{algorithm}
+
+{\ABF} remedies a weakness of {\UBF} by making each boostrap filter adapted to the data.
+However, this benefit carries a cost, since adapted simulation is not immune from the curse of dimensionality.
+Therefore, we also consider an algorithm called {\ABFIR} which uses an intermediate resampling technique to carry out the adapted simulation.
 Intermediate resampling involves assessing the satisfactory progress of particles toward the subsequent observation at a collection of times between observations.
-This is well defined when the latent system is a continuous time process $\{\myvec{X}(t)\}$ with observation times $t_{1:\Time}$ such that $\myvec{X}_\time=\myvec{X}(t_{\time})$.
-For a continuous time latent process model, intermediate resampling can have favorable scaling properties when the intermediate times 
+This is well defined when the latent process has a continuous time representation, $\{\myvec{X}(t)\}$, with observation times $t_{1:\Time}$.
+We write $\Ninter$ intermediate resampling times as 
 \begin{equation}
 \nonumber
-t_{\time-1}=t_{\time,0}<t_{\time,1}<\dots<t_{\time,\Ninter}=t_{\time} 
+t_{\time-1}=t_{\time,0}<t_{\time,1}<\dots<t_{\time,\Ninter}=t_{\time}.
 \end{equation}
-scale with $\Ninter\approx \Unit$ \citep{park20}.
+Carrying out an intermediate resampling procedure can have favorable scaling properties when $\Ninter$ is proportional to $\Unit$ \citep{park20}.
 In the case $\Ninter=1$, {\ABFIR} reduces to {\ABF}.
@@
 This is a generalization of the popular auxiliary particle filter \citep{pitt99}.
-The implementation in this pseudocode constructs the guide $g_{\time,\ninter,\rep,\np}$ using a simulated moment method proposed by \citet{park20}.
+Only an ideal guide function fully addresses COD \citep{park20} and on nontrivial problems this is not available.
+However, practical guide functions can nevertheless improve performance.
+
+The implementation in the {\ABFIR} pseudocode constructs the guide $g_{\time,\ninter,\rep,\np}$ using a simulated moment method proposed by \citet{park20}.
 The quantities  $\myvec{X}_{\time,\rep,\npgir}^{G}$, $V_{\unit,\time,\rep}$, $\myvec{\mu}^{\GP}_{\time,\ninter,\rep,\np}$, $V^{\mathrm{meas}}_{\unit,\time,\ninter,\rep,\np}$, $V^{\mathrm{proc}}_{\unit,\time,\ninter,\rep}$ and $\theta_{\unit,\time,\ninter,\rep,\np}$ constructed in {\ABFIR} are used only to construct $g_{\time,\ninter,\rep,\np}$.
@@
 \end{eqnarray}
-Also required for {\ABFIR} is an inverse function ${\VtoTheta}_{\unit\comma\time}$ such that
+Also required for {\ABFIR} is an inverse function ${\VtoTheta}_{\unit\comma\time}\big(V,x_{\unit\comma\time},\theta\big)$ such that ${\VtoTheta}_{\unit\comma\time}\big(V,x_{\unit\comma\time},\theta\big)=\phi$ implies
 \begin{equation}
 \nonumber
-{\thetaToV}_{\unit\comma\time} \big( x_{\unit\comma\time},\VtoTheta_{\unit\comma\time}(V,x_{\unit\comma\time},\theta) \big) = V.
+{\thetaToV}_{\unit\comma\time} \big( x_{\unit\comma\time},\phi \big) = V.
 \end{equation}
@@
 
-\begin{center}
-\noindent\begin{tabular}{l}
-\hline
-{\bf {\ABFIR}. Adapted bagged filter with intermediate resampling.} 
-\firstLineSpace \\  
-\vspace{0.4mm} \\
-\hline
-\firstLineSpace
-Initialize adapted simulation: $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\myvec{x}_0)$
-\\
-For $\time\ \mathrm{in}\ \seq{1}{\Time}$
-\\
-\asp Guide simulations:
+% ABF-IR PSEUDOCODE iiiiiiiiii
+\begin{algorithm}[H]
+  \caption{\bf ABF with intermediate resampling (ABF-IR)
+  }\label{alg:abfir}
+  \KwIn{same as Algorithm~\ref{alg:abf} (ABF) plus:
+%    simulator for $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1})$ and $f_{\myvec{X}_0}(\myvec{x}_0)$;
+%    evaluator for $f_{{Y}_{\unit,\time}|{X}_{\unit,\time}}({y}_{\unit,\time}\given {x}_{\unit,\time})$;
+%    data, $\data{\myvec{y}}_{1:\Time}$;
+%    number of particles per replicate, $\Np$;
+%    number of replicates, $\Rep$;
+%    neighborhood structure, $B_{\unit,\time}$;
+    intermediate timesteps, $\Ninter$;
+    measurement variance functions, ${\VtoTheta}_{\unit\comma\time}$ and ${\thetaToV}_{\unit\comma\time}$;
+    approximate mean functions, $\myvec{\mu}$ and $h_{\unit\comma\time}$
+  }
+\For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{  
+initialize adapted simulation, $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\cdot)$
+  \;
+  \nllabel{alg:abfir:for:n}  
+\For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{
+  guide simulations:
     $\myvec{X}_{\time,\rep,\npgir}^{G} \sim 
@@
     \big( \myvec{x}_{\time}\given \myvec{X}^{\IF}_{\time-1,\rep} \big)$
-\\
-\asp Guide sample variance: $V_{\unit,\time,\rep}=
+  for $\np$ in $\seq{1}{\Np}$
+  \;
+% Guide sample variance:
+  $V_{\unit,\time,\rep}=
     \var \big\{
-        h_{\unit\comma\time}\big( {X}_{\unit,\time,\rep,\npgir}^{G}\big), \npgir \mbox{ in } \seq{1}{\Npgir}
+    h_{\unit\comma\time}\big( {X}_{\unit,\time,\rep,\npgir}^{G}\big), \ 
+      \npgir \mbox{ in } \seq{1}{\Npgir}
       \big\}$
-\\
-\asp $\guideFunc^{\resample}_{\time,0,\rep,\np}=1 \; \; $ and
+  \;
+  $\guideFunc^{\resample}_{\time,0,\rep,\np}=1 \; \; $ and
     $\; \myvec{X}_{\time,0,\rep,\np}^{\GR}=\myvec{X}^{\IF}_{\time-1,\rep}$
-\\
-\asp For $\ninter  \,\, \mathrm{in} \,\, \seq{1}{\Ninter}$
-\\
-\asp\asp Intermediate proposals:
+  for $\np$ in $\seq{1}{\Np}$
+  \;
+%%% intermediate loop sssss  
+  \For{$\ninter\ \mathrm{in}\ \seq{1}{\Ninter}$}{
+%Intermediate proposals:
      ${\myvec{X}}_{\time,\ninter,\rep,\np}^{\GP}
@@
        \big(\mydot|{\myvec{X}}_{\time,\ninter-1,\rep,\np}^{\GR}\big)$
-\\
-\asp\asp 
+     for $\np$ in $\seq{1}{\Np}$
+     \;
      $\myvec{\mu}^{\GP}_{\time,\ninter,\rep,\np} 
-           = \myvec{\mu}\big( \myvec{X}^{\GP}_{\time,\ninter,\rep,\np},t_{\time,\ninter},t_{\time} \big)$
-\\
-\asp\asp      %Measurement variance at skeleton: 
+       = \myvec{\mu}\big( \myvec{X}^{\GP}_{\time,\ninter,\rep,\np},
+       t_{\time,\ninter},t_{\time} \big)$
+     for $\np$ in $\seq{1}{\Np}$
+     \;
+%Measurement variance at skeleton: 
      $V^{\mathrm{meas}}_{\unit,\time,\ninter,\rep,\np}
         = \thetaToV_{\unit}(\theta,\mu^{\GP}_{\unit,\time,\ninter,\rep,\np})$
-\\
-\asp\asp  %Process variance:
+     for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$
+     \;
+%Process variance:
      $V^{\mathrm{proc}}_{\unit,\time,\ninter,\rep}
@@
        \big(t_{\time}-t_{\time,0}\big)$ 
-\\
-\asp\asp
-%      Moment matching:
+     for $\unit$ in $\seq{1}{\Unit}$
+     \;
+%Moment matching:
      $\theta_{\unit,\time,\ninter,\rep,\np}= 
        \VtoTheta_{\unit}\big(
-            V^{\mathrm{meas}}_{\unit,\time,\ninter,\rep,\np} + V^{\mathrm{proc}}_{\unit,\time,\ninter,\rep}, 
+       V^{\mathrm{meas}}_{\unit,\time,\ninter,\rep,\np} +
+       V^{\mathrm{proc}}_{\unit,\time,\ninter,\rep}, 
        \, \mu^{\GP}_{\unit,\time,\ninter,\rep,\np}
        \big)$
-\\
-\asp\asp  % Guide function: 
-        $
-\guideFunc_{\time,\ninter,\rep,\np}=
+     for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$
+     \;
+% Guide function: 
+     $\guideFunc_{\time,\ninter,\rep,\np}=
        \prod_{\unit=1}^{\Unit}
@@
        \big(
-            \data{y}_{\unit,\time}\given \mu^{\GP}_{\unit,\time,\ninter,\rep,\np} \giventh \theta_{\unit,\time,\ninter,\rep,\np} 
+         \data{y}_{\unit,\time}\given \mu^{\GP}_{\unit,\time,\ninter,\rep,\np}
+	 \giventh \theta_{\unit,\time,\ninter,\rep,\np} 
          \big)$
-\\
-\asp\asp Guide weights:
+     for $\np$ in $\seq{1}{\Np}$
+     \;
+guide weights:
      $w^G_{\time,\ninter,\rep,\np}= \guideFunc^{}_{\time,\ninter,\rep,\np}
        \big/ \guideFunc^{\resample}_{\time,\ninter-1,\rep,\np}$     
-\\
-\asp\asp
-      Resampling:
-        $\prob\big[\resampleIndex({\rep,\np})=a \big] = w^G_{\time,\ninter,\rep,a}
+     for $\np$ in $\seq{1}{\Np}$
+     \;
+resampling:
+     $\prob\big[\resampleIndex({\rep,\np})=a \big] =
+       w^G_{\time,\ninter,\rep,a}
        \Big( \sum_{\altNp=1}^{\Np}w^G_{\time,\ninter,\rep,\altNp}\Big)^{-1}$
-\\
-\asp\asp
-        $\myvec{X}_{\time,\ninter,\rep,\np}^{\GR}=\myvec{X}_{\time,\ninter,\rep,\resampleIndex({\rep,\np})}^{\GP}\; \; $ and
-        $\; \guideFunc^{\resample}_{\time,\ninter,\rep,\np}= \guideFunc^{}_{\time,\ninter,\rep,\resampleIndex({\rep,\np})}\,$
-\\
-\asp
-End For
-\\
-\asp
-  Set $\myvec{X}^{\IF}_{\time,\rep}=\myvec{X}^{\GR}_{\time,\Ninter,\rep,1}$ 
-\\ 
-\asp Measurement weights:
+     for $\np$ in $\seq{1}{\Np}$       
+     \;
+     $\myvec{X}_{\time,\ninter,\rep,\np}^{\GR}=
+       \myvec{X}_{\time,\ninter,\rep,\resampleIndex({\rep,\np})}^{\GP}\; \; $
+     and
+     $\; \guideFunc^{\resample}_{\time,\ninter,\rep,\np}=
+       \guideFunc^{}_{\time,\ninter,\rep,\resampleIndex({\rep,\np})}\,$
+     for $\np$ in $\seq{1}{\Np}$       
+  } 
+  $\myvec{X}^{\IF}_{\time,\rep}=\myvec{X}^{\GR}_{\time,\Ninter,\rep,1}$
+  \;
+%  Measurement weights:
   $w^M_{\unit,\time,\rep,\npgir} = 
@@
     \big (\data{y}_{\unit,\time}\given X^{G}_{\unit,\time,\rep,\npgir} \big)$
-\\
-\asp % Prediction weights:
+  for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$
+  \;
+% Prediction weights:
   $w^{\LCP}_{\unit,\time,\rep,\npgir}= \displaystyle
@@
         w^M_{\altUnit,\time,\rep,\npgir}$
-\\
-End for
-\\
-$\displaystyle \MC{\loglik}_{\unit,\time}= 
+  for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$      	
+  \;
+}
+}
+\KwOut{
+$\displaystyle \MC{\loglik}=
+\sum_{\unit=1}^{\Unit} \sum_{\time=1}^{\Time}
 \log\Bigg(
@@
 $
-\vspace{1mm}
-\\
-\hline
-\end{tabular}
-\end{center}
-
-\clearpage
+%for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$
+}
+\end{algorithm}
 
@@
 Other guide functions could be developed and inserted into the {\ABFIR} algorithm, including other constructions considered by \citet{park20}.
-The number of particles for the guide simulations in {\ABFIR} does not necessarily have to be set equal to the number of intermediate proposals. 
-We make that simplification here since there is no compelling reason against it.
 
 One might wonder why it is appropriate to keep many particle representations at intermediate timesteps while resampling down to a single representative at each observation time.
-Part of the answer is that adaptive simulation can fail when one resamples down to a single particle too often (Sec.~\SuppSecAdaptedSimulation).
+An answer is that adaptive simulation can fail to track the observation sequence when one resamples down to a single particle too often (Sec.~\SuppSecAdaptedSimulation).
 
@@
 
-\section{Likelihood factorizations and their approximations}
-
-The approximation error for the algorithms we study here can be divided into two sources: a localization bias due to conditioning on a finite neighborhood, and Monte Carlo error. 
-The localization bias does not disappear in the limit as Monte Carlo effort increases.
-It does become small as the conditioning neighborhood increases, but the Monte Carlo effort grows exponentially in the size of this neighborhood.
-The class of problems for which these algorithms are useful are ones where a relatively small neighborhood is adequate. 
-Although the filtering inference is carried out using localization, the simulation of the process is carried out globally which avoids the introduction of additional boundary effects and ensures that the simulations comply with any constraint properties of the global process simulator.
-
-In this subsection, we consider the deterministic limit of each algorithm for infinite Monte Carlo effort.
-We explain why the {\UBF}, {\ABF} and {\ABFIR} algorithms approximately target the likelihood function, subject to suitable mixing behavior.
+%\section{Likelihood factorizations and their approximations}
+
+\subsection{{\ABFIR} theory}
+\label{sec:abf:theory}
+
+We  start by considering a deterministic limit for infinite Monte Carlo effort and explaining why the {\ABF} and {\ABFIR} algorithms approximately target the likelihood function, subject to suitable mixing behavior.
 Subsequently, we consider the scaling properties as Monte Carlo effort increases.
 We adopt a convention that densities involving $Y_{\unit\comma\time}$ are implicitly evaluated at the data, $\data{y}_{\unit\comma\time}$, and densities involving $X_{\unit\comma\time}$ are implicitly evaluated at $x_{\unit\comma\time}$ unless otherwise specified.
-We write $A^{+}_{\unit\comma\time}=A_{\unit\comma\time}\cup (\unit,\time)$ and $B^{+}_{\unit\comma\time}=B_{\unit\comma\time}\cup (\unit,\time)$.
+We write $A^{+}_{\unit\comma\time}=A_{\unit\comma\time}\cup (\unit,\time)$, matching the defintion $B^{+}_{\unit\comma\time}=B_{\unit\comma\time}\cup (\unit,\time)$.
 The essential ingredient in all the algorithms is a localization of the likelihood, which may be factorized sequentially as
@@
 \end{eqnarray}
-For 
-$\unitTimeSubset \subset \UnitSet\times 1{\mycolon}\Time$, 
-define 
-$\unitTimeSubset^{[m]}=\unitTimeSubset \cap \big(\UnitSet\times \{m\}\big)$ 
-and set
+Using the convention that an empty density $f_{Y_{\emptyset}}$ evaluates to 1, we define
 \begin{equation}
@@
 = 
-\prod_{m=1}^{\Time} f_{Y_{\unitTimeSubset^{[m]}}|\myvec{X}^{}_{m-1}} \big( \data{y}_{\unitTimeSubset^{[m]}} \given \myvec{X}^{}_{m-1} \big),
+\prod_{m=1}^{\Time} f_{Y_{\unitTimeSubset^{[m]}}|\myvec{X}^{}_{m-1}} \big( \data{y}_{\unitTimeSubset^{[m]}} \given \myvec{X}^{}_{m-1} \big).
 \end{equation}
-using the convention that an empty density $f_{Y_{\emptyset}}$ evaluates to 1.
-Denoting $\E_{g}$ for expectation for $(\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time})$ having density $g_{\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time}}$, we have
+Denoting $\E_{g}$ for expectation for $(\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time})$ having density $g_{\myvec{X}_{0:\Time},\myvec{X}^P_{1:\Time}}$, we have $ f_{Y_{A_{u,n}}} = \E_g \big[ \adapted^{}_{A^{}_{\unit\comma\time}}\big]$ and thus
 \begin{equation}
@@
 where $B^c_{\unit\comma\time}$ is the complement of $B_{\unit\comma\time}$ in $A_{\unit\comma\time}$.
-As long as $\adapted_{B^c_{\unit\comma\time}}$ is approximately independent of $X_{\unit\comma\time}$ under $g$, this term approximately cancels in the numerator and denominator of the right hand side of \myeqref{eq:AB:ratio2}.
-
-\section{{\UBF} theory}
-
-For each pair $(\Unit,\Time)$ we suppose there is a dataset $\data{\myvec{y}}_{1:\Time}$ and a model $f_{\myvec{X}_{0:\Time},\myvec{Y}_{1:\Time}}$.
-We are interested in results that hold for all $(\Unit,\Time)$, but we do not require that the models for $(\Unit_1,\Time_1)\neq(\Unit_2,\Time_2)$ are nested in any way. 
-The following conditions impose requirements that distant regions of space-time behave similarly and have only weak dependence.
-The conditions are written non-asymptotically in terms of $\epsilon>0$ and $Q>1$ which are used to bound the asymptotic bias and variance in Theorem~\ref{thm:tif}.
-Stronger bounds are obtained when the conditions hold for small $\eone$, $\etwo$ and $Q$.
-
-\Ai  %% Assumption 1  Assumption~\ref{A1}
-
-\Aii  %% Assumption 2  Assumption~\ref{A1b}
-
-\Aiii   %% Assumption 3  Assumption~\ref{A2}
-
-\Aiv    %% Assumption 4  Assumption~\ref{A:unconditional:mix}
-
-The two mixing conditions in Assumptions~\ref{A1} and~\ref{A:unconditional:mix} are subtly different. 
-Assumption~\ref{A1} is a conditional mixing property, dependent on the data, whereas~\ref{A:unconditional:mix} asserts a form of unconditional mixing.
-Although both capture a similar concept of weak coupling, conditional and unconditional mixing properties do not readily imply one another.
-Assumption~\ref{A2} is a compactness condition of a type that has proved useful in the theory of particle filters despite the rarity of its holding exactly.
-Theorem~\ref{thm:tif} shows that these conditions let {\UBF} compute the likelihood with a Monte Carlo variance of order $\Unit\Time\Rep^{-1}$ with a bias of order $\Unit\Time\epsilon$.
-
-\TheoremI %% Theorem~\ref{thm:tif}
-
-\begin{proof}
-A complete proof is given in Sec.~\SuppSecThmI. 
-Briefly, the assumptions imply a multivariate central limit theorem for $\{\MC{\loglik}_{\unit\comma\time}, (\unit,\time)\in \UnitSet{\times}\ObsTimeSet\}$ as $\Rep\to\infty$.
-The limiting variances and covariances are uniformly bounded, using Assumptions~\ref{A1b} and~\ref{A2}. 
-Assumption~\ref{A1} provides a uniform bound on the discrepancy between ${\loglik}_{\unit\comma\time}$ and mean of the Gaussian limit.
-This is enough to derive \myeqref{th1:lik:bound}.
-Assumption~\ref{A:unconditional:mix} gives a stronger bound on covariances between sufficiently distant units, leading to \myeqref{th1:lik:bound2}.
-\end{proof}
-
-\section{{\ABFIR} theory}
-
-Since {\ABF} is {\ABFIR} with $\Ninter=1$, we focus attention on {\ABFIR}.
-At a conceptual level, the localized likelihood estimate in {\ABFIR} has the same structure as its {\UBF} counterpart.
-However, {\ABFIR} additionally requires the capability to satisfactorily implement adapted simulation.
+Under our assumptions, the term corresponding to $\adapted_{B^c_{\unit\comma\time}}$ approximately cancels in the numerator and denominator of the right hand side of \myeqref{eq:AB:ratio2}.
+
+The localized likelihood estimate in  {\ABF} and {\ABFIR} has similar structure to {\UBF}.
+However,  {\ABF} and {\ABFIR} additionally require the capability to satisfactorily implement adapted simulation.
 Adapted simulation is a local calculation, making it an easier task than the global operation of filtering.
-Nevertheless, adapted simulation via importance sampling is vulnerable to COD for sufficiently large values of $\Unit$.
-For a continuous time model, the use of $\Ninter>1$ is motivated by a result that guided intermediate resampling can reduce, or even remove, the COD in the context of a particle filtering algorithm \cite{park20}.
-Assumptions~\ref{B1}--\ref{B:unconditional:mix} are analogous to ~\ref{A1}--\ref{A:unconditional:mix} and are non-asymptotic assumptions involving $\ethree>0$, $\efourA>0$ and $Q>1$ which are required to hold uniformly over space and time.
-Assumption~\ref{B:temporal:mix}--\ref{B:XG_XA_ind} control the Monte~Carlo error arising from adapted simulation.
+Nevertheless, adapted simulation via importance sampling (as carried out by ABF) is vulnerable to COD.
+For a continuous time model, the use of intermediate resampling  in ABF-IR is motivated by a result that this can reduce the COD, or avoid it entirely for an ideal guide function \citep{park20}.
+Without intermediate resampling even an ideal proposal distribution does not avoid COD for a particle filter \citep{snyder15}.
+Assumptions~\ref{B1}--\ref{B:unconditional:mix} below are analogous to ~\ref{A1}--\ref{A:unconditional:mix} and are non-asymptotic assumptions involving $\ethree>0$, $\efourA>0$ and $Q>1$ which are required to hold uniformly over space and [time.
+Assumptions~\ref{B:temporal:mix}--\ref{B:XG_XA_ind} control the Monte~Carlo error arising from adapted simulation.
 ~\ref{B:temporal:mix} is a stability property which asserts that the effect of the latent process on the future of the adapted process decays over time. 
 Assumption~\ref{B:girf} is a non-asymptotic bound on Monte~Carlo error for a single step of adapted simulation.
-The scaling of the constant $\BvConstant$ with  $\Unit$,  $\Time$ and $\Ninter$ in  Assumption~\ref{B:girf} has been studied by \citet{park20}, where it was established that setting $\Ninter=\Unit$ can lead to $\BvConstant$ being constant, when using an ideal guide function, or slowly growing with $\Unit$ otherwise.
-The $\efive^{-3}$ error rate in Assumption~\ref{B:girf} follows from balancing the two sources of error defined in the statement of Theorem~2 of \citet{park20}.
+The scaling of the constant $\BvConstant$ with  $\Unit$,  $\Time$ and $\Ninter$ in  Assumption~\ref{B:girf} has been studied by \citet{park20}, where it was established that setting $\Ninter=\Unit$ can lead to $\BvConstant$ being polynomial in $\Unit$ and $\Time$ when using an ideal guide function.
+This property is critical to enable ABF-IR to avoid COD. Since ABF has $\Ninter=1$ it suffers from COD, albeit at an empirically slower rate than PF.
+The $\efive^{-3}$ error rate in Assumption~\ref{B:girf} follows from balancing the two sources of error defined in the statement of Theorem~2 of \citet{park20} (details are provided in Sec.~{\SuppSecAssumption}).
 Assumption~\ref{B:XG_XA_ind} can be guaranteed by the construction of the algorithm, if independently generated Monte~Carlo random variables are used for building the guide function and the one-step prediction particles.
@@
 
-The theoretical results foreshadow our empirical observations that the relative performance of {\UBF}, {\ABF} and {\ABFIR} is situation-dependent.
+The theoretical results foreshadow our empirical observations (Sec.~\ref{sec:examples}) that the relative performance of {\UBF}, {\ABF} and {\ABFIR} is situation-dependent.
 Assumption~\ref{A:unconditional:mix} is a mixing assumption for the unconditional latent process, whereas Assumption~\ref{B:unconditional:mix} replaces this with a mixing assumption for the adapted process conditional on the data.
@@
 \section{Examples}
-
-Likelihood evaluation via filtering does not by itself enable parameter estimation for POMP models, however it provides a foundation for Bayesian \citep{andrieu10} and likelihood-based \citep{ionides15} inference.
-We therefore investigate likelihood evaluation with the expectation that improved methods for likelihood evaluation will lead to improved statistical methodology. 
-Monte Carlo methods for computing the log likelihood generally suffer from both bias and variance, both of which can be considerable for large datasets and complex models.
-Appropriate inference methodology, such as Monte Carlo adjusted profile (MCAP) confidence intervals, can accommodate substantial Monte Carlo variance so long as the bias is slowly varying across the statistically plausible region of the parameter space \citep{ionides17}.
-
-We compare the performance of the three bagged filters ({\UBF}, {\ABF} and {\ABFIR}) against each other and against alternative approaches.
-Ensemble Kalman filter (EnKF) methods propagate the ensemble members by simulation from the dynamic model and then update the ensemble to assimilate observations using a Gaussian-inspired rule \citep{evensen96,lei10}.
-The block particle filter (BPF) \citep{rebeschini15,ng02} partitions the latent space and combines independently drawn components from each partition.
+\label{sec:examples}
+
+We compare the performance of the three bagged filters ({\UBF}, {\ABF} and {\ABFIR}) against each other and against alternative plug-and-play approaches.
+The plug-and-play property facilitates numerical implementation for general classes of models, and all the algorithms and models under consideration are implemented in the R packages \code{pomp} \citep{king16} and \code{spatPomp} \citep{asfaw21arxiv}.
+Ensemble Kalman filter (EnKF) methods propagate the ensemble members by simulation from the dynamic model and then update the ensemble to assimilate observations using a Gaussian-inspired rule \citep{evensen09book,lei10}.
+The block particle filter \citep[BPF,][]{rebeschini15,ng02} partitions the latent space and combines independently drawn components from each partition.
 BPF overcomes COD under weak coupling assumptions \citep{rebeschini15}.
 Unlike these two methods, our bagged filters modify particles only according to the latent dynamics.
-Thus, our methods satisfy any conservation laws, continuity or smoothness that arise when simulating from the dynamic model.
-
-We first consider a spatiotemporal Gaussian process for which the exact likelihood is available via a Kalman filter. 
+Thus, our methods respect conservation laws and continuity or smoothness conditions obeyed by the dynamic model.
+We also compare with a guided intermediate resampling filter \citep[GIRF,][]{park19}, one of many variants of the particle filter designed to scale to larger numbers of units than are possible with a basic particle filter.
+
+First, in Sec.~\ref{sec:bm}, we consider a spatiotemporal Gaussian process for which the exact likelihood is available via a Kalman filter. 
 We see in Fig.~\ref{fig:bm_alt_plot} that {\ABFIR} can have a considerable advantages over {\UBF} and {\ABF} for problems with an intermediate level of coupling.
-We then develop a model for measles transmission within and between cities.
-The model is weakly coupled, leading to successful performance for all three bagged filters.
+Then, in Sec.~\ref{sec:measles}, we develop a model for measles transmission within and between cities.
+The measles model is weakly coupled, leading to successful performance for all three bagged filters.
 This class of metapopulation models was the primary motivation for the development of these methodologies.
-However, for a highly coupled system of moderate dimensions, a global particle filter such as GIRF may outperform our bagged methods;
-an example is the Lorenz-96 model demonstrated in Sec.~\SuppSecLorenz.
+In Sec.~\ref{sec:profile} we demonstrate an extension from likelihood evaluation to likelihood maximization for the measles model.
+Additionally, in  Sec.~\SuppSecLorenz, we compare performance on the Lorenz-96 model, a highly coupled system used to test inference methods for geophysical applications.
 
 \subsection{Correlated Brownian motion}
+\label{sec:bm}
 
@@
     system.time(
-      girf(bm_list[[job$U_id]], method='adams',
+      girf(bm_list[[job$U_id]],
         Np=bm_girf_Np,
         Ninter = job$U,
+        kind="moment",
+#        kind="bootstrap",
         lookahead = bm_girf_lookahead,
@@
 bm_jobs$girf_time <- vapply(bm_girf_list,function(x) x$time,numeric(1))
-@
-
-<<bm_apf,cache=F,echo=F,eval=F>>=
-bm_apf <- stew(file=paste0(bm_files_dir,"bm_apf.rda"),seed=5981724,{
-  foreach(job=iter(bm_jobs,"row")) %dopar% {
-    system.time(
-      girf(bm_list[[job$U_id]], method='adams',
-        Np=bm_apf_Np,
-        Ninter = 1,
-        lookahead = 2,
-        Nguide = bm_apf_nguide,
-        tol = bm_tol 
-      )-> bm_apf_out 
-    ) -> bm_apf_time
-#    uncomment for debugging 
-#    list(logLik=logLik(bm_apf_out),time=bm_apf_time["elapsed"],bm_apf_out)
-    list(logLik=logLik(bm_apf_out),time=bm_apf_time["elapsed"])
-  } -> bm_apf_list
-})
-
-bm_jobs$apf_logLik <- vapply(bm_apf_list,function(x)x$logLik,numeric(1))
-bm_jobs$apf_time <- vapply(bm_apf_list,function(x) x$time,numeric(1))
-
 @
@@
     system.time(
-      abfir(bm_list[[job$U_id]], method='adams',
+      abfir(bm_list[[job$U_id]], 
         Nrep = as.integer(bm_abfir_Nrep),
@@
     system.time(
-      pfilter(bm_list[[job$U_id]],Np=bm_pfilter_Np, tol=bm_tol) -> bm_pfilter_out
+      pfilter(bm_list[[job$U_id]],Np=bm_pfilter_Np) -> bm_pfilter_out
     ) -> bm_pfilter_time
@@
 
-if(0){
-  ggplot(bm_results,mapping = aes(x = U, y = logLik_per_unit)) +
-  geom_point(aes(color = Method)) +
-  scale_y_continuous(limits=c(-108,NA)) +
-# scale_y_continuous() +
-  ylab("log likelihood per unit")
-}
-
 bm_max <- max(bm_results$logLik_per_obs)
@@
   theme(legend.key.width = unit(1,"cm"))+
+  theme(axis.text.x = element_text(size = 12),
+         axis.text.y = element_text(size = 12),
+         axis.title.x = element_text(size = 14),
+         axis.title.y = element_text(size = 14),
+         panel.grid.major = element_blank(),
+         panel.grid.minor = element_blank(),
+         panel.background = element_blank(),
+         panel.border = element_rect(colour = "black", fill=NA, size=0.8),
+         axis.line = element_line(colour = "black"))+
   ylab("log likelihood per unit per time")
@@
 Thus, GIRF here represents a state-of-the-art auxiliary particle filter that targets the complete joint filter density for all units.
-We use the general-purpose, plug-and-play implementation of GIRF provided by the \texttt{spatPomp} R package \citep{asfaw20github};
- this implementation of GIRF is thus not tailored in any way to exploit special features of this toy model.
-PF works well for small values of $\Unit$ in Fig.~\ref{fig:bm_alt_plot}, but rapidly starts struggling as $\Unit$ increases.
+We use the general-purpose, plug-and-play implementation of GIRF provided by the \texttt{spatPomp} R package \citep{asfaw21github}; for a Gaussian model, one can calculate an ideal guide function for GIRF but that was not used.
+PF works well for small values of $\Unit$ in Fig.~\ref{fig:bm_alt_plot} and rapidly starts struggling as $\Unit$ increases.
 GIRF behaves comparably to PF for small $\Unit$ but its performance is maintained for larger $\Unit$.
@@
 \subsection{Spatiotemporal measles epidemics}
+\label{sec:measles}
 
@@
 The population dynamics are written in terms of counting processes $N_{\bullet\bullet,\unit}(t)$ enumerating cumulative transitions in city $\unit$, up to time $t$, between compartments identified by the subscripts.
+We model the $\Unit$ largest cities in the UK, ordered in decreasing size so that $\unit=1$ corresponds to London.
+We vary $\Unit$ to test methodologies on a hierarchy of filtering challenges.
 Our model is described by the following system of stochastic differential equations, for $\unit=1,\dots, \Unit$,
@@
 Here, we model $v_{\unit\altUnit}$ as fixed through time and symmetric between any two arbitrary cities, though a natural extension would allow for temporal variation and asymmetric movement between two cities.
-The transition processes $N_{EI,\unit}(t)$, $N_{EI,\unit}(t)$ and $N_{\bullet D,\unit}(t)$ are modeled as conditional Poisson processes with per-capita rates $\mu_{EI}$, $\mu_{IR}$ and $\mu_{\bullet D}$ respectively, and we fix $\mu_{\bullet D}=50 \mbox{ year}^{-1}$.
+The transition processes $N_{EI,\unit}(t)$, $N_{IR,\unit}(t)$ and $N_{\bullet D,\unit}(t)$ are modeled as conditional Poisson processes with per-capita rates $\mu_{EI}$, $\mu_{IR}$ and $\mu_{\bullet D}$ respectively, and we fix $\mu_{\bullet D}=50 \mbox{ year}^{-1}$.
 The birth process $N_{BS,\unit}(t)$ is an inhomogeneous Poisson processes with rate $\mu_{BS,\unit}(t)$, given by interpolated census data.
@@
     system.time(
-      girf(mscale_list[[job$U_id]], method='adams',
+      girf(mscale_list[[job$U_id]],
+        kind="moment",
         Np=mscale_girf_Np,
@@
 mscale_jobs$girf_time <- vapply(mscale_girf_list,function(x) x$time,numeric(1))
-@
-
-<<mscale_bootgirf,cache=F,echo=F>>=
-mscale_bootgirf <- stew(file=paste0(mscale_files_dir,"mscale_bootgirf.rda"),seed=5981724,{
-### begin test
-#mt <- mscale_list[[2]]
-#bootgirf2(mt, method='adams',
-#        Np=100,
-#        Ninter = 4,
-#        lookahead = 2,
-#        Nguide = 10) -> bt
-### end test
-  foreach(job=iter(mscale_jobs,"row")) %dopar% {
-    system.time(
-# job <- mscale_jobs[1,] # for debugging
-      bootgirf2(mscale_list[[job$U_id]], method='adams',
-        Np=mscale_bootgirf_Np,
-        Ninter = job$U,
-        lookahead = mscale_bootgirf_lookahead,
-        Nguide = mscale_bootgirf_nguide,
-        tol = mscale_tol) -> mscale_bootgirf_out
-   ) -> mscale_bootgirf_time   
-#    uncomment for debugging 
-#    list(logLik=logLik(mscale_bootgirf_out),time=mscale_bootgirf_time["elapsed"],mscale_bootgirf_out)
-    list(logLik=logLik(mscale_bootgirf_out),time=mscale_bootgirf_time["elapsed"])
-  } -> mscale_bootgirf_list
-})
-mscale_jobs$bootgirf_logLik <- vapply(mscale_bootgirf_list,function(x)x$logLik,numeric(1))
-mscale_jobs$bootgirf_time <- vapply(mscale_bootgirf_list,function(x) x$time,numeric(1))
 @
@@
     system.time(
-      abfir(mscale_list[[job$U_id]], method='adams',
+      abfir(mscale_list[[job$U_id]], 
         Nrep = as.integer(mscale_abfir_Nrep),
@@
 
-<<mscale_loglik_plot, echo=F,  fig.height=4, fig.width=5.5, out.width="4.5in", fig.cap = paste('log likelihood estimates for simulated data from the measles model of various dimensions.
+<<mscale_loglik_plot, echo=F,  fig.height=3, fig.width=4.5, out.width="4in", fig.cap = paste('log likelihood estimates for simulated data from the measles model of various dimensions.
 {\\UBF}, {\\ABF} and {\\ABFIR} are compared with a guided intermediate resampling filter (GIRF), a standard particle filter (PF), a block particle filter (BPF) and an ensemble Kalman filter (EnKF).')>>=
@@
 
-#mscale_methods <- c("ABF", "ABF-IR","UBF","GIRF","PF","BPF","EnKF")
-mscale_methods <- c("ABF", "ABF-IR","UBF","GIRF","PF","BPF","EnKF","bootgirf")
+mscale_methods <- c("ABF", "ABF-IR","UBF","GIRF","PF","BPF","EnKF")
+# mscale_methods <- c("ABF", "ABF-IR","UBF","GIRF","PF","BPF","EnKF","bootgirf")
 mscale_results <- data.frame(
@@
     mscale_jobs$bpf_logLik,
-    mscale_jobs$enkf_logLik,
-    mscale_jobs$bootgirf_logLik
+    mscale_jobs$enkf_logLik
+#    mscale_jobs$bootgirf_logLik
   ),
@@
 
+mscale_linetype <- c(1,1,1,1,2,2,2,3)
+mscale_shape <- c(1,2,4,5,1,2,4,3)
+#mscale_color <- cbPalette[c(2:8,1)]
+mscale_color <- cbPalette[2:8]
+# mscale_color <- c(1,2,3,4,5,6,8)
+
 ggplot(mscale_results,mapping = aes(x = U, y = logLik_per_obs, group=Method,color=Method,linetype=Method,shape=Method)) +
-  scale_linetype_manual(values=c(1,1,1,1,2,2,2,3)) +
-  scale_shape_manual(values=c(1,2,4,5,1,2,4,3)) +
-#  scale_color_manual(values=c(1,2,3,4,5,6,8))+
-  scale_color_manual(values=cbPalette[c(2:8,1)])+
+  scale_linetype_manual(values=mscale_linetype) +
+  scale_shape_manual(values=mscale_shape) +
+  scale_color_manual(values=mscale_color)+
   geom_point() +
   stat_summary(fun=mean, geom="line") +
-  coord_cartesian(ylim=c(mscale_max-2,mscale_max))+
+  coord_cartesian(ylim=c(mscale_max-2.5,mscale_max))+
   theme(legend.key.width = unit(1,"cm"))+
-  ylab("log likelihood per observation")
+  theme(axis.text.x = element_text(size = 12),
+         axis.text.y = element_text(size = 12),
+         axis.title.x = element_text(size = 14),
+         axis.title.y = element_text(size = 14),
+         panel.grid.major = element_blank(),
+         panel.grid.minor = element_blank(),
+         panel.background = element_blank(),
+         panel.border = element_rect(colour = "black", fill=NA, size=0.8),
+         axis.line = element_line(colour = "black"))+
+  ylab("log likelihood per unit per time")
 
@@
 Fig.~\ref{fig:measles_image_plot} shows a simulation from our model, showing that trajectories from this model can capture some features of the system that have been hard to understand: how can it be that disease transmission dynamics between locations have important levels of interaction yet are not locked in synchrony \citep{becker20}?
-Here, we are developing statistical tools rather than engaging directly in the scientific debate.
+Here, we are testing statistical tools rather than engaging directly in the scientific debate so we test methods on the simulated data.
 
@@
 We suspect that the limitations of EnKF on this problem are due to the nonlinearity, non-Gaussianity, and discreteness of fadeout and reintroduction dynamics.
-All the algorithms have various tuning parameters that could influence the results. 
-Some investigations of alternatives are presented in the Sec.~\SuppSecMeasles.
+Thus, EnKF is relatively effective with small ensemble size but soon reaches the limit of its capabilities (Sec.~{\SuppSecFixedU}).
+By contrast, the bagged filters and block particle filter perform substantially better than EnKF for larger ensemble size (Sec.~{\SuppSecFixedU}).
+All the algorithms have various other tuning parameters that could influence the results.
+Some investigations of alternatives are presented in Secs.~{\SuppSecMeasles},~{\SuppSecMeaslesNbhd} and~{\SuppSecIvsJ}.
 Generalizable conclusions are hard to infer from numerical comparisons of complex algorithms on complex models.
@@
     slice_abf_Np_per_replicate <- 100
-    slice_bpf_units_per_block <- 2
+    slice_bpf_units_per_block <- 1
     slice_bpf_Np <- 40000
@@
     system.time(
-      abfir(slice_sim,params=theta,  method='adams',
+      abfir(slice_sim,params=theta,
         Nrep = slice_abfir_Nrep,
@@
 
-#pdf(file="tmp.pdf",height=7,width=3.5)
-
 par(omi=c(0.5,0.1,0.1,0.1))
@@
 This slice varies $\gravity$ while fixing the other parameters at the values used for the simulation.
-Scientifically, this gives us an upper bound on the identifiability of $\gravity$ from such data, since the likelihood slice provides statistically efficient inference when all other parameters are known.
-In this case, the Monte Carlo error is comparable to the change in the log likelihood over a wide range of values of $\gravity$. 
 Fig.~\ref{fig:slice_combined_plot}(B) shows a similar plot calculated using BPF with comparable computational effort.
@@
 By contrast, Fig.~\ref{fig:slice_combined_plot}(C) shows that EnKF has substantial bias in estimating $\gravity$, as well as considerably lower likelihood. 
-
-%Fig~\ref{fig:slice_abf_plot} took $\Sexpr{myround(mean(slice_abf_results[,"time.elapsed"]*gl_cores/60),1)}$ core minutes to evaluate each point using a parallelized implementation on a computing node with $\Sexpr{gl_cores}$ cores, corresponding to a run time of  $\Sexpr{myround(mean(slice_abf_results[,"time.elapsed"]/60),1)}$ per likelihood evaluation.
-%By contrast, Fig~\ref{fig:slice_bpf_plot} took $\Sexpr{myround(mean(slice_bpf_results[,"time.elapsed"]/60),1)}$ minutes per likelihood evaluation on a single core.
-%
-%, equivalent in effort to $\Sexpr{myround(mean(slice_bpf_results[,"time.elapsed"]/60/gl_cores),1)}$ minutes for a parallel implementation.
-%
-%and Fig~\ref{fig:slice_ubf_plot} took $\Sexpr{myround(mean(slice_ubf_results[,"time.elapsed"]/60),1)}$ minutes,
+Likelihood slices have less inferential value than likelihood profiles, but provide a computationally and conceptually simpler setting that can be insightful.
+Scientifically, the slices in Fig.~\ref{fig:slice_combined_plot} give an upper bound on the identifiability of $\gravity$ from such data, since the likelihood slice provides statistically efficient inference when all other parameters are known.
 
@@
 
+\subsection{Likelihood maximization and profile likelihood}
+\label{sec:profile}
+<<profile_plot, echo=F, fig.height=3, fig.width=5, out.width="4in", fig.cap = 'An iterated bagged filter used to maximize the likelihood, compute a profile likelihood, and hence construct a confidence interval.
+The profiling is carried out over the coupling parameter, $G$.'>>=
+
+# see ./measles_iubf/README for how these .rda files were generated
+iubf_dir <- './measles_iubf/'
+rdanames <- system(paste('ls ', iubf_dir, 'generated/abf_outs_3/*.rda', sep = ''), intern = TRUE) # run level 3
+g_prof <- data.frame()
+for(fn in rdanames){
+  load(fn)
+  g_prof_row <- c('g' = final_params['g'],
+                  'loglik' = ll[1],
+                  'loglik_se' = ll[2])
+  g_prof <- rbind(g_prof, g_prof_row)
+}
+colnames(g_prof) <- c('g', 'loglik', 'loglik_se')
+
+plot.profile <- function(mcap1,ylab,xline=2.5,yline=2.5,xlab="",quadratic=FALSE,...){
+   if(missing(ylab)) ylab <- "profile log likelihood"
+   ggplot() + geom_point(aes(x = mcap1$parameter, y = mcap1$lp), shape = 1) +
+      geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$smoothed),
+                color = "red") +
+      {if(quadratic) geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$quadratic),
+                               color = "blue",
+                               lwd = 1.25)} +
+      labs(x = xlab, y = ylab) +
+      geom_vline(xintercept = mcap1$ci, color = "red") +
+      geom_hline(yintercept = max(mcap1$fit$smoothed, na.rm = T) - mcap1$delta, color = "red") +
+      theme(panel.border = element_rect(colour = "black", fill=NA))
+}
+g_prof %>% dplyr::group_by(g) %>% dplyr::filter(loglik==min(loglik)) -> mins
+g_prof[which(!(g_prof$loglik %in% mins$loglik)),] -> top_g_prof
+
+g_prof_mcap <- spatPomp::mcap(
+   lp=top_g_prof[,"loglik"],
+   parameter=top_g_prof[,"g"],
+   lambda = 0.90
+)
+plot.profile(g_prof_mcap,xlab='G', quadratic = FALSE) +
+   geom_vline(xintercept=400,color="blue",linetype="dashed") +
+   theme(axis.text.x = element_text(size = 12),
+         axis.text.y = element_text(size = 12),
+         axis.title.x = element_text(size = 14),
+         axis.title.y = element_text(size = 14),
+         panel.grid.major = element_blank(),
+         panel.grid.minor = element_blank(),
+         panel.background = element_blank(),
+         panel.border = element_rect(colour = "black", fill=NA, size=0.8),
+         axis.line = element_line(colour = "black"))
+
+@
+
+Likelihood evaluation via filtering does not by itself enable parameter estimation for POMP models, however it provides a foundation for Bayesian and likelihood-based inference.
+In particular, filtering algorithms can be modified to carry out likelihood maximization by stochastically perturbing parameters in a sequence of filtering operations with decreasing perturbation variance \citep{ionides15}.
+We demonstrate this for the measles model in Fig.~\ref{fig:profile_plot} using an iterated bagged filter algorithm which is fully described in Sec.~{\SuppSecParameterEst}.
+
+Monte Carlo methods for computing and maximizing the log likelihood suffer from  bias and variance, both of which can be considerable for large datasets and complex models.
+Appropriate inference methodology, such as Monte Carlo adjusted profile (MCAP) confidence intervals, can accommodate substantial Monte Carlo variance so long as the bias is slowly varying across the statistically plausible region of the parameter space \citep{ionides17,ning21}.
+Fig.~\ref{fig:profile_plot} constructs an MCAP 95\% confidence interval for the coupling parameter, $G$, using an iterated unadapted bagged filter to maximize over the parameters, $a$, $\bar\beta$, $\sigma_{SE}$, $\psi$, $\mu_{EI}$ and $\mu_{IR}$.
+This simulation study, carried out with $\Unit=20$ and $\Time=208$, shows that $G$ is identifiable via likelihood-based inference in the absence of assumptions about these parameters.
+
+The likelihood estimate provided by bagged filters could be viewed as a composite likelihood \citep{varin11} rather than an approximation to the likelihood.
+However, in situations where the likelihood approximation is found to be adequate, it is convenient to take advantage of the tools of likelihood-based inference.
+
 \section{Discussion}
-
-The pseudocode presented for the bagged filters describes how the outputs are calculated given the inputs, but does not prescribe details of how these quantities are calculated. 
-There is scope for implementations to trade off memory, computation and communication by making differing decisions on how the various loops defined in the pseudocode are coded, such as decisions on memory over-writing and parallelization.
-This article focuses on  the logical structure of the algorithms, leaving room for future research on implementation-specific considerations, though some supplementary discussion of memory-efficient implementation is given in Sec.~\SuppSecMemoryEfficient.
-
-This article has not exhausted the possibilities for bagging different PF algorithms.
-Bagging a standard PF is one possibility, but the algorithm becomes rapidly degenerate due to COD.
-Our approach has been to bag PF variants that are already degenerate, so there is less scope for COD to damage them.
-Alternatively, one could bag PF variants that avoid the COD by using local weights themselves.
-
-We have focused on likelihood evaluation, a critical quantity for testing scientific hypotheses about model structure and the values of unknown parameters.
-However, our bagged filters also provide localized solutions to the filtering problem, using the local weights to combine estimates across replicates.
-A solution to the filtering problem, in turn, provides a foundation for forecasting.
-Effective filtering and likelihood evaluation can enable likelihood maximization via iterated filtering algorithms \citep{ionides11,ionides15}.
-
-Plug-and-play inference based on sequential Monte Carlo likelihood evaluation has proved successful for investigating highly nonlinear partially observed dynamic systems of low dimension arising in analysis of epidemiological and ecological time series data \citep{breto18statSci,pons-salort18,decelles18,marino18}.
-A benefit of the plug-and-play property is that it facilitates development of broadly applicable software, which in turn promotes creative scientific model development \citep{breto09,he10}.
-
-Geophysical data assimilation has similarities to inference on spatiotemporal biological systems.
+\label{sec:discussion}
+
+The nested loops used in the pseudocode for the bagged filters can be computed in various different orders to give mathematically equivalent results.
+There is scope for implementations to trade off memory, computation and communication by varying decisions on how the loops defined in the pseudocode are coded, including decisions on memory over-writing and parallelization.
+This article focuses on  the properties of the quantities calculated by the algorithms, leaving room for future research on implementation-specific considerations, though some supplementary discussion of memory-efficient implementation is given in Sec.~\SuppSecMemoryEfficient.
+
+Plug-and-play inference based on sequential Monte Carlo likelihood evaluation has proved successful for investigating highly nonlinear partially observed dynamic systems of low dimension arising in analysis of epidemiological and ecological population dynamics \citep{breto18statSci,pons-salort18,decelles18,marino18}.
+This article develops a methodological extension motivated by the analysis of interacting biological populations.
+Similar challenges related to nonlinear non-Gaussian dynamic models arise in geophysical modeling.
 Relative to biological systems, geophysical applications are characterized by a greater number of spatial locations, better mathematical understanding of the underlying processes, and lower stochasticity.
-From this literature, the locally weighted particle filter of \citep{poterjoy16} is perhaps closest to our approach, but those local weights are used to construct a localized Kalman gain which is motivated by a Gaussian approximation.
-
-The EnKF algorithm arose originally via geophysical research \citep{evensen96} and has since become used more widely for inference on SpatPOMP models \citep{katzfuss19,lei10}.
+From this literature, the locally weighted particle filter of \citet{poterjoy16,poterjoy19} is perhaps closest to our approach, but the local weights of \citet{poterjoy16,poterjoy19} are used to construct a localized Kalman gain which is motivated by a Gaussian approximation comparable to EnKF.
+EnKF arose originally via geophysical research \citep{evensen94} and has since become used more widely for inference on SpatPOMP models \citep{katzfuss19,li20}.
 However, EnKF can fail entirely even on simple POMP models if the structure is sufficiently non-Gaussian.
@@
 
-Latent state trajectories constructed in our BF algorithms are all generated from the model simulator, appropriately reweighted and resampled, and so are necessarily valid sample paths of the model.
+Latent state trajectories constructed in our BF algorithms are all generated from the model simulator, appropriately reweighted and resampled, and so they are necessarily valid sample paths of the model.
 For example, spatial smoothness properties of the model through space, or conservation properties where some function of the system remains unchanged through time, are maintained in the BF trajectories.
-This is not generally true for the block particle filter (since resampling blocks can lead to violations at block boundaries) or for EnKF (since the filter procedure perturbs particles using a linear update rule that cannot respect nonlinear relationships).
+This is not true for the block particle filter, due to the indepdent resampling of the blocks (see Sec.~\SuppSecConstraint).
+EnKF preserves linear constraints, since the filter procedure perturbs particles using a linear update rule, but cannot respect nonlinear relationships.
 The practical importance of smoothness and conservation considerations will vary with the system under investigation, but this property of BF gives the scientific investigator one less thing to worry about.
@@
 The algorithms {\UBF}, {\ABF}, {\ABFIR}, GIRF, PF, BPF, and EnKF compared in this article all enjoy the plug-and-play property, facilitating their implementations in general-purpose software.
-The numerical results for this paper use the \code{abf}, \code{abfir}, \code{girf}, \code{pfilter}, \code{bpfilter} and \code{enkf} functions via the open-source R package \code{spatPomp} \citep{asfaw20github} that provides a spatiotemporal extension of the R package \code{pomp} \citep{king16}.
+The numerical results for this paper use the \code{abf}, \code{abfir}, \code{girf}, \code{pfilter}, \code{bpfilter} and \code{enkf} functions via the open-source R package \code{spatPomp} \citep{asfaw21arxiv} that provides a spatiotemporal extension of the R package \code{pomp} \citep{king16}.
 {\UBF} was implemented using \code{abf} with $\Np=1$ particles per replicate.
-The source code for this paper will be contributed to an open-source scientific archive upon acceptance for publication.
+The source code for this article is available as supplementary material.
 

### 2021-09-02 si/si.Rnw
code to make unformatted accepted version

--- 2020-11-28 si.Rnw
@@
 #run_level <- "jasa-sept-2020"
-run_level <- 2
+ run_level <- "manual"
+#run_level <- 2
+
+if(run_level=="manual"){
+  abfNbhd_run_level <- 3
+  lz_run_level <- 5
+  ij_run_level <- 2
+  bm_run_level <- 4
+  mscale_run_level <- 5
+  slice_run_level <- 3
+  fixedU_run_level <- 4
+  vsenkf_run_level <- 2
+}
+
+if(is.numeric(run_level)){
   abfNbhd_run_level <- run_level
   lz_run_level <- run_level
+  ij_run_level <- run_level
   bm_run_level <- run_level
@@
   slice_run_level <- run_level
+  fixedU_run_level <- run_level
+  vsenkf_run_level <- run_level
+
+}
 
@@
   lz_run_level <- 5
+  ij_run_level <- NA
   bm_run_level <- 4
@@
 }
-
 @
@@
 
-\input{header-ms.tex}
-\input{theorems.tex}
+\input{../header-ms.tex}
+\input{../theorems.tex}
 
@@
 \renewcommand\thelemma{S\arabic{lemma}}
+\renewcommand\thealgocf{S\arabic{algocf}}
 
@@
 
-\begin{center}
-\noindent\begin{tabular}{l}
-\hline
-{\bf Algorithm~G-{\UBF} (Generalized unadapted bagged filter).}\rule[-1.5mm]{0mm}{6mm}\\
-\hline
-{\bf input:}\rule[-1.5mm]{0mm}{6mm} \\
-Simulator for $f_{{X}_{\time}|{X}_{\time-1}}({x}_{\time}|{x}_{\time-1})$\\
-Evaluator for $f_{Y_{\unit,\time}|{X}_{\time}}(\data{y}_{\unit,\time}\given {x}_{\time})$\\
-Number of bootstrap filters, $\Rep$\\
-Neighborhood structure, $B_{\unit,\time}$, for $\unit\in 1\mycolon\Unit$ and $\time\in 1\mycolon\Time$\\
-Data, $\data{y}_{\unit,\time}$ for $\unit\in 1\mycolon\Unit$ and $\time\in 1\mycolon\Time$ \\
-{\bf output:}\rule[-1.5mm]{0mm}{5mm} \\
-Log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit,\time}$  \rule[-2mm]{0mm}{3mm}
-\\
-\hline
-For $\rep$ in $1\mycolon\Rep$ \rule[-1.5mm]{0mm}{6mm}
-\\
-\asp simulate ${X}_{\time,\rep}^{\simulation}$ from the dynamic model, for $\time\in 1{\mycolon}\Time$\\
-End For\\
-Prediction weights, 
-$w^P_{\unit,\time,\rep}= f_{Y_{B_{\unit,\time}}|{X}_{1:\time}}(\data{y}_{B_{\unit,\time}}\given {X}^{\simulation}_{1:\time,\rep})$\\
-Measurement weights,
-$w^M_{\unit,\time,\rep}=f_{Y_{\unit,\time}|X_{\time},Y_{B_{\unit,\time}}}(\data{y}_{\unit,\time}\given X^{\simulation}_{\time,\rep},\data{y}_{B_{\unit,\time}})$
-\\
-Conditional log likelihood estimate,
+\begin{algorithm}[H]
+  \caption{\bf Generalized unadapted bagged filter (G-UBF).
+   }\label{alg:gubf}
+  \KwIn{
+    simulator for $f_{{X}_{\time}|{X}_{\time-1}}({x}_{\time}\given {x}_{\time-1})$ and $f_{{X}_0}({x}_0)$;
+    evaluator for $f_{{Y}_{\unit,\time}|{X}_{\time}}({y}_{\unit,\time}\given {x}_{\time})$;
+    data, $\data{\myvec{y}}_{1:\Time}$;
+    number of replicates, $\Rep$;
+    neighborhood structure, $B_{\unit,\time}$
+  }
+\For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{
+simulate
+  ${X}_{0:\Time,\rep}^{\simulation}$
+  from the dynamic model, for $\time$ in $\seq{1}{\Time}$
+  \;
+prediction weights, 
+  $w^P_{\unit,\time,\rep}= f_{Y_{B_{\unit,\time}}|{X}_{1:\time}}
+    (\data{y}_{B_{\unit,\time}}\given {X}^{\simulation}_{1:\time,\rep})$
+  for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$ 
+  \;
+measurement weights,
+  $w^M_{\unit,\time,\rep}=f_{Y_{\unit,\time}|X_{\time},Y_{B_{\unit,\time}}}
+    (\data{y}_{\unit,\time}\given
+    X^{\simulation}_{\time,\rep},\data{y}_{B_{\unit,\time}})$
+  for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$ 
+}
   $\MC{\loglik}_{\unit,\time}= 
@@
         \sum_{\tilde \rep=1}^\Rep w^P_{\unit,\time,\tilde \rep}
-\right)
-$
-\\
-\hline
-\end{tabular}
-\end{center}
+    \right)$
+  for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$ 
+\;
+\KwOut{
+log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit\comma\time}$\\
+}
+\end{algorithm}
 
@@
 \end{eqnarray*}
-Call the adapted simulation process $\{A_n\}$, defined conditionally on $\{Y_n\}$. 
-We see from the above calculation that 
+Call the adapted simulation process $\{A_{\time},\time=1,2,\dots\}$, defined conditionally on $\{Y_{\time},\time=1,2,\dots\}$. 
+We see from the above calculation that $A_{\time}$ can be constructed by the recursion
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-$\{A_n\}$ tracks $\{X_n\}$ when the process $\{A_n-X_n\}$ is stable.
+$A_{\time}$ tracks $X_{\time}$ when the process $\{A_{\time}-X_{\time},\time=1,2,\dots\}$ is stable.
 This happens when $\mu(x)-\mu(y)$ is negative when $x$ is sufficiently larger than $y$. 
@@
 \end{equation}
-Together, the results in [\ref{thm1:loglik:linearization}], [\ref{var:bound:without:assumption}], [\ref{var:bound:with:assumption}] and [\ref{thm1:bias:bound}] confirm the assertions of the theorem.
+Together, the results in \myeqref{thm1:loglik:linearization}, \myeqref{var:bound:without:assumption}, \myeqref{var:bound:with:assumption} and \myeqref{thm1:bias:bound} confirm the assertions of the theorem.
 \end{proof}
@@
 Lower variance than $O(\Unit \Time)$ cannot be anticipated for any sequential Monte Carlo method since the log likelihood estimate can be written as a sum of $\Unit\Time$ terms each of which involves its own sequential Monte Carlo calculation.
-The following Proposition~\ref{prop:cov:mix} gives an implication of Assumption~\ref{B:unconditional:mix}
+The following Proposition~\ref{prop:cov:mix} gives an implication of Assumption~\ref{B:unconditional:mix}.
 
@@
 
-bm_files_dir <- paste0("bm_",bm_run_level,"/")
+bm_files_dir <- paste0("../bm_",bm_run_level,"/")
 if(!dir.exists(bm_files_dir)) dir.create(bm_files_dir)
@@
 
-measles_model_dir <- "measlesModel/"
+measles_model_dir <- "../measlesModel/"
 load(file=paste0(measles_model_dir,"measles_spatPomp.rda"))
@@
 
-mscale_files_dir <- paste0("mscale_",mscale_run_level,"/")
+mscale_files_dir <- paste0("../mscale_",mscale_run_level,"/")
 
@@
     $\psi$ & \Sexpr{measles_params["psi"]} &---& reporting overdispersion \\
-    $G$ & \Sexpr{measles_params["g"]} &---& gravitation constant \\  \hline
+    $G$ & \Sexpr{measles_params["g"]} &---& gravitation constant \\
+    $S_{\unit}(0), \unit \in 1:\Unit$ & \Sexpr{measles_params["S1_0"]} &---& initial susceptible fraction \\
+    $E_{\unit}(0), \unit \in 1:\Unit$ & \Sexpr{measles_params["E1_0"]} &---& initial exposed fraction \\
+    $I_{\unit}(0), \unit \in 1:\Unit$ & \Sexpr{measles_params["I1_0"]} &---& initial infectious fraction \\
+\hline
   \end{tabular}
-  \caption{\label{tab:measles_parameters} Table of parameters for the spatiotemporal measles transmission model}
+  \caption{\label{tab:measles_parameters} Parameters for the spatiotemporal measles transmission model}
 \end{table}
@@
 
-Table~\ref{tab:measles_parameters} gives the model parameter values and Table~\ref{tab:mscale} gives the algorithmic settings used for the filters.
+Table~\ref{tab:measles_parameters} gives the model parameter values.
+The parameters were selected based on values found in time series analysis of single cities \citep{he10,bjornstad02}.
+The coupling parameter, $G$, was chosen so that the number of cities with simulated epidemics outside the dominant 2-year cycle is visually similar to the data.
+Initial values were chosen to initialize the system with the 2-year cycles observed in the UK during this period.
+The model code was parameterized via the basic reproduction number, $\mathcal{R}_0=\bar\beta \big(\mu_{\bullet D}+\mu_{IR}\big)^{-1}$.
+We used $\mathcal{R}_0=30$, corresponding to $\bar\beta = 1560.6 = 30 \times (52+0.02)$.
+$\mathcal{R}_0=30$ is higher than most estimates for measles but consistent with estimates previously found via time series analysis of pre-vaccine UK measles in single cities \citep{he10,bjornstad02}.
+
+Table~\ref{tab:mscale} gives the algorithmic settings used for the filters.
 The times in Table~\ref{tab:mscale} give the total time required by each algorithm to calculate all its results for Figure~{\MainFigureMeaslesScaling} in the main text, using \Sexpr{mscale_cores} cores.
@@
 The limit studied in this experiment is a balance between the two: the assumption is that one is prepared to invest a growing amount of computational effort as the data grow, but this should not grow too fast.
-{\ABFIR} was permitted the greatest computational effort, but two considerations balance this.
+{\ABFIR} was permitted the greatest computational effort, but the following two considerations balance this:
 \begin{enumerate}
@@
 
-  abfNbhd_model_dir <- "measlesModel/"
+  abfNbhd_model_dir <- "../measlesModel/"
   # note: care is required if abfNbhd_model_dir is set to something other than the default measlesModel
@@
 \section{\secTitleSpace A Lorenz-96 example}
+\label{sec:lorenz}
 
@@
       U=u, N=lz_N,
-      dt=lz_dt,dt_obs=lz_dt_obs,
+      delta_t=lz_dt,delta_obs=lz_dt_obs,
       regular_params=lz_params)
@@
     system.time(
-      girf(lz_list[[job$U_id]], method='adams',
+      girf(lz_list[[job$U_id]], 
         Np=lz_girf_Np,
         Ninter = job$U,
+	kind="moment",
         lookahead = lz_girf_lookahead,
@@
 
-<<lz_ubf2,cache=F,echo=F,eval=F>>=
-lz_ubf2 <- stew(file=paste0(lz_files_dir,"lz_ubf2.rda"),seed=844424,{
-  foreach(job=iter(lz_jobs,"row")) %do% {
-    system.time(
-      abf(lz_list[[job$U_id]],
-        Nrep = lz_ubf_Nrep,
-        Np=1,
-        nbhd = lz_nbhd2, tol=lz_tol) -> lz_ubf2_out
-    ) -> lz_ubf2_time
-#    uncomment for debugging
-#    list(logLik=logLik(lz_ubf2_out),time=lz_ubf2_time["elapsed"],lz_ubf2_out)
-    list(logLik=logLik(lz_ubf2_out),time=lz_ubf2_time["elapsed"])
-  } -> lz_ubf2_list
-})
-lz_jobs$ubf2_logLik <- vapply(lz_ubf2_list,function(x)x$logLik,numeric(1))
-lz_jobs$ubf2_time <- vapply(lz_ubf2_list,function(x) x$time,numeric(1))
-
-@
-
 <<lz_abfir,cache=F,echo=F,eval=T>>=
@@
     system.time(
-      abfir(lz_list[[job$U_id]], method='adams',
+      abfir(lz_list[[job$U_id]],
         Nrep = as.integer(lz_abfir_Nrep),
@@
       enkf(lz_list[[job$U_id]], 
-        Np=lz_enkf_Np,
-        h = lz_h,
-        R = diag((coef(lz_list[[job$U_id]])["tau"])^2, nrow = job$U)  
-      ) -> lz_enkf_out 
+        Np=lz_enkf_Np) -> lz_enkf_out 
     ) -> lz_enkf_time
 #    uncomment for debugging 
+#    enkf(lz_list[[1]],Np=lz_enkf_Np)-> lztmp 
 #    list(logLik=logLik(lz_enkf_out),time=lz_enkf_time["elapsed"],lz_enkf_out)
@@
 
-<<lz_image_plot, echo=F, fig.height=3.5, fig.width=7, out.width="6.5in", fig.cap = paste('Lorenz \'96 simulation used in the main text')>>=
+<<lz_image_plot, echo=F, fig.height=3.5, fig.width=7, out.width="6.5in", fig.cap = paste('Lorenz \'96 simulation used for Figure \\ref{fig:lz_loglik_plot}')>>=
 library(fields)
@@
 
-\begin{center}
-\noindent\begin{tabular}{l}
-\hline
-{\bf 
-{{\ABF}$_2$}. Adapted bagged filter, representation $2$.}\inputSpace\\
-\hline
-%{\bf input:} From Table~1 \inputSpace \\
-%\hline
-\firstLineSpace
-Initialize adapted simulation: $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\myvec{x}_0)$
-\\
-For $\time\ \mathrm{in}\ \seq{1}{\Time}$
-\\
-\asp  Proposals:
+\begin{algorithm}[H]
+  \caption{\bf Adapted bagged filter, representation $2$ (ABF$_2$).
+   }\label{alg:abf2}
+\For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{
+initialize adapted simulation:
+  $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\myvec{x}_0)$
+\;
+\For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{
+proposals:
     $\myvec{X}_{\time,\rep,\np}^{\IP} \sim 
@@
     \big( \myvec{x}_{\time}\given \myvec{X}^{\IF}_{\time-1,\rep}\big)$
-\\
-\asp Measurement weights:
+    for $\np$ in $\seq{1}{\Np}$
+\;
+measurement weights:
   $w^M_{\unit,\time,\rep,\np} = 
@@
     \big (\data{y}_{\unit,\time}\given X^{\IP}_{\unit,\time,\rep,\np}\big)$
-\\
-\asp  Adapted resampling weights:
+    for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$
+\;
+adapted resampling weights:
   $w^{\IF}_{\time,\rep,\np} = 
     \prod_{\unit=1}^{\Unit} w^M_{\unit,\time,\rep,\np}$
-\\
-\asp
-      Resampling:
+    for $\np$ in $\seq{1}{\Np}$
+\;
+resampling:
   $\prob\big[\resampleIndex({\rep})=a \big] = w^{\IF}_{\time,\rep,a}
@@
     \Big)^{-1}$
-\\
-\asp 
+\;
 $\myvec{X}^{\IF}_{\time,\rep} = \myvec{X}^{\IP}_{\time,\rep,r(\rep)}$ 
-\\
-\asp Compute $\gamma^{}_{\unit,\time+k,\rep,k}$  for $k\ \mathrm{in}\ \seq{0}{K}$,
-\\
-\asp Compute $\gamma^{+}_{\unit,\time,\rep,0}$
-\\
-End for
-\\
+  \;
+  compute $\gamma^{}_{\unit,\time+k,\rep,k}$  for $\unit$ in $\seq{1}{\Unit}$, $k\ \mathrm{in}\ \seq{0}{K}$,
+  \;
+  compute $\gamma^{+}_{\unit,\time,\rep,0}$ for $\unit$ in $\seq{1}{\Unit}$
+}
+}
 $\displaystyle \MC{\loglik}_{\unit,\time}= 
@@
 $
-\vspace{1mm}
-\\
-\hline
-\end{tabular}
-\end{center}
+for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$
+\end{algorithm}
 
@@
 
-\bibliography{bib-iif}
+%%%%%%%%%%%%%%%%%%%%%% fls %%%%%%%%%%%%%%%%
+
+\section{\secTitleSpace Bagged filters for functions of the latent states}
+
+The theory and methodology in the main article focused on filtering for likelihood estimation.
+Here, we describe extensions to other filtering problems.
+Let $\{\filtFunc_{\filts}:\myvec{X}^{\Unit}\to\R, \mbox{ $\filts$ in $\seq{1}{\Filts}$}\}$ be a collection of functions where $\filtFunc_{\filts}$ depends only on a subset of units $\filtFuncSet_{\filts}\subset \seq{1}{\Unit}$.
+We suppose that there exist neighborhoods
+\begin{equation}
+B^\prime_{\filts,\time}\subset A^\prime_{\time}=(\seq{1}{\Unit})\times(\seq{0}{\time})
+\end{equation}
+such that $\filtFunc_{\filts}(\myvec{X}_{\time})$ is approximately independent of $\{Y_{\unit,\time}:(\unit,\time) \in B^{\prime c}_{\filts,\time}\}$ given  $\{Y_{\unit,\time}: (\unit,\time) \in B^\prime_{\filts,\time}\}$, where $B^{\prime c}_{\filts,\time}$ is a complement in $A^\prime_{\time}$.
+Unlike the sets $A_{\unit,\time}$ and $B_{\unit,\time}$ defined for likelihood estimation, the sets $A^\prime_{\time}$ and $B^\prime_{\filts,\time}$ can include any locations at time $\time$.
+We consider Monte Carlo estimation of
+\begin{equation}
+\label{eq:filt1}
+\overline{\filtFunc}_{k,\time}= \E\big[\filtFunc_{\filts}(\myvec{X}_\time) \big| \myvec{Y}_{1:\time}\big], \quad \filts \mbox{ in } \seq{1}{\Filts}.
+\end{equation}
+For example, if we set $\filtFunc_{\unit}(\myvec{x}_\time)=x_{\unit,\time}$ and $\Filts=\Unit$, the collection of quantities $\{\overline{\filtFunc}_{\unit,\time}\}$ estimated in \eqref{eq:filt1} corresponds to a vector of filter means.
+Setting $\filtFunc_{\filts}(\myvec{x}_\time)=x_{\unit_{\filts},\time}\, x_{\altUnit_{\filts},\time}$ enables calculation of a collection of filter covariances between units $\unit_k$ and $\altUnit_k$ for $k$ in $\seq{1}{\Filts}$.
+
+We consider bagged filtering approaches to estimation of $\{\overline{\filtFunc}_{\filts,\time}, \filts \mbox{ in } \seq{1}{\Filts}\}$.
+Although each $\overline{\filtFunc}_{\filts,\time}$ is required to have only local dependence, some global quantities such as filter means and their variances across all units, can be expressed in terms of collections of such quantities.
+For bagged filtering to operate successfully, the neighborhoods $\{B^\prime_{\filts,\time}\}$ should not be large.
+Since  $B^\prime_{\filts,\time}$ will typically be larger than $\{\filtFuncSet_\filts\}$, this rules out estimation of filtered quantities that cannot be adequately represented by a collection of localized filtering calculations.
+We now present variants of the pseudocode in the main text, targeted at estimation of $\{\overline{\filtFunc}_{k,\time},k \mbox{ in } 1{\mycolon}K\}$.
+We do not prove theorems about these algorithms, but we conjecture from their similarity to the algorithms in the main text that comparable theoretical results should exist.
+Table~\ref{tab:filter:inputs} lists the inputs, outputs and ranges of the implicit loops for the following algorithms.
+
+\begin{table}[htbp]
+\begin{center}
+\noindent\begin{tabular}{l}
+\hline
+\inputSpace {\bf Latent state estimation via bagged filters.}\\
+\hline
+\inputSpace {\bf input:}
+\\
+collection of functions, $\filtFunc_{\filts}$\\
+neighborhoods, $B^\prime_{\filts,\time}$\\
+simulator for $f_{\myvec{X}_0}(\myvec{x}_0)$ and $f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}(\myvec{x}_{\time}\given \myvec{x}_{\time-1})$\\
+    evaluator for $f_{{Y}_{\unit\comma\time}|{X}_{\unit\comma\time}}({y}_{\unit\comma\time}\given {x}_{\unit\comma\time})$\\
+    number of replicates, ${\Rep}$\\
+    data, $\data{\myvec{y}}_{1:\Time}$\\
+{\ABF} and {\ABFIR}:    particles per replicate,  $\Np$\\
+{\ABFIR}: number of intermediate timesteps, $\Ninter$ \\
+{\ABFIR}: measurement variance parameterizations, ${\VtoTheta}_{\unit\comma\time}$ and ${\thetaToV}_{\unit\comma\time}$\\
+{\ABFIR}: approximate process and observation mean functions, $\myvec{\mu}$ and $h_{\unit\comma\time}$\\
+\inputSpace {\bf output:}\\
+  filter estimate, $\MC{\overline{\filtFunc}}_{\filts,\time}$ for $\filts$ in $\seq{1}{\Filts}$\\
+\lastLineSpace \\
+\hline
+\end{tabular}
+\end{center}
+\caption{\label{tab:filter:inputs}
+Notation for bagged filter for latent state estimation: inputs, outputs and implicit loops.}
+\end{table}
+
+\clearpage
+
+\begin{algorithm}[H]
+\caption{\bf Unadapted bagged filter for latent state estimation.}
+\For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{
+initialize simulation, $\myvec{X}_{0,\rep} \sim f_{\myvec{X}_0}(\cdot)$
+  \;
+\For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{
+simulate
+    $\myvec{X}_{\time,\rep} \sim
+      f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}}
+      \big( \mydot \given \myvec{X}_{\time-1,\rep}
+    \big)$
+  \;
+measurement weights,
+  $w^M_{\unit,\time,\rep}=
+    f_{Y_{\unit,\time}|X_{\unit,\time}}(\data{y}_{\unit,\time}
+    \given X^{\tif}_{\unit,\time,\rep})$
+  for $\unit$ in $\seq{1}{\Unit}$
+  \;
+filtering weights, 
+  $w^F_{\filts,\time,\rep}=
+    \prod_{(\tilde \unit,\tilde n)\in B^\prime_{\filts,\time}}
+    w^M_{\tilde\unit,\tilde n,\rep}$
+  for $\filts$ in $\seq{1}{\Filts}$
+}
+}
+$\displaystyle
+\MC{\overline{\filtFunc}}_{\filts,\time}= 
+\frac{
+  \sum_{\rep=1}^\Rep \filtFunc_\filts(\myvec{X}^{\tif}_{\time,\rep}) \,
+  w^F_{\filts,\time,\rep}
+}{
+  \sum_{\rep=1}^\Rep 
+  w^F_{\filts,\time,\rep}
+}$  for $\filts$ in $\seq{1}{\Filts}$, $\time$ in $\seq{1}{\Time}$
+\end{algorithm}
+
+\begin{algorithm}
+\caption{\bf 
+Adapted bagged filter for latent state estimation.}
+\For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{
+initialize adapted simulation, $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\cdot)$
+  \;
+\For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{
+proposals:
+    $\myvec{X}_{\time,\rep,\np}^{\IP} \sim 
+    f_{\myvec{X}_{\time}|X_{1:\Unit,\time-1}} 
+    \big( \myvec{x}_{\time}\given \myvec{X}^{\IF}_{\time-1,\rep}\big)$
+  for $\np$ in $\seq{1}{\Np}$
+  \;
+measurement weights:
+  $w^M_{\unit,\time,\rep,\np} = 
+    f_{Y_{\unit,\time}|X_{\unit\comma\time}} 
+    \big (\data{y}_{\unit\comma\time}\given
+    X^{\IP}_{\unit\comma\time,\rep,\np}\big)$
+  for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$
+  \;
+adapted resampling weights:
+  $w^{\IF}_{\time,\rep,\np} = 
+    \prod_{\unit=1}^{\Unit} w^M_{\unit,\time,\rep,\np}$
+  for $\np$ in $\seq{1}{\Np}$
+  \;
+resampling:
+  $\prob\big[\resampleIndex({\rep})=a \big] = w^{\IF}_{\time,\rep,a}
+    \Big( 
+      \sum_{q=1}^{\Np} w^{\IF}_{\time,\rep,q}
+    \Big)^{-1}$
+\;
+$\myvec{X}^{\IF}_{\time,\rep} = \myvec{X}^{\IP}_{\time,\rep,r(\rep)}$ 
+\;
+% Prediction weights:
+  $w^{F}_{\filts,\time,\rep,\np}= \displaystyle
+    \prod_{\altTime=1}^{\time-1}
+    \Big[
+      \frac{1}{\Np}\sum_{q=1}^{\Np}
+      \hspace{1mm}
+         \prod_{\altUnit:(\altUnit,\altTime)\in B^\prime_{\filts,\time}} 
+      \hspace{-1mm}
+          w^M_{\altUnit,\altTime,\rep,q}
+    \Big] \prod_{\altUnit:(\altUnit,\time)\in B^\prime_{\filts,\time}} 
+      \hspace{-1mm}
+    w^M_{\altUnit,\time,\rep,\np}$
+  for $\np$ in $\seq{1}{\Np}$,  $\filts$ in $\seq{1}{\Filts}$
+\;
+}
+}
+$\displaystyle \MC{\overline{\filtFunc}}_{\filts,\time}=
+\frac{
+\sum_{\rep=1}^\Rep \sum_{\np=1}^{\Np} \filtFunc_\filts(\myvec{X}^{\IP}_{\time,\rep,\np}) \,
+w^F_{\filts,\time,\rep,\np}
+}{
+\sum_{\rep=1}^\Rep \sum_{\np=1}^{\Np} w^F_{\filts,\time,\rep,\np}
+}
+$
+   for $\filts$ in $\seq{1}{\Filts}$, $\time$ in $\seq{1}{\Time}$
+\end{algorithm}
+
+\begin{algorithm}
+\caption{\bf ABF-IR for latent state estimation.}
+\For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{  
+initialize adapted simulation, $\myvec{X}^{\IF}_{0,\rep} \sim f_{\myvec{X}_0}(\cdot)$
+  \;
+  \nllabel{alg:abfir:for:n}  
+\For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{
+  guide simulations:
+    $\myvec{X}_{\time,\rep,\npgir}^{G} \sim 
+    f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}} 
+    \big( \myvec{x}_{\time}\given \myvec{X}^{\IF}_{\time-1,\rep} \big)$
+  for $\np$ in $\seq{1}{\Np}$
+  \;
+% Guide sample variance:
+  $V_{\unit,\time,\rep}=
+    \var \big\{
+    h_{\unit\comma\time}\big( {X}_{\unit,\time,\rep,\npgir}^{G}\big), \ 
+      \npgir \mbox{ in } \seq{1}{\Npgir}
+      \big\}$
+  \;
+  $\guideFunc^{\resample}_{\time,0,\rep,\np}=1 \; \; $ and
+    $\; \myvec{X}_{\time,0,\rep,\np}^{\GR}=\myvec{X}^{\IF}_{\time-1,\rep}$
+  for $\np$ in $\seq{1}{\Np}$
+  \;
+%%% intermediate loop sssss  
+  \For{$\ninter\ \mathrm{in}\ \seq{1}{\Ninter}$}{
+%Intermediate proposals:
+     ${\myvec{X}}_{\time,\ninter,\rep,\np}^{\GP}
+       \sim {f}_{{\myvec{X}}_{\time,\ninter}|{\myvec{X}}_{\time,\ninter-1}}
+       \big(\mydot|{\myvec{X}}_{\time,\ninter-1,\rep,\np}^{\GR}\big)$
+     for $\np$ in $\seq{1}{\Np}$
+     \;
+     $\myvec{\mu}^{\GP}_{\time,\ninter,\rep,\np} 
+       = \myvec{\mu}\big( \myvec{X}^{\GP}_{\time,\ninter,\rep,\np},
+       t_{\time,\ninter},t_{\time} \big)$
+     for $\np$ in $\seq{1}{\Np}$
+     \;
+%Measurement variance at skeleton: 
+     $V^{\mathrm{meas}}_{\unit,\time,\ninter,\rep,\np}
+        = \thetaToV_{\unit}(\theta,\mu^{\GP}_{\unit,\time,\ninter,\rep,\np})$
+     for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$
+     \;
+%Process variance:
+     $V^{\mathrm{proc}}_{\unit,\time,\ninter,\rep}
+       = V_{\unit,\time,\rep} \,
+       \big(t_{\time}-t_{\time,\ninter}\big) \Big/
+       \big(t_{\time}-t_{\time,0}\big)$ 
+     for $\unit$ in $\seq{1}{\Unit}$
+     \;
+%Moment matching:
+     $\theta_{\unit,\time,\ninter,\rep,\np}= 
+       \VtoTheta_{\unit}\big(
+       V^{\mathrm{meas}}_{\unit,\time,\ninter,\rep,\np} +
+       V^{\mathrm{proc}}_{\unit,\time,\ninter,\rep}, 
+       \, \mu^{\GP}_{\unit,\time,\ninter,\rep,\np}
+       \big)$
+     for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$
+     \;
+% Guide function: 
+     $\guideFunc_{\time,\ninter,\rep,\np}=
+       \prod_{\unit=1}^{\Unit}
+       f_{Y_{\unit,\time}|X_{\unit,\time}}
+       \big(
+         \data{y}_{\unit,\time}\given \mu^{\GP}_{\unit,\time,\ninter,\rep,\np}
+	 \giventh \theta_{\unit,\time,\ninter,\rep,\np} 
+         \big)$
+     for $\np$ in $\seq{1}{\Np}$
+     \;
+guide weights:
+     $w^G_{\time,\ninter,\rep,\np}= \guideFunc^{}_{\time,\ninter,\rep,\np}
+       \big/ \guideFunc^{\resample}_{\time,\ninter-1,\rep,\np}$     
+     for $\np$ in $\seq{1}{\Np}$
+     \;
+resampling:
+     $\prob\big[\resampleIndex({\rep,\np})=a \big] =
+       w^G_{\time,\ninter,\rep,a}
+       \Big( \sum_{\altNp=1}^{\Np}w^G_{\time,\ninter,\rep,\altNp}\Big)^{-1}$
+     for $\np$ in $\seq{1}{\Np}$       
+     \;
+     $\myvec{X}_{\time,\ninter,\rep,\np}^{\GR}=
+       \myvec{X}_{\time,\ninter,\rep,\resampleIndex({\rep,\np})}^{\GP}\; \; $
+     and
+     $\; \guideFunc^{\resample}_{\time,\ninter,\rep,\np}=
+       \guideFunc^{}_{\time,\ninter,\rep,\resampleIndex({\rep,\np})}\,$
+     for $\np$ in $\seq{1}{\Np}$       
+  } 
+  $\myvec{X}^{\IF}_{\time,\rep}=\myvec{X}^{\GR}_{\time,\Ninter,\rep,1}$
+  \;
+%  Measurement weights:
+  $w^M_{\unit,\time,\rep,\npgir} = 
+    f_{Y_{\unit,\time}|X_{\unit,\time}} 
+    \big (\data{y}_{\unit,\time}\given X^{G}_{\unit,\time,\rep,\npgir} \big)$
+  for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$
+  \;
+% Filter weights:
+  $w^{F}_{\filts,\time,\rep,\npgir}= \displaystyle
+  \prod_{\altTime=1}^{\time-1}
+  \Big[
+    \frac{1}{\Npgir}\sum_{q=1}^{\Npgir}
+    \hspace{1mm}
+       \prod_{\altUnit:(\altUnit,\altTime)\in B^\prime_{\filts,\time}} 
+    \hspace{-1mm}
+        w^M_{\altUnit,\altTime,\rep,q}
+  \Big] \prod_{\altUnit:(\altUnit,\time)\in B^\prime_{\filts,\time}} 
+    \hspace{-1mm}
+        w^M_{\altUnit,\time,\rep,\npgir}$
+  for $\filts$ in $\seq{1}{\Filts}$, $\unit$ in $\seq{1}{\Unit}$
+  \;
+}
+}
+$\displaystyle \MC{\overline{\filtFunc}}_{\filts,\time}=
+\frac{
+\sum_{\rep=1}^\Rep \sum_{\np=1}^{\Np} \filtFunc_\filts(\myvec{X}^{\IP}_{\time,\rep,\np}) \,
+w^F_{\filts,\time,\rep,\np}
+}{
+\sum_{\rep=1}^\Rep \sum_{\np=1}^{\Np} w^F_{\filts,\time,\rep,\np}
+}
+$
+for $\filts$ in $\seq{1}{\Filts}$, $\time$ in $\seq{1}{\Time}$
+\end{algorithm}
+
+\clearpage
+
+%%%%%%%%%%%%%%%%%%%%%% ifif %%%%%%%%%%%%%%%%
+
+\section{\secTitleSpace An iterated bagged filter for parameter estimation}
+
+This paper focuses on evaluating the likelihood function for SpatPOMP models via filtering.
+Although the likelihood function is fundamental for inference, evaluation alone is not sufficient.
+Likelihood maximization enables calculation of the maximum likelihood estimate, profile likelihood confidence intervals, likelihood ratio tests and likelihood-based model selection critera.
+Iterated filtering methodology \citep{ionides06-pnas,ionides11,ionides15} provides an approach to extending filtering algorithms to likelihood maximization algorithms. 
+Here, we demonstrate one such extension in the context of bagged filters.
+This demonstration provides a proof of concept to motivate future work.
+
+Iterated filtering approaches apply filtering to a modified version of the model where parameters are perturbed at each time point.
+The filtering procedure directs the perturbed parameters toward values consistent with the data.
+At the end of each filtering operation, a parameter updating rule is applied and a new filtering iteration is started with reduced perturbation variance.
+Under suitable conditions, iterative procedures of this type converge to a neighborhood of a maximum likelihood parameter despite the presence of Monte Carlo filtering error.
+We implemented an iterated bagged filter procedure, described by the pseudocode below.
+The pseudocode is presented for an iterated adapted bagged filter (IABF) but the iterated unadapted filter (IUBF) corresponds to the case $\Np=1$ and the iterated bagged filter with intermediate resampling (IABF-IR) follows by adding the intermediate resampling procedure used by ABF-IR.
+The filtering step of IABF uses ABF to estimate the likelihood at the $K$ perturbed parameter sets.
+The selection step does not resample parameters based on these estimated likelihood.
+Rather, it selects the parameters with the top $p$ quantile of likelihoods and copies them appropriately to get $K$ new parameters for the next filtering step.
+This quantile-based resampling allows us to maintain the diversity of the $K$ parameter sets and avoid \textit{parameter degeneracy}, whereby very few parameters are resampled, leading to an inefficient search of parameter space.
+
+For simplicity, in this description, we assume that parameters are transformed so that their values are unconstrained.
+Our software implementation, provided the R package \texttt{spatPomp} \citep{asfaw21arxiv}, provides facilities for carrying out such transformations.
+The Gaussian distribution used for perturbations, and the geometric perturbation variance reduction factor,  $\alpha$, are convenient specifications but are not required in theory \citep{ionides15}.
+As another simplification, the pseudocode for IABF represents the logical structure of the algorithm without attending to issues of memory management and parallelization.
+For implementation issues, we refer to \texttt{spatPomp} \citep{asfaw21arxiv}.
+
+In Figure~5 of the main text, we use this IABF implementation to construct a profile likelihood for the measles model.
+We use $\Np=1$, $\Rep=30000$, $K=250$, $p=0.8$, $\alpha=0.5$, $M=15$ and $\Sigma$ set to be a diagonal matrix with perturbation variance for each non-initial value parameter set to 0.02.
+For this exercise, we fix the initial value parameters at their true values.
+
+The resulting Monte Carlo parameter estimates have Monte Carlo uncertainty in both likelihood evaluation and maximization.
+Therefore, we used a Monte Carlo adjusted profile likelihood \citep{ionides17,ning21} that accounts for this uncertainty.
+Fig.~\MainFigureMeaslesProfile gives empirical demonstration of this procedure on the measles model.
+
+\begin{algorithm}[H]
+  \caption{\bf Iterated adapted bagged filter (IABF).} \label{alg:iabf}
+  \KwIn{ same as ABF table in main text plus:
+  maximization iterations, $M$;
+  number of parameter vectors, $K$;
+  perturbation variance, $\Sigma$;
+  variance reduction factor after 50 iterations, $\alpha$;
+  starting parameters, $\theta_{1:K}^{(0)}$;
+  resampling proportion, $p$
+}
+\For{$m\ \mathrm{in}\ \seq{1}{M}$}{
+  $\theta_{0,k}^{F} = \theta_{k}^{(m-1)}$ for $k$ in $\seq{1}{K}$
+  \;
+initialize adapted simulation:
+  $\myvec{X}^{F}_{0,\rep,k} \sim f_{\myvec{X}_0}\big(\myvec{x}_0
+    \giventh \theta_{0,k}^{F}\big)$
+  for $k$ in $\seq{1}{K}$
+  \;
+  \For{$\time\ \mathrm{in}\ \seq{1}{\Time}$}{
+    $\theta_{\time,k}^{P}\sim\normal\big[\theta_{\time-1,k}^{F},
+      \alpha^{2m/50} \Sigma\big]$
+    for  $k$ in $\seq{1}{K}$
+    \;
+    \For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{
+    $\myvec{X}_{\time,\rep,\np,k}^{\IP} \sim 
+      f_{\myvec{X}_{\time}|\myvec{X}_{\time-1}} 
+      \big( \myvec{x}_{\time}\given \myvec{X}^{F}_{\time-1,\rep,k}
+      \giventh \theta_{\time,k}^{P}\big)$
+    for  $\np$ in $\seq{1}{\Np}$, $k$ in $\seq{1}{K}$
+    \;
+% measurement weights:
+    $w^M_{\unit,\time,\rep,\np,k} = 
+      f_{Y_{\unit,\time}|X_{\unit\comma\time}} 
+      \big (\data{y}_{\unit\comma\time}\given
+      X^{\IP}_{\unit\comma\time,\rep,\np,k}
+      \giventh \theta_{\time,k}^{P} \big)$
+    for $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$, $k$ in $\seq{1}{K}$
+    \;
+% adapted resampling weights:
+    $w^{\IF}_{\time,\rep,\np,k} = 
+       \prod_{\unit=1}^{\Unit} w^M_{\unit,\time,\rep,\np,k}$
+    for $\np$ in $\seq{1}{\Np}$, $k$ in $\seq{1}{K}$
+    \;
+    resampling:
+    $\prob\big[\resampleIndex({\rep},k)=a \big] = w^{\IF}_{\time,\rep,a,k}
+      \Big( 
+        \sum_{\xi=1}^{\Np} w^{\IF}_{\time,\rep,\xi,k}
+      \Big)^{-1}$
+    for $k$ in $\seq{1}{K}$
+    \;
+    $\myvec{X}^{\IF}_{\time,\rep,k} = \myvec{X}^{\IP}_{\time,\rep,r(\rep,k),k}$ 
+    for $k$ in $\seq{1}{K}$
+    \;
+% prediction weights:
+    $w^{\LCP}_{\unit,\time,\rep,\np,k}= \displaystyle
+      \prod_{\altTime=1}^{\time-1}
+      \Big[
+        \frac{1}{\Np}\sum_{\xi=1}^{\Np}
+        \hspace{1mm}
+        \prod_{\altUnit:(\altUnit,\altTime)\in B_{\unit,\time}} 
+        \hspace{-1mm}
+        w^M_{\altUnit,\altTime,\rep,\xi,k}
+      \Big] \prod_{\altUnit:(\altUnit,\time)\in B_{\unit,\time}} 
+      \hspace{-1mm}
+      w^M_{\altUnit,\time,\rep,\np,k}$ \hspace{3cm}
+    for  $\unit$ in $\seq{1}{\Unit}$, $\np$ in $\seq{1}{\Np}$, $k$ in $\seq{1}{K}$
+    \;
+    }
+    $\displaystyle \MC{\loglik}_{\time,k}= 
+      \sum_{\unit=1}^\Unit\log\Bigg(
+      \frac{
+      \sum_{\rep=1}^\Rep \sum_{\np=1}^{\Np} w^M_{\unit,\time,\rep,\np,k}
+      w^P_{\unit,\time,\rep,\np,k}
+      }{
+      \sum_{\rep=1}^\Rep \sum_{\np=1}^{\Np} w^P_{\unit,\time,\rep,\np,k}
+      }
+      \Bigg)$
+    for  $k$ in $\seq{1}{K}$
+    \;
+select the highest $p K$ likelihoods:
+    find $s$ with \hspace{4cm}
+    $\{s(1),\dots,s(\lceil p K\rceil)\}=
+      \big\{k: \sum_{\tilde{k}=1}^{K}{\mathbf{1}} 
+      \{\MC{\loglik}_{\time,\tilde{k}} > \MC{\loglik}_{\time,k} \} < (1-p) K
+      \big\}$
+    \;
+make $1/p$ copies of successful parameters,
+    $\theta_{\time,k}^{F}=\theta_{\time,s(\lceil p k\rceil)}^{P}$
+    for  $k$ in $\seq{1}{K}$
+    \;
+    $\myvec{X}^{F}_{\time,\rep,k} = \myvec{X}^{A}_{\time,\rep,
+      s(\lceil p k\rceil)}$
+    for $\rep$ in $\seq{1}{\Rep}$, $k$ in $\seq{1}{K}$
+    \;
+  }
+  $\theta_{k}^{(m)}=\theta_{\Time,k}^{F}$
+  for $k$ in $\seq{1}{K}$
+}
+\KwOut{Parameter estimates approaching the maximum likelihood estimate, $\theta^{(M)}_{1:K}$}
+\end{algorithm}
+
+%%%%%%%%%%%%%%%%%%%%%%%% ttttttt %%%%%
+
+\section{\secTitleSpace Replicates versus particles for the measles model}
+
+<<ij_settings,cache=FALSE,echo=F>>=
+
+ij_files_dir <- paste0("ij_",ij_run_level,"/")
+if(!dir.exists(ij_files_dir)) dir.create(ij_files_dir)
+
+stew(file=paste0(ij_files_dir,"ij_settings.rda"),{
+
+  ij_tol <- 1e-300
+  ij_cores <- cores
+  ij_nbhd <- function(object, time, unit) {
+      nbhd_list <- list()
+      if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+      if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+      return(nbhd_list)
+  }
+
+  if(ij_run_level==1){ #######
+    ij_U <- 2
+    ij_N <- 5
+    ij_replicates <- 2 
+    ij1_Np <- 50
+    ij1_I <- 2
+    ij2_Np <- 33
+    ij2_I <- 3
+    ij3_Np <- 25
+    ij3_I <- 4
+    ij4_Np <- 20
+    ij4_I <- 5
+    ij5_Np <- 15
+    ij5_I <- 7
+    ij6_Np <- 10
+    ij6_I <- 10
+    ij7_Np <- 7
+    ij7_I <- 15
+    ij8_Np <- 5
+    ij8_I <- 20
+    ij9_Np <- 4
+    ij9_I <- 25
+    ij10_Np <- 3
+    ij10_I <- 33
+    ij11_Np <- 2
+    ij11_I <- 50
+} else if(ij_run_level==2){ #######
+    ij_U <- 40
+    ij_N <- 104
+    ij_replicates <- 5
+    ij1_Np <- 2000
+    ij1_I <- 500
+    ij2_Np <- 1000
+    ij2_I <- 1000
+    ij3_Np <- 500
+    ij3_I <- 2000
+    ij4_Np <- 250
+    ij4_I <- 4000
+    ij5_Np <- 100
+    ij5_I <- 10000
+    ij6_Np <- 50
+    ij6_I <- 20000
+    ij7_Np <- 25
+    ij7_I <- 40000
+    ij8_Np <- 10
+    ij8_I <- 40000
+    ij9_Np <- 5
+    ij9_I <- 40000
+    ij10_Np <- 1
+    ij10_I <- 40000
+    ij11_Np <- 2
+    ij11_I <- 40000
+  } else if(ij_run_level==3){
+  } else if(ij_run_level==4){
+  } else if(ij_run_level==5){
+  }
+
+})
+
+ij_jobs <- data.frame(reps=1:ij_replicates,U=ij_U,N=ij_N)
+
+@
+
+<<ij_spatPomp,eval=T,echo=F>>=
+ij_spatPomp <- stew(file=paste0(ij_files_dir,"ij_spatPomp.rda"),{
+  ij_model_dir <- "../measlesModel/"
+  # note: care is required if ij_model_dir is set to something other than
+  # the default measlesModel.
+  # However, it may be useful to do that while testing model variations.  
+  load(file=paste0(ij_model_dir,"measles_spatPomp.rda"))
+  ij_measles <- measles_subset(m_U=ij_U, m_N=ij_N)
+})
+
+@
+
+<<ij_nbhd,echo=F>>=
+ij_nbhd <- function(object, time, unit) {
+  nbhd_list <- list()
+  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+  return(nbhd_list)
+}
+@
+
+<<ij1,cache=F,echo=F>>=
+ij1 <- stew(file=paste0(ij_files_dir,"ij1.rda"),seed=844424,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles, 
+        Nrep = ij1_I,
+        Np=ij1_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij1_out 
+    ) -> ij1_time
+#    uncomment for debugging 
+#    list(logLik=logLik(ij1_out),time=ij1_time["elapsed"],ij1_out)
+    list(logLik=logLik(ij1_out),time=ij1_time["elapsed"])
+  } -> ij1_list
+})
+ij1_jobs <- ij_jobs
+ij1_jobs$logLik <- vapply(ij1_list,function(x)x$logLik,numeric(1))
+ij1_jobs$time <- vapply(ij1_list,function(x) x$time,numeric(1))
+ij1_jobs$Method <- ij1_Np
+@
+
+<<ij2,cache=F,echo=F>>=
+ij2 <- stew(file=paste0(ij_files_dir,"ij2.rda"),seed=389855,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles, 
+        Nrep = ij2_I,
+        Np=ij2_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij2_out 
+    ) -> ij2_time
+#    uncomment for debugging 
+#    list(logLik=logLik(ij2_out),time=ij2_time["elapsed"],ij2_out)
+    list(logLik=logLik(ij2_out),time=ij2_time["elapsed"])
+  } -> ij2_list
+})
+ij2_jobs <- ij_jobs
+ij2_jobs$logLik <- vapply(ij2_list,function(x)x$logLik,numeric(1))
+ij2_jobs$time <- vapply(ij2_list,function(x) x$time,numeric(1))
+ij2_jobs$Method <-  ij2_Np
+@
+
+<<ij3,cache=F,echo=F>>=
+ij3 <- stew(file=paste0(ij_files_dir,"ij3.rda"),seed=186711,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles,
+        Nrep = ij3_I,
+        Np=ij3_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij3_out
+    ) -> ij3_time
+#    uncomment for debugging
+#    list(logLik=logLik(ij3_out),time=ij3_time["elapsed"],ij3_out)
+    list(logLik=logLik(ij3_out),time=ij3_time["elapsed"])
+  } -> ij3_list
+})
+ij3_jobs <- ij_jobs
+ij3_jobs$logLik <- vapply(ij3_list,function(x)x$logLik,numeric(1))
+ij3_jobs$time <- vapply(ij3_list,function(x) x$time,numeric(1))
+ij3_jobs$Method <-  ij3_Np
+@
+
+<<ij4,cache=F,echo=F>>=
+ij4 <- stew(file=paste0(ij_files_dir,"ij4.rda"),seed=186711,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles,
+        Nrep = ij4_I,
+        Np=ij4_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij4_out
+    ) -> ij4_time
+#    uncomment for debugging
+#    list(logLik=logLik(ij4_out),time=ij4_time["elapsed"],ij4_out)
+    list(logLik=logLik(ij4_out),time=ij4_time["elapsed"])
+  } -> ij4_list
+})
+ij4_jobs <- ij_jobs
+ij4_jobs$logLik <- vapply(ij4_list,function(x)x$logLik,numeric(1))
+ij4_jobs$time <- vapply(ij4_list,function(x) x$time,numeric(1))
+ij4_jobs$Method <-  ij4_Np
+@
+
+<<ij5,cache=F,echo=F>>=
+ij5 <- stew(file=paste0(ij_files_dir,"ij5.rda"),seed=186711,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles,
+        Nrep = ij5_I,
+        Np=ij5_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij5_out
+    ) -> ij5_time
+#    uncomment for debugging
+#    list(logLik=logLik(ij5_out),time=ij5_time["elapsed"],ij5_out)
+    list(logLik=logLik(ij5_out),time=ij5_time["elapsed"])
+  } -> ij5_list
+})
+ij5_jobs <- ij_jobs
+ij5_jobs$logLik <- vapply(ij5_list,function(x)x$logLik,numeric(1))
+ij5_jobs$time <- vapply(ij5_list,function(x) x$time,numeric(1))
+ij5_jobs$Method <- ij5_Np
+@
+
+<<ij6,cache=F,echo=F>>=
+ij6 <- stew(file=paste0(ij_files_dir,"ij6.rda"),seed=186711,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles,
+        Nrep = ij6_I,
+        Np=ij6_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij6_out
+    ) -> ij6_time
+#    uncomment for debugging
+#    list(logLik=logLik(ij6_out),time=ij6_time["elapsed"],ij6_out)
+    list(logLik=logLik(ij6_out),time=ij6_time["elapsed"])
+  } -> ij6_list
+})
+ij6_jobs <- ij_jobs
+ij6_jobs$logLik <- vapply(ij6_list,function(x)x$logLik,numeric(1))
+ij6_jobs$time <- vapply(ij6_list,function(x) x$time,numeric(1))
+ij6_jobs$Method <- ij6_Np
+@
+
+<<ij7,cache=F,echo=F>>=
+ij7 <- stew(file=paste0(ij_files_dir,"ij7.rda"),seed=186711,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles,
+        Nrep = ij7_I,
+        Np=ij7_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij7_out
+    ) -> ij7_time
+#    uncomment for debugging
+#    list(logLik=logLik(ij7_out),time=ij7_time["elapsed"],ij7_out)
+    list(logLik=logLik(ij7_out),time=ij7_time["elapsed"])
+  } -> ij7_list
+})
+ij7_jobs <- ij_jobs
+ij7_jobs$logLik <- vapply(ij7_list,function(x)x$logLik,numeric(1))
+ij7_jobs$time <- vapply(ij7_list,function(x) x$time,numeric(1))
+ij7_jobs$Method <- ij7_Np
+@
+
+<<ij8,cache=F,echo=F>>=
+ij8 <- stew(file=paste0(ij_files_dir,"ij8.rda"),seed=186711,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles,
+        Nrep = ij8_I,
+        Np=ij8_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij8_out
+    ) -> ij8_time
+#    uncomment for debugging
+#    list(logLik=logLik(ij8_out),time=ij8_time["elapsed"],ij8_out)
+    list(logLik=logLik(ij8_out),time=ij8_time["elapsed"])
+  } -> ij8_list
+})
+ij8_jobs <- ij_jobs
+ij8_jobs$logLik <- vapply(ij8_list,function(x)x$logLik,numeric(1))
+ij8_jobs$time <- vapply(ij8_list,function(x) x$time,numeric(1))
+ij8_jobs$Method <- ij8_Np
+@
+
+<<ij9,cache=F,echo=F>>=
+ij9 <- stew(file=paste0(ij_files_dir,"ij9.rda"),seed=186711,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles,
+        Nrep = ij9_I,
+        Np=ij9_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij9_out
+    ) -> ij9_time
+#    uncomment for debugging
+#    list(logLik=logLik(ij9_out),time=ij9_time["elapsed"],ij9_out)
+    list(logLik=logLik(ij9_out),time=ij9_time["elapsed"])
+  } -> ij9_list
+})
+ij9_jobs <- ij_jobs
+ij9_jobs$logLik <- vapply(ij9_list,function(x)x$logLik,numeric(1))
+ij9_jobs$time <- vapply(ij9_list,function(x) x$time,numeric(1))
+ij9_jobs$Method <- ij9_Np
+@
+
+<<ij10,cache=F,echo=F>>=
+ij10 <- stew(file=paste0(ij_files_dir,"ij10.rda"),seed=186711,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles,
+        Nrep = ij10_I,
+        Np=ij10_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij10_out
+    ) -> ij10_time
+#    uncomment for debugging
+#    list(logLik=logLik(ij10_out),time=ij10_time["elapsed"],ij10_out)
+    list(logLik=logLik(ij10_out),time=ij10_time["elapsed"])
+  } -> ij10_list
+})
+ij10_jobs <- ij_jobs
+ij10_jobs$logLik <- vapply(ij10_list,function(x)x$logLik,numeric(1))
+ij10_jobs$time <- vapply(ij10_list,function(x) x$time,numeric(1))
+ij10_jobs$Method <- ij10_Np
+@
+
+<<ij11,cache=F,echo=F>>=
+ij11 <- stew(file=paste0(ij_files_dir,"ij11.rda"),seed=186711,{
+  foreach(job=iter(ij_jobs,"row")) %do% {
+    system.time(
+      abf(ij_measles,
+        Nrep = ij11_I,
+        Np=ij11_Np,
+        nbhd = ij_nbhd, tol=ij_tol) -> ij11_out
+    ) -> ij11_time
+#    uncomment for debugging
+#    list(logLik=logLik(ij11_out),time=ij11_time["elapsed"],ij11_out)
+    list(logLik=logLik(ij11_out),time=ij11_time["elapsed"])
+  } -> ij11_list
+})
+ij11_jobs <- ij_jobs
+ij11_jobs$logLik <- vapply(ij11_list,function(x)x$logLik,numeric(1))
+ij11_jobs$time <- vapply(ij11_list,function(x) x$time,numeric(1))
+ij11_jobs$Method <- ij11_Np
+@
+
+<<ij_loglik_plot, echo=F, fig.height=4, fig.width=4, out.width="3.5in", fig.align='center', fig.cap = paste('log likelihood estimates for simulated data from the measles model using {\\ABF}, with varying algorithmic parameters.')>>=
+
+ij_results <- rbind(ij1_jobs,ij2_jobs,ij3_jobs,ij4_jobs,ij5_jobs,ij6_jobs,ij7_jobs,ij8_jobs,ij9_jobs,ij10_jobs,ij11_jobs)
+#ij_results <- rbind(ij1_jobs,ij2_jobs,ij3_jobs)
+ij_results$logLik_per_unit <- ij_results$logLik/ij_results$U
+ij_results$logLik_per_obs <- ij_results$logLik_per_unit/ij_N
+
+save(file=paste0(ij_files_dir,"ij_results.rda"),ij_results)
+
+ij_max <- max(ij_results$logLik_per_obs)
+
+#ggplot(ij_results,mapping = aes(x = Method, y = logLik_per_obs)) +
+#  geom_point() +
+#  coord_cartesian(ylim=c(ij_max-2,ij_max))+
+#  ylab("log likelihood per observation")
+# qplot(Method,logLik,data=ij_results,geom="boxplot")
+qplot(Method,logLik_per_obs,data=ij_results,log="x",
+  ylab="log likelihood per unit per time",
+  xlab="Number of particles per replicate")+stat_summary(fun=mean, geom="line")
+
+#ggplot(bm_results,mapping = aes(x = U, y = logLik_per_obs, group=Method,color=Method)) +
+#  geom_point() +
+#  stat_summary(fun=mean, geom="line") +
+#  coord_cartesian(ylim=c(bm_max-0.35,bm_max))+
+#  ylab("log likelihood per unit per time")
+
+@
+
+It is necessary in practice to decide whether computational resources are best directed toward a large number of replicates, $\Rep$, or a large number of particles per replicate, $\Np$.
+Computational effort for ABF and ABF-IR is approximately proportional to $\Np\Rep$, and UBF corresponds to ABF with $\Np=1$.
+Suitable algorithmic parameters may depend on the model under consideration, and here we consider resource allocation for the measles model studied in the main text, using simulated data with $U=\Sexpr{ij_U}$ and $N=\Sexpr{ij_N}$.
+From Figure~\MainFigureMeaslesSlice, we know that this implementation of the model is well suited to ABF and UBF.
+ABF-IR performs less well, and the weak performance of GIRF suggests that the weakness may be due to an inadequate guide function.
+Figure~\ref{fig:ij_loglik_plot} investigates the tradeoff between UBF and ABF by plotting evaluated log likelihood against $\Rep$ with $\Np$ chosen to give approximately a constant computational effort.
+Algorithmic settings and run times are reported in Table~\ref{tab:ij}.
+For our implementation, choosing $\Np$ very low and $\Rep$ correspondingly high led to greater computation time, perhaps because the code was written to parallelize nicely when $\Np$ is relatively large.
+
+We interpret the bimodal curve as follows.
+When ABF is carried out with an inadequate number of particles for each bootstrap replicate, the algorithm cannot make a good representation of a draw from the adapted distribution.
+In that case, there is an advantage to using UBF, which does not attempt to carry out adapted simulation.
+For very small numbers of particles per replicate, the ABF algorithm behaves like a not-quite-properly-weighted version of UBF.
+Large numbers of particles per replicate presumably lead to improved Monte Carlo representation of draws from the adapted distribution, but computational cost constraints prevent combining this with a large number of replicates.
+We see a mode around 500 particles per replicate where ABF out-performs UBF.
+On this problem, UBF is relatively successful, presumably because the measles dynamics in each city are strongly attracted toward relatively few stable cycles (annual epidemics, or peaks in odd years, or peaks in even years) and a tractable number of simulations can represent all these scenarios.
+The Lorenz model of Sec.~\ref{sec:lorenz} provides an alternative situation, where adaptation has more advantages.
+Also, the plug-and-play guide function appears to operate successfully in this case, as evidenced by relatively strong performance from ABF-IR and GIRF.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{ccc}
+\hline
+  $\Np$ & $\Island$ & time\\
+\hline
+\Sexpr{ij1_Np} & \Sexpr{ij1_I}  & \Sexpr{myround(sum(ij1_jobs$time)/60,1)} \\
+\Sexpr{ij2_Np} & \Sexpr{ij2_I}  & \Sexpr{myround(sum(ij2_jobs$time)/60,1)} \\
+\Sexpr{ij3_Np} & \Sexpr{ij3_I}  & \Sexpr{myround(sum(ij3_jobs$time)/60,1)} \\
+\Sexpr{ij4_Np} & \Sexpr{ij4_I}  & \Sexpr{myround(sum(ij4_jobs$time)/60,1)} \\
+\Sexpr{ij5_Np} & \Sexpr{ij5_I}  & \Sexpr{myround(sum(ij5_jobs$time)/60,1)} \\
+\Sexpr{ij6_Np} & \Sexpr{ij6_I}  & \Sexpr{myround(sum(ij6_jobs$time)/60,1)} \\
+\Sexpr{ij7_Np} & \Sexpr{ij7_I}  & \Sexpr{myround(sum(ij7_jobs$time)/60,1)} \\
+\Sexpr{ij8_Np} & \Sexpr{ij8_I}  & \Sexpr{myround(sum(ij8_jobs$time)/60,1)} \\
+\Sexpr{ij9_Np} & \Sexpr{ij9_I}  & \Sexpr{myround(sum(ij9_jobs$time)/60,1)} \\
+\Sexpr{ij11_Np} & \Sexpr{ij11_I}  & \Sexpr{myround(sum(ij11_jobs$time)/60,1)} \\
+\Sexpr{ij10_Np} & \Sexpr{ij10_I}  & \Sexpr{myround(sum(ij10_jobs$time)/60,1)} \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Bootstrap replications, $\Rep$, particles per replicate, $\Np$, and computational time (total minutes for the five points) for the results presented in Figure~\ref{fig:ij_loglik_plot}}\label{tab:ij}
+\end{table}
+
+\clearpage
+
+%%%%%%%%%%%% uuuuuuuuu
+
+\section{\secTitleSpace Varying Monte Carlo effort for measles with fixed $\Unit$}
+
+<<fixedU_settings,cache=FALSE,echo=F>>=
+fixedU_files_dir <- paste0("fixedU_",fixedU_run_level,"/")
+if(!dir.exists(fixedU_files_dir)) dir.create(fixedU_files_dir)
+
+stew(file=paste0(fixedU_files_dir,"fixedU_settings.rda"),{
+  fixedU_tol <- 1e-300
+  fixedU_cores <- cores
+  fixedU_nbhd <- function(object, time, unit) {
+      nbhd_list <- list()
+      if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+      if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+      return(nbhd_list)
+  }
+  if(fixedU_run_level==1){ #######
+    fixedU_effort <- c(4,3,2,1)
+    fixedU_replicates <- 2
+    fixedU_enkf_Np <- c(10,10,10,10)
+    fixedU_pfilter_Np <- c(10,10,10,10)
+    fixedU_bpf_Np <- c(10,10,10,10)
+    fixedU_bpf_units_per_block <- 1
+    fixedU_girf_Np <- c(10,10,10,10)
+    fixedU_girf_Nguide <- c(10,10,10,10)
+    fixedU_girf_lookahead <- 1
+    fixedU_abf_I <- c(10,10,10,10)
+    fixedU_abf_Np <- c(10,10,10,10)
+    fixedU_ubf_I <- c(10,10,10,10)
+    fixedU_abfir_I <- c(10,10,10,10)
+    fixedU_abfir_Np <- c(10,10,10,10)
+  } else if(fixedU_run_level==2){
+    fixedU_effort <- c(4,3,2,1)
+    fixedU_replicates <- 3
+    fixedU_enkf_Np <- c(1600, 800, 400, 200)
+    fixedU_pfilter_Np <- c(1600, 800, 400, 200)
+    fixedU_bpf_Np <- c(1600, 800, 400, 200)
+    fixedU_bpf_units_per_block <- 1
+    fixedU_girf_Np <- c(800, 800, 400, 200)
+    fixedU_girf_Nguide <- c(40,20,20,20)
+    fixedU_girf_lookahead <- 1
+    fixedU_abf_I <- c(200,200,100,50)
+    fixedU_abf_Np <- c(20,10,10,10)
+    fixedU_ubf_I <- c(1600, 800, 400, 200)
+    fixedU_abfir_I <- c(200,200,100,50)
+    fixedU_abfir_Np <- c(20,10,10,10)
+  } else if(fixedU_run_level==3){
+    fixedU_effort <- c(10,9,8,7,6,5,4,3,2,1)
+    fixedU_replicates <- 3
+    fixedU_enkf_Np <- c(51200,25600,12800,6400,3200,1600,800,400,200,100)
+    fixedU_pfilter_Np <- c(51200,25600,12800,6400,3200,1600,800,400,200,100)
+    fixedU_bpf_Np <- c(51200,25600,12800,6400,3200,1600,800,400,200,100)
+    fixedU_bpf_units_per_block <- 2
+    fixedU_girf_Np <- c(1280,640,320,320,160,160,80,40,20,10)
+    fixedU_girf_Nguide <- c(40,40,40,20,20,10,10,10,10,10)
+    fixedU_girf_lookahead <- 1
+    fixedU_abf_I <- c(1280,640,320,320,160,160,80,40,20,10)
+    fixedU_abf_Np <- c(40,40,40,20,20,10,10,10,10,10)
+    fixedU_ubf_I <- c(51200,25600,12800,6400,3200,1600,800,400,200,100)
+    fixedU_abfir_I <- c(1280,640,320,320,160,160,80,40,20,10)
+    fixedU_abfir_Np <- c(40,40,40,20,20,10,10,10,10,10)
+  } else if(fixedU_run_level==4){
+    fixedU_effort <- c(10,9,8,7,6,5,4,3,2,1)
+    fixedU_replicates <- 5
+    fixedU_enkf_Np <- c(51200,25600,12800,6400,3200,1600,800,400,200,100)
+    fixedU_pfilter_Np <- c(51200,25600,12800,6400,3200,1600,800,400,200,100)
+    fixedU_bpf_Np <- c(51200,25600,12800,6400,3200,1600,800,400,200,100)
+    fixedU_bpf_units_per_block <- 2
+    fixedU_girf_Np <- c(1280,640,320,320,160,160,80,40,20,10)
+    fixedU_girf_Nguide <- c(40,40,40,20,20,10,10,10,10,10)
+    fixedU_girf_lookahead <- 1
+    fixedU_abf_I <- c(1280,640,320,320,160,160,80,40,20,10)
+    fixedU_abf_Np <- c(40,40,40,20,20,10,10,10,10,10)
+    fixedU_ubf_I <- c(51200,25600,12800,6400,3200,1600,800,400,200,100)
+    fixedU_abfir_I <- c(1280,640,320,320,160,160,80,40,20,10)
+    fixedU_abfir_Np <- c(40,40,40,20,20,10,10,10,10,10)
+  } else if(fixedU_run_level==5){
+  } 
+})
+fixedU_jobs <- expand.grid(effort=fixedU_effort,reps=1:fixedU_replicates)
+@
+
+<<fixedU_spatPomp,eval=T,echo=F>>=
+fixedU_spatPomp <- stew(file=paste0(fixedU_files_dir,"fixedU_spatPomp.rda"),{
+  fixedU_model_dir <- "../mscale_5/"
+  # note: the ij simulation study used ../measlesModel/measles_spatPomp.rda instea
+  load(file=paste0(fixedU_model_dir,"mscale_spatPomp.rda"))
+  # U = c(40,36,32,28,24,20,16,14,12,10,8,6,4,2) so mscale_list[[10]] implies U = 10
+  fixedU_measles <- mscale_list[[6]]
+})
+@
+
+The theoretical results in our manuscript concern sub-exponential computational requirements as $\Unit$ increases.
+However, the behavior as effort varies for fixed $\Unit$ has practical consequences for the problems on which different methods are applicable.
+We therefore investigated empirically how the variance of our estimates scales with effort across multiple methods, on the measles model.
+We fixed our data to be the \Sexpr{length(fixedU_measles@unit_names)}-city measles data from Figure~{\MainFigureMeaslesScaling} in the main text.
+Table~\ref{tab:meffortscale} below shows the schedule of  Monte Carlo effort used to compute the likelihood estimates in Figure~\ref{fig:fixedU-loglik-plot}.
+For ABF and ABF-IR we vary $\Rep\Np$; for UBF we vary $\Rep$; for BPF and EnKF we vary $\Np$.
+
+If computational considerations prevented an ensemble size larger than 100, EnKF or UBF would be a good choice on this problem.
+EnKF cannot benefit much from a larger ensemble size, since its limitation is the Gaussian approximation in its update rule, rather than Monte Carlo variability. 
+For an ensemble size of 1000 or more, the bagged filters and block particle filter provide a substantial improvement over EnKF.
+ABF-IR is relatively computationally demanding on this problem, perhaps because further research is needed into the choice of guide function.
+
+<<fixedU_abf,cache=F,echo=F>>=
+fixedU_abf <- stew(file=paste0(fixedU_files_dir,"fixedU_abf.rda"),seed=844423,{
+  foreach(i=1:nrow(fixedU_jobs)) %do% {
+      system.time(
+        abf(fixedU_measles, 
+          Nrep = fixedU_abf_I[which(fixedU_jobs$effort[i] == fixedU_effort)],
+          Np=fixedU_abf_Np[which(fixedU_jobs$effort[i] == fixedU_effort)],
+          nbhd = fixedU_nbhd,
+          tol=fixedU_tol) -> fixedU_abf_out 
+      ) -> fixedU_abf_time
+      list(logLik=logLik(fixedU_abf_out),time=fixedU_abf_time["elapsed"])
+    # }
+  } -> fixedU_abf_list
+})
+fixedU_jobs$abf_logLik <- vapply(fixedU_abf_list,function(x)x$logLik,numeric(1))
+fixedU_jobs$abf_time <- vapply(fixedU_abf_list,function(x)x$time,numeric(1))
+@
+
+<<fixedU_ubf,cache=F,echo=F>>=
+fixedU_ubf <- stew(file=paste0(fixedU_files_dir,"fixedU_ubf.rda"),seed=844423,{
+  foreach(i=1:nrow(fixedU_jobs)) %do% {
+      system.time(
+        abf(fixedU_measles, 
+          Nrep = fixedU_ubf_I[which(fixedU_jobs$effort[i] == fixedU_effort)],
+          Np=1,
+          nbhd = fixedU_nbhd,
+          tol=fixedU_tol) -> fixedU_ubf_out 
+      ) -> fixedU_ubf_time
+      list(logLik=logLik(fixedU_ubf_out),time=fixedU_ubf_time["elapsed"])
+    # }
+  } -> fixedU_ubf_list
+})
+fixedU_jobs$ubf_logLik <- vapply(fixedU_ubf_list,function(x)x$logLik,numeric(1))
+fixedU_jobs$ubf_time <- vapply(fixedU_ubf_list,function(x)x$time,numeric(1))
+@
+
+<<fixedU_abfir,cache=F,echo=F>>=
+fixedU_abfir <- stew(file=paste0(fixedU_files_dir,"fixedU_abfir.rda"),seed=844423,{
+  foreach(i=1:nrow(fixedU_jobs)) %do% {
+      system.time(
+        abfir(fixedU_measles, 
+          Nrep = fixedU_abfir_I[which(fixedU_jobs$effort[i] == fixedU_effort)],
+          Np=fixedU_abfir_Np[which(fixedU_jobs$effort[i] == fixedU_effort)],
+          Ninter=length(fixedU_measles@unit_names)/2,
+          nbhd = fixedU_nbhd,
+          tol=fixedU_tol) -> fixedU_abfir_out 
+      ) -> fixedU_abfir_time
+      list(logLik=logLik(fixedU_abfir_out),time=fixedU_abfir_time["elapsed"])
+    # }
+  } -> fixedU_abfir_list
+})
+fixedU_jobs$abfir_logLik <- vapply(fixedU_abfir_list,function(x)x$logLik,numeric(1))
+fixedU_jobs$abfir_time <- vapply(fixedU_abfir_list,function(x)x$time,numeric(1))
+@
+
+<<fixedU_girf,cache=F,echo=F>>=
+fixedU_girf <- stew(file=paste0(fixedU_files_dir,"fixedU_girf.rda"),seed=844424,{
+  foreach(i=1:nrow(fixedU_jobs)) %do% {
+      system.time(
+        girf(fixedU_measles, 
+          Np=fixedU_girf_Np[which(fixedU_jobs$effort[i] == fixedU_effort)],
+          Ninter=length(fixedU_measles@unit_names),
+          Nguide=fixedU_girf_Nguide[which(fixedU_jobs$effort[i] == fixedU_effort)],
+          kind="moment",
+          tol=fixedU_tol) -> fixedU_girf_out 
+      ) -> fixedU_girf_time
+      list(logLik=logLik(fixedU_girf_out),time=fixedU_girf_time["elapsed"])
+    # }
+  } -> fixedU_girf_list
+})
+fixedU_jobs$girf_logLik <- vapply(fixedU_girf_list,function(x)x$logLik,numeric(1))
+fixedU_jobs$girf_time <- vapply(fixedU_girf_list,function(x)x$time,numeric(1))
+@
+
+<<fixedU_pfilter,cache=F,echo=F>>=
+fixedU_pfilter <- stew(file=paste0(fixedU_files_dir,"fixedU_pfilter.rda"),seed=844424,{
+  foreach(i=1:nrow(fixedU_jobs)) %do% {
+      system.time(
+        pfilter(fixedU_measles, 
+          Np=fixedU_pfilter_Np[which(fixedU_jobs$effort[i] == fixedU_effort)]) -> fixedU_pfilter_out 
+      ) -> fixedU_pfilter_time
+      list(logLik=logLik(fixedU_pfilter_out),time=fixedU_pfilter_time["elapsed"])
+    # }
+  } -> fixedU_pfilter_list
+})
+fixedU_jobs$pfilter_logLik <- vapply(fixedU_pfilter_list,function(x)x$logLik,numeric(1))
+fixedU_jobs$pfilter_time <- vapply(fixedU_pfilter_list,function(x)x$time,numeric(1))
+@
+
+<<fixedU_bpfilter,cache=F,echo=F>>=
+fixedU_bpfilter <- stew(file=paste0(fixedU_files_dir,"fixedU_bpfilter.rda"),seed=844424,{
+  foreach(i=1:nrow(fixedU_jobs)) %do% {
+      system.time(
+        bpfilter(fixedU_measles, 
+          Np=fixedU_bpf_Np[which(fixedU_jobs$effort[i] == fixedU_effort)],
+          block_size=fixedU_bpf_units_per_block,
+          ) -> fixedU_bpfilter_out 
+      ) -> fixedU_bpfilter_time
+      list(logLik=logLik(fixedU_bpfilter_out),time=fixedU_bpfilter_time["elapsed"])
+    # }
+  } -> fixedU_bpfilter_list
+})
+fixedU_jobs$bpf_logLik <- vapply(fixedU_bpfilter_list,function(x)x$logLik,numeric(1))
+fixedU_jobs$bpf_time <- vapply(fixedU_bpfilter_list,function(x)x$time,numeric(1))
+@
+
+<<fixedU_enkf,cache=F,echo=F>>=
+fixedU_enkf <- stew(file=paste0(fixedU_files_dir,"fixedU_enkf.rda"),seed=844424,{
+  foreach(i=1:nrow(fixedU_jobs)) %do% {
+      system.time(
+        enkf(fixedU_measles, 
+          Np=fixedU_enkf_Np[which(fixedU_jobs$effort[i] == fixedU_effort)]) -> fixedU_enkf_out 
+      ) -> fixedU_enkf_time
+      list(logLik=logLik(fixedU_enkf_out),time=fixedU_enkf_time["elapsed"])
+    # }
+  } -> fixedU_enkf_list
+})
+fixedU_jobs$enkf_logLik <- vapply(fixedU_enkf_list,function(x)x$logLik,numeric(1))
+fixedU_jobs$enkf_time <- vapply(fixedU_enkf_list,function(x)x$time,numeric(1))
+@
+
+<<fixedU_extract_times,echo=F>>=
+fixedU_ubf_times <-  myround(fixedU_jobs %>% dplyr::group_by(effort) %>% dplyr::summarize(meanubf = mean(ubf_time*fixedU_cores/60)) %>% dplyr::pull(meanubf),1)
+fixedU_abf_times <-  myround(fixedU_jobs %>% dplyr::group_by(effort) %>% dplyr::summarize(meanabf = mean(abf_time*fixedU_cores/60)) %>% dplyr::pull(meanabf),1)
+fixedU_abfir_times <-  myround(fixedU_jobs %>% dplyr::group_by(effort) %>% dplyr::summarize(meanabfir = mean(abfir_time*fixedU_cores/60)) %>% dplyr::pull(meanabfir),1)
+fixedU_enkf_times <-  myround(fixedU_jobs %>% dplyr::group_by(effort) %>% dplyr::summarize(meanenkf = mean(enkf_time*fixedU_cores/60)) %>% dplyr::pull(meanenkf),1)
+fixedU_bpf_times <-  myround(fixedU_jobs %>% dplyr::group_by(effort) %>% dplyr::summarize(meanbpf = mean(bpf_time*fixedU_cores/60)) %>% dplyr::pull(meanbpf),1)
+@
+
+\begin{table}
+\begin{center}
+\begin{tabular}{|p{0.22\linewidth}|p{0.1\linewidth}|p{0.1\linewidth}|p{0.1\linewidth}|p{0.1\linewidth}|p{0.1\linewidth}|}
+  \hline & 
+  {\UBF}  & 
+  {\ABF} & 
+  {\ABFIR}  & 
+  EnKF &
+  BPF 
+\\ 
+  \hline
+  Monte Carlo &
+  \Sexpr{fixedU_ubf_I[10]} &
+  \Sexpr{fixedU_abf_I[10]} $\times$ \Sexpr{fixedU_abf_Np[10]} &
+  \Sexpr{fixedU_abfir_I[10]} $\times$ \Sexpr{fixedU_abfir_Np[10]} &
+  \Sexpr{fixedU_enkf_Np[10]} &
+  \Sexpr{fixedU_bpf_Np[10]} 
+\\
+   sample size, &
+  \Sexpr{fixedU_ubf_I[9]} &
+  \Sexpr{fixedU_abf_I[9]} $\times$ \Sexpr{fixedU_abf_Np[9]} &
+  \Sexpr{fixedU_abfir_I[9]} $\times$ \Sexpr{fixedU_abfir_Np[9]} &
+  \Sexpr{fixedU_enkf_Np[9]} &
+  \Sexpr{fixedU_bpf_Np[9]} 
+\\
+  $\Rep\times\Np$ &
+  \Sexpr{fixedU_ubf_I[8]} &
+  \Sexpr{fixedU_abf_I[8]} $\times$ \Sexpr{fixedU_abf_Np[8]} &
+  \Sexpr{fixedU_abfir_I[8]} $\times$ \Sexpr{fixedU_abfir_Np[8]} &
+  \Sexpr{fixedU_enkf_Np[8]} &
+  \Sexpr{fixedU_bpf_Np[8]} 
+\\
+   &
+  \Sexpr{fixedU_ubf_I[7]} &
+  \Sexpr{fixedU_abf_I[7]} $\times$ \Sexpr{fixedU_abf_Np[7]} &
+  \Sexpr{fixedU_abfir_I[7]} $\times$ \Sexpr{fixedU_abfir_Np[7]} &
+  \Sexpr{fixedU_enkf_Np[7]} &
+  \Sexpr{fixedU_bpf_Np[7]} 
+\\
+   &
+  \Sexpr{fixedU_ubf_I[6]} &
+  \Sexpr{fixedU_abf_I[6]} $\times$ \Sexpr{fixedU_abf_Np[6]} &
+  \Sexpr{fixedU_abfir_I[6]} $\times$ \Sexpr{fixedU_abfir_Np[6]} &
+  \Sexpr{fixedU_enkf_Np[6]} &
+  \Sexpr{fixedU_bpf_Np[6]} 
+\\
+   &
+  \Sexpr{fixedU_ubf_I[5]} &
+  \Sexpr{fixedU_abf_I[5]} $\times$ \Sexpr{fixedU_abf_Np[5]} &
+  \Sexpr{fixedU_abfir_I[5]} $\times$ \Sexpr{fixedU_abfir_Np[5]} &
+  \Sexpr{fixedU_enkf_Np[5]} &
+  \Sexpr{fixedU_bpf_Np[5]} 
+\\
+  &
+  \Sexpr{fixedU_ubf_I[4]} &
+  \Sexpr{fixedU_abf_I[4]} $\times$ \Sexpr{fixedU_abf_Np[4]} &
+  \Sexpr{fixedU_abfir_I[4]} $\times$ \Sexpr{fixedU_abfir_Np[4]} &
+  \Sexpr{fixedU_enkf_Np[4]} &
+  \Sexpr{fixedU_bpf_Np[4]} 
+\\
+   &
+  \Sexpr{fixedU_ubf_I[3]} &
+  \Sexpr{fixedU_abf_I[3]} $\times$ \Sexpr{fixedU_abf_Np[3]} &
+  \Sexpr{fixedU_abfir_I[3]} $\times$ \Sexpr{fixedU_abfir_Np[3]} &
+  \Sexpr{fixedU_enkf_Np[3]} &
+  \Sexpr{fixedU_bpf_Np[3]} 
+\\
+  &
+  \Sexpr{fixedU_ubf_I[2]} &
+  \Sexpr{fixedU_abf_I[2]} $\times$ \Sexpr{fixedU_abf_Np[2]} &
+  \Sexpr{fixedU_abfir_I[2]} $\times$ \Sexpr{fixedU_abfir_Np[2]} &
+  \Sexpr{fixedU_enkf_Np[2]} &
+  \Sexpr{fixedU_bpf_Np[2]} 
+\\
+  &
+  \Sexpr{fixedU_ubf_I[1]} &
+  \Sexpr{fixedU_abf_I[1]} $\times$ \Sexpr{fixedU_abf_Np[1]} &
+  \Sexpr{fixedU_abfir_I[1]} $\times$ \Sexpr{fixedU_abfir_Np[1]} &
+  \Sexpr{fixedU_enkf_Np[1]} &
+  \Sexpr{fixedU_bpf_Np[1]} 
+\\
+  \hline
+  intermediate steps, $S$ & 
+  --- & 
+  --- &
+  $\Unit/2$ & 
+  %$\Unit$ & 
+  --- &
+  %--- &
+  ---
+\\
+\hline
+  effort, in core mins.  \rule{0mm}{4.5mm}  &
+  \Sexpr{fixedU_ubf_times[1]} &
+  \Sexpr{fixedU_abf_times[1]} &
+  \Sexpr{fixedU_abfir_times[1]} &
+  \Sexpr{fixedU_enkf_times[1]} &
+  \Sexpr{fixedU_bpf_times[1]} 
+\\
+   &
+  \Sexpr{fixedU_ubf_times[2]} &
+  \Sexpr{fixedU_abf_times[2]} &
+  \Sexpr{fixedU_abfir_times[2]} &
+  \Sexpr{fixedU_enkf_times[2]} &
+  \Sexpr{fixedU_bpf_times[2]}  
+\\
+   &
+  \Sexpr{fixedU_ubf_times[3]} &
+  \Sexpr{fixedU_abf_times[3]} &
+  \Sexpr{fixedU_abfir_times[3]} &
+  \Sexpr{fixedU_enkf_times[3]} &
+  \Sexpr{fixedU_bpf_times[3]}  
+\\
+   &
+  \Sexpr{fixedU_ubf_times[4]} &
+  \Sexpr{fixedU_abf_times[4]} &
+  \Sexpr{fixedU_abfir_times[4]} &
+  \Sexpr{fixedU_enkf_times[4]} &
+  \Sexpr{fixedU_bpf_times[4]}  
+\\
+   &
+  \Sexpr{fixedU_ubf_times[5]} &
+  \Sexpr{fixedU_abf_times[5]} &
+  \Sexpr{fixedU_abfir_times[5]} &
+  \Sexpr{fixedU_enkf_times[5]} &
+  \Sexpr{fixedU_bpf_times[5]}  
+\\
+   &
+  \Sexpr{fixedU_ubf_times[6]} &
+  \Sexpr{fixedU_abf_times[6]} &
+  \Sexpr{fixedU_abfir_times[6]} &
+  \Sexpr{fixedU_enkf_times[6]} &
+  \Sexpr{fixedU_bpf_times[6]}  
+\\
+   &
+  \Sexpr{fixedU_ubf_times[7]} &
+  \Sexpr{fixedU_abf_times[7]} &
+  \Sexpr{fixedU_abfir_times[7]} &
+  \Sexpr{fixedU_enkf_times[7]} &
+  \Sexpr{fixedU_bpf_times[7]}  
+\\
+   &
+  \Sexpr{fixedU_ubf_times[8]} &
+  \Sexpr{fixedU_abf_times[8]} &
+  \Sexpr{fixedU_abfir_times[8]} &
+  \Sexpr{fixedU_enkf_times[8]} &
+  \Sexpr{fixedU_bpf_times[8]}  
+\\
+   &
+  \Sexpr{fixedU_ubf_times[9]} &
+  \Sexpr{fixedU_abf_times[9]} &
+  \Sexpr{fixedU_abfir_times[9]} &
+  \Sexpr{fixedU_enkf_times[9]} &
+  \Sexpr{fixedU_bpf_times[9]}  
+\\
+   &
+  \Sexpr{fixedU_ubf_times[10]} &
+  \Sexpr{fixedU_abf_times[10]} &
+  \Sexpr{fixedU_abfir_times[10]} &
+  \Sexpr{fixedU_enkf_times[10]} &
+  \Sexpr{fixedU_bpf_times[10]}  
+\\
+\hline
+\end{tabular}
+\caption{Algorithmic parameters and run times (in core-minutes) that correspond to the effort levels shown in Figure \ref{fig:fixedU-loglik-plot}.}
+\label{tab:meffortscale}
+\end{center}
+\end{table}
+
+<<fixedU-loglik-plot, echo=F,  fig.height=3.5, fig.width=5, out.width="5in", fig.cap = paste("Exploring the efficiency of the bagged filter methods with varying Monte Carlo effort.
+For each of ", length(fixedU_effort)," levels of computational effort, we run our methods on a simulated measles data set from ", length(fixedU_measles@unit_names)," cities.
+The variability of our estimates decreases as effort increases and we observe that the unadapted bagged filter has a competitive advantage over BPF in a low computational effort regime.
+Methods that perform less well are clipped off to focus on the best methods for this example.
+The algorithmic parameters and run times are given in Table \\ref{tab:meffortscale}.")>>=
+
+# put output in tall format for plotting
+# fixedU_methods <- c("ABF", "ABF-IR","UBF","GIRF","PF","BPF","EnKF")
+fixedU_methods <- c("ABF", "ABF-IR", "UBF", "BPF", "EnKF")
+fixedU_results <- data.frame(
+  Method=rep(fixedU_methods,each=nrow(fixedU_jobs)),
+  logLik=c(
+    fixedU_jobs$abf_logLik,
+    fixedU_jobs$abfir_logLik,
+    fixedU_jobs$ubf_logLik,
+    # fixedU_jobs$girf_logLik,
+    # fixedU_jobs$pfilter_logLik,
+    fixedU_jobs$bpf_logLik,
+    fixedU_jobs$enkf_logLik
+  ),
+  effortLevel=rep(fixedU_jobs$effort,reps=length(fixedU_methods))
+)
+fixedU_point_perturbation <- 0.25
+fixedU_results <- fixedU_results %>% 
+  dplyr::mutate(effortLevel2 = 100*2^(effortLevel - 1))
+fixedU_results$effortLevel3 <- fixedU_results$effortLevel2+
+  rep( fixedU_point_perturbation*seq(from=-1,to=1,length=length(fixedU_methods)),
+    each=nrow(fixedU_jobs))
+fixedU_results$logLik_per_unit <- fixedU_results$logLik/length(fixedU_measles@unit_names)
+fixedU_results$logLik_per_obs <- fixedU_results$logLik_per_unit/length(fixedU_measles@times)
+# fixedU_results <- fixedU_results %>% dplyr::mutate(effortCat = paste('LEVEL',effortLevel,sep=''))
+save(file=paste0(fixedU_files_dir,"fixedU_results.rda"),fixedU_results,fixedU_jobs)
+
+fixedU_max <- max(fixedU_results$logLik_per_obs)
+cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+fixedU_linetype <- c(1,1,1,2,2)
+fixedU_shape <- c(1,2,4,2,4)
+fixedU_color <- cbPalette[c(2,3,4,5,7)]
+fixedU_results_plot <- fixedU_results %>% dplyr::filter(Method %in% c("ABF", "ABF-IR", "UBF", "BPF", "EnKF"))
+ggplot(fixedU_results_plot,
+       mapping = aes(x = effortLevel3, y = logLik_per_obs, group=Method,color=Method,linetype=Method,shape=Method)) +
+  scale_linetype_manual(values=fixedU_linetype) +
+  scale_shape_manual(values=fixedU_shape) +
+  scale_color_manual(values=fixedU_color)+
+  scale_x_continuous(trans = "log2",
+                     breaks = c(100, 400, 1600, 6400, 25600)
+                     #breaks = c(100,200,400,800,1600,3200,6400,12800,25600,51600),
+                     #labels = c(100,200,400,800,1600,3200,6400,12800,25600,51600)
+                     ) +
+  geom_point() +
+  stat_summary(fun=mean, geom="line") +
+  coord_cartesian(ylim=c(fixedU_max-3,fixedU_max))+
+  theme(legend.key.width = unit(1,"cm"))+
+  theme(axis.text.x = element_text(size = 8),
+         axis.text.y = element_text(size = 12),
+         axis.title.x = element_text(size = 14),
+         axis.title.y = element_text(size = 14),
+         panel.grid.major = element_blank(),
+         panel.grid.minor = element_blank(),
+         panel.background = element_blank(),
+         panel.border = element_rect(colour = "black", fill=NA, size=0.8),
+         axis.line = element_line(colour = "black"))+
+  ylab("Log likelihood per unit per time")+
+  xlab("Monte Carlo sample size")
+
+@
+
+\clearpage
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{\secTitleSpace A constraint that breaks the block particle filter}
+
+<<con_setup,echo=F>>=
+source("con.R")
+
+con_run_level <- 3
+library(doParallel)
+library(doRNG)
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores)) cores <- detectCores()  
+registerDoParallel(cores)
+
+con_files_dir <- paste0("con_",con_run_level,"/")
+if(!dir.exists(con_files_dir)) dir.create(con_files_dir)
+
+stew(file=paste0(con_files_dir,"con_settings.rda"),{
+
+  # copy variables that should be included in the stew
+  con_run_level <- con_run_level 
+  con_cores <- cores
+
+  con_tol <- 1e-300
+
+  # run_level 1 for debugging; 2 for quick run; 3 for long run; 4 for production(?)
+
+  if(con_run_level==1){
+    con_U <- 5
+    con_N <- 10
+    con_replicates <- 2 # number of Monte Carlo replicates
+    con_pfilter_Np <- 100
+    con_abf_Nrep <- 3
+    con_abf_Np_per_replicate <- 10
+    con_ubf_Nrep <- 3
+    con_bpf_Np <- 50
+    con_bpf_units_per_block <- 1
+    con_enkf_Np <- 20
+  } else if(con_run_level==2){
+    con_U <- 5
+    con_N <- 10
+    con_replicates <- 5 # number of Monte Carlo replicates
+    con_pfilter_Np <- 1000
+    con_abf_Nrep <- 50
+    con_abf_Np_per_replicate <- 50
+    con_ubf_Nrep <- 1000
+    con_bpf_Np <- 1000
+    con_bpf_units_per_block <- 1
+    con_enkf_Np <-1000
+  }
+ else if(con_run_level==3){
+    con_U <- 5
+    con_N <- 10
+    con_replicates <- 5 # number of Monte Carlo replicates
+    con_pfilter_Np <- 10000
+    con_abf_Nrep <- 100
+    con_abf_Np_per_replicate <- 100
+    con_ubf_Nrep <- 10000
+    con_bpf_Np <- 10000
+    con_bpf_units_per_block <- 1
+    con_enkf_Np <-10000
+  }
+
+  set.seed(54321)
+  con1 <- con(U=con_U,N=con_N)
+})
+
+con_jobs <- data.frame(Rep=1:con_replicates)
+
+@
+
+<<con_nbhd,echo=F>>=
+con_nbhd <- function(object, time, unit) {
+  nbhd_list <- list()
+  if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+  if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+  return(nbhd_list)
+}
+@
+
+<<con_kf,cache=F,echo=F>>=
+con_kf <- stew(file=paste0(con_files_dir,"con_kf.rda"),seed=25487,{
+  con1_kf <- con1
+  foreach(i=seq_along(con_U),.combine=c)%do%{
+    con_dmat <- matrix(0,con_U,con_U)
+    for(u in 1:con_U) {
+      con_dmat[u,u] <- 1
+      for(v in 1:con_U) {
+        con_dmat[u,v] <- con_dmat[u,v] - 1/con_U
+      }
+    }
+    rootQ <- con_dmat * coef(con1_kf)["sigma"]
+    kf <- pomp:::kalmanFilter(
+      t=1:con_N,
+      y=obs(con1_kf),
+      X0=rinit(con1_kf),
+      A= diag(con_U),
+      Q= rootQ %*% rootQ,
+      C=diag(con_U),
+      R=diag(coef(con1_kf)["tau"]^2, nrow=con_U)
+    )$loglik
+  } -> con_kf_results
+})
+con_jobs$logLik <- rep(con_kf_results,times=con_replicates)
+@
+
+<<con_girf,cache=F,echo=F,eval=F>>=
+con_girf <- stew(file=paste0(con_files_dir,"con_girf.rda"),seed=5981724,{
+foreach(job=iter(con_jobs,"row")) %dopar% {
+    system.time(
+      girf(con_list[[job$U_id]],
+        Np=con_girf_Np,
+        Ninter = job$U,
+        kind="moment",
+#        kind="bootstrap",
+        lookahead = con_girf_lookahead,
+        Nguide = con_girf_nguide,
+        tol = con_tol) -> con_girf_out 
+    ) -> con_girf_time
+#    uncomment for debugging 
+#    list(logLik=logLik(con_girf_out),time=con_girf_time["elapsed"],con_girf_out)
+    list(logLik=logLik(con_girf_out),time=con_girf_time["elapsed"])
+  } -> con_girf_list
+})
+con_jobs$girf_logLik <- vapply(con_girf_list,function(x)x$logLik,numeric(1))
+con_jobs$girf_time <- vapply(con_girf_list,function(x) x$time,numeric(1))
+@
+
+<<con_abf,cache=F,echo=F>>=
+con_abf <- stew(file=paste0(con_files_dir,"con_abf.rda"),seed=844424,{
+  foreach(job=iter(con_jobs,"row")) %do% {
+    system.time(
+      abf(con1,
+        Nrep = con_abf_Nrep,
+        Np=con_abf_Np_per_replicate,
+        nbhd = con_nbhd, tol=con_tol) -> con_abf_out 
+    ) -> con_abf_time
+#    uncomment for debugging 
+#    list(logLik=logLik(con_abf_out),time=con_abf_time["elapsed"],con_abf_out)
+    list(logLik=logLik(con_abf_out),time=con_abf_time["elapsed"])
+  } -> con_abf_list
+})
+con_jobs$abf_logLik <- vapply(con_abf_list,function(x)x$logLik,numeric(1))
+con_jobs$abf_time <- vapply(con_abf_list,function(x) x$time,numeric(1))
+
+@
+
+<<con_ubf,cache=F,echo=F>>=
+ con_ubf <- stew(file=paste0(con_files_dir,"con_ubf.rda"),seed=844424,{
+  foreach(job=iter(con_jobs,"row")) %do% {  
+    system.time(
+      abf(con1, 
+        Nrep = con_ubf_Nrep,
+        Np=1,
+        nbhd = con_nbhd, tol=con_tol) -> con_ubf_out 
+    ) -> con_ubf_time
+    list(logLik=logLik(con_ubf_out),time=con_ubf_time["elapsed"])
+  } -> con_ubf_list
+})
+con_jobs$ubf_logLik <- vapply(con_ubf_list,function(x)x$logLik,numeric(1))
+con_jobs$ubf_time <- vapply(con_ubf_list,function(x) x$time,numeric(1))
+
+@
+
+<<con_abfir,cache=F,echo=F,eval=F>>=
+con_abfir <- stew(file=paste0(con_files_dir,"con_abfir.rda"),seed=53398,{
+  foreach(job=iter(con_jobs,"row")) %do% {
+    system.time(
+      abfir(con1,
+        Nrep = as.integer(con_abfir_Nrep),
+        Np=con_abfir_Np_per_replicate,
+        Ninter = as.integer(job$U/2),
+        nbhd = con_nbhd, tol=con_tol) -> con_abfir_out 
+    ) -> con_abfir_time
+    list(logLik=logLik(con_abfir_out),time=con_abfir_time["elapsed"])
+  } -> con_abfir_list
+})
+con_jobs$abfir_logLik <- vapply(con_abfir_list,function(x)x$logLik,numeric(1))
+con_jobs$abfir_time <- vapply(con_abfir_list,function(x) x$time,numeric(1))
+@
+
+<<con_pfilter,cache=F,echo=F>>=
+con_pfilter <- stew(file=paste0(con_files_dir,"con_pfilter.rda"),seed=53285,{
+  foreach(job=iter(con_jobs,"row")) %dopar% {
+    system.time(
+      pfilter(con1,
+        Np=con_pfilter_Np) -> con_pfilter_out
+    ) -> con_pfilter_time
+    list(logLik=logLik(con_pfilter_out),time=con_pfilter_time["elapsed"])
+  } -> con_pfilter_list
+})
+con_jobs$pfilter_logLik <- vapply(con_pfilter_list,function(x)x$logLik,numeric(1))
+con_jobs$pfilter_time <- vapply(con_pfilter_list,function(x) x$time,numeric(1))
+@
+
+<<con_bpf,cache=F,echo=F>>=
+con_bpf <- stew(file=paste0(con_files_dir,"con_bpf.rda"),seed=53285,{
+  foreach(job=iter(con_jobs,"row")) %dopar% {
+    system.time(
+      bpfilter(con1,
+        Np=con_bpf_Np,
+	block_size=con_bpf_units_per_block) -> con_bpf_out
+    ) -> con_bpf_time
+#
+# replace using logLik(con_bpf_out) when that method exists
+#
+    list(logLik=con_bpf_out@loglik,time=con_bpf_time["elapsed"])
+  } -> con_bpf_list
+})
+con_jobs$bpf_logLik <- vapply(con_bpf_list,function(x)x$logLik,numeric(1))
+con_jobs$bpf_time <- vapply(con_bpf_list,function(x) x$time,numeric(1))
+@
+
+<<con_enkf,cache=F,echo=F>>=
+con_enkf <- stew(file=paste0(con_files_dir,"con_enkf.rda"),seed=53285,{
+  foreach(job=iter(con_jobs,"row")) %dopar% {
+    system.time(
+      enkf(con1,
+        Np=con_enkf_Np) -> con_enkf_out
+    ) -> con_enkf_time
+    list(logLik=con_enkf_out@loglik,time=con_enkf_time["elapsed"])
+  } -> con_enkf_list
+})
+con_jobs$enkf_logLik <- vapply(con_enkf_list,function(x)x$logLik,numeric(1))
+con_jobs$enkf_time <- vapply(con_enkf_list,function(x) x$time,numeric(1))
+@
+
+A potential advantage of the bagged filters over the block particle filter is that the former have filter trajectories preserving constraints on the system, whereas the latter does not.
+Potentially, conservation constraints could be an important part of a coupled model: physical matter, or money, or members of a biological population can move from location to location but cannot so easily be spontaneously created.
+We present a toy example to demonstrate the issue.
+
+Suppose $\myvec{W}(t)=W_{1:\Unit}(t)$ is a collection of $\Unit$ independent standard Brownian motions, and let $\myvec{X}(t)$ be the Ito solution to a system of stochastic differential equations,
+\begin{equation} \label{model:constrained}
+dX_{\unit}(t) = \sigma\left(
+  dW_{\unit}(t) - \frac{1}{\Unit}\sum_{\altUnit=1}^{\Unit}dW_{\altUnit}(t)
+  \right) +  \sum_{\altUnit=1}^{\Unit} X_{\altUnit}(t) \, dt,
+\end{equation}
+for $\unit \in \seq{1}{\Unit}$, with initial condition $X_{\unit}(0)=0$.
+We consider additive Gaussian measurement error,
+\begin{equation}\label{model:constrained:measurement}
+Y_{\unit,\time}=X_{\unit}(\time)+ \tau \epsilon_{\unit,\time},
+\end{equation}
+where $\{\epsilon_{\unit,\time}\}$ is a collection of $\iid$ standard normal random variables.
+This toy model has stable trajectories constrained to the linear space $\sum_{\altUnit=1}^{\Unit} X_{\altUnit}(t)=0$, but trajectories diverge if that constraint is broken.
+A small demonstration is enough to make the point that the block particle filter can behave very poorly on this problem.
+Results are reported in Table~\ref{tab:constrained} with with $\Unit=5$, $\Time=10$, $\sigma=1$, $\tau=1$, a block size of 1 for BPF, neighborhoods for UBF and ABF consisting of two co-located lags, and an Euler solution to \eqref{model:constrained} with time step \Sexpr{con1@rprocess@delta.t}.
+Any block size less than a single block of size $\Unit$ gives the same qualitative conclusion of poor performance of BPF.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{|p{0.4\linewidth}|p{0.1\linewidth}|p{0.1\linewidth}|p{0.1\linewidth}|p{0.1\linewidth}|p{0.1\linewidth}|}
+  \hline & 
+  PF &
+  EnKF &
+  {\UBF}  & 
+  {\ABF} & 
+  BPF
+  \\
+  \hline 
+Log likelihood per observation &
+ \Sexpr{myround(logmeanexp(con_jobs$pfilter_logLik)/(con_U*con_N),2)} &
+ \Sexpr{myround(logmeanexp(con_jobs$enkf_logLik)/(con_U*con_N),2)} &
+ \Sexpr{myround(logmeanexp(con_jobs$ubf_logLik)/(con_U*con_N),2)} &
+ \Sexpr{myround(logmeanexp(con_jobs$abf_logLik)/(con_U*con_N),2)} &
+ \Sexpr{myround(logmeanexp(con_jobs$bpf_logLik)/(con_U*con_N),2)} 
+\\ 
+  Monte Carlo effort, $\Rep\times\Np$    \rule{0mm}{4.5mm} & 
+  \Sexpr{con_pfilter_Np} &
+  \Sexpr{con_enkf_Np} &
+  \Sexpr{con_ubf_Nrep} &
+      \Sexpr{con_abf_Nrep} $\times$ \Sexpr{con_abf_Np_per_replicate}&
+        \Sexpr{con_bpf_Np} 
+\\
+\hline
+\end{tabular}
+\caption{Numerical results for likelihood evaluation on the model with a linear constraint given in equations~\eqref{model:constrained} and~\eqref{model:constrained:measurement}.}
+\label{tab:constrained}
+\end{center}
+\end{table}
+
+The Monte Carlo effort in Table~\ref{tab:constrained} is large for all methods compared to the size of the problem, so the results primarily reflect inherent bias in the approaches rather than Monte Carlo variability.
+On this small problem, PF can provide an accurate likelihood estimate.
+EnKF preserves linear constraints, but does not preserve nonlinear constraints.
+Also, the system is Gaussian, which matches the approximation inherent in EnKF. 
+Therefore, EnKF closely matches PF here, but an example with a nonlinear constraint could also break EnKF.
+PF and the bagged filters preserve linear and nonlinear constraints.
+
+\clearpage
+
+\setcounter{assumptionB}{5}
+
+\section{\secTitleSpace Deriving Assumption~\ref{B:girf} from a previous intermediate resampling result}
+
+Intermediate resampling has been studied in the context of filtering highly informative observations \citep{delmoral15} and high-dimensional observations \citep{park20}.
+Adapted simulation is a special case of filtering with a single observation $(N=1)$ and a known initial condition.
+We show that a theorem from \citet{park20} implies Assumption~\ref{B:girf} holds with the constant  $\BvConstant$ being polynomial in the number of units, $\Unit$, when carrying out intermediate resampling with an ideal guide function.
+Recall the statement of Assumption~\ref{B:girf}:
+
+\Bv
+
+\medskip
+
+We check this follows as a consequence of the following theorem, in which $\myvec{X}_{t_N}^{F,j}$ is the j$th$ filter particle at the $N$th observation time, $t_N$, constructed using the guided intermediate resampling filter (GIRF) algorithm of \citet{park20} with an ideal guide function.
+We do not introduce the regularity conditions required for this theorem.
+
+\medskip
+
+\noindent{\bf Theorem~2} \citep{park20}.
+{\it 
+Suppose multinomial resampling is used in GIRF, with an ideal guide function.
+Under regularity conditions, there is a constant $C^*_1>0$ such that for any real-valued function $h(\myvec{x})$ with $\lVert h \rVert_\infty \leq 1$ and any constant $a>1$,
+\begin{equation}
+  \left\lvert \frac{1}{J} \sum_{j=1}^J h(\myvec{X}_{t_N}^{F,j}) - \mathbb{E} \big[ h(\myvec{X}_{t_N}) | \myvec{Y}_{1:N}=\myvec{y}_{1:N} \big] \right\rvert
+  \leq \frac{ 4 a (C^*_1+1)}{\sqrt{J}} (NS+1)
+  \label{extension:eqn:mainbound}
+\end{equation}
+with probability at least 
+\begin{equation}
+\label{extension:prob:bound}
+1-\frac{(2NS+1)(NS+1)}{a^2},
+\end{equation}
+given that $\sqrt J \geq 8 a (C^*_1+1) NS$.
+}
+
+In our context, $N=1$ since we are carrying out guided sampling over one timestep. 
+Now, set 
+\begin{equation}
+\frac{(2NS+1)(NS+1)}{a^2} < \epsilon/2,
+\end{equation}
+say, 
+\begin{equation}
+\label{eq:extension:1}
+a = \sqrt{\frac{2(2S+1)(S+1)}{\epsilon}}.
+\end{equation}
+We want  $J$ large enough that
+\begin{equation}
+\label{eq:extension:2}
+\frac{ 4 a (C^*_1+1)}{\sqrt{J}} (NS+1) < \epsilon/2,
+\end{equation}
+since then the bounds in \myeqref{extension:eqn:mainbound} and \myeqref{extension:prob:bound}, together with the bound on $h$, give the bound in expectation in Assumption~\ref{B:girf}.
+Putting \myeqref{eq:extension:1} into \myeqref{eq:extension:2} gives
+\begin{equation}
+\label{eq:extension:3}
+\frac{ 4(C^*_1+1)\sqrt{2(2S+1)(S+1)}}{\sqrt{\epsilon J}} (S+1) < \epsilon/2.
+\end{equation}
+and so
+\begin{equation}
+\label{eq:extension:4}
+J > \frac{128(2S+1)(S+1)^3(C^*_1+1)^2}{\epsilon^3}
+\end{equation}
+which implies that we can pick
+\begin{equation}
+\BvConstant =  128(2S+1)(S+1)^3 (C^*_1+1)^2 
+\end{equation}
+in Assumption~\ref{B:girf}.
+
+The constant $C^*_1$ depends on the number of intermediate resapling intervals, $S$.
+Proposition~1 of \citet{park20} shows that $C^*_1$ in Theorem~2 can be bounded as a function of $\Unit$ if intermediate resampling is carried out with $S=\Unit$.
+
+In practical situations, one can only approximate an ideal guide function, leading to an exponential bound on $\BvConstant$ which is small if the guide function is effective.
+ABF-IR is less sensitive to the choice of guide function than GIRF, since the likelihood is calculated using local weights and the intermediate resampling is used only for the adapted simulation.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% iiiiiiiiiiiiiii
+
+\clearpage
+
+\bibliography{../bib-iif}
 

### 2021-09-02 theorems.tex
code to make unformatted accepted version

@@
 \begin{assumptionB}\label{B1}
-  There is an $\ethree>0$, independent of $\Unit$ and $\Time$, and a collection of neighborhoods $\{B_{\unit\comma\time}\subset A_{\unit\comma\time}, \unit\in 1\mycolon\Unit,\time\in 1\mycolon\Time\}$ such that the following holds for all $\unit$ and $\time$, and any bounded real-valued function $|h(x)|\le 1$\emph{:} if we write $A=A_{u,n}$, $B=B_{u,n}$, $f_{A}(x_A) = f_{Y_{A}|X_{A}}(\data y_{A}|x_A)$, and $f_{B}(x_B) = f_{Y_{B}|X_{B}}(\data y_{B}|x_B)$,
+  There is an $\ethree>0$, independent of $\Unit$ and $\Time$, and a collection of neighborhoods $\{B_{\unit\comma\time}\subset A_{\unit\comma\time}, \unit\in 1\mycolon\Unit,\time\in 1\mycolon\Time\}$ such that the following holds for all $\unit$ and $\time$, and any bounded real-valued function $|h(x)|\le 1$.
+  Setting $A=A_{u,n}$, $B=B_{u,n}$, $f_{A}(x_A) = f_{Y_{A}|X_{A}}(\data y_{A}|x_A)$, and $f_{B}(x_B) = f_{Y_{B}|X_{B}}(\data y_{B}|x_B)$, so that we have the identity
+    \[
+    f_{X_{u,n}|Y_A}(x | \data{y}_A) = 
+    \frac{
+      \E_{g} \! \big[ 
+      f_{A}\big(X_{A}^P\big) f_{X_{u,n}|X_{A^{[n]}},\myvec X_{n-1}} \big(x | X_{A^{[n]}}^P, \myvec X_{n-1}\big) 
+      \big]
+    } {
+      \E_{g} \! \big[f_A\big(X_A^P\big)\big]
+    },
+  \]
+we require that
 %\begin{equation}
@@
 \nonumber
-%\label{asif:eq:weak_coupling2}
-&&\hspace{-7mm}
+%\label{abf:eq:weak_coupling2}
+&&\hspace{-30mm}
 \left| 
@@
 dx
-\right| \hspace{0mm} < \hspace{0mm} \ethree.
+\right| \hspace{1mm} < \hspace{1mm} \ethree.
 %\end{equation}
@@
 \begin{assumptionB}\label{B:unconditional:mix}
-Define $g_{{X}_{A}|{X}_{B}}$ via \myeqref{eq:g}.
+  We use subscripts of $g$ to denote marginal and conditional densities derived from \myeqref{eq:g}.
 Suppose there is an $\efourA$, independent of $\Unit$ and $\Time$, such that the following holds.
@@
 \big| 
-   g^{}_{{X}^{P}_{B^{}_{\altUnit\comma\altTime}}{X}^{P}_{B^{}_{\unit\comma\time}}} 
+   g^{}_{{X}^{P}_{B^{}_{\altUnit\comma\altTime} \cup B^{}_{\unit\comma\time}}} 
    - g^{}_{{X}^{P}_{B^{}_{\altUnit\comma\altTime}}} \, g^{}_{{X}^{P}_{B^{}_{\unit\comma\time}}}
@@
 %&&\hspace{-30mm}
-(1/2) \, \efourA \, \,  g^{}_{{X}^{P}_{B^{}_{\altUnit\comma\altTime}}} \, g^{}_{{X}^{P}_{B^{}_{\unit\comma\time}}}
+(1/2) \, \efourA \, \,  g^{}_{{X}^{P}_{{B^{}_{\altUnit\comma\altTime}} \cup {B^{}_{\unit\comma\time}}}}
 %\\
@@
 -
-g^{}_{X^P_{B^{}_{\altUnit\comma\altTime}},X^P_{B^{}_{\unit\comma\time}}|\myvec{X}_{0:\Time}}
+g^{}_{X^P_{B^{}_{\altUnit\comma\altTime} \cup B^{}_{\unit\comma\time}}|\myvec{X}_{0:\Time}}
 \big|
@@
 &&\hspace{-30mm} < (1/2) \, \efourA \, \, 
- g^{}_{X^P_{B^{}_{\altUnit\comma\altTime}},X^P_{B^{}_{\unit\comma\time}}|\myvec{X}_{0:\Time}}
+ g^{}_{X^P_{B^{}_{\altUnit\comma\altTime} \cup B^{}_{\unit\comma\time}}|\myvec{X}_{0:\Time}}
       \end{eqnarray}
@@
 \hspace{-0.5mm}
+\EMC
 \left|  
 \, 
-  \EMC\Big[ 
     \frac{1}{\Np} \sum_{\np=1}^{\Np} h(\myvec{X}^{\GR}_{\time,\Ninter,\np,\rep}) 
-   \Big] \!
 -  \E_g \! \big[h(\myvec{X}_\time)\given \myvec{X}_{\time-1}=\myvec{x}^{\IF}_{\time-1,\Ninter,\rep}\big] 
@@
 \newcommand\TheoremII{
-\begin{theorem}  \label{thm:asif}
+\begin{theorem}  \label{thm:abf}
 Let $\MC{\loglik}$ denote the Monte Carlo likelihood approximation constructed by {\ABFIR}, or by {\ABF} since this is the special case of {\ABFIR} with $\Ninter=1$.
@@
 \begin{equation}
-\siOnly{\label{th:asif:lik:bound}}
+\siOnly{\label{th:abf:lik:bound}}
 \msOnly{\nonumber}
@@
 \begin{equation}
-\siOnly{\label{th:asif:lik:bound2}}
+\siOnly{\label{th:abf:lik:bound2}}
 \msOnly{\nonumber}

### 2021-09-02 measles_iubf/analyses/abf_run.R

@@ -1,0 +1,131 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+ run_level <- 1
+} else run_level <- as.numeric(args)
+
+out_files_dir <- paste0("../generated/abf_outs_",run_level,"/")
+if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-20;N<-2*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+
+if(run_level==1){ # personal machine testing
+  abf_nbhd <- function(object, time, unit) {
+    nbhd_list <- list()
+    if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+    if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+    # if(unit==1 && time > 1) nbhd_list <- c(nbhd_list, list(c(2, time-1)))
+    # if(unit==2) nbhd_list <- c(nbhd_list, list(c(1, time)))
+    # if(unit==3) nbhd_list <- c(nbhd_list, list(c(1, time)))
+    # if(unit==4) nbhd_list <- c(nbhd_list, list(c(3, time)))
+    # if(unit==5) nbhd_list <- c(nbhd_list, list(c(4, time)))
+    # if(unit==6) nbhd_list <- c(nbhd_list, list(c(1, time)))
+    # if(unit==6) nbhd_list <- c(nbhd_list, list(c(4, time)))
+    # if(unit==7) nbhd_list <- c(nbhd_list, list(c(1, time)))
+    # if(unit==8) nbhd_list <- c(nbhd_list, list(c(1, time)))
+    # if(unit==9) nbhd_list <- c(nbhd_list, list(c(1, time)))
+    # if(unit==10) nbhd_list <- c(nbhd_list, list(c(5, time)))
+
+    return(nbhd_list)
+  }
+  abf_nabf <- 1
+  abf_nrep <- 35000
+  abf_np <- 1
[…]

### 2021-09-02 measles_iubf/analyses/abf_slice_10/abf_slice_R0.R

@@ -1,0 +1,62 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+library("ggplot2")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+ run_level <- 1
+} else run_level <- as.numeric(args)
+
+out_files_dir <- paste0("../generated/abf_slice_",run_level,"/u10/")
+if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-10;N<-8*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+R0s <- seq(from=15, to=60, by = 5)
+slice_params <- as.matrix(replicate(length(R0s),theta))
+slice_params['R0',] <- R0s
+abf_reps_per_param <- 4
+abf_nrep <- 40000
+abf_np <- 1
+slice_runs <- vector(mode="list", length=length(R0s))
+abf_nbhd <- function(object, time, unit) {
+ nbhd_list <- list()
+ if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+ if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+ return(nbhd_list)
+}
+stew(file=paste0(out_files_dir,"abf_slice_R0",".rda"),seed=5981724,{
+  for(i in seq_len(ncol(slice_params))){
+   slice_runs[[i]] <- vector(mode="list", length = abf_reps_per_param)
+   for(j in seq_len(abf_reps_per_param)){
+    slice_runs[[i]][[j]] <- abf(m,
+                                params = slice_params[,i],
+                                Nrep = abf_nrep,
+                                Np = abf_np,
+                                nbhd = abf_nbhd
+                            )
+   }
+  }
+})
+slice_logliks <- vector(mode="list", length=length(R0s))
+for(i in seq_len(ncol(slice_params))){
+  for(j in seq_len(abf_reps_per_param)){
[…]

### 2021-09-02 measles_iubf/analyses/abf_slice_10/abf_slice_g.R

@@ -1,0 +1,61 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+library("ggplot2")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+ run_level <- 1
+} else run_level <- as.numeric(args)
+
+out_files_dir <- paste0("../generated/abf_slice_",run_level,"/u10/")
+if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-10;N<-8*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+gs <- seq(from=100, to=1100, by = 100)
+slice_params <- as.matrix(replicate(length(gs),theta))
+slice_params['g',] <- gs
+abf_reps_per_param <- 4
+abf_nrep <- 40000
+abf_np <- 1
+slice_runs <- vector(mode="list", length=length(gs))
+abf_nbhd <- function(object, time, unit) {
+ nbhd_list <- list()
+ if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+ if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+ return(nbhd_list)
+}
+stew(file=paste0(out_files_dir,"abf_slice_g",".rda"),seed=5981724,{
+  for(i in seq_len(ncol(slice_params))){
+   slice_runs[[i]] <- vector(mode="list", length = abf_reps_per_param)
+   for(j in seq_len(abf_reps_per_param)){
+    slice_runs[[i]][[j]] <- abf(m,
+                                params = slice_params[,i],
+                                Nrep = abf_nrep,
+                                Np = abf_np,
+                                nbhd = abf_nbhd
+                            )
+   }
+  }
+})
+slice_logliks <- vector(mode="list", length=length(gs))
+for(i in seq_len(ncol(slice_params))){
+  for(j in seq_len(abf_reps_per_param)){
[…]

### 2021-09-02 measles_iubf/analyses/abf_slice_10/abf_slice_psi.R

@@ -1,0 +1,100 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+library("ggplot2")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+ run_level <- 1
+} else run_level <- as.numeric(args)
+
+print(paste("run level: ", run_level))
+out_files_dir <- paste0("../generated/abf_slice_",run_level,"/u10/")
+if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-10;N<-8*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+psis <- seq(from=0.1, to=0.4, by = 0.05)
+slice_params <- as.matrix(replicate(length(psis),theta))
+slice_params['psi',] <- psis
+
+if(run_level==1){ # personal machine testing
+  abf_nbhd <- function(object, time, unit) {
+    nbhd_list <- list()
+    if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+    if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+    return(nbhd_list)
+  }
+  job_num <- 0
+  abf_reps_per_param <- 3
+  abf_nrep <- 40000
+  abf_np <- 1
+} else if(run_level==2){ # first draft results
+  abf_nbhd <- function(object, time, unit) {
+    nbhd_list <- list()
+    if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+    if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+    return(nbhd_list)
+  }
+  job_num <- as.integer(Sys.getenv('SLURM_ARRAY_TASK_ID'))
+  abf_reps_per_param <- 3
+  abf_nrep <- 10000*job_num
+  abf_np <- 1
+}
+
+print(paste("job_num: ", job_num))
+
+slice_runs <- vector(mode="list", length=length(psis))
[…]

### 2021-09-02 measles_iubf/analyses/abf_slice_20/abf_slice_R0.R

@@ -1,0 +1,100 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+library("ggplot2")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+  run_level <- 1
+} else run_level <- as.numeric(args)
+
+print(paste("run level: ", run_level))
+out_files_dir <- paste0("../generated/abf_slice_",run_level,"/u20/")
+if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-20;N<-8*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+R0s <- seq(from=15, to=60, by = 5)
+slice_params <- as.matrix(replicate(length(R0s),theta))
+slice_params['R0',] <- R0s
+
+if(run_level==1){ # personal machine testing
+  abf_nbhd <- function(object, time, unit) {
+    nbhd_list <- list()
+    if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+    if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+    return(nbhd_list)
+  }
+  job_num <- 0
+  abf_reps_per_param <- 3
+  abf_nrep <- 40000
+  abf_np <- 1
+} else if(run_level==2){ # first draft results
+  abf_nbhd <- function(object, time, unit) {
+    nbhd_list <- list()
+    if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+    if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+    return(nbhd_list)
+  }
+  job_num <- as.integer(Sys.getenv('SLURM_ARRAY_TASK_ID'))
+  abf_reps_per_param <- 3
+  abf_nrep <- 10000*job_num
+  abf_np <- 1
+}
+
+print(paste("job_num: ", job_num))
+
+slice_runs <- vector(mode="list", length=length(R0s))
+stew(file=paste0(out_files_dir,"abf_slice_R0",job_num,".rda"),seed=5981724,{
[…]

### 2021-09-02 measles_iubf/analyses/abf_slice_20/abf_slice_g.R

@@ -1,0 +1,100 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+library("ggplot2")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+  run_level <- 1
+} else run_level <- as.numeric(args)
+
+print(paste("run level: ", run_level))
+out_files_dir <- paste0("../generated/abf_slice_",run_level,"/u20/")
+if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-20;N<-8*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+gs <- seq(from=100, to=1000, by = 100)
+slice_params <- as.matrix(replicate(length(gs),theta))
+slice_params['g',] <- gs
+
+if(run_level==1){ # personal machine testing
+  abf_nbhd <- function(object, time, unit) {
+    nbhd_list <- list()
+    if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+    if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+    return(nbhd_list)
+  }
+  job_num <- 0
+  abf_reps_per_param <- 3
+  abf_nrep <- 40000
+  abf_np <- 1
+} else if(run_level==2){ # first draft results
+  abf_nbhd <- function(object, time, unit) {
+    nbhd_list <- list()
+    if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+    if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+    return(nbhd_list)
+  }
+  job_num <- as.integer(Sys.getenv('SLURM_ARRAY_TASK_ID'))
+  abf_reps_per_param <- 3
+  abf_nrep <- 10000*job_num
+  abf_np <- 1
+}
+
+print(paste("job_num: ", job_num))
+
+slice_runs <- vector(mode="list", length=length(gs))
[…]

### 2021-09-02 measles_iubf/analyses/abf_slice_20/abf_slice_psi.R

@@ -1,0 +1,100 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+library("ggplot2")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+ run_level <- 1
+} else run_level <- as.numeric(args)
+
+print(paste("run level: ", run_level))
+out_files_dir <- paste0("../generated/abf_slice_",run_level,"/u20/")
+if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-20;N<-8*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+psis <- seq(from=0.1, to=0.4, by = 0.05)
+slice_params <- as.matrix(replicate(length(psis),theta))
+slice_params['psi',] <- psis
+
+if(run_level==1){ # personal machine testing
+  abf_nbhd <- function(object, time, unit) {
+    nbhd_list <- list()
+    if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+    if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+    return(nbhd_list)
+  }
+  job_num <- 0
+  abf_reps_per_param <- 3
+  abf_nrep <- 40000
+  abf_np <- 1
+} else if(run_level==2){ # first draft results
+  abf_nbhd <- function(object, time, unit) {
+    nbhd_list <- list()
+    if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+    if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+    return(nbhd_list)
+  }
+  job_num <- as.integer(Sys.getenv('SLURM_ARRAY_TASK_ID'))
+  abf_reps_per_param <- 3
+  abf_nrep <- 10000*job_num
+  abf_np <- 1
+}
+
+print(paste("job_num: ", job_num))
+
+slice_runs <- vector(mode="list", length=length(psis))
[…]

### 2021-09-02 measles_iubf/analyses/abf_slice_5/abf_slice_R0.R

@@ -1,0 +1,62 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+library("ggplot2")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+  run_level <- 1
+} else run_level <- as.numeric(args)
+
+out_files_dir <- paste0("../generated/abf_slice_",run_level, "/u5/")
+if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-5;N<-8*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+R0s <- seq(from=15, to=60, by = 5)
+slice_params <- as.matrix(replicate(length(R0s),theta))
+slice_params['R0',] <- R0s
+abf_reps_per_param <- 4
+abf_nrep <- 30000
+abf_np <- 1
+slice_runs <- vector(mode="list", length=length(R0s))
+abf_nbhd <- function(object, time, unit) {
+ nbhd_list <- list()
+ if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+ if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+ return(nbhd_list)
+}
+stew(file=paste0(out_files_dir,"abf_slice_R0",".rda"),seed=5981724,{
+  for(i in seq_len(ncol(slice_params))){
+   slice_runs[[i]] <- vector(mode="list", length = abf_reps_per_param)
+   for(j in seq_len(abf_reps_per_param)){
+    slice_runs[[i]][[j]] <- abf(m,
+                                params = slice_params[,i],
+                                Nrep = abf_nrep,
+                                Np = abf_np,
+                                nbhd = abf_nbhd
+                            )
+   }
+  }
+})
+slice_logliks <- vector(mode="list", length=length(R0s))
+for(i in seq_len(ncol(slice_params))){
+  for(j in seq_len(abf_reps_per_param)){
[…]

### 2021-09-02 measles_iubf/analyses/abf_slice_5/abf_slice_g.R

@@ -1,0 +1,61 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+library("ggplot2")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+ run_level <- 1
+} else run_level <- as.numeric(args)
+
+out_files_dir <- paste0("../generated/abf_slice_",run_level, "/u5/")
+if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-5;N<-8*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+gs <- seq(from=100, to=1100, by = 100)
+slice_params <- as.matrix(replicate(length(gs),theta))
+slice_params['g',] <- gs
+abf_reps_per_param <- 4
+abf_nrep <- 30000
+abf_np <- 1
+slice_runs <- vector(mode="list", length=length(gs))
+abf_nbhd <- function(object, time, unit) {
+ nbhd_list <- list()
+ if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+ if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+ return(nbhd_list)
+}
+stew(file=paste0(out_files_dir,"abf_slice_g",".rda"),seed=5981724,{
+  for(i in seq_len(ncol(slice_params))){
+   slice_runs[[i]] <- vector(mode="list", length = abf_reps_per_param)
+   for(j in seq_len(abf_reps_per_param)){
+    slice_runs[[i]][[j]] <- abf(m,
+                                params = slice_params[,i],
+                                Nrep = abf_nrep,
+                                Np = abf_np,
+                                nbhd = abf_nbhd
+                            )
+   }
+  }
+})
+slice_logliks <- vector(mode="list", length=length(gs))
+for(i in seq_len(ncol(slice_params))){
+  for(j in seq_len(abf_reps_per_param)){
[…]

### 2021-09-02 measles_iubf/analyses/abf_slice_5/abf_slice_psi.R

@@ -1,0 +1,61 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+library("ggplot2")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+  run_level <- 1
+} else run_level <- as.numeric(args)
+
+out_files_dir <- paste0("../generated/abf_slice_",run_level, "/u5/")
+if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-5;N<-8*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+psis <- seq(from=0.1, to=0.4, by = 0.05)
+slice_params <- as.matrix(replicate(length(psis),theta))
+slice_params['psi',] <- psis
+abf_reps_per_param <- 4
+abf_nrep <- 30000
+abf_np <- 1
+slice_runs <- vector(mode="list", length=length(psis))
+abf_nbhd <- function(object, time, unit) {
+ nbhd_list <- list()
+ if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+ if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+ return(nbhd_list)
+}
+stew(file=paste0(out_files_dir,"abf_slice_psi",".rda"),seed=5981724,{
+  for(i in seq_len(ncol(slice_params))){
+   slice_runs[[i]] <- vector(mode="list", length = abf_reps_per_param)
+   for(j in seq_len(abf_reps_per_param)){
+    slice_runs[[i]][[j]] <- abf(m,
+                                params = slice_params[,i],
+                                Nrep = abf_nrep,
+                                Np = abf_np,
+                                nbhd = abf_nbhd
+                            )
+   }
+  }
+})
+slice_logliks <- vector(mode="list", length=length(psis))
+for(i in seq_len(ncol(slice_params))){
+  for(j in seq_len(abf_reps_per_param)){
[…]

### 2021-09-02 measles_iubf/analyses/data_analysis.R

@@ -1,0 +1,93 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+ run_level <- 1
+} else run_level <- as.numeric(args)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-10;N<-2*26
+m <- measles_sim_w_params(U=U,N=N)
+md <- as.data.frame(m)
+
+library(ggplot2)
+ggplot(data = md, mapping = aes(x = year, y = cases)) + 
+ geom_line() +
+ facet_wrap(~city, scales = "free_y")
+
+m2 <- measles_sim_w_params(U=U,N=N,mup=c(alpha=1,iota=0,R0=41.23672,cohort=0,amplitude=0.5354223, gamma=45.61916, sigma=50.27219,mu=0.02,sigmaSE=0.1351309, rho=0.5,psi=0.2071051,g=1100))
+md2 <- as.data.frame(m2)
+ggplot(data = md2, mapping = aes(x = year, y = cases)) + 
+ geom_line() +
+ facet_wrap(~city, scales = "free_y")
+
+m3 <- measles_sim_w_params(U=U,N=N,mup=c(alpha=1,iota=0,R0=46.77458,cohort=0,amplitude=0.6040571, gamma=44.81039, sigma=37.18639,mu=0.02,sigmaSE=0.1439365, rho=0.5,psi=0.2084648,g=455.9014))
+md3 <- as.data.frame(m3)
+ggplot(data = md3, mapping = aes(x = year, y = cases)) + 
+ geom_line() +
+ facet_wrap(~city, scales = "free_y")
+p1 <- coef(m)
+p2 <- coef(m)
+p2[['g']] <- 10000
+# p1[c("g","R0", "amplitude", "gamma", "sigma", "sigmaSE", "psi")] <- c(1100,
+#                                                                      41.23672,
+#                                                                      0.5354223,
+#                                                                      45.61916,
+#                                                                      50.27219,
+#                                                                      0.1351309,
+#                                                                      0.2071051)
+# p2[c("g","R0", "amplitude", "gamma", "sigma", "sigmaSE", "psi")] <- c(455.9014,
+#                                                                       46.77458,
+#                                                                       0.6040571,
+#                                                                       44.81039,
[…]

### 2021-09-02 measles_iubf/analyses/iubf_measles_sim.R

@@ -1,0 +1,35 @@
+measles_sim_w_params <- function(U = 40,
+                                 N=15*26,
+                                 mup=c(alpha=1,
+                                       iota=0,
+                                       R0=30,
+                                       amplitude=0.5,
+                                       cohort=0,
+                                       gamma=52,
+                                       sigma=52,
+                                       mu=0.02,
+                                       sigmaSE=0.15,
+                                       rho=0.5,
+                                       psi=0.15,
+                                       g=400)){
+ measles_sim_U <- U
+ measles_uk <- measles(measles_sim_U)
+ measles_RPnames <- c("alpha","iota","R0","amplitude","cohort",
+                      "gamma","sigma","mu","sigmaSE","rho","psi","g")
+ measles_paramnames <- measles_RPnames
+ 
+ measles_params <- rep(NA,length(measles_paramnames))
+ names(measles_params) <- measles_paramnames
+ measles_unit_params <- mup
+ measles_params[measles_RPnames] <- measles_unit_params[measles_RPnames]
+
+ set.seed(34)
+ measles_sim <- simulate(measles_uk,params=measles_params)
+ measles_sim@data <- measles_sim@data[1:U,1:N]
+ time(measles_sim) <- measles_sim@times[1:N]
+ 
+ m_paramnames <- measles_RPnames
+ m_params <- measles_params[names(measles_params)%in%m_paramnames]
+ coef(measles_sim) <- m_params
+ return(measles_sim)
+}

### 2021-09-02 measles_iubf/analyses/iubf_output_profile.R

@@ -1,0 +1,61 @@
+library("spatPomp")
+library("ggplot2")
+read.csv('../generated/iubf_out.csv', row.names = NULL) -> g_prof
+g_prof <- g_prof %>% dplyr::filter(U==20)
+plot.profile <- function(mcap1,ylab,xline=2.5,yline=2.5,xlab="",quadratic=FALSE,...){
+ if(missing(ylab)) ylab <- "profile log likelihood"
+ ggplot() + geom_point(aes(x = mcap1$parameter, y = mcap1$lp)) +
+  geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$smoothed),
+            color = "red") +
+  {if(quadratic) geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$quadratic),
+                           color = "blue",
+                           lwd = 1.25)} +
+  labs(x = xlab, y = ylab) +
+  geom_vline(xintercept = mcap1$ci, color = "red") +
+  geom_hline(yintercept = max(mcap1$fit$smoothed, na.rm = T) - mcap1$delta, color = "red") +
+  theme(panel.border = element_rect(colour = "black", fill=NA))
+}
+
+g_prof_mcap <- spatPomp::mcap(
+ lp=g_prof[,"loglik"],
+ parameter=g_prof[,"g"],
+ lambda = 0.8
+)
+plot.profile(g_prof_mcap,xlab='g', quadratic = FALSE) +
+ geom_vline(xintercept=400,color="blue",linetype="dashed") +
+ theme(axis.text.x = element_text(size = 12),
+       axis.text.y = element_text(size = 12),
+       axis.title.x = element_text(size = 14),
+       axis.title.y = element_text(size = 14))
+
+# ggplot(data = g_prof, mapping = aes(x = g, y = loglik))+
+#  geom_point() + 
+#   geom_vline(xintercept=400,color="blue",linetype="dashed")
+#   
+
+# measles_unit_params <- c(
+#   alpha=1,
+#   iota=0,  # set to zero for a closed population
+#   R0=30,
+#   cohort=0,
+#   amplitude=0.5, gamma=52, sigma=52,mu=0.02,
+#   sigmaSE=0.15, rho=0.5,
+#   psi=0.15,
+#   g=400
+# )
+
+# p1<-ggplot(data = g_prof, mapping = aes(x = g, y = gamma, color = loglik))+
[…]

### 2021-09-02 measles_iubf/analyses/iubf_profile_design.R

@@ -1,0 +1,70 @@
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+
+theta_start_t <- partrans(m,theta,"toEst")
+theta_t_hi <- theta_t_lo <- theta_start_t
+theta_t_lo[c("sigma",
+             "gamma",
+             "sigmaSE",
+             "psi",
+             "amplitude",
+             "R0")] <- theta_t_lo[c("sigma",
+                                    "gamma",
+                                    "sigmaSE",
+                                    "psi",
+                                    "amplitude",
+                                    "R0")] - log(2)
+theta_t_hi[c("sigma",
+             "gamma",
+             "sigmaSE",
+             "psi",
+             "amplitude",
+             "R0")] <- theta_t_hi[c("sigma",
+                                    "gamma",
+                                    "sigmaSE",
+                                    "psi",
+                                    "amplitude",
+                                    "R0")] + log(2)
+
+if(!file.exists('../generated/iubf_out.csv')) 
+        writeLines(text = paste0(
+                                c(
+                                        c("U","N","loglik","Nabf","Nparam","Nrep_per_param","prop"),
+                                        names(theta_start_t),
+                                        "notes"
+                                ),
+                                collapse = ','),
+                   con = '../generated/iubf_out.csv')
+if(!file.exists('../generated/abf_out.csv')) 
+        writeLines(text = paste0(
+                c(
+                        c("U","N","run_level","Nabf_runs","Nrep"),
+                        names(theta_start_t),
+                        "abf_loglik",
+                        "abf_loglik_se",
+                        "notes"
+                ),
+                collapse = ','),
+                con = '../generated/abf_out.csv')
+
+
+theta_t_lo <- partrans(m,theta_t_lo,"fromEst")
+theta_t_lo <- theta_t_lo[!names(theta_t_lo)=='g']
+theta_t_hi <- partrans(m,theta_t_hi,"fromEst")
+theta_t_hi <- theta_t_hi[!names(theta_t_hi)=='g']
+
+profile_design(
+ g=all_g,
+ lower=theta_t_lo,
+ upper=theta_t_hi,
+ nprof=g_nprof
+) -> pd
+
+row_order = c()
+for(i in seq_len(g_nprof)){
+ to_add <- which(seq_len(nrow(pd)) %% g_nprof == i-1)
+ row_order <- c(row_order, to_add)
+}
+
+pd <- pd[row_order,]
+rownames(pd) <- NULL

### 2021-09-02 measles_iubf/analyses/iubf_run.R

@@ -1,0 +1,220 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+
+dd <- function(){
+   rm(list = ls())
+}
+dd()
+
+args = commandArgs(trailingOnly=TRUE)
+if(length(args)==0){
+   run_level <- 1
+} else run_level <- as.numeric(args)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+if(run_level==1){ # personal machine testing
+   U <- 5; N <- 10
+   iubf_nbhd <- function(object, time, unit) {
+      nbhd_list <- list()
+      if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+      if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+      return(nbhd_list)
+   }
+   iubf_nubf <- 5
+   iubf_nrep_per_param <- 5000
+   iubf_nparam <- 200
+   iubf_prop <- 0.80
+   g_lo = 200
+   g_hi = 1100
+   g_num = 5
+   g_nprof = 2
+} else if(run_level==2){ # first draft results
+   U <- 10; N <- 8*26
+   iubf_nbhd <- function(object, time, unit) {
+      nbhd_list <- list()
+      if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+      if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+      return(nbhd_list)
+   }
+   iubf_nubf <- 15
+   iubf_nrep_per_param <- 35000
+   abf_nreps <- 60000
+   iubf_nparam <- 250
+   iubf_prop <- 0.80
+   g_lo = 100
+   g_hi = 1100
+   g_num = 11
+   g_nprof = 3
+   all_g = seq(from = g_lo, to = g_hi, length = g_num)
+   all_g = append(all_g, 150, after = 1)
+   all_g = c(50,all_g)
+   abf_nruns = 8
+} else if(run_level==3){ # production run
+   U <- 20; N <- 8*26
+   iubf_nbhd <- function(object, time, unit) {
+      nbhd_list <- list()
[…]

### 2021-09-02 measles_iubf/analyses/ubf_g_bias_exp.R

@@ -1,0 +1,69 @@
+library("spatPomp")
+library("doRNG")
+library("doParallel")
+library("ggplot2")
+
+# args = commandArgs(trailingOnly=TRUE)
+# if(length(args)==0){
+#  run_level <- 1
+# } else run_level <- as.numeric(args)
+# 
+# out_files_dir <- paste0("../generated/abf_slice_",run_level,"/")
+# if(!dir.exists(out_files_dir)) dir.create(out_files_dir)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+if(is.na(cores) & run_level >= 2) cores <- 36
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+registerDoRNG(1598260027L)
+
+source('../analyses/iubf_measles_sim.R')
+set.seed(1)
+U<-3;N<-2*26
+m <- measles_sim_w_params(U=U,N=N)
+theta <- coef(m)
+theta_alt <- theta
+theta_alt['g'] <- 1100
+
+abf_nbhd <- function(object, time, unit) {
+ nbhd_list <- list()
+ if(time>1) nbhd_list <- c(nbhd_list, list(c(unit, time-1)))
+ if(time>2) nbhd_list <- c(nbhd_list, list(c(unit, time-2)))
+ return(nbhd_list)
+}
+abf_nrep <- 1500
+abf_np <- 1
+set.seed(1)
+a <- replicate(5, abf(m,
+                      params = theta,
+                      nbhd = abf_nbhd,
+                      Nrep = abf_nrep,
+                      Np = abf_np)
+)
+a_alt <- replicate(5, abf(m,
+                          params = theta_alt,
+                          nbhd = abf_nbhd,
+                          Nrep = abf_nrep,
+                          Np = abf_np)
+)
+
+p <- replicate(5, pfilter(m,
+                          params = theta,
+                          Np = 20000))
+p_alt <- replicate(5, pfilter(m,
+                              params = theta_alt,
+                              Np = 20000))
+
+g <- replicate(5, girf(m,
+                       params = theta,
+                       Np = 2000,
+                       Nguide = 30,
+                       lookahead = 1))
+g_alt <- replicate(5, girf(m,
+                       params = theta_alt,
+                       Np = 2000,
+                       Nguide = 30,
+                       lookahead = 1))
+
+
+

### 2021-09-02 measles_iubf/bash/run_measles_iubf.sh

@@ -1,0 +1,4 @@
+#!/bin/bash
+# nohup bash run_measles_iubf.sh run_level &
+export LC_ALL=en_US.UTF-8
+Rscript ../analyses/iubf_run.R $1

### 2021-09-02 si/con.R

@@ -1,0 +1,160 @@
+library(spatPomp)
+#' Constrained motion spatPomp simulator
+#'
+#' Generate a class \sQuote{spatPomp} object representing a \code{U}-dimensional
+#' Brownian motion constrained to sum to zero across units, with a nonlinear
+#' drift that applies if this constraint is broken.
+#'
+#' @param U A length-one numeric signifying dimension of the process.
+#' @param N A length-one numeric signifying the number of observation time steps to evolve the process.
+#' @param delta_t Process simulations are performed every \code{delta_t} time units
+#' whereas observations occur every one time unit
+#' @importFrom utils data
+#' @return An object of class \sQuote{spatPomp} representing a simulation from a \code{U}-dimensional
+#' Constrained motion
+#' @examples
+#' co <- con(U=4, N=20)
+#' # See all the model specifications of the object
+#' spy(co)
+#' @export
+
+con <- function(U=10,N=20,delta_t=0.2){
+# debug
+# U=10 ; N=20; delta_t=0.2
+  obs_names <- paste0("U",1:U)
+  con_data <- data.frame(time=rep(1:N,U),unit=rep(obs_names,each=N),Y=rep(NA,U*N),stringsAsFactors=F)
+  con_unitnames <- unique(con_data[["unit"]])
+  con_unitnames_level <- paste("U",sort(as.numeric(stringr::str_remove(con_unitnames, "U"))),sep='')
+
+  con_unit_statenames <- c("X")
+  con_statenames <- paste0(con_unit_statenames,1:U)
+
+  con_IVPnames <- paste0(con_statenames,"_0")
+  con_RPnames <- c("sigma","tau")
+  con_paramnames <- c(con_RPnames,con_IVPnames)
+
+  con_rprocess <- spatPomp_Csnippet("
+    double dW[U];
+    double dWsum = 0;
+    double Xsum = 0;
+    int u;
+
+    for (u = 0 ; u < U ; u++) {
+      dW[u] = rnorm(0,sigma*sqrt(dt));
+      dWsum += dW[u];
+      Xsum += X[u];
+    }
+    for (u = 0 ; u < U ; u++) {
+      dW[u] -= dWsum/U;
+    }
+    for (u = 0 ; u < U ; u++) {
+      X[u] += dW[u] + Xsum*dt;
+    }
+  ", unit_statenames = c("X"))
+
+
[…]

### 2021-11-24 ms.Rnw
adding reference to published article

@@
 
+\vspace{3mm}
+
+\noindent This document is a reproducible version of \citet{ionides21} compiled on \today, using R version \Sexpr{getRversion()}, \code{pomp} version \Sexpr{packageVersion("pomp")} and \code{spatPomp} version \Sexpr{packageVersion("spatPomp")}.
+
 %\vfill
@@
 
-\spacingset{1.45} % DON'T change the spacing!
+%\spacingset{1.45} % DON'T change the spacing!
+\spacingset{1}
 
@@
 
-slice_abf_mcap <- panelPomp::mcap(
+slice_abf_mcap <- mcap(
   lp=slice_abf_results[,"logLik"],
@@
 
-slice_ubf_mcap <- panelPomp::mcap(
+slice_ubf_mcap <- mcap(
   lp=slice_ubf_results[,"logLik"],
@@
 
-slice_bpf_mcap <- panelPomp::mcap(
+slice_bpf_mcap <- mcap(
   lp=slice_bpf_results[,"logLik"],
@@
 
-slice_enkf_mcap <- panelPomp::mcap(
+slice_enkf_mcap <- mcap(
   lp=slice_enkf_results[,"logLik"],

### 2022-11-21 ms.Rnw
update for pomp 4.3.4.2

@@
 # run_level <- "jasa-sept-2020"
- run_level <- "manual"
-# run_level <- 2
+# run_level <- "manual"
+ run_level <- 1
 
@@
     rootQ <- coef(bm)["rho"]^bm_dmat * coef(bm)["sigma"]
-    kf <- pomp:::kalmanFilter(
-      t=1:bm_N,
-      y=obs(bm),
+    kf <- pomp:::kalmanFilter(bm,
       X0=rinit(bm),
@@
       R=diag(coef(bm)["tau"]^2, nrow=bm_units)
-    )$loglik
+    )$logLik
   } -> bm_kf_results
@@
 slice_abf_mcap <- mcap(
-  lp=slice_abf_results[,"logLik"],
+  logLik=slice_abf_results[,"logLik"],
   parameter=slice_abf_results[,"g"],
-  lambda = 1
+  span = 1
 )
@@
 slice_ubf_mcap <- mcap(
-  lp=slice_ubf_results[,"logLik"],
+  logLik=slice_ubf_results[,"logLik"],
   parameter=slice_ubf_results[,"g"],
-  lambda = 0.8
+  span = 0.8
 )
@@
 slice_bpf_mcap <- mcap(
-  lp=slice_bpf_results[,"logLik"],
+  logLik=slice_bpf_results[,"logLik"],
   parameter=slice_bpf_results[,"g"],
-  lambda = 1
+  span = 1
 )
@@
 slice_enkf_mcap <- mcap(
-  lp=slice_enkf_results[,"logLik"],
+  logLik=slice_enkf_results[,"logLik"],
   parameter=slice_enkf_results[,"g"],
-  lambda = 1
+  span = 1
 )
@@
   # par(mai=c(0.7,0.7,0.3,0.3))
-  ggplot() + geom_point(aes(x = mcap1$parameter, y = mcap1$lp)) + 
+  ggplot() + geom_point(aes(x = mcap1$parameter, y = mcap1$logLik)) + 
     geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$smoothed),
@@
   if(missing(ylab)) ylab <- "log likelihood"
-  plot(mcap1$lp ~ mcap1$parameter, xlab="",ylab="",xaxt="n",...)
+  plot(mcap1$logLik ~ mcap1$parameter, xlab="",ylab="",xaxt="n",...)
   mtext(side=2,text=ylab,line=yline)
@@
 text(xloc,
-  (1-yloc)*min(slice_abf_mcap$lp) + yloc*max(slice_abf_mcap$lp),
+  (1-yloc)*min(slice_abf_mcap$logLik) + yloc*max(slice_abf_mcap$logLik),
   cex=1.4,"ABF")
@@
 text(xloc,
-  (1-yloc)*min(slice_bpf_mcap$lp) + yloc*max(slice_bpf_mcap$lp),
+  (1-yloc)*min(slice_bpf_mcap$logLik) + yloc*max(slice_bpf_mcap$logLik),
   cex=1.4,"BPF")
@@
 text(xloc,
-  (1-yloc)*min(slice_enkf_mcap$lp) + yloc*max(slice_enkf_mcap$lp),
+  (1-yloc)*min(slice_enkf_mcap$logLik) + yloc*max(slice_enkf_mcap$logLik),
   cex=1.4,"EnKF")
@@
    if(missing(ylab)) ylab <- "profile log likelihood"
-   ggplot() + geom_point(aes(x = mcap1$parameter, y = mcap1$lp), shape = 1) +
+   ggplot() + geom_point(aes(x = mcap1$parameter, y = mcap1$logLik), shape = 1) +
       geom_line(mapping = aes(x = mcap1$fit$parameter, y = mcap1$fit$smoothed),
@@
 
-g_prof_mcap <- spatPomp::mcap(
-   lp=top_g_prof[,"loglik"],
+g_prof_mcap <- pomp::mcap(
+   logLik=top_g_prof[,"loglik"],
    parameter=top_g_prof[,"g"],
-   lambda = 0.90
+   span = 0.90
 )
